### 2024-12-08 cvm_unbiased/main.tex
Add proof for version of CVM that is unbiased.

\documentclass{article}
\usepackage[left=1.2in, right=1.2in]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{algorithmicx}
\newcommand{\prob}{\mathcal P}
\newcommand{\expectation}{\mathrm{E}}
\newcommand{\eps}{\varepsilon}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{lemma}{Lemma}

\begin{document}
\title{Unbiased CVM Algorithm}
\maketitle
\section{Notation}
For a finite set $S$, the probability space of uniformly sampling from the set is denoted by $U(S)$, i.e., for each $s \in S$ we have
$\prob_{U(S)}({s}) = |S|^{-1}$.
For the bernoulli probability space, over the set $\{\mathrm{true},\mathrm{false}\}$ we'll write $\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{\mathrm{true}\}) = p$.
$I(P)$ is the indicator function for a predicate $P$, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
\section{Algorithm}
Let us fix 
\begin{itemize}
  \item $a_1,\ldots,a_l$ for the stream of elements, and
  \item $n$ for the maximal number of elements in the buffer, and
  \item $f$ for the fraction of elements to keep in the buffer, when we subsample.
\end{itemize}
Note that: $nf$ must be an integer $\frac{1}{2} \leq f < 1$.

\begin{algorithm}
\caption{Unbiased CVM algorithm}
\begin{algorithmic}
\State $\chi \gets \emptyset$; $p \gets 1$
\For{$i \gets 1$ to $l$}
  \State $\chi \gets \chi - \{a_i\}$
  \State $\mathrm{coin} \gets \mathrm{Ber}(p)$
  \If{$\mathrm{coin}$}
    \State $\chi \gets \chi \cup \{a_i\}$
  \EndIf
  \If{$|\chi| = n$}
    \State $\chi \gets U(\{ S \subseteq \chi \,|\, |S|=nf \})$
    \State $p \gets p f$
  \EndIf
\EndFor
\State \Return{$p^{-1} |\chi|$}
\end{algorithmic}
\end{algorithm}
We will denote the first five lines of the loop as step 1, the last four lines as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
The elements of the probability spaces are pairs composed of a set and a fraction, representing $\chi$ and $p$. 
For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, $\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state. 
We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, in particular, $\chi$ (resp. $p$) is the projection to the first (resp. second) component.

The state of the algorithm after processing only step 1 of the $i$-th loop iteration is denoted by $\Omega'_i$.
So the sequence of states is represented by the distributions $\Omega_0, \Omega'_1, \Omega_1, \ldots, \Omega'_l, \Omega_l$.
A few easy-to-see observations without proof for any $0 \leq i \leq l$:
\begin{itemize}
\item $p(\omega) \leq 1$ a.s. (almost surely) for $\omega \in \Omega_i$,
\item $\chi(\omega) \subseteq \{a_1,\ldots,a_i\}$ a.s. for $\omega \in \Omega_i$,
\item $|\chi(\omega)| < n$ a.s. for $\omega \in \Omega_i$.
\end{itemize}
Also for the intermediate states, we have:
\begin{itemize}
\item $p(\omega) \leq 1$ a. s. for $\omega \in \Omega'_i$,
\item $\chi(\omega) \subseteq \{a_1,\ldots,a_i\}$ a.s. for $\omega \in \Omega'_i$,
\item $|\chi(\omega)| \leq n$ a.s. for $\omega \in \Omega'_i$.
\end{itemize}

\begin{lemma}
\label{le:neg_cor}
For all $k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$:
\begin{align*}
  \expectation_{\Omega_k}\left[ \prod_{s \in S} p^{-1} I(s \in \chi) \right] & \leq 1 & 
  \expectation_{\Omega'_k}\left[ \prod_{s \in S} p^{-1} I(s \in \chi) \right] & \leq 1
\end{align*}
\end{lemma}
\begin{proof}
We show the result using induction over $k$. Note that we show the statement for arbitrary $S$, e.g., the induction statements are:
\begin{eqnarray*}
P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[p^{-|S|} I(S \subseteq \chi)\right] \leq 1\right) \\
Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[p^{-|S|} I(S \subseteq \chi)\right] \leq 1\right)
\end{eqnarray*}
and we will show $P(0),Q(1),P(1),Q(2),P(2),\ldots,Q(l),P(l)$ successively.
\paragraph{Induction start $P(0)$:} \phantom{.}\\
We have $p=1$ $\chi=\emptyset$ and $S \subseteq \emptyset$, and hence
\[
\expectation_{\Omega_0}\left[p^{-|S|} I(S \subseteq \chi)\right] = \expectation_{\Omega_0}\left[1^{-|\emptyset|} I(\emptyset \subseteq \emptyset))\right] = 1 \leq 1 \textrm{.}
\]
\paragraph{Induction step $P(k) \rightarrow Q(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$ and define $S' := S - \{ a_{k+1} \}$.
Note that $\Omega’_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where $a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable over the space $\Omega_k$.

In particular, for example:
\[
\prob_{\Omega'_{k+1}}( P(\chi, p) ) = \int_{\Omega_k} \int_{Ber(p(\omega))} P(\chi(\omega)-\{a_{k+1}\}\cup f(\tau), p(\omega)) \, d \tau \, d \omega  
\]
for all predicates $P$ where we define $f(\mathrm{true}) = \{a_{k+1}\}$ and $f(\mathrm{false}) = \emptyset$.

We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.
\\
If $a_k \in S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{ p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} I(\tau) \, d \tau \, d \omega \\ 
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|+1)} I(S' \subseteq \chi(\omega)) p(\omega) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}
If $a_k \notin S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} 1 \, d \tau \, d \omega \\ 
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|)} I(S' \subseteq \chi(\omega)) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}
\paragraph{Induction step $Q(k+1) \rightarrow P(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$.
Let us also introduce the abbreviation $C(X) := \{T \subset X \, | \, |T| = n f \}$, i.e., the set of $nf$-subsets of $X$.

First we estimate the probability that $S$ is in a random subset of $C(T)$:
\begin{equation}
  \label{eq:prob_c_t}
  \int_{U(C(T))} I(S \subseteq \tau) \, d \tau = I(S \subseteq T) \binom{|T-S|}{nf-|S|} \binom{|T|}{nf}^{-1} \leq I(S \subseteq T) \left(\frac{nf}{|T|}\right)^{|S|}
\end{equation}
For the above it is best to consider the two cases $S \subseteq T$ and the converse separately. Note that if $S \subseteq \tau$ for $\tau \in C(S)$ then it must be true that: 
$S \subseteq T$.

Let us again note that $\Omega_{k+1}$ is a compound distribution over $\Omega_k$. In general, for all predicates $P$:
\[
  \prob_{\Omega_{k+1}}( P(\chi, p) )  = \int_{\Omega'_{k+1}} I( |\chi(\omega)|<n ) P(\chi(\omega),p(\omega)) + I( |\chi(\omega)|=n ) \int_{U(\chi(\omega))} P(\tau, f p(\omega)) \, d \tau  \, d \omega \textrm{.}
\]

With this we can can now verify the induction step:
\begin{eqnarray*}
&  & \textstyle\expectation_{\Omega_{k+1}}\left[p^{-|S|} I(S \subseteq \chi)\right]  \\
& = &\textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega 
+ \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)}{(p(\omega)f)^{|S|}} \int_{U(C(\chi(\omega)))} I(S \subseteq \tau) d \tau \, d \omega  \\
& \underset{\textrm{Eq.~\ref{eq:prob_c_t}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega
+ \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)}{(p(\omega)f)^{|S|}} I(S \subseteq \chi(\omega)) \left(\frac{nf}{|\chi(\omega)|}\right)^{|S|} \, d \omega  \\
& = & \textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)+I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega \\
& = & \textstyle \expectation_{\Omega'_{k+1}}\left[p^{-|S|} I(S \subseteq \chi) \right] \leq_{IH} 1
\end{eqnarray*}

\end{proof}
The dual version of Lemma~\ref{le:neg_cor} would be: $\expectation[ (1-p)^{-|S|} I(S \cap \chi=\emptyset) ] \leq 1$, but this is somewhat undefined for $p=1$.
This can easily be resolved, by taking into account that the first sub-sampling step happens deterministically, exactly at the iteration $k$, when $\{a_1,\ldots,a_k\} = n$,
i.e., we have $p \leq f < 1$ a.s. for $\Omega_l$ if $\{a_1,\ldots,a_l\} \geq n$ and $p=1$ a.s. otherwise.

\begin{lemma}
\label{le:neg_cor_2}
If $\{a_1,\ldots,a_l\} \geq n$, then for all $S \subseteq \{a_1,\ldots,a_l\}$:
\[
\expectation_{\Omega_l} \left[ (1-p)^{-|S|} I(S \cap \chi=\emptyset) \right] \leq 1 \textrm{.}
\]
\end{lemma}
\begin{proof}
The proof is analogous to the proof of Lemma~\ref{le:neg_cor} using induction (with special handling for the $p=1$ case).
\end{proof}

\section{Concentration}
This section establishes that the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$. This will be done by Chernoff bounds for the probability that the estimate is above $(1+\eps)|A|$ (resp. below $(1-\eps)|A|$ assuming $p$ is not too small and a tail estimate for $p$ being too small.

It should be noted that concentration is trivial, if $|A| \leq n$, i.e., if we never need to do sub-sampling.
So we assume $|A| \geq n$.

We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$. Moreover we'll introduce the notation $\tilde{p} := \max(p,q)$.
Basically $\tilde{p}$ is a clamped version of $p$, and in particular, $\tilde{p}^{-1}$ is bounded from above by $q^{-1}$.

The following is a helper lemma:
\begin{lemma}
\label{le:helper}
For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
\[
\prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.

Moreover, we define 
\[
  a(t) := \sup_{p \in (0,1]} p (e^{t/ \tilde{p}}-1) = q (e^{t/q}-1)
\]

To get a tail estimate, we use the Cram\'{e}r-Chernoff method: 
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\  
  & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} (1 + I(s \in \chi) (e^{t \tilde{p}^{-1}}-1) ) \right] \\  
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \left(1 + \frac{a(t)}{p} I(s \in \chi) \right) \right] \\  
 & = & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \expectation_\Omega \left[ \prod_{s \in T} (p^{-1} I(s \in \chi) ) \right] \\  
 & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A'| ) \sum_{T \subseteq A'} a(t)^{|T|} \\  
 & = & \left(\exp( -t (1+\eps) |A| ) (1+a(t))\right)^{|A'|} \\ & \leq & \exp ( |A| (\ln(1+a(t))-t (1+\eps)))
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$) for some $t > 0$:
\[
  \ln(1+a(t))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
which can be verified for $t = q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) > 0$.
The latter can be established by analyzing the function $f(q,\eps) = -\ln (1+a(t)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

Using the previous we can estimate bounds for $p$ becoming too small, as well as, an upper tail bound.
\begin{lemma}\label{le:low_p}
\[
\prob_{\Omega_l}(p < q) \leq l \exp\left( - \frac{n}{12}\right)
\]
\end{lemma}
\begin{proof}
We'll use a similar strategy as in the $\mathrm{Bad}_2$ bound in the original CVM paper. 
Let $j$ be maximal, s.t., $q \leq f^j$.

Hence $f^{j+1} < q$ and: 
\begin{equation}
\label{eq:f_j}
  f^j \leq 2f f^j < 2q = \frac{n}{2|A|} \textrm{.}
\end{equation}

First, we bound the probability of jumping from $p=f^j$ to $p=f^j+1$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$.

Then
\begin{eqnarray*}
  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( \tilde{p}^{-1} |\chi| \geq f^{-j} n ) \\
    & \underset{Eq.~\ref{eq:f_j}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| ) \\
    & \underset{Lem.~\ref{le:helper}}{\leq} & \exp( - n/12)
\end{eqnarray*}

The probability that this happens ever in the entire process is then at most $l$ times the above which proves the lemma.
\end{proof}

Now we come to the tail bounds. The upper tail bound follows directly from Lemma~\ref{le:helper}:
\begin{lemma}\label{le:upper_tail}
Let $0 < \eps \leq 1$ then:
\[
  L = \prob_{\Omega_l} ( p^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
\begin{eqnarray*}  
  L & =&  \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \\
  & \leq & \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| ) \\
  & \underset{Lemma~\ref{le:helper}}{\leq}  &  \exp\left(-\frac{n}{12} \eps^2\right)
\end{eqnarray*}
\end{proof}

\begin{lemma}\label{le:lower_tail}
Let $0 < \eps \leq 1$ then:
\[
  L := \prob_{\Omega_l} ( p^{-1} |\chi| \leq (1-\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
\]
\end{lemma}
\begin{proof}
Let us define 
\begin{align*}
  a & := \frac{\eps}{1-\eps} &
  t & := (1-q+\eps q) \ln \left(1 + \frac{\eps}{(1-q)(1-\eps)} \right) \\
  f(p) & = \begin{cases} 1-p+\eps p & \textrm{ if } p \geq q \\ \frac{t}{\ln \left(1+\frac{3}{4}a\right)} & \textrm{ otherwise.} \end{cases}
\end{align*}

We first observe that $t > 0$. Moreover, 
\begin{equation}\label{eq:f_p}
  \exp\left( \frac{t}{f(p)}\right)-1 \leq a (1-p) \textrm{ for } 0 < p < 1  
\end{equation}

With these definitions we again follow the Cram\'{e}r-Chernoff method: 
{\allowdisplaybreaks
\begin{eqnarray*}
  L & = & \prob_{\Omega_l} \left( |A-\chi| \geq (1-(1-\eps)p) |A| \wedge p \geq q \right) \\
    & = & \prob_{\Omega_l} \left( |A-\chi| \geq f(p)|A| \wedge p \geq q \right) \\
    & \leq & \prob_{\Omega_l} \left( |A-\chi| \geq f(p)|A| \right)  \\
    & \leq & \exp( -t |A| ) \expectation_{\Omega_l} \left[ \exp \left( \frac{t}{f(p)} |A-\chi| \right) \right] \\
    & = & \exp( -t |A| ) \expectation_{\Omega_l} \left[ \prod_{s \in A} 1 + (e^{t/f(p)} - 1) I(s \notin \chi) \right] \\
    & \underset{Eq.~\ref{eq:f_p}}{\leq}  & \exp( -t |A| ) \sum_{T \subseteq A} \expectation_{\Omega_l} \left[ \prod_{s \in T} \frac{a}{1-p} I(s \notin \chi) \right] \\
    & \underset{Lemma~\ref{le:neg_cor_2}}{\leq} & \exp( -t |A| ) (1+a)^{|A|}
\end{eqnarray*}
}

Substituting $t$ and $a$ and using $|A| = \frac{q}{4n}$, we can see that the lemma is true if
\[
  f(q,\eps) = (1-q+\eps q) \ln\left(1+\frac{\eps}{(1-q)(1-\eps)}\right) - \ln\left(1+\frac{\eps}{1-\eps}\right) - \frac{q}{3} \eps^2
\]
is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
This can be verified by checking that $f(q,0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
\end{proof}
We can now establish the concentration result:
\begin{theorem}
Let $0 < \eps < 1$ and $0 < \delta < 1$ and $n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
\[
  L = \prob_{\Omega_l} \left( | p^{-1} |\chi| - |A| | \geq \eps |A| \right) \leq \delta
\]
\end{theorem}
\begin{proof}
Note that the theorem is trivial if $|A| < n$. If not:
\begin{eqnarray*}
  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right) + 
    \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1+\eps) |A| \wedge p \geq q \right) + 
    \prob_{\Omega_l} \left( p < q \right) \\
    & \underset{Le.~\ref{le:low_p}-\ref{le:upper_tail}}{\leq} &
    \exp\left( - \frac{n}{12} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
    & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3} 
\end{eqnarray*}
\end{proof}

\section{Unbiasedness}
It is easy to conclude using Lemma~\ref{le:neg_cor} and \ref{le:neg_cor_2} that for each $s \in A$ that:
$\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
By linearity of expectation we can conclude that
\[
  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = |A| \textrm{.}
\]
\end{document}

### 2024-12-09 cvm_unbiased/build.sh
Add build script

@@ -1,0 +1,1 @@
+latexmk -interaction=nonstopmode -pdf main.tex

### 2024-12-16 cvm_unbiased/main.tex
fix some minor typos

@@
 \caption{Unbiased CVM algorithm}
-\begin{algorithmic}
+\begin{algorithmic}[1]
 \State $\chi \gets \emptyset$; $p \gets 1$
@@
 \end{algorithm}
-We will denote the first five lines of the loop as step 1, the last four lines as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
+
+We will denote the first five lines of the loop ($4$--$7$) as step 1, the last four lines ($8$--$11$) as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
 The elements of the probability spaces are pairs composed of a set and a fraction, representing $\chi$ and $p$.
@@
 
-We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.
-\\
-If $a_k \in S$ then:
+We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.\\
+
+If $a_{k+1} \in S$ then:
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-If $a_k \notin S$ then:
+
+If $a_{k+1} \notin S$ then:
 \begin{eqnarray*}
@@
 Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$.
-Let us also introduce the abbreviation $C(X) := \{T \subset X \, | \, |T| = n f \}$, i.e., the set of $nf$-subsets of $X$.
+Let us also introduce the abbreviation $C(X) := \{\tau \subset X \, | \, |\tau| = n f \}$, i.e., the set of $nf$-subsets of $X$.
 
@@
  & = & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \expectation_\Omega \left[ \prod_{s \in T} (p^{-1} I(s \in \chi) ) \right] \\
- & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A'| ) \sum_{T \subseteq A'} a(t)^{|T|} \\  
+ & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \\
  & = & \left(\exp( -t (1+\eps) |A| ) (1+a(t))\right)^{|A'|} \\ & \leq & \exp ( |A| (\ln(1+a(t))-t (1+\eps)))
@@
 
-First, we bound the probability of jumping from $p=f^j$ to $p=f^j+1$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
+First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
 It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$.

### 2024-12-18 cvm_unbiased/main.tex
Fix proof for the CVM' paper.
Thanks to Yong-Kiam for spotting the error in Lemma 2 (of the previous paper).

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{amsthm}
@@
 \usepackage{algorithmicx}
+\usepackage{hyperref}
 \newcommand{\prob}{\mathcal P}
@@
 $\prob_{U(S)}({s}) = |S|^{-1}$.
-For the bernoulli probability space, over the set $\{\mathrm{true},\mathrm{false}\}$ we'll write $\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{\mathrm{true}\}) = p$.
+For the bernoulli probability space, over the set $\{0,1\}$ we'll write $\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{1\}) = p$.
 $I(P)$ is the indicator function for a predicate $P$, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
@@
 \end{itemize}
+Given a set $T$ with $n$ elements, we'll write $C(T)$ for the $nf$-subsets of $T$, i.e.:
+\[
+  C(T) = \{ \tau \subseteq T \, | \, |\tau| = nf \} \textrm{.}
+\]
+\begin{lemma}\label{le:samp_prod}
+Let $T$ be a set with $n$ elements, $S \subseteq T$ and $g: \{0,1\} \rightarrow \mathbb R^+$ then
+\[
+  \int_{U(C(T))} \prod_{s \in S} g(I(s \in \tau)) \leq \prod_{s \in S} \int_{\mathrm{Ber}(f)} g(\tau) d \tau
+\]
+\end{lemma}
+\begin{proof}
+We'll verify the cases $g(0) \leq g(1)$ and the converse seperately. For the case $g(0) \leq g(1)$:
+We want to use Theorem 2.11 by Joag-Dev and Proschan~\cite{joagdev1983}, which requires us to view $C(T)$ as a permutation distribution.
+
+Let us for that consider $P := U(\{f : T \rightarrow \{0,..,|T|-1\}, f \textrm{bij.}\})$, i.e., we are assigning randomly distinct positions from $0$ to $|T|-1$ to the elements
+of $T$. Then $P$ is clearly a permuation distribution and the random variables describing the positions of each element are negatively associated according to Theorem 2.11.
+
+Now we can represent $C(T)$ as the choice of all the elements whose associated position is greater or equal to $(1-f)n$, i.e.:
+\begin{equation*}
+  \int_{U(C(T))} \prod_{s \in S} g(I(s \in \tau)) = \int_P \prod_{s \in S} g(I(\tau(s) \geq (1-f)n)) \, d\tau := R
+\end{equation*}
+Now we can use the Property P2~\cite[Page 288]{joagdev1983} since the $\tau(s)$ are negatively associated:
+\begin{equation}\label{eq:samp_prod_2}
+  R \leq \prod_{s \in S} \int_P g(I(\tau(s) \geq (1-f)n)) \, d\tau
+\end{equation}
+Note that this works because $x \rightarrow g(I(x \geq (1-f)n))$ is a monotone increasing function.
+Moreover the right hand side of Eq.~\ref{eq:samp_prod_2} is easy to be seen to be equal to the right hand side of the lemma we want to show.
+
+For the case $g(0) > g(1)$ we can do a similar trick, except in this case we associate with the choice of $\tau$ the elements with positions $\{0,\ldots,nf-1\}$.
+\end{proof}
 
@@
 \label{le:neg_cor}
-For all $k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$:
+Let $h : \mathbb R \rightarrow \mathbb R^+$ be a positive concave function, then for all 
+$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$:
 \begin{align*}
-  \expectation_{\Omega_k}\left[ \prod_{s \in S} p^{-1} I(s \in \chi) \right] & \leq 1 &
-  \expectation_{\Omega'_k}\left[ \prod_{s \in S} p^{-1} I(s \in \chi) \right] & \leq 1
+  \expectation_{\Omega_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & \leq h(1)^{|S|} &
+  \expectation_{\Omega'_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & \leq h(1)^{|S|}
 \end{align*}
@@
 \begin{proof}
+Note because $h$ is concave we can use the reverse Jensen's inequality
+\begin{equation}\label{eq:rev_jensen}
+  E [ h(X(\omega)) ] \leq h(E X)
+\end{equation}
+for all integrable random variables and probability spaces.
+
 We show the result using induction over $k$. Note that we show the statement for arbitrary $S$, e.g., the induction statements are:
 \begin{eqnarray*}
-P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[p^{-|S|} I(S \subseteq \chi)\right] \leq 1\right) \\
-Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[p^{-|S|} I(S \subseteq \chi)\right] \leq 1\right)
+P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} \right) \\
+Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} \right)
 \end{eqnarray*}
@@
 \paragraph{Induction start $P(0)$:} \phantom{.}\\
-We have $p=1$ $\chi=\emptyset$ and $S \subseteq \emptyset$, and hence
+We have $S \subseteq \emptyset$, and hence
 \[
-\expectation_{\Omega_0}\left[p^{-|S|} I(S \subseteq \chi)\right] = \expectation_{\Omega_0}\left[1^{-|\emptyset|} I(\emptyset \subseteq \emptyset))\right] = 1 \leq 1 \textrm{.}
+\expectation_{\Omega_0}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] = \expectation_{\Omega_0}\left[1\right] = 1 \leq h(1)^0 \textrm{.}
 \]
@@
 Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$ and define $S' := S - \{ a_{k+1} \}$.
-Note that $\Omega’_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where $a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable over the space $\Omega_k$.
+Note that $\Omega'_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where $a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable over the space $\Omega_k$.
 
@@
 \[
-\prob_{\Omega'_{k+1}}( P(\chi, p) ) = \int_{\Omega_k} \int_{Ber(p(\omega))} P(\chi(\omega)-\{a_{k+1}\}\cup f(\tau), p(\omega)) \, d \tau \, d \omega
+\prob_{\Omega'_{k+1}}( P(\chi, p) ) = \int_{\Omega_k} \int_{\textrm{Ber}(p(\omega))} P(\chi(\omega)-\{a_{k+1}\}\cup f(\tau), p(\omega)) \, d \tau \, d \omega
 \]
-for all predicates $P$ where we define $f(\mathrm{true}) = \{a_{k+1}\}$ and $f(\mathrm{false}) = \emptyset$.
+for all predicates $P$ where we define $f(1) = \{a_{k+1}\}$ and $f(0) = \emptyset$.
 
-We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.\\
-
-If $a_{k+1} \in S$ then:
+We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$. If $a_{k+1} \in S$:
 \begin{eqnarray*}
-  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{ p(\omega)^{|S'|+1}}
-  \int_{\mathrm{Ber}(p(\omega))} I(\tau) \, d \tau \, d \omega \\
-  & = & \int_{\Omega_k} p(\omega)^{-(|S'|+1)} I(S' \subseteq \chi(\omega)) p(\omega) \, d \omega \\
-  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
+  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & = & \int_{\Omega_k} \left(\prod_{s \in S'} h(p^{-1} I(s \in \chi))\right)
+  \int_{\mathrm{Ber}(p(\omega))} h(p^{-1} \tau) \, d \tau \, d \omega \\
+  & \underset{\textrm{Eq.~\ref{eq:rev_jensen}}}{\leq} & \int_{\Omega_k} \left( \prod_{s \in S'} h(p^{-1} I(s \in \chi)) \right)
+  h\left(\int_{\mathrm{Ber}(p(\omega))} p^{-1} \tau \, d \tau\right) \, d \omega \\
+  & = & \int_{\Omega_k} \left( \prod_{s \in S'} h(p^{-1} I(s \in \chi)) \right) h(1) \, d \omega \\
+  & \underset{\textrm{IH}}{\leq} & h(1)^{|S'|} h(1) = h(1)^{|S|}
 \end{eqnarray*}
 
-If $a_{k+1} \notin S$ then:
-\begin{eqnarray*}
-  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{p(\omega)^{|S'|+1}}
-  \int_{\mathrm{Ber}(p(\omega))} 1 \, d \tau \, d \omega \\
-  & = & \int_{\Omega_k} p(\omega)^{-(|S'|)} I(S' \subseteq \chi(\omega)) \, d \omega \\
-  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
-\end{eqnarray*}
+If $a_{k+1} \notin S$ then $S' = S$ and:
+\begin{equation}
+  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] = \int_{\Omega_k} \prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \leq_{IH} h(1)^{|S'|} = h(1)^{|S|}
+\end{equation}
 \paragraph{Induction step $Q(k+1) \rightarrow P(k+1)$:} \phantom{.}\\
 Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$.
-Let us also introduce the abbreviation $C(X) := \{\tau \subset X \, | \, |\tau| = n f \}$, i.e., the set of $nf$-subsets of $X$.
-
-First we estimate the probability that $S$ is in a random subset of $C(T)$:
-\begin{equation}
-  \label{eq:prob_c_t}
-  \int_{U(C(T))} I(S \subseteq \tau) \, d \tau = I(S \subseteq T) \binom{|T-S|}{nf-|S|} \binom{|T|}{nf}^{-1} \leq I(S \subseteq T) \left(\frac{nf}{|T|}\right)^{|S|}
-\end{equation}
-For the above it is best to consider the two cases $S \subseteq T$ and the converse separately. Note that if $S \subseteq \tau$ for $\tau \in C(S)$ then it must be true that:
-$S \subseteq T$.
 
@@
 \begin{eqnarray*}
-&  & \textstyle\expectation_{\Omega_{k+1}}\left[p^{-|S|} I(S \subseteq \chi)\right]  \\
-& = &\textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega
-+ \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)}{(p(\omega)f)^{|S|}} \int_{U(C(\chi(\omega)))} I(S \subseteq \tau) d \tau \, d \omega  \\
-& \underset{\textrm{Eq.~\ref{eq:prob_c_t}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega
-+ \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)}{(p(\omega)f)^{|S|}} I(S \subseteq \chi(\omega)) \left(\frac{nf}{|\chi(\omega)|}\right)^{|S|} \, d \omega  \\
-& = & \textstyle \int_{\Omega'_{k+1}} \frac{I(|\chi(\omega)| = n)+I(|\chi(\omega)| < n)}{p(\omega)^{|S|}} I(S \subseteq \chi(\omega)) \, d \omega \\
-& = & \textstyle \expectation_{\Omega'_{k+1}}\left[p^{-|S|} I(S \subseteq \chi) \right] \leq_{IH} 1
+&  & \textstyle\expectation_{\Omega_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi))\right]  \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \int_{U(C(\chi))} \prod_{s \in S \cap \chi} h((pf)^{-1}I(s \in \tau )) d \tau \, d \omega  \\
+& \underset{\textrm{Le.~\ref{le:samp_prod}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} \int_{\mathrm{Ber(f)}} h((pf)^{-1}\tau) d \tau \, d \omega  \\
+& \underset{\textrm{Eq.~\ref{eq:rev_jensen}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} h\left(\int_{\mathrm{Ber(f)}} (pf)^{-1}\tau d \tau\right) \, d \omega  \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} h(p^{-1}) \, d \omega  \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega  \\
+& = & \textstyle \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \underset{\textrm{IH}}{\leq} h(1)^{|S|}
 \end{eqnarray*}
 
-\end{proof}
-The dual version of Lemma~\ref{le:neg_cor} would be: $\expectation[ (1-p)^{-|S|} I(S \cap \chi=\emptyset) ] \leq 1$, but this is somewhat undefined for $p=1$.
-This can easily be resolved, by taking into account that the first sub-sampling step happens deterministically, exactly at the iteration $k$, when $\{a_1,\ldots,a_k\} = n$,
-i.e., we have $p \leq f < 1$ a.s. for $\Omega_l$ if $\{a_1,\ldots,a_l\} \geq n$ and $p=1$ a.s. otherwise.
-
-\begin{lemma}
-\label{le:neg_cor_2}
-If $\{a_1,\ldots,a_l\} \geq n$, then for all $S \subseteq \{a_1,\ldots,a_l\}$:
-\[
-\expectation_{\Omega_l} \left[ (1-p)^{-|S|} I(S \cap \chi=\emptyset) \right] \leq 1 \textrm{.}
-\]
-\end{lemma}
-\begin{proof}
-The proof is analogous to the proof of Lemma~\ref{le:neg_cor} using induction (with special handling for the $p=1$ case).
 \end{proof}
@@
 
-It should be noted that concentration is trivial, if $|A| \leq n$, i.e., if we never need to do sub-sampling.
+It should be noted that concentration is trivial, if $|A| < n$, i.e., if we never need to do sub-sampling.
 So we assume $|A| \geq n$.
@@
 Moreover, we define
-\[
-  a(t) := \sup_{p \in (0,1]} p (e^{t/ \tilde{p}}-1) = q (e^{t/q}-1)
-\]
+\begin{align*}
+  t & := q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) \\
+  h(x) & := 1+qx(e^{t/q}-1)  
+\end{align*}
 
@@
   \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
-  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
+  & \underset{\textrm{Markov,}t>0}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
  & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
-  & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} (1 + I(s \in \chi) (e^{t \tilde{p}^{-1}}-1) ) \right] \\
- & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \left(1 + \frac{a(t)}{p} I(s \in \chi) \right) \right] \\
- & = & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \expectation_\Omega \left[ \prod_{s \in T} (p^{-1} I(s \in \chi) ) \right] \\
- & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \\
- & = & \left(\exp( -t (1+\eps) |A| ) (1+a(t))\right)^{|A'|} \\ & \leq & \exp ( |A| (\ln(1+a(t))-t (1+\eps)))
+ & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} h( p^{-1} I(s \in \chi) ) \right] \\
+ & \underset{\textrm{Le.~\ref{le:neg_cor}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
+ & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
 \end{eqnarray*}}
-So we just need to show that (using $|A|=\frac{n}{4q}$) for some $t > 0$:
+So we just need to show that (using $|A|=\frac{n}{4q}$):
 \[
-  \ln(1+a(t))-t (1+\eps) \leq \frac{-q \eps^2}{3}
+  \ln(h(1))-t (1+\eps) \leq \frac{-q \eps^2}{3}
 \]
-which can be verified for $t = q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) > 0$.
-The latter can be established by analyzing the function $f(q,\eps) = -\ln (1+a(t)) +t (1+\eps) - \frac{q\eps^2}{3}$.
+The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
 For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
@@
   \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( \tilde{p}^{-1} |\chi| \geq f^{-j} n ) \\
-    & \underset{Eq.~\ref{eq:f_j}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| ) \\
-    & \underset{Lem.~\ref{le:helper}}{\leq} & \exp( - n/12)
+    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| ) \\
+    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp( - n/12)
 \end{eqnarray*}
@@
   & \leq & \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| ) \\
-  & \underset{Lemma~\ref{le:helper}}{\leq}  &  \exp\left(-\frac{n}{12} \eps^2\right)
+  & \underset{\textrm{Le.~\ref{le:helper}}}{\leq}  &  \exp\left(-\frac{n}{12} \eps^2\right)
 \end{eqnarray*}
@@
 \begin{lemma}\label{le:lower_tail}
-Let $0 < \eps \leq 1$ then:
+Let $0 < \eps < 1$ then:
 \[
@@
 \begin{align*}
-  a & := \frac{\eps}{1-\eps} &
-  t & := (1-q+\eps q) \ln \left(1 + \frac{\eps}{(1-q)(1-\eps)} \right) \\
-  f(p) & = \begin{cases} 1-p+\eps p & \textrm{ if } p \geq q \\ \frac{t}{\ln \left(1+\frac{3}{4}a\right)} & \textrm{ otherwise.} \end{cases}
+  t & := - q \ln\left( \frac{(1-\eps)^{-1}-q}{1-q}\right) < 0 \\
+  h(x) & := 1+q \min(x,q^{-1}) (e^{t/q}-1)  
 \end{align*}
-
-We first observe that $t > 0$. Moreover,
-\begin{equation}\label{eq:f_p}
-  \exp\left( \frac{t}{f(p)}\right)-1 \leq a (1-p) \textrm{ for } 0 < p < 1
-\end{equation}
 
@@
 \begin{eqnarray*}
-  L & = & \prob_{\Omega_l} \left( |A-\chi| \geq (1-(1-\eps)p) |A| \wedge p \geq q \right) \\
-    & = & \prob_{\Omega_l} \left( |A-\chi| \geq f(p)|A| \wedge p \geq q \right) \\
-    & \leq & \prob_{\Omega_l} \left( |A-\chi| \geq f(p)|A| \right)  \\
-    & \leq & \exp( -t |A| ) \expectation_{\Omega_l} \left[ \exp \left( \frac{t}{f(p)} |A-\chi| \right) \right] \\
-    & = & \exp( -t |A| ) \expectation_{\Omega_l} \left[ \prod_{s \in A} 1 + (e^{t/f(p)} - 1) I(s \notin \chi) \right] \\
-    & \underset{Eq.~\ref{eq:f_p}}{\leq}  & \exp( -t |A| ) \sum_{T \subseteq A} \expectation_{\Omega_l} \left[ \prod_{s \in T} \frac{a}{1-p} I(s \notin \chi) \right] \\
-    & \underset{Lemma~\ref{le:neg_cor_2}}{\leq} & \exp( -t |A| ) (1+a)^{|A|}
+  L & = & \prob_{\Omega_l} \left( \tilde{p}^{-1} |\chi| \leq (1-\eps)|A| \right) \\
+    & \underset{\textrm{Markov,}t<0}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
+    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
+    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} h( p^{-1} I(s \in \chi) ) \right] \\
+    & \underset{\textrm{Le.~\ref{le:neg_cor}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
+    & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
 \end{eqnarray*}
@@
 
-Substituting $t$ and $a$ and using $|A| = \frac{q}{4n}$, we can see that the lemma is true if
+Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
 \[
-  f(q,\eps) = (1-q+\eps q) \ln\left(1+\frac{\eps}{(1-q)(1-\eps)}\right) - \ln\left(1+\frac{\eps}{1-\eps}\right) - \frac{q}{3} \eps^2
+  f(q,\eps) = t (1-\eps) - \ln(1+q(e^{t/q}-1)) - \frac{q}{3} \eps^2
 \]
@@
 \section{Unbiasedness}
-It is easy to conclude using Lemma~\ref{le:neg_cor} and \ref{le:neg_cor_2} that for each $s \in A$ that:
-$\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
+Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor} using $h(x) = x$ and $h(x) = \max(M-x,0)$ that:
+\begin{eqnarray*}
+  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] \leq 1
+  \expectation_{\Omega_l} [ M - p^{-1} I(s \in S) ] = \expectation_{\Omega_l} [ M - \max(0,p^{-1} I(s \in S)) ] \leq M-1  
+\end{eqnarray*}
+which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
 By linearity of expectation we can conclude that
@@
 \]
+\bibliographystyle{plainurl}
+\bibliography{main}
+
 \end{document}

### 2024-12-18 cvm_unbiased/build.sh

@@ -1,1 +1,1 @@
-latexmk -interaction=nonstopmode -pdf main.tex
+latexmk -bibtex -interaction=nonstopmode -pdf main.tex

### 2024-12-19 cvm_unbiased/main.tex
Minor improvements to Lemma 3 and 5.
Fix bug with concavity.

@@
 \usepackage{hyperref}
+
 \newcommand{\prob}{\mathcal P}
@@
 \newcommand{\eps}{\varepsilon}
+\renewcommand{\phi}{\varphi}
 \newtheorem{theorem}{Theorem}
@@
 \begin{lemma}\label{le:samp_prod}
-Let $T$ be a set with $n$ elements, $S \subseteq T$ and $g: \{0,1\} \rightarrow \mathbb R^+$ then
+Let $T$ be a set with $n$ elements, $S \subseteq T$ and $g: \{0,1\} \rightarrow \mathbb R_{\geq 0}$ then
 \[
@@
 For the case $g(0) > g(1)$ we can do a similar trick, except in this case we associate with the choice of $\tau$ the elements with positions $\{0,\ldots,nf-1\}$.
+
+Important Note: In the paper by Joag-Dev and Proschan ``positive'' means non-negative. (See last paragraph of Section 1.)
 \end{proof}
@@
 \begin{lemma}
-\label{le:neg_cor}
-Let $h : \mathbb R \rightarrow \mathbb R^+$ be a positive concave function, then for all 
-$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$:
-\begin{align*}
-  \expectation_{\Omega_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & \leq h(1)^{|S|} &
-  \expectation_{\Omega'_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & \leq h(1)^{|S|}
-\end{align*}
+\label{le:neg_cor_prelim}
+Let $\phi : (0,1] \times {0,1} \rightarrow \mathbb R_{\geq 0}$ be function, fulfilling the following conditions:
+\begin{enumerate}
+\item \label{cond:phi_1} $(1-\alpha) \phi(x,0) + \alpha \phi(x,1) \leq \phi(x/\alpha,1)$ for all $0 < \alpha < 1$, $0 < x \leq 1$, and
+\item \label{cond:phi_2} $\phi(x,0) \leq \phi(y,0)$ for all $0 < x < y \leq 1$.
+\end{enumerate}
+Then for all 
+$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
+\[
+  \expectation_{\Omega}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} 
+\]
 \end{lemma}
 \begin{proof}
-Note because $h$ is concave we can use the reverse Jensen's inequality
-\begin{equation}\label{eq:rev_jensen}
-  E [ h(X(\omega)) ] \leq h(E X)
-\end{equation}
-for all integrable random variables and probability spaces.
-
-We show the result using induction over $k$. Note that we show the statement for arbitrary $S$, e.g., the induction statements are:
+We show the result using induction over $k$. Note that we show the statement for arbitrary $S$, i.e., the induction statements are:
 \begin{eqnarray*}
-P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} \right) \\
-Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} \right)
+P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} \right) \\
+Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} \right)
 \end{eqnarray*}
@@
 \[
-\expectation_{\Omega_0}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] = \expectation_{\Omega_0}\left[1\right] = 1 \leq h(1)^0 \textrm{.}
+\expectation_{\Omega_0}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] = \expectation_{\Omega_0}\left[1\right] = 1 \leq \phi(1,1)^0 \textrm{.}
 \]
@@
 \begin{eqnarray*}
-  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] & = & \int_{\Omega_k} \left(\prod_{s \in S'} h(p^{-1} I(s \in \chi))\right)
-  \int_{\mathrm{Ber}(p(\omega))} h(p^{-1} \tau) \, d \tau \, d \omega \\
-  & \underset{\textrm{Eq.~\ref{eq:rev_jensen}}}{\leq} & \int_{\Omega_k} \left( \prod_{s \in S'} h(p^{-1} I(s \in \chi)) \right)
-  h\left(\int_{\mathrm{Ber}(p(\omega))} p^{-1} \tau \, d \tau\right) \, d \omega \\
-  & = & \int_{\Omega_k} \left( \prod_{s \in S'} h(p^{-1} I(s \in \chi)) \right) h(1) \, d \omega \\
-  & \underset{\textrm{IH}}{\leq} & h(1)^{|S'|} h(1) = h(1)^{|S|}
+  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] & = & \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right)
+  \int_{\mathrm{Ber}(p(\omega))} \phi(p, \tau) \, d \tau \, d \omega \\
+  & = & \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right) ((1-p) \phi(p,0) + p \phi(p,1)) \, d \omega\\
+  & \underset{\textrm{Cond~\ref{cond:phi_1}}}{\leq} & \int_{\Omega_k} \left( \prod_{s \in S'} \phi(p, I(s \in \chi)) \right)
+  \phi(1,1) \, d \omega \\
+  & \underset{\textrm{IH}}{\leq} & \phi(1,1)^{|S'|} \phi(1,1) = \phi(1,1)^{|S|}
 \end{eqnarray*}
@@
 \begin{equation}
-  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] = \int_{\Omega_k} \prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \leq_{IH} h(1)^{|S'|} = h(1)^{|S|}
+  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] = \int_{\Omega_k} \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \leq_{IH} \phi(1,1)^{|S'|} = \phi(1,1)^{|S|}
 \end{equation}
@@
 \begin{eqnarray*}
-&  & \textstyle\expectation_{\Omega_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi))\right]  \\
-& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \int_{U(C(\chi))} \prod_{s \in S \cap \chi} h((pf)^{-1}I(s \in \tau )) d \tau \, d \omega  \\
-& \underset{\textrm{Le.~\ref{le:samp_prod}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} \int_{\mathrm{Ber(f)}} h((pf)^{-1}\tau) d \tau \, d \omega  \\
-& \underset{\textrm{Eq.~\ref{eq:rev_jensen}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} h\left(\int_{\mathrm{Ber(f)}} (pf)^{-1}\tau d \tau\right) \, d \omega  \\
-& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} h(0) \prod_{s \in S \cap \chi} h(p^{-1}) \, d \omega  \\
-& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S} h(p^{-1} I(s \in \chi)) \, d \omega  \\
-& = & \textstyle \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \underset{\textrm{IH}}{\leq} h(1)^{|S|}
+&  & \textstyle\expectation_{\Omega_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi))\right]  \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(pf,0) \int_{U(C(\chi))} \prod_{s \in S \cap \chi} \phi(pf,I(s \in \tau )) d \tau \, d \omega  \\
+& \underset{\textrm{Le.~\ref{le:samp_prod}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(pf,0) \prod_{s \in S \cap \chi} \int_{\mathrm{Ber(f)}} \phi(pf,\tau) d \tau \, d \omega  \\
+& \underset{\textrm{Cond~\ref{cond:phi_2}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(p,0) \prod_{s \in S \cap \chi} ((1-f)\phi(pf,0)+f\phi(pf,1)) \, d \omega  \\
+& \underset{\textrm{Cond~\ref{cond:phi_1}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(p,0) \prod_{s \in S \cap \chi} \phi(p,1) \, d \omega \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega  \\
+& = & \textstyle \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] \underset{\textrm{IH}}{\leq} \phi(1,1)^{|S|}
 \end{eqnarray*}
 
+\end{proof}
+
+More practical versions of the previous lemma:
+\begin{lemma}\label{le:neg_cor_pos}
+Let $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
+$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
+\[
+  \expectation_{\Omega}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
+\]
+\end{lemma}
+\begin{proof}
+Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
+\[
+  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) = (1-\alpha) h(0) + \alpha h(x^{-1}) \leq h(\alpha x^{-1}) = \phi(x/\alpha,1)
+\]
+for $0 < \alpha < 1$ and $0< x \leq 1$, and
+\[
+  \phi(x,0) = h(0) \leq h(0) = \phi(y,0)
+\]
+for $0 < x < y \leq 1$.
+\end{proof}
+
+Unfortunately the above function is only applicable for monotone increasing functions, because a monotone decreasing function
+will eventually be negative. The following modified version is for that case:
+
+\begin{lemma}\label{le:neg_cor_neg}
+Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
+$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
+\[
+  \expectation_{\Omega}\left[ \prod_{s \in S} I(p > q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
+\]
+\end{lemma}
+\begin{proof}
+Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
+\begin{eqnarray*}
+  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x > q) h(x^{-1}) \\
+   & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
+\end{eqnarray*}
+for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$. And
+\[
+  \phi(x,0) = I(x>q) h(0) \leq I(y > q) h(0) = \phi(y,0)
+\]
+for $0 < x < y \leq 1$, where we used that $x > q$ implies $y > q$.
 \end{proof}
@@
 \begin{align*}
-  t & := q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) \\
+  t & := q \ln( 1 + \eps) \\
   h(x) & := 1+qx(e^{t/q}-1)  
@@
  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} h( p^{-1} I(s \in \chi) ) \right] \\
- & \underset{\textrm{Le.~\ref{le:neg_cor}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
+ & \underset{\textrm{Le.~\ref{le:neg_cor_pos}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
  & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
@@
 \[
-  L := \prob_{\Omega_l} ( p^{-1} |\chi| \leq (1-\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
+  L := \prob_{\Omega_l} ( p^{-1} |\chi| \leq (1-\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{8} \eps^2\right)
 \]
@@
 \begin{align*}
-  t & := - q \ln\left( \frac{(1-\eps)^{-1}-q}{1-q}\right) < 0 \\
-  h(x) & := 1+q \min(x,q^{-1}) (e^{t/q}-1)  
+  t & := - q \ln(1-\eps) < 0 \\
+  h(x) & := 1+q x (e^{t/q}-1)  
 \end{align*}
+
+Note that $h(x) \geq 0$ for $0 \leq x \leq q^{-1}$ (can be checked by verifying it is true for $h(0)$ and $h(q^{-1})$ and the fact that the function is affine.)
 
@@
 \begin{eqnarray*}
-  L & = & \prob_{\Omega_l} \left( \tilde{p}^{-1} |\chi| \leq (1-\eps)|A| \right) \\
-    & \underset{\textrm{Markov,}t<0}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
-    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
-    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} h( p^{-1} I(s \in \chi) ) \right] \\
-    & \underset{\textrm{Le.~\ref{le:neg_cor}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
+  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
+    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p > q \right) \\
+    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
+    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
+    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
+    & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
     & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
@@
 \[
-  f(q,\eps) = t (1-\eps) - \ln(1+q(e^{t/q}-1)) - \frac{q}{3} \eps^2
+  f(q,\eps) = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
 \]
@@
     & \underset{Le.~\ref{le:low_p}-\ref{le:upper_tail}}{\leq} &
-    \exp\left( - \frac{n}{12} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
+    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
     & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
@@
 \section{Unbiasedness}
-Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor} using $h(x) = x$ and $h(x) = \max(M-x,0)$ that:
+Let $M$ be large enough such that $p^{-1} \leq M$ a.s.
+(e.g.
+we can choose $M = f^{-l}$).
+Then we can derive from Lemma~\ref{le:neg_cor_pos} using $h(x) = x$ and Lemma~\ref{le:neg_cor_neg} $h(x) = M+1-x$ that:
 \begin{eqnarray*}
-  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] \leq 1
-  \expectation_{\Omega_l} [ M - p^{-1} I(s \in S) ] = \expectation_{\Omega_l} [ M - \max(0,p^{-1} I(s \in S)) ] \leq M-1  
+  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & \leq 1 & \\
+  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  
 \end{eqnarray*}
@@
 \]
+
 \bibliographystyle{plainurl}

### 2024-12-20 cvm_unbiased/main.tex
Minor improvements.

@@
 More practical versions of the previous lemma:
-\begin{lemma}\label{le:neg_cor_pos}
-Let $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
-$k \in \{0,\ldots,l\}$, $S \subseteq \{a_1,..,a_k\}$, $\Omega \in \{\Omega_k,\Omega_k'\}$:
-\[
-  \expectation_{\Omega}\left[ \prod_{s \in S} h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
-\]
-\end{lemma}
-\begin{proof}
-Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
-\[
-  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) = (1-\alpha) h(0) + \alpha h(x^{-1}) \leq h(\alpha x^{-1}) = \phi(x/\alpha,1)
-\]
-for $0 < \alpha < 1$ and $0< x \leq 1$, and
-\[
-  \phi(x,0) = h(0) \leq h(0) = \phi(y,0)
-\]
-for $0 < x < y \leq 1$.
-\end{proof}
-
-Unfortunately the above function is only applicable for monotone increasing functions, because a monotone decreasing function
-will eventually be negative. The following modified version is for that case:
-
 \begin{lemma}\label{le:neg_cor_neg}
@@
 
-We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$. Moreover we'll introduce the notation $\tilde{p} := \max(p,q)$.
-Basically $\tilde{p}$ is a clamped version of $p$, and in particular, $\tilde{p}^{-1}$ is bounded from above by $q^{-1}$.
+We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$.
 
@@
 \[
-\prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
+\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
 \]
@@
 \begin{eqnarray*}
-  \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
-  & \underset{\textrm{Markov,}t>0}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
- & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
- & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} h( p^{-1} I(s \in \chi) ) \right] \\
- & \underset{\textrm{Le.~\ref{le:neg_cor_pos}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
+  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A'|) \right \wedge p \geq q) \\
+  & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
+  & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
+ & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
+ & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
+ & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
  & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
@@
 \begin{eqnarray*}
-  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( \tilde{p}^{-1} |\chi| \geq f^{-j} n ) \\
-    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| ) \\
-    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp( - n/12)
+  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
+    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wegde p \geq q ) \\
+    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp(- n/12)
 \end{eqnarray*}
@@
     & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p > q \right) \\
-    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
-    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
+    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
+    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
     & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\

### 2024-12-20 cvm_unbiased/main.tex
Minor corrections.

@@
 
-The following is a helper lemma:
+Let us start with the upper tail bound:
 \begin{lemma}
-\label{le:helper}
+\label{le:upper_tail}
 For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
 \[
-\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
+\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
 \]
@@
 \begin{eqnarray*}
-  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \right \wedge p \geq q) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A'|) \right \wedge p \geq q) \\
+  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
   & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
@@
   \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
-    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wegde p \geq q ) \\
-    & \underset{\textrm{Le.~\ref{le:helper}}}{\leq} & \exp(- n/12)
+    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wedge p \geq q ) \\
+    & \underset{\textrm{Le.~\ref{le:upper_tail}}}{\leq} & \exp(- n/12)
 \end{eqnarray*}
@@
 The probability that this happens ever in the entire process is then at most $l$ times the above which proves the lemma.
-\end{proof}
-
-Now we come to the tail bounds. The upper tail bound follows directly from Lemma~\ref{le:helper}:
-\begin{lemma}\label{le:upper_tail}
-Let $0 < \eps \leq 1$ then:
-\[
-  L = \prob_{\Omega_l} ( p^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{12} \eps^2\right)
-\]
-\end{lemma}
-\begin{proof}
-\begin{eqnarray*}
-  L & =&  \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| \wedge p \geq q) \\
-  & \leq & \prob_{\Omega_l} ( \tilde{p}^{-1} |\chi| \geq (1+\eps)|A| ) \\
-  & \underset{\textrm{Le.~\ref{le:helper}}}{\leq}  &  \exp\left(-\frac{n}{12} \eps^2\right)
-\end{eqnarray*}
 \end{proof}
@@
     \prob_{\Omega_l} \left( p < q \right) \\
-    & \underset{Le.~\ref{le:low_p}-\ref{le:upper_tail}}{\leq} &
+    & \underset{Le.~\ref{le:upper_tail}-\ref{le:lower_tail}}{\leq} &
     \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
@@
 \section{Unbiasedness}
-Let $M$ be large enough such that $p^{-1} \leq M$ a.s.
-(e.g.
-we can choose $M = f^{-l}$).
-Then we can derive from Lemma~\ref{le:neg_cor_pos} using $h(x) = x$ and Lemma~\ref{le:neg_cor_neg} $h(x) = M+1-x$ that:
+Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that:
 \begin{eqnarray*}
-  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & \leq 1 & \\
+  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in S) ] \leq 1 \\
   \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  

### 2024-12-20 cvm_unbiased/main.tex
Fixed some typos

@@
 \[
-\prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
+L := \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
 \]
@@
 \begin{eqnarray*}
-  \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
+  L & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
   & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
@@
 \]
-The latter can be established by analyzing the function $f(q,\eps) = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3}$.
-For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
-$f(q,\eps) \geq 0$.
+The latter can be established by analyzing the function
+\[
+  f(\eps) := -\ln(1+q \eps) + q \ln(1+\eps) (1+\eps) - \frac{q \eps^2}{3} = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3} \textrm{.}
+\]
+For which it is easy to check $f(0) = 0$ and the derivative with respect to $\eps$ is non-negative in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
+$f(\eps) \geq 0$.
 \end{proof}
 
-Using the previous we can estimate bounds for $p$ becoming too small, as well as, an upper tail bound.
+Using the previous we can estimate bounds for $p$ becoming too small:
 \begin{lemma}\label{le:low_p}
@@
 We'll use a similar strategy as in the $\mathrm{Bad}_2$ bound in the original CVM paper.
-Let $j$ be maximal, s.t., $q \leq f^j$.
-
-Hence $f^{j+1} < q$ and:
+Let $j$ be maximal, s.t., $q \leq f^j$. Hence $f^{j+1} < q$ and:
 \begin{equation}
@@
 \end{equation}
-
 First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
-It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$.
-
-Then
+It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$. Then
 \begin{eqnarray*}
   \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
-    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( \tilde{p}^{-1} |\chi| \geq 2|A| \wedge p \geq q ) \\
+    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( p^{-1} |\chi| \geq 2|A| \wedge p \geq q ) \\
     & \underset{\textrm{Le.~\ref{le:upper_tail}}}{\leq} & \exp(- n/12)
 \end{eqnarray*}
-
 The probability that this happens ever in the entire process is then at most $l$ times the above which proves the lemma.
@@
 \begin{align*}
-  t & := - q \ln(1-\eps) < 0 \\
+  t & := q \ln(1-\eps) < 0 \\
   h(x) & := 1+q x (e^{t/q}-1)  
@@
 \begin{eqnarray*}
-  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
-    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq exp(t (1-\eps)|A|) \wedge p > q \right) \\
+  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
+    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1-\eps)|A|) \wedge p > q \right) \\
     & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
@@
 }
-
 Substituting $t$ and $h$ and using $|A| = \frac{n}{4q}$, we can see that the lemma is true if
 \[
-  f(q,\eps) = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
+  f(\eps) := -q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
 \]
 is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
-This can be verified by checking that $f(q,0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
+This can be verified by checking that $f(0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
 \end{proof}
@@
   L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right) +
-    \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1+\eps) |A| \wedge p \geq q \right) +
+    \prob_{\Omega_l} \left( | p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q \right) +
     \prob_{\Omega_l} \left( p < q \right) \\

### 2025-01-06 cvm_unbiased/main.tex
Add counter example

@@
 \begin{eqnarray*}
-  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in S) ] \leq 1 \\
-  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  
+  \expectation_{\Omega_l} [ p^{-1} I(s \in \chi) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in \chi) ] \leq 1 \\
+  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in \chi) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in \chi)) ] \leq M  
 \end{eqnarray*}
-which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
+which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in \chi) ] = 1$.
 By linearity of expectation we can conclude that
 \[
-  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = |A| \textrm{.}
+  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in \chi) ] = |A| \textrm{.}
 \]
+
+\section{Other approaches}
+Initially I had tried to show that the RV's $I(s \in \chi)$ might be negatively associated. Unfortunately that is wrong, and because of the closure properties of
+negative association the same is true of $p^{-1} I(s \in \chi)$.
+
+The best counter-example (so far) I have is: 
+\begin{itemize}
+\item $a_1 = 1, \ldots, a_{15} = 15$ 
+\item $n = 10$
+\item $f = \frac{1}{2}$
+\end{itemize}
+then the RV's $f(\chi) = I(|\chi \cap \{1,..,10\}| \geq 5)$ and $g(\chi) = |\chi \cap \{11,\ldots,15\}|$ then 
+\[
+  E f g > E f E g
+\]
+with
+\begin{eqnarray*}
+  E f E g & = & \frac{59537}{24576} < 2.4225667318 \\
+  E f g & = & \frac{78625}{32256} > 2.4375310019 
+\end{eqnarray*}
+
+Note: I verified this using Haskell with exact computation, i.e, we branch on every coin flip and big integer rational arithmetic. So these are exact values.
 

### 2025-01-14 paper/main.tex
add initial paper stub in LIPIcs

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles.
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Formalization of CVM} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%NOTE: we have to inline all of these into one file for the final version if accepted.

\input{sec/abstract}

\input{sec/intro}

\input{sec/negdep}

\input{sec/cvmalt}

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{main}

\end{document}

### 2025-01-14 paper/sec/intro.tex
add initial paper stub in LIPIcs

\section{Introduction}
\label{sec:intro}

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{DBLP:conf/esa/0001VM22} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community.
Indeed, Knuth later published a note on the algorithm, pointing out its simplicity and also christening it the \emph{CVM} algorithm (which we will follow for the rest of this paper).
One striking property of the algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Indeed, the algorithm could theoretically be implemented in a setting where the objects in the data stream only allow for equality comparisons.
Another property is its simplicity, which is why the authors called it ``an algorithm for the text-book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.

\begin{algorithm}[h!]
	\caption{CVM algorithm for distinct elements estimation.}\label{alg:cvm}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A randomized cardinality approximation $c$ for set $A = \{ a_1,\dots,a_l \}$ with $$\Pr_c \left( | c - |A| | > \varepsilon |A| \right) \leq \delta$$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil{\frac{12}{\varepsilon^2} \ln{(\frac{81}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State {TODO}
  \EndFor
  \Return $\frac{|\chi|}{p}$
  \end{algorithmic}
\end{algorithm}

The algorithm's state is a buffer $\chi$ and a fraction $p > 0$.
The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
The algorithm iterates over stream elements, adding each one to the buffer with probability $p$, and conversely removing them with probability $(1-p)$.
If the buffer gets too large, approximately half of the elements are removed by discarding each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.

The pen-and-paper proof of correctness for~\cref{alg:cvm} is remarkably simple, requiring only undergraduate-level exposure to the concepts of randomized algorithms.
We set out to formalize the proof as described in the literature~\cite{DBLP:conf/esa/0001VM22}, hoping that it would be an easy exercise for the formalized algorithms textbook.

%TODO: a bit too long in details, we should trim this and move to later section
\subparagraph*{The road not taken:}
One of the difficult aspects of the proof is that it relies on an eager-lazy coin flip conversion.
To understand that, we should note that none of the observable random variables, such as the presence of a stream element in the buffer or conditions on the value of $p$, are independent of the other state variables, which makes the algorithm hard to analyze and makes the application of standard techniques from probability theory, such as Hoeffding's theorem impossible.
The authors resolved that problem by a simulation argument---they show that~\cref{alg:cvm} behaves stochastically identically to a different algorithm, which makes the relevant coin flips in a different order.
That modified algorithm performs a column of coin flips for each stream element.
An element is kept in the buffer if the first $k=log2(p-1)$ rows of the column are heads.
At each sub-sampling step, when p is divided by two, i.e., if k increases by one.
The algorithm examines the newly activated $k$-th row of the previous sequence elements to decide whether the element should be kept in the buffer.
This preserves the invariant that the buffer consists of exactly those sequence elements whose associated coin flip column starts with $log2(p-1)$ heads.
Of course, the new algorithm is not practical for actual implementation, but one can verify its correctness using standard Chernoff bounds, and on the other hand, it is possible to show that its behavior is equivalent to Algorithm 1.
To summarize, while the algorithm is marvelous, the proof was still very technical.
The simulation argument, in particular, is not so elegant to formalize.

\subparagraph*{A more direct proof:}
We set out to try to find a more direct proof, which also eases the formalization effort.
For the following discussion, we will analyze Algorithm 1 with line 8 removed, i.e., the algorithm does not output $\bot$, nor performs a second check of $|\chi|=n$.
Note that this happens if the very improbable event where none of the elements in X are removed during a subsampling step, which happens with probability at most $2^{-n/2}$.
Overall, the probability of it happening during the course of the algorithm is at most $\frac{\delta}{2}$. (Note that removing line 8 does not affect the correctness of the algorithm, but it loses its space consumption bound.)
It is easy to see that any probability established about the algorithm missing line 8 will be true for the original algorithm with a possible correction by, at most,  $\frac{\delta}{2}$.
We will remember and correct this at the end of this section.

Let us consider an imaginary situation where, somehow, $p$ is fixed at some point in the algorithm.
For example, we could imagine a final sub-sampling loop, which is run until a fixed $p$ is reached.
Then, the indicator random variables representing the presence of a stream element TODO

### 2025-01-14 paper/sec/negdep.tex
add initial paper stub in LIPIcs

\section{Negative Dependence}
\label{sec:negdep}

TODO

### 2025-01-14 paper/sec/intro.tex
minor edits

@@
 In 2022, Chakraborty, Vinodchandran, and Meel~\cite{DBLP:conf/esa/0001VM22} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community.
-Indeed, Knuth later published a note on the algorithm, pointing out its simplicity and also christening it the \emph{CVM} algorithm (which we will follow for the rest of this paper).
+Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its simplicity and christening it the \emph{CVM} algorithm (which we will use for the rest of this paper).
 One striking property of the algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
-Indeed, the algorithm could theoretically be implemented in a setting where the objects in the data stream only allow for equality comparisons.
-Another property is its simplicity, which is why the authors called it ``an algorithm for the text-book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
+Instead, the algorithm could theoretically be implemented in a setting where the objects in the data stream only allow for equality comparisons.
+Another property is its simplicity, which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
 
@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A randomized cardinality approximation $c$ for set $A = \{ a_1,\dots,a_l \}$ with $$\Pr_c \left( | c - |A| | > \varepsilon |A| \right) \leq \delta$$
+  \Ensure A cardinality estimate $c$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\Pr_c \left( | c - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, p \gets 1, n = \ceil{\frac{12}{\varepsilon^2} \ln{(\frac{81}{\delta})} }$
   \For{$i \gets 1$ to $l$}
-    \State {TODO}
+    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
+    \If{$b$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State $\chi \getsr \chi_{\downarrow \frac{1}{2}}$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
+      \State $p \gets \frac{p}{2}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \Return $\bot$ \Comment fail if $\chi$ remains too large
+    \EndIf
   \EndFor
-  \Return $\frac{|\chi|}{p}$
+  \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
   \end{algorithmic}
@@
 The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
-The algorithm iterates over stream elements, adding each one to the buffer with probability $p$, and conversely removing them with probability $(1-p)$.
+The algorithm iterates over stream elements, adding each one to the buffer with probability $p$ or conversely removing them with probability $(1-p)$.
 If the buffer gets too large, approximately half of the elements are removed by discarding each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.

### 2025-01-14 paper/sec/intro.tex
more minor tweaks

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{DBLP:conf/esa/0001VM22} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community.
-Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its simplicity and christening it the \emph{CVM} algorithm (which we will use for the rest of this paper).
-One striking property of the algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{DBLP:conf/esa/0001VM22} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
+Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its simplicity and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
+One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
 Instead, the algorithm could theoretically be implemented in a setting where the objects in the data stream only allow for equality comparisons.
-Another property is its simplicity, which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
+Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
 
@@
 
+%TODO: I think this should be later, we can give a forward reference
 The algorithm's state is a buffer $\chi$ and a fraction $p > 0$.
@@
 
-The pen-and-paper proof of correctness for~\cref{alg:cvm} is remarkably simple, requiring only undergraduate-level exposure to the concepts of randomized algorithms.
-We set out to formalize the proof as described in the literature~\cite{DBLP:conf/esa/0001VM22}, hoping that it would be an easy exercise for the formalized algorithms textbook.
+The pen-and-paper proof of correctness for~\cref{alg:cvm} is remarkably straightforward requiring only undergraduate-level exposure to the analysis of randomized algorithms.
+We set out to formalize the proof as described in the literature~\cite{DBLP:conf/esa/0001VM22}, hoping to show that it would also be an easy formalization exercise.
 

### 2025-01-15 isabelle/edit.sh
Add dummy doc files to fix building, tweak sess

@@ -1,10 +1,0 @@
-export CVC5_SOLVER="/usr/sbin/cvc5"
-export CVC5_VERSION="1.2.0"
-
-export Z3_SOLVER="/usr/sbin/z3"
-export Z3_VERSION="4.13.0"
-
-export ZIPPERPOSITION_HOME="/home/joe/.opam/5.2.0/bin"
-export ZIPPERPOSITION_VERSION="2.1"
-
-isabelle vscode -R CVM

### 2025-01-22 isabelle/CVM_Subsample/document/root.tex
WIP: Document preparation for CVM subsample.

\documentclass[11pt,a4paper]{article}
\usepackage{isabelle,isabellesym}
\usepackage{amsfonts,amsmath, amssymb}

\usepackage[only,bigsqcap]{stmaryrd}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}

\newcommand{\getsr}{\xleftarrow{\$}}

\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclareMathOperator{\prob}{\mathcal P}
\DeclareMathOperator{\expect}{\mathbb E}
\DeclareMathOperator{\Ber}{\mathrm{Ber}}
\DeclareMathOperator{\indicator}{\mathrm{I}}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

\begin{document}

\title{Verification of the CVM algorithm with a new Recursive Analysis Technique}
\author{Emin Karayel, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan,\\and Joe Watt}
\maketitle

\begin{abstract}
In 2022, Chakraborty et al.~\cite{chakraborty22} published a streaming algorithm for the distinct
elements problem, that deviated considerably from the state-of-the art, both due to its simplicity
and avoidance of standard derandomization techniques, while still maintaining a close to optimal
logarithmic space complexity.

In this entry, we verify its correctness using a new verification technique we discovered during the
effort, that simplifies the analysis considerably compared to the orignal proof by Chakraborty et
al. The main idea is based on the well-known notion of probabilistic recurrence relations. However,
instead of recurrence relations with respect to the state, we investigate general functions of the
state and derive conditions on the function, which preserve the reccurence relation.

The latter allows us to derive concentration bounds which is usually not possible with 
recurrence relations.

This new technique also opened up the possible design space, and we introduce a new variant of the 
CVM algorithm, that uses less space, fewer steps, and also has an additional property in addition 
to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
the desired result. The latter is also a new property, that classic algorithms for the distinct
elements problem do not posses.
\end{abstract}

\tableofcontents

\parindent 0pt\parskip 0.5ex

\input{session}

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

### 2025-01-23 isabelle/CVM_Subsample/document/root.tex
Document preparation: More minor refinements.

@@
 \usepackage{amsfonts,amsmath, amssymb}
+\usepackage[top=1.4in, bottom=1.4in, left=1.4in, right=1.4in]{geometry}
 
@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
+\let\oldsection\section
+\renewcommand\section{\clearpage\oldsection}
 
@@
 \begin{abstract}
-In 2022, Chakraborty et al.~\cite{chakraborty22} published a streaming algorithm for the distinct
+In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming algorithm for the distinct
 elements problem, that deviated considerably from the state-of-the art, both due to its simplicity

### 2025-01-24 isabelle/CVM_Subsample/document/root.tex
slightly edit abstract

@@
 \begin{abstract}
-In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming algorithm for the distinct
-elements problem, that deviated considerably from the state-of-the art, both due to its simplicity
-and avoidance of standard derandomization techniques, while still maintaining a close to optimal
-logarithmic space complexity.
+In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming
+algorithm (henceforth, the CVM algorithm) for the distinct elements problem
+that deviated considerably from the state-of-the art, due to its simplicity
+and avoidance of standard derandomization techniques, while still maintaining
+a close-to-optimal logarithmic space complexity.
 
-In this entry, we verify its correctness using a new verification technique we discovered during the
-effort, that simplifies the analysis considerably compared to the orignal proof by Chakraborty et
-al. The main idea is based on the well-known notion of probabilistic recurrence relations. However,
-instead of recurrence relations with respect to the state, we investigate general functions of the
-state and derive conditions on the function, which preserve the reccurence relation.
+In this entry, we verify the CVM algorithm's correctness using a new technique
+which simplifies the formal analysis considerably compared to the original
+proof by Chakraborty et al.. The main idea is based on the well-known notion of
+probabilistic recurrence relations. However, instead of recurrence relations
+with respect to the state, we investigate general functions of the state and
+derive conditions on the function which preserve the relevant recurrence
+relation. This generalization allows us to derive concentration bounds which is
+usually not possible with recurrence relations.
 
-The latter allows us to derive concentration bounds which is usually not possible with 
-recurrence relations.
-
-This new technique also opened up the possible design space, and we introduce a new variant of the 
-CVM algorithm, that uses less space, fewer steps, and also has an additional property in addition 
-to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
-the desired result. The latter is also a new property, that classic algorithms for the distinct
-elements problem do not posses.
+This new technique opens up the possible algorithm design space, and we
+introduce a new variant of the CVM algorithm that uses less space, fewer steps,
+and also has an additional property in addition to concentration: unbiasedness.
+The latter property means that the new variant returns the correct distinct
+element count in expectation, which is a property that neither the original CVM
+algorithm nor classic algorithms for the distinct elements problem posses.
+% TODO(YK): in what sense does it use less space and fewer steps?
 \end{abstract}

### 2025-01-24 isabelle/CVM_Subsample/document/appendix.tex
Added Appendix * Merge changes to root.tex

--- 2024-12-20 cvm_unbiased/main.tex
@@
-\documentclass{article}
-\usepackage[left=1.2in, right=1.2in]{geometry}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{mathtools}
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-\usepackage{algorithmicx}
-\usepackage{hyperref}
+\section{Informal Proof}
+This section includes an informal version of the proof for the tail bounds and unbiasedness
+of the abstract algorithm (Algorithm~\ref{alg:cvm_abs}) for interested readers.
 
-\newcommand{\prob}{\mathcal P}
-\newcommand{\expectation}{\mathrm{E}}
-\newcommand{\eps}{\varepsilon}
-\renewcommand{\phi}{\varphi}
-\newtheorem{theorem}{Theorem}
-\theoremstyle{definition}
-\newtheorem{lemma}{Lemma}
+This means we assume for the $\mathrm{subsample}(\chi)$ operation, that it fulfills 
+Eq.~\ref{eq:subsample_condition} and returns allways a subset of $\chi$.
 
-\begin{document}
-\title{Unbiased CVM Algorithm}
-\maketitle
-\section{Notation}
-For a finite set $S$, the probability space of uniformly sampling from the set is denoted by $U(S)$, i.e., for each $s \in S$ we have
-$\prob_{U(S)}({s}) = |S|^{-1}$.
-For the bernoulli probability space, over the set $\{0,1\}$ we'll write $\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{1\}) = p$.
-$I(P)$ is the indicator function for a predicate $P$, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
-\section{Algorithm}
-Let us fix
-\begin{itemize}
-  \item $a_1,\ldots,a_l$ for the stream of elements, and
-  \item $n$ for the maximal number of elements in the buffer, and
-  \item $f$ for the fraction of elements to keep in the buffer, when we subsample.
-\end{itemize}
-Note that: $nf$ must be an integer $\frac{1}{2} \leq f < 1$.
+\paragraph{Notation:} For the bernoulli probability space, over the set $\{0,1\}$ we'll write 
+$\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{1\}) = p$. $I(P)$ is the indicator function for a 
+predicate $P$, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
 
-\begin{algorithm}
-\caption{Unbiased CVM algorithm}
-\begin{algorithmic}[1]
-\State $\chi \gets \emptyset$; $p \gets 1$
-\For{$i \gets 1$ to $l$}
-  \State $\chi \gets \chi - \{a_i\}$
-  \State $\mathrm{coin} \gets \mathrm{Ber}(p)$
-  \If{$\mathrm{coin}$}
-    \State $\chi \gets \chi \cup \{a_i\}$
-  \EndIf
-  \If{$|\chi| = n$}
-    \State $\chi \gets U(\{ S \subseteq \chi \,|\, |S|=nf \})$
-    \State $p \gets p f$
-  \EndIf
-\EndFor
-\State \Return{$p^{-1} |\chi|$}
-\end{algorithmic}
-\end{algorithm}
+Like in the formalization, we will denote the first five lines of the loop ($3$--$7$) as step 1, 
+the last four lines ($8$--$10$) as step 2. For the distribution of the state of the algorithm after
+processing $i$ elements of the sequence, we will write $\Omega_i$.
+The elements of the probability spaces are pairs composed of a set and the number of subsampling, 
+steps representing $\chi$ and $p$.
 
-We will denote the first five lines of the loop ($4$--$7$) as step 1, the last four lines ($8$--$11$) as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
-The elements of the probability spaces are pairs composed of a set and a fraction, representing $\chi$ and $p$.
-For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, $\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state.
-We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, in particular, $\chi$ (resp. $p$) is the projection to the first (resp. second) component.
+For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, 
+$\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state.
+We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, 
+in particular, $\chi$ (resp. $p$) is the projection to the first (resp. second) component.
 
-The state of the algorithm after processing only step 1 of the $i$-th loop iteration is denoted by $\Omega'_i$.
-So the sequence of states is represented by the distributions $\Omega_0, \Omega'_1, \Omega_1, \ldots, \Omega'_l, \Omega_l$.
-A few easy-to-see observations without proof for any $0 \leq i \leq l$:
-\begin{itemize}
-\item $p(\omega) \leq 1$ a.s. (almost surely) for $\omega \in \Omega_i$,
-\item $\chi(\omega) \subseteq \{a_1,\ldots,a_i\}$ a.s. for $\omega \in \Omega_i$,
-\item $|\chi(\omega)| < n$ a.s. for $\omega \in \Omega_i$.
-\end{itemize}
-Also for the intermediate states, we have:
-\begin{itemize}
-\item $p(\omega) \leq 1$ a. s. for $\omega \in \Omega'_i$,
-\item $\chi(\omega) \subseteq \{a_1,\ldots,a_i\}$ a.s. for $\omega \in \Omega'_i$,
-\item $|\chi(\omega)| \leq n$ a.s. for $\omega \in \Omega'_i$.
-\end{itemize}
-Given a set $T$ with $n$ elements, we'll write $C(T)$ for the $nf$-subsets of $T$, i.e.:
-\[
-  C(T) = \{ \tau \subseteq T \, | \, |\tau| = nf \} \textrm{.}
-\]
-\begin{lemma}\label{le:samp_prod}
-Let $T$ be a set with $n$ elements, $S \subseteq T$ and $g: \{0,1\} \rightarrow \mathbb R_{\geq 0}$ then
-\[
-  \int_{U(C(T))} \prod_{s \in S} g(I(s \in \tau)) \leq \prod_{s \in S} \int_{\mathrm{Ber}(f)} g(\tau) d \tau
-\]
-\end{lemma}
-\begin{proof}
-We'll verify the cases $g(0) \leq g(1)$ and the converse seperately. For the case $g(0) \leq g(1)$:
-We want to use Theorem 2.11 by Joag-Dev and Proschan~\cite{joagdev1983}, which requires us to view $C(T)$ as a permutation distribution.
+Important to note is that the formalized version of the algorithm keeps track of the number of
+subsampling iterations $k$, e.g., the number of times the condition in line 8 in step 2 is true,
+instead of $p$ directly. This is, however, an equivalent way of representing the state, because 
+$p = f^k$.
 
-Let us for that consider $P := U(\{f : T \rightarrow \{0,..,|T|-1\}, f \textrm{bij.}\})$, i.e., we are assigning randomly distinct positions from $0$ to $|T|-1$ to the elements
-of $T$. Then $P$ is clearly a permuation distribution and the random variables describing the positions of each element are negatively associated according to Theorem 2.11.
+The state of the algorithm after processing only step 1 of the $i$-th loop iteration is denoted by 
+$\Omega'_i$. So the sequence of states is represented by the distributions 
+$\Omega_0, \Omega'_1, \Omega_1, \ldots, \Omega'_l, \Omega_l$.
 
-Now we can represent $C(T)$ as the choice of all the elements whose associated position is greater or equal to $(1-f)n$, i.e.:
-\begin{equation*}
-  \int_{U(C(T))} \prod_{s \in S} g(I(s \in \tau)) = \int_P \prod_{s \in S} g(I(\tau(s) \geq (1-f)n)) \, d\tau := R
-\end{equation*}
-Now we can use the Property P2~\cite[Page 288]{joagdev1983} since the $\tau(s)$ are negatively associated:
-\begin{equation}\label{eq:samp_prod_2}
-  R \leq \prod_{s \in S} \int_P g(I(\tau(s) \geq (1-f)n)) \, d\tau
-\end{equation}
-Note that this works because $x \rightarrow g(I(x \geq (1-f)n))$ is a monotone increasing function.
-Moreover the right hand side of Eq.~\ref{eq:samp_prod_2} is easy to be seen to be equal to the right hand side of the lemma we want to show.
-
-For the case $g(0) > g(1)$ we can do a similar trick, except in this case we associate with the choice of $\tau$ the elements with positions $\{0,\ldots,nf-1\}$.
-
-Important Note: In the paper by Joag-Dev and Proschan ``positive'' means non-negative. (See last paragraph of Section 1.)
-\end{proof}
+\subsection{Loop Invariant}
+After these preliminiaries, we can go to the main proof, whose core is an probabilistic loop
+invariant for Algorithm~\ref{alg:cvm_abs} that can be verified inductively.
 
@@
 \label{le:neg_cor_prelim}
-Let $\phi : (0,1] \times {0,1} \rightarrow \mathbb R_{\geq 0}$ be function, fulfilling the following conditions:
+Let $\phi : (0,1] \times {0,1} \rightarrow \mathbb R_{\geq 0}$ be function, fulfilling the following
+  conditions:
 \begin{enumerate}
-\item \label{cond:phi_1} $(1-\alpha) \phi(x,0) + \alpha \phi(x,1) \leq \phi(x/\alpha,1)$ for all $0 < \alpha < 1$, $0 < x \leq 1$, and
+\item \label{cond:phi_1} $(1-\alpha) \phi(x,0) + \alpha \phi(x,1) \leq \phi(x/\alpha,1)$ for all
+  $0 < \alpha < 1$, $0 < x \leq 1$, and
 \item \label{cond:phi_2} $\phi(x,0) \leq \phi(y,0)$ for all $0 < x < y \leq 1$.
@@
 \[
-  \expectation_{\Omega}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} 
+  \expect_{\Omega}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} 
 \]
@@
 \begin{proof}
-We show the result using induction over $k$. Note that we show the statement for arbitrary $S$, i.e., the induction statements are:
+We show the result using induction over $k$. Note that we show the statement for arbitrary $S$,
+i.e., the induction statements are:
 \begin{eqnarray*}
-P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} \right) \\
-Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; \expectation_{\Omega'_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] \leq \phi(1,1)^{|S|} \right)
+P(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \;
+  \expect_{\Omega_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
+    \leq \phi(1,1)^{|S|} \right) \\
+Q(k) & :\leftrightarrow & \left(\forall S \subseteq \{a_1,..,a_k\}. \; 
+  \expect_{\Omega'_k}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right]
+    \leq \phi(1,1)^{|S|} \right)
 \end{eqnarray*}
@@
 \[
-\expectation_{\Omega_0}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] = \expectation_{\Omega_0}\left[1\right] = 1 \leq \phi(1,1)^0 \textrm{.}
+\expect_{\Omega_0}\left[ \prod_{s \in S} \phi(p, I(s \in \chi)) \right] =
+\expect_{\Omega_0}\left[1\right] = 1 \leq \phi(1,1)^0 \textrm{.}
 \]
@@
 Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$ and define $S' := S - \{ a_{k+1} \}$.
-Note that $\Omega'_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where $a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable over the space $\Omega_k$.
+Note that $\Omega'_{k+1}$ can be constructed from $\Omega_k$ as a compound distribution, where
+$a_{k+1}$ is included in the buffer, with the probability $p$, which is itself a random variable
+over the space $\Omega_k$.
 
@@
 \[
-\prob_{\Omega'_{k+1}}( P(\chi, p) ) = \int_{\Omega_k} \int_{\textrm{Ber}(p(\omega))} P(\chi(\omega)-\{a_{k+1}\}\cup f(\tau), p(\omega)) \, d \tau \, d \omega
+\prob_{\Omega'_{k+1}}( P(\chi, p) ) = \int_{\Omega_k} \int_{\textrm{Ber}(p(\omega))} 
+P(\chi(\omega)-\{a_{k+1}\}\cup f(\tau), p(\omega)) \, d \tau \, d \omega
 \]
@@
 \begin{eqnarray*}
-  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] & = & \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right)
+  & & \textstyle \expect_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] \\
+ & = & \textstyle \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right)
   \int_{\mathrm{Ber}(p(\omega))} \phi(p, \tau) \, d \tau \, d \omega \\
-  & = & \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right) ((1-p) \phi(p,0) + p \phi(p,1)) \, d \omega\\
-  & \underset{\textrm{Cond~\ref{cond:phi_1}}}{\leq} & \int_{\Omega_k} \left( \prod_{s \in S'} \phi(p, I(s \in \chi)) \right)
-  \phi(1,1) \, d \omega \\
-  & \underset{\textrm{IH}}{\leq} & \phi(1,1)^{|S'|} \phi(1,1) = \phi(1,1)^{|S|}
+  & = & \textstyle \int_{\Omega_k} \left(\prod_{s \in S'} \phi(p, I(s \in \chi))\right) ((1-p) \phi(p,0) + 
+  p \phi(p,1)) \, d \omega\\
+  & \underset{\textrm{Cond~\ref{cond:phi_1}}}{\leq} & \textstyle \int_{\Omega_k} \left( \prod_{s \in S'} 
+  \phi(p, I(s \in \chi)) \right) \phi(1,1) \, d \omega \\
+  & \underset{\textrm{IH}}{\leq} & \textstyle \phi(1,1)^{|S'|} \phi(1,1) = \phi(1,1)^{|S|}
 \end{eqnarray*}
@@
 If $a_{k+1} \notin S$ then $S' = S$ and:
-\begin{equation}
-  \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] = \int_{\Omega_k} \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \leq_{IH} \phi(1,1)^{|S'|} = \phi(1,1)^{|S|}
-\end{equation}
+\[
+  \textstyle \expect_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] =
+  \int_{\Omega_k} \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \leq_{IH} \phi(1,1)^{|S'|} =
+  \phi(1,1)^{|S|}
+\]
 \paragraph{Induction step $Q(k+1) \rightarrow P(k+1)$:} \phantom{.}\\
@@
 
-Let us again note that $\Omega_{k+1}$ is a compound distribution over $\Omega_k$. In general, for all predicates $P$:
-\[
-  \prob_{\Omega_{k+1}}( P(\chi, p) )  = \int_{\Omega'_{k+1}} I( |\chi(\omega)|<n ) P(\chi(\omega),p(\omega)) + I( |\chi(\omega)|=n ) \int_{U(\chi(\omega))} P(\tau, f p(\omega)) \, d \tau  \, d \omega \textrm{.}
-\]
+Let us again note that $\Omega_{k+1}$ is a compound distribution over $\Omega_k$. In general, for 
+all predicates $P$:
+\begin{align*}
+  & \prob_{\Omega_{k+1}}( P(\chi, p) ) = \\
+  & \int_{\Omega'_{k+1}} I( |\chi(\omega)|<n ) P(\chi(\omega),p(\omega)) +
+  I( |\chi(\omega)|=n ) \int_{\mathrm{subsample}(\chi(\omega))}
+  P(\tau, f p(\omega)) \, d \tau \, d \omega \textrm{.}
+\end{align*}
 
@@
 \begin{eqnarray*}
-&  & \textstyle\expectation_{\Omega_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi))\right]  \\
-& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(pf,0) \int_{U(C(\chi))} \prod_{s \in S \cap \chi} \phi(pf,I(s \in \tau )) d \tau \, d \omega  \\
-& \underset{\textrm{Le.~\ref{le:samp_prod}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(pf,0) \prod_{s \in S \cap \chi} \int_{\mathrm{Ber(f)}} \phi(pf,\tau) d \tau \, d \omega  \\
-& \underset{\textrm{Cond~\ref{cond:phi_2}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(p,0) \prod_{s \in S \cap \chi} ((1-f)\phi(pf,0)+f\phi(pf,1)) \, d \omega  \\
-& \underset{\textrm{Cond~\ref{cond:phi_1}}}{\leq} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} \phi(p,0) \prod_{s \in S \cap \chi} \phi(p,1) \, d \omega \\
-& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
-& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega  \\
-& = & \textstyle \expectation_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] \underset{\textrm{IH}}{\leq} \phi(1,1)^{|S|}
+&  & \textstyle\expect_{\Omega_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi))\right]  \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi|<n)\prod_{s\in S} \phi(p, I(s \in \chi)) \, d \omega\\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi|=n)
+  \prod_{s \in S \setminus \chi(\omega)} \phi(pf,0) 
+    \int_{\mathrm{subsample}(\chi)} \prod_{s \in S \cap \chi} \phi(pf,I(s \in \tau )) d \tau \, d \omega \\
+& \leq \marginnote{\footnotesize Eq.~\ref{eq:subsample_condition}} & \textstyle \int_{\Omega'_{k+1}} 
+  I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)}
+  \phi(pf,0) \prod_{s \in S \cap \chi} \int_{\mathrm{Ber(f)}} \phi(pf,\tau) d \tau \, d \omega  \\
+& \leq \marginnote{\footnotesize Cond~\ref{cond:phi_2}}  &
+  \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n)\prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} 
+  \phi(p,0) \prod_{s \in S \cap \chi} ((1-f)\phi(pf,0)+f\phi(pf,1)) \, d \omega  \\
+& \leq \marginnote{\footnotesize Cond~\ref{cond:phi_1}} & \textstyle \int_{\Omega'_{k+1}} I(|\chi| < n) 
+  \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi| = n) \prod_{s \in S \setminus \chi(\omega)} 
+  \phi(p,0) \prod_{s \in S \cap \chi} \phi(p,1) \, d \omega \\
+& = & \textstyle \int_{\Omega'_{k+1}} I(|\chi|<n)\prod_{s \in S} \phi(p,I(s\in\chi)) \, d\omega \\
+& + & \textstyle \int_{\Omega'_{k+1}} I(|\chi|=n)\prod_{s \in S} \phi(p,I(s\in\chi)) \, d\omega \\
+& = & \textstyle \expect_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] 
+  \leq  \marginnote{\footnotesize IH} \phi(1,1)^{|S|}
 \end{eqnarray*}
-
 \end{proof}
 
-More practical versions of the previous lemma:
+A corrolary and more practical version of the previous lemma is:
 \begin{lemma}\label{le:neg_cor_neg}
@@
 \[
-  \expectation_{\Omega}\left[ \prod_{s \in S} I(p > q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
+  \expect_{\Omega}\left[ \prod_{s \in S} I(p>q) h(p^{-1} I(s \in \chi)) \right] \leq h(1)^{|S|} 
 \]
@@
 \begin{proof}
-Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just need to check Conditions 1/2, Indeed
+Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
+need to check Conditions 1/2, Indeed
 \begin{eqnarray*}
-  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x > q) h(x^{-1}) \\
+  (1-\alpha) \phi(x,0) + \alpha \phi(x,1) & = & (1-\alpha) I(x>q) h(0) + \alpha I(x>q) h(x^{-1}) \\
    & \leq & I(x > q) h(\alpha x^{-1}) \leq I(x > q \alpha) h(\alpha x^{-1}) = \phi(x/\alpha,1)
@@
 
-\section{Concentration}
-This section establishes that the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
-This will be done by Chernoff bounds for the probability that the estimate is above $(1+\eps)|A|$ (resp.
-below $(1-\eps)|A|$ assuming $p$ is not too small and a tail estimate for $p$ being too small.
+It should be noted that this is a probabilistic reccurence relation, but the main innovation is that
+we establish a relation, with respect to general classes of functions of the state variables.
 
-It should be noted that concentration is trivial, if $|A| < n$, i.e., if we never need to do sub-sampling.
-So we assume $|A| \geq n$.
+\subsection{Concentration}
+Let us now see, how we can obtain concentration bounds using the latter, i.e., that the result of 
+the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
+This will be done using the Cram\'{e}r-Chernoff method for the probability that the estimate is 
+above $(1+\eps)|A|$ (resp. below $(1-\eps)|A|$ assuming $p$ is not too small and a tail estimate 
+for $p$ being too small.
+
+It should be noted that concentration is trivial, if $|A| < n$, i.e., if we never need to do
+sub-sampling. So we assume $|A| \geq n$.
 
@@
 \label{le:upper_tail}
-For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and $0 < \eps \leq 1$:
+For any $\Omega \in \{\Omega_0,\ldots,\Omega_l\} \cup \{\Omega'_1,\ldots,\Omega'_l\}$ and
+$0 < \eps \leq 1$:
 \[
-L := \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq \exp\left(-\frac{n}{12} \eps^2\right)
+L := \prob_{\Omega} \left( p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q\right) \leq
+  \exp\left(-\frac{n}{12} \eps^2\right)
 \]
@@
 \begin{proof}
-By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$. Let $A' = A \cap \{a_1,\ldots,a_k\}$.
+By assumption there exists a $k$ such that $\Omega \in \{\Omega_k, \Omega'_k\}$.
+Let $A' = A \cap \{a_1,\ldots,a_k\}$.
 
@@
 \begin{eqnarray*}
-  L & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
-  & \leq & \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
-  & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
- & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
- & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
+  L & \underset{t > 0}{\leq} & \prob_{\Omega} \left( \exp(t p^{-1} |\chi|) \geq
+    \exp(t (1+\eps) |A|) \wedge p \geq q \right) \\
+  & \leq &
+    \prob_{\Omega} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1+\eps) |A|) \right) \\
+  & \underset{\textrm{Markov}}{\leq} &
+    \exp( -t (1+\eps) |A| ) \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
+ & \leq & \exp( -t (1+\eps) |A| ) 
+  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
+ & \leq & \exp( -t (1+\eps) |A| ) 
+  \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
  & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
@@
 \[
-  f(\eps) := -\ln(1+q \eps) + q \ln(1+\eps) (1+\eps) - \frac{q \eps^2}{3} = -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3} \textrm{.}
+  f(\eps) := -\ln(1+q \eps) + q \ln(1+\eps) (1+\eps) - \frac{q \eps^2}{3} =
+    -\ln (h(1)) +t (1+\eps) - \frac{q\eps^2}{3} \textrm{.}
 \]
-For which it is easy to check $f(0) = 0$ and the derivative with respect to $\eps$ is non-negative in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
-$f(\eps) \geq 0$.
+For which it is easy to check $f(0) = 0$ and the derivative with respect to $\eps$ is non-negative
+in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e., $f(\eps) \geq 0$.
 \end{proof}
@@
 \end{equation}
-First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
-It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$. Then
+First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the
+algorithm, e.g., after processing $k$ stream elements. It can only happen if $|\chi|=n$, $p=f^j$
+in $\Omega'_k$. Then
 \begin{eqnarray*}
-  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq & \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
-    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( p^{-1} |\chi| \geq 2|A| \wedge p \geq q ) \\
+  \prob_{\Omega'_k} ( |\chi| \geq n \wedge p=f^j) & \leq &
+    \prob( p^{-1} |\chi| \geq f^{-j} n \wedge p \geq q) \\
+    & \underset{\textrm{Eq.~\ref{eq:f_j}}}{\leq} & \prob( p^{-1} |\chi| \geq 2|A| \wedge p \geq q)\\
     & \underset{\textrm{Le.~\ref{le:upper_tail}}}{\leq} & \exp(- n/12)
 \end{eqnarray*}
-The probability that this happens ever in the entire process is then at most $l$ times the above which proves the lemma.
+The probability that this happens ever in the entire process is then at most $l$ times the above
+which proves the lemma.
 \end{proof}
@@
 \[
-  L := \prob_{\Omega_l} ( p^{-1} |\chi| \leq (1-\eps)|A| \wedge p \geq q) \leq \exp\left(-\frac{n}{8} \eps^2\right)
+  L := \prob_{\Omega_l} ( p^{-1} |\chi| \leq (1-\eps)|A| \wedge p \geq q) \leq
+    \exp\left(-\frac{n}{8} \eps^2\right)
 \]
@@
 
-Note that $h(x) \geq 0$ for $0 \leq x \leq q^{-1}$ (can be checked by verifying it is true for $h(0)$ and $h(q^{-1})$ and the fact that the function is affine.)
+Note that $h(x) \geq 0$ for $0 \leq x \leq q^{-1}$ (can be checked by verifying it is true for
+$h(0)$ and $h(q^{-1})$ and the fact that the function is affine.)
 
@@
 \begin{eqnarray*}
-  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq \exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
-    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq \exp(t (1-\eps)|A|) \wedge p > q \right) \\
-    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
-    & = & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
-    & \leq & \exp( -t (1-\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
+  L & \underset{t<0}{=} & \prob_{\Omega_l} \left( \exp(t p^{-1} |\chi|) \geq
+    \exp(t (1-\eps)|A|) \wedge p \geq q \right) \\
+    & \leq & \prob_{\Omega_l} \left( I(p \geq q) \exp(t p^{-1} |\chi|) \geq
+    \exp(t (1-\eps)|A|) \wedge p > q \right) \\
+    & \underset{\textrm{Markov}}{\leq} & \exp( -t (1-\eps) |A| )
+    \expect_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
+    & = & \exp( -t (1-\eps) |A| ) \expect_\Omega \left[ \prod_{s \in A} I(p \geq q) 
+      \exp ( t p^{-1} I(s \in \chi) ) \right] \\
+    & \leq & \exp( -t (1-\eps) |A| ) \expect_\Omega 
+      \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
     & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
@@
 \[
-  f(\eps) := -q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
+  f(\eps) := q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 =
+  t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
 \]
 is non-negative for $0 \leq q \leq \frac{1}{4}$ and $0 < \eps < 1$.
-This can be verified by checking that $f(0) = 0$ and that the derivative with respect to $\eps$ is non-negative.
+This can be verified by checking that $f(0) = 0$ and that the derivative with respect to $\eps$ is
+non-negative.
 \end{proof}
 We can now establish the concentration result:
-\begin{theorem}
-Let $0 < \eps < 1$ and $0 < \delta < 1$ and $n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
+\begin{theorem}\label{th:concentration}
+Let $0 < \eps < 1$ and $0 < \delta < 1$ and
+$n \geq \frac{12}{\eps^2} \ln\left(\frac{3l}{\delta}\right)$ then:
 \[
@@
 \begin{eqnarray*}
-  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right) +
+  L & \leq & \prob_{\Omega_l} \left( | p^{-1} |\chi| \leq (1-\eps) |A| \wedge p \geq q \right)\\
+  & + &
     \prob_{\Omega_l} \left( | p^{-1} |\chi| \geq (1+\eps) |A| \wedge p \geq q \right) +
@@
     & \underset{Le.~\ref{le:upper_tail}-\ref{le:lower_tail}}{\leq} &
-    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + l \exp\left( - \frac{n}{12} \right) \\
+    \exp\left( - \frac{n}{8} \eps^2 \right)  + \exp\left( - \frac{n}{12} \eps^2 \right) + 
+    l \exp\left( - \frac{n}{12} \right) \\
     & \leq & \frac{\delta}{3} + \frac{\delta}{3} + \frac{\delta}{3}
@@
 \end{proof}
-
-\section{Unbiasedness}
-Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$). Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that:
+\subsection{Unbiasedness\label{sec:unbiasedness}}
+Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$).
+Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that:
 \begin{eqnarray*}
-  \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in S) ] \leq 1 \\
-  \expectation_{\Omega_l} [ M + 1 - p^{-1} I(s \in S) ] & = & \expectation_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in S)) ] \leq M  
+  \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] & = &
+    \expect_{\Omega_l} [ I(p \geq M^{-1}) p^{-1} I(s \in \chi) ] \leq 1 \\
+  \expect_{\Omega_l} [ M + 1 - p^{-1} I(s \in \chi) ] & = &
+    \expect_{\Omega_l} [ I(p \geq M^{-1}) (M + 1 - p^{-1} I(s \in \chi)) ] \leq M
 \end{eqnarray*}
-which implies $\expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = 1$.
+which implies $\expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = 1$.
 By linearity of expectation we can conclude that
 \[
-  \expectation_{\Omega_l} [ p^{-1} |\chi| ] = \sum_{s \in A} \expectation_{\Omega_l} [ p^{-1} I(s \in S) ] = |A| \textrm{.}
+  \expect_{\Omega_l} [ p^{-1} |\chi| ] =
+    \sum_{s \in A} \expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = |A| \textrm{.}
 \]
-
-\bibliographystyle{plainurl}
-\bibliography{main}
-
-\end{document}

### 2025-01-24 isabelle/CVM_Subsample/document/root.tex
Added Appendix * Merge changes to root.tex

--- 2025-01-23 isabelle/CVM_Subsample/document/root.tex
@@
 \usepackage{isabelle,isabellesym}
-\usepackage{amsfonts,amsmath, amssymb}
+\usepackage{amsfonts,amsmath, amssymb,amsthm}
 \usepackage[top=1.4in, bottom=1.4in, left=1.4in, right=1.4in]{geometry}
@@
 
+\usepackage{marginnote}
 \usepackage{algorithm}
@@
 
+\newcommand{\eps}{\varepsilon}
+\renewcommand{\phi}{\varphi}
+\newtheorem{theorem}{Theorem}
+\theoremstyle{definition}
+\newtheorem{lemma}{Lemma}
+
 % this should be the last package used
@@
 \begin{abstract}
-In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming algorithm for the distinct
-elements problem, that deviated considerably from the state-of-the art, both due to its simplicity
+In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming 
+algorithm (henceforth, the CVM algorithm) for the distinct
+elements problem, that deviated considerably from the state-of-the art, due to its simplicity
 and avoidance of standard derandomization techniques, while still maintaining a close to optimal
@@
 
-In this entry, we verify its correctness using a new verification technique we discovered during the
-effort, that simplifies the analysis considerably compared to the orignal proof by Chakraborty et
-al. The main idea is based on the well-known notion of probabilistic recurrence relations. However,
-instead of recurrence relations with respect to the state, we investigate general functions of the
-state and derive conditions on the function, which preserve the reccurence relation.
+In this entry, we verify the CVM algorithm's correctness using a new technique which simplifies
+the formal analysis considerably compared to the orignal proof by Chakraborty et
+al. The main idea is based on a probabilistic invariant that allows us to derive concentration using
+Cram\'{e}r Chernoff method.
 
-The latter allows us to derive concentration bounds which is usually not possible with 
-recurrence relations.
-
-This new technique also opened up the possible design space, and we introduce a new variant of the 
-CVM algorithm, that uses less space, fewer steps, and also has an additional property in addition 
+This new technique opens up the possible algorithm design space, and we introduce a new variant of the 
+CVM algorithm, that is total, and also has an additional property in addition 
 to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
-the desired result. The latter is also a new property, that classic algorithms for the distinct
-elements problem do not posses.
+the desired result. The latter is also a new property, that neither the original CVM algorithm
+nor classic algorithms for the distinct elements problem posses.
 \end{abstract}
@@
 
+\appendix
+\input{appendix}
+
 \end{document}

### 2025-01-25 isabelle/CVM_Subsample/document/root.tex
Update root.tex
fix minor typo

@@
 the formal analysis considerably compared to the orignal proof by Chakraborty et
-al. The main idea is based on a probabilistic invariant that allows us to derive concentration using
-Cram\'{e}r Chernoff method.
+al. The main idea is based on a probabilistic invariant that allows us to derive concentration bounds
+using the Cram\'{e}r-Chernoff method.
 
@@
 the desired result. The latter is also a new property, that neither the original CVM algorithm
-nor classic algorithms for the distinct elements problem posses.
+nor classic algorithms for the distinct elements problem possess.
 \end{abstract}

### 2025-01-25 isabelle/CVM_Subsample/document/appendix.tex
Merge Utils_Basic/PMF.   + Minor improvments.

@@
 
-\paragraph{Notation:} For the bernoulli probability space, over the set $\{0,1\}$ we'll write 
-$\mathrm{Ber}(p)$, i.e., $P_{\mathrm{Ber}(p)}(\{1\}) = p$. $I(P)$ is the indicator function for a 
-predicate $P$, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
+\paragraph{Notation:} For a finite set $S$, the probability space of uniformly sampling from the set
+is denoted by $U(S)$, i.e., for each $s \in S$ we have $\prob_{U(S)}({s}) = |S|^{-1}$.
+For the bernoulli probability space, over the set $\{0,1\}$ we'll write $\mathrm{Ber}(p)$, i.e.,
+$P_{\mathrm{Ber}(p)}(\{1\}) = p$. $I(P)$ is the indicator function for a predicate $P$, i.e.,
+$I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.
 
@@
 \subsection{Loop Invariant}
-After these preliminiaries, we can go to the main proof, whose core is an probabilistic loop
+After these preliminiaries, we can go to the main proof, whose core is a probabilistic loop
 invariant for Algorithm~\ref{alg:cvm_abs} that can be verified inductively.

### 2025-01-25 isabelle/CVM_Subsample/document/root.tex
Merge Utils_Basic/PMF.   + Minor improvments.
Update root.tex
some very minor edits (minus appendix)

@@
 
-\title{Verification of the CVM algorithm with a new Recursive Analysis Technique}
+\title{Verification of the CVM algorithm with a New Recursive Analysis Technique}
 \author{Emin Karayel, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan,\\and Joe Watt}
@@
 al. The main idea is based on a probabilistic invariant that allows us to derive concentration bounds
-using the Cram\'{e}r-Chernoff method.
+using the Cram\'{e}r--Chernoff method.
 

### 2025-01-25 isabelle/CVM_Subsample/document/appendix.tex
minor fixes in appendix

@@
 
-This means we assume for the $\mathrm{subsample}(\chi)$ operation, that it fulfills 
-Eq.~\ref{eq:subsample_condition} and returns allways a subset of $\chi$.
+This means we assume the $\mathrm{subsample}(\chi)$ operation fulfills
+Eq.~\ref{eq:subsample_condition} and always returns a subset of $\chi$.
 
@@
 is denoted by $U(S)$, i.e., for each $s \in S$ we have $\prob_{U(S)}({s}) = |S|^{-1}$.
-For the bernoulli probability space, over the set $\{0,1\}$ we'll write $\mathrm{Ber}(p)$, i.e.,
+We write $\mathrm{Ber}(p)$ for the Bernoulli probability space, over the set $\{0,1\}$, i.e.,
 $P_{\mathrm{Ber}(p)}(\{1\}) = p$. $I(P)$ is the indicator function for a predicate $P$, i.e.,
@@
 processing $i$ elements of the sequence, we will write $\Omega_i$.
-The elements of the probability spaces are pairs composed of a set and the number of subsampling, 
-steps representing $\chi$ and $p$.
+The elements of the probability spaces are pairs composed of a set and the number of subsampling
+steps, representing $\chi$ and $p$ respectively.
 
 For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, 
-$\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state.
+$\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc., and $\Omega_l$ denotes the final state.
 We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, 
@@
 
-Important to note is that the formalized version of the algorithm keeps track of the number of
+Note that the formalized version of the algorithm keeps track of the number of
 subsampling iterations $k$, e.g., the number of times the condition in line 8 in step 2 is true,
@@
 \subsection{Loop Invariant}
-After these preliminiaries, we can go to the main proof, whose core is a probabilistic loop
+After these preliminaries, we can go to the main proof, whose core is a probabilistic loop
 invariant for Algorithm~\ref{alg:cvm_abs} that can be verified inductively.
@@
 \label{le:neg_cor_prelim}
-Let $\phi : (0,1] \times {0,1} \rightarrow \mathbb R_{\geq 0}$ be function, fulfilling the following
+Let $\phi : (0,1] \times \{0,1\} \rightarrow \mathbb R_{\geq 0}$ be a function, fulfilling the following
   conditions:
@@
   \textstyle \expect_{\Omega'_{k+1}}\left[\prod_{s \in S} \phi(p, I(s \in \chi)) \right] =
-  \int_{\Omega_k} \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \leq_{IH} \phi(1,1)^{|S'|} =
+  \int_{\Omega_k} \prod_{s \in S} \phi(p, I(s \in \chi)) \, d \omega \leq_{\textrm{IH}} \phi(1,1)^{|S'|} =
   \phi(1,1)^{|S|}
@@
 
-Let us again note that $\Omega_{k+1}$ is a compound distribution over $\Omega_k$. In general, for 
+Let us again note that $\Omega_{k+1}$ is a compound distribution over $\Omega'_{k+1}$. In general, for 
 all predicates $P$:
@@
 
-A corrolary and more practical version of the previous lemma is:
+A corollary and more practical version of the previous lemma is:
 \begin{lemma}\label{le:neg_cor_neg}
@@
 Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
-need to check Conditions 1/2, Indeed
+need to check Conditions 1 and 2. Indeed,
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$. And
+for $0 < \alpha < 1$ and $0< x \leq 1$, where we used that $x > q$ implies $x > q \alpha$; and
 \[
@@
 
-It should be noted that this is a probabilistic reccurence relation, but the main innovation is that
+It should be noted that this is a probabilistic recurrence relation, but the main innovation is that
 we establish a relation, with respect to general classes of functions of the state variables.
@@
 \subsection{Concentration}
-Let us now see, how we can obtain concentration bounds using the latter, i.e., that the result of 
-the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
-This will be done using the Cram\'{e}r-Chernoff method for the probability that the estimate is 
-above $(1+\eps)|A|$ (resp. below $(1-\eps)|A|$ assuming $p$ is not too small and a tail estimate 
+Let us now see how we can obtain concentration bounds using Lemma~\ref{le:neg_cor_neg}, i.e., that
+the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
+This will be done using the Cram\'{e}r--Chernoff method for the probability that the estimate is 
+above $(1+\eps)|A|$ (resp.~below $(1-\eps)|A|$) assuming $p$ is not too small and a tail estimate 
 for $p$ being too small.
@@
 It should be noted that concentration is trivial, if $|A| < n$, i.e., if we never need to do
-sub-sampling. So we assume $|A| \geq n$.
+sub-sampling, so we assume $|A| \geq n$.
 
-We define $q := n/(4|A|)$ - notice that $q \leq \frac{1}{4}$.
+Define $q := n/(4|A|)$ and notice that $q \leq \frac{1}{4}$.
 
@@
 Let $A' = A \cap \{a_1,\ldots,a_k\}$.
-
-Moreover, we define
+Moreover, we define:
 \begin{align*}
@@
 
-To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
+To get a tail estimate, we use the Cram\'{e}r--Chernoff method:
 {\allowdisplaybreaks
@@
 
-Using the previous we can estimate bounds for $p$ becoming too small:
+Using the previous result we can also estimate bounds for $p$ becoming too small:
 \begin{lemma}\label{le:low_p}
@@
 \begin{proof}
-We'll use a similar strategy as in the $\mathrm{Bad}_2$ bound in the original CVM paper.
+We will use a similar strategy as in the $\mathrm{Bad}_2$ bound from the original CVM paper~\cite{chakraborty2022}.
 Let $j$ be maximal, s.t., $q \leq f^j$. Hence $f^{j+1} < q$ and:
@@
 First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the
-algorithm, e.g., after processing $k$ stream elements. It can only happen if $|\chi|=n$, $p=f^j$
+algorithm, e.g., while processing $k$ stream elements. It can only happen if $|\chi|=n$, $p=f^j$
 in $\Omega'_k$. Then
@@
 
-With these definitions we again follow the Cram\'{e}r-Chernoff method:
+With these definitions we again follow the Cram\'{e}r--Chernoff method:
 {\allowdisplaybreaks
@@
 \subsection{Unbiasedness\label{sec:unbiasedness}}
-Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g. we can choose $M = f^{-l}$).
-Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that:
+Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g., we can choose $M = f^{-l}$).
+Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that
+for all $s \in A$:
 \begin{eqnarray*}
@@
 which implies $\expect_{\Omega_l} [ p^{-1} I(s \in \chi) ] = 1$.
-By linearity of expectation we can conclude that
+By linearity of expectation we conclude
 \[

### 2025-01-25 isabelle/CVM_Subsample/document/appendix.tex
Also verify that state_chi allways contains below thresh elements.  + Fix minor typos.

@@
-\section{Informal Proof}
+\section{Informal Proof\label{apx:informal_proof}}
 This section includes an informal version of the proof for the tail bounds and unbiasedness
@@
 \begin{lemma}
-\label{le:neg_cor_prelim}
+\label{le:prob_invariant}
 Let $\phi : (0,1] \times \{0,1\} \rightarrow \mathbb R_{\geq 0}$ be a function, fulfilling the following
@@
 A corollary and more practical version of the previous lemma is:
-\begin{lemma}\label{le:neg_cor_neg}
+\begin{lemma}\label{le:prob_invariant_simple}
 Let $q \leq 1$ and $h : \mathbb [0,q^{-1}] \rightarrow \mathbb R_{\geq 0}$ be concave then for all 
@@
 \begin{proof}
-Follows from Lemma~\ref{le:neg_cor_prelim} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
+Follows from Lemma~\ref{le:prob_invariant} for $\phi(p,\tau) := I(p > q) h(\tau p^{-1})$. We just
 need to check Conditions 1 and 2. Indeed,
@@
 \subsection{Concentration}
-Let us now see how we can obtain concentration bounds using Lemma~\ref{le:neg_cor_neg}, i.e., that
+Let us now see how we can obtain concentration bounds using Lemma~\ref{le:prob_invariant_simple}, i.e., that
 the result of the algorithm is concentrated around the cardinality of $A = \{ a_1, \ldots, a_l \}$.
@@
   \expect_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
- & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
+ & \underset{\textrm{Le.~\ref{le:prob_invariant_simple}}}{\leq} & \exp( -t (1+\eps) |A| ) h(1)^{|A'|} \\
  & \underset{h(1)\geq 1}{\leq} & \left(\exp( \ln(h(1)) -t (1+\eps))\right) ^ {|A|} 
@@
       \left[ \prod_{s \in A} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
-    & \underset{\textrm{Le.~\ref{le:neg_cor_neg}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
+    & \underset{\textrm{Le.~\ref{le:prob_invariant_simple}}}{\leq} & \exp( -t (1-\eps) |A| ) (h(1))^{|A|} \\
     & = & \exp( \ln(h(1)) -t (1-\eps) )^{|A|}
@@
 Let $M$ be large enough such that $p^{-1} \leq M$ a.s. (e.g., we can choose $M = f^{-l}$).
-Then we can derive from Lemma~\ref{le:neg_cor_neg} using $h(x) = x$ and $h(x) = M+1-x$ that
+Then we can derive from Lemma~\ref{le:prob_invariant_simple} using $h(x) = x$ and $h(x) = M+1-x$ that
 for all $s \in A$:

### 2025-01-29 paper/main.tex
minor edits
more minor tweaks
WIP: Paper   * Saving current state

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}
 %This is a template for producing LIPIcs articles.
@@
 
+\usepackage{bigints}
 \usepackage{algorithm}
@@
 
+\newcommand{\getsr}{\xleftarrow{\$}}
+\newcommand{\todo}[1]{\textcolor{red}{#1}}
+\newcommand{\ift}[3]{\mathrm{if} \; #1 \; \mathrm{then} \; #2 \; \mathrm{else} \; #3}
+
 \usepackage{mathtools}
 \DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
+\DeclareMathOperator{\Ber}{\mathrm{Ber}}
+\DeclareMathOperator{\prob}{\mathcal P}
+\DeclareMathOperator{\expect}{\mathrm{E}}
+\DeclareMathOperator{\bigo}{\mathcal O}
+%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
 
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+\usepackage{isabelle}
+\usepackage{isabellesym}
+\isabellestyle{it}
 
@@
 
-\title{Formalization of CVM} %TODO Please add
+\title{Verification of the CVM algorithm with a New Recursive Analysis Technique} %TODO Please add
 
@@
 Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
+
+\author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+
+\author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+
+\author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
 
@@
 
-\acknowledgements{I want to thank \dots}%optional
+%\acknowledgements{I want to thank \dots}%optional
 
@@
 
+\input{sec/semra}
+
+\input{sec/invariants}
+
 \input{sec/negdep}
 
-\input{sec/cvmalt}
+\input{sec/formalization}
+
+\input{sec/original_proof}
 

### 2025-01-29 paper/sec/formalization.tex
WIP: Paper   * Saving current state

\section{Formalization}
\todo{Overview of the two AFP entries. Include main statement of correctness for the CVM algorithm.}

### 2025-01-29 paper/sec/intro.tex
WIP: Paper   * Saving current state

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{DBLP:conf/esa/0001VM22} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its simplicity and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $c$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\Pr_c \left( | c - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil{\frac{12}{\varepsilon^2} \ln{(\frac{81}{\delta})} }$
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
   \For{$i \gets 1$ to $l$}
@@
     \If{$|\chi| = n$}
-      \State $\chi \getsr \chi_{\downarrow \frac{1}{2}}$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
       \State $p \gets \frac{p}{2}$
@@
   \EndFor
-  \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
+  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
   \end{algorithmic}
@@
 %TODO: I think this should be later, we can give a forward reference
-The algorithm's state is a buffer $\chi$ and a fraction $p > 0$.
-The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
-The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
-The algorithm iterates over stream elements, adding each one to the buffer with probability $p$ or conversely removing them with probability $(1-p)$.
-If the buffer gets too large, approximately half of the elements are removed by discarding each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
-After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
-This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
+%% The algorithm's state is a buffer $\chi$  (initially empty) and a fraction $p > 0$ (initially set to $1$).
+%% The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
+%% The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
+%% The algorithm iterates over stream elements, adding each one to the buffer with probability $p$ or conversely -- if the current stream element is already in the buffer -- removing them with probability $(1-p)$.
+%% If the buffer gets too large, approximately half of the elements are removed by discarding each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
+%% After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
+%% This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
+The output of the algorithm is probably-approximately correct, i.e., the probability that the relative error of its output exceeds $\varepsilon$ is at most $\delta$.
+Let us assume the space usage for the elements in the stream is $b$ bits, then the algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, while the stream the algorithm processes requires far more space to store.%
+\footnote{The optimal algorithm requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+It would not be possible to present using such elementary steps as used in \cref{alg:cvm}, and involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}
 
-The pen-and-paper proof of correctness for~\cref{alg:cvm} is remarkably straightforward requiring only undergraduate-level exposure to the analysis of randomized algorithms.
-We set out to formalize the proof as described in the literature~\cite{DBLP:conf/esa/0001VM22}, hoping to show that it would also be an easy formalization exercise.
+The pen-and-paper proof relies on multiple conversions of the algorithm.
+The main reason for that is that the standard methods, such as Chernoff-Hoeffding bounds, usually make statements about independent random variables, such as for example, tail estimates on the deviation of their sum from their mean.
+For the original version of the algorithm, the random operations are far from being independent:
+For example in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations.
+Similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
+Indeed, it seems impossible to analyze \cref{alg:cvm} directly using known methods.
+A brief description of the reduction arguments Chakraborty et al. use to circumvent the problem is presented in Section~\todo{x}. % Maybe we should skip this?
 
-%TODO: a bit too long in details, we should trim this and move to later section
-\subparagraph*{The road not taken:}
-One of the difficult aspects of the proof is that it relies on an eager-lazy coin flip conversion.
-To understand that, we should note that none of the observable random variables, such as the presence of a stream element in the buffer or conditions on the value of $p$, are independent of the other state variables, which makes the algorithm hard to analyze and makes the application of standard techniques from probability theory, such as Hoeffding's theorem impossible.
-The authors resolved that problem by a simulation argument---they show that~\cref{alg:cvm} behaves stochastically identically to a different algorithm, which makes the relevant coin flips in a different order.
-That modified algorithm performs a column of coin flips for each stream element.
-An element is kept in the buffer if the first $k=log2(p-1)$ rows of the column are heads.
-At each sub-sampling step, when p is divided by two, i.e., if k increases by one.
-The algorithm examines the newly activated $k$-th row of the previous sequence elements to decide whether the element should be kept in the buffer.
-This preserves the invariant that the buffer consists of exactly those sequence elements whose associated coin flip column starts with $log2(p-1)$ heads.
-Of course, the new algorithm is not practical for actual implementation, but one can verify its correctness using standard Chernoff bounds, and on the other hand, it is possible to show that its behavior is equivalent to Algorithm 1.
-To summarize, while the algorithm is marvelous, the proof was still very technical.
-The simulation argument, in particular, is not so elegant to formalize.
+This is the point, where our new analysis technique comes in, which enables analysis, very similar to how deterministic algorithms are verified, using a loop invariant.
+Except that the loop invariant is a functional probabilistic inequality.
+For example, we consider invariants of the form:
+\[
+  \expect [ h ] \leq h(c)
+\]
+where $h$ may range over a class of functions, and the expectation is over the distribution of the state of the algorithm at each loop iteration.
+Note that we are able to establish tail bounds using these invariants, which has, so far, not been possible using a simple loop invariant for randomized algorithms.
+The loop invariants are established, essentially, using the key property of the Giry monad:
+\[
+  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f (x)} [h] \, d x \textrm{.}
+\]
+Here $m \isa{\isasymbind} f$ denotes distribution of a randomized algorithm, which represents the sequential composition of $m$ with $f$.
 
-\subparagraph*{A more direct proof:}
-We set out to try to find a more direct proof, which also eases the formalization effort.
-For the following discussion, we will analyze Algorithm 1 with line 8 removed, i.e., the algorithm does not output $\bot$, nor performs a second check of $|\chi|=n$.
-Note that this happens if the very improbable event where none of the elements in X are removed during a subsampling step, which happens with probability at most $2^{-n/2}$.
-Overall, the probability of it happening during the course of the algorithm is at most $\frac{\delta}{2}$.
+The verification of algorithm using our new technique requires only 1044 lines, while the orignal proof required \todo{x} lines.
+To show that our new technique is more powerful, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$, instead of independently deciding for each element, whether to keep it.
+This new version has the benefit, that it is total, note that the second check in line 11 now becomes obsolete.
+Indeed, more interestingly, it can be shown that the resulting algorithm is unbiased, i.e., the expected result is exactly the cardinality of the elements in the stream.
+
+To analyze the new variant, we use results from the theory of negatively dependent random variables.
+This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.
+The concept of negative association is a generalization of independence.
+Such variables observe closure properties and fulfill Chernoff-Hoeffding bounds, similar to independent random variables.
+It should however, be stressed that the theory of negatively associated RVs is orthogonal to our new technique, but the formalization accompanying this work also contains a development of the latter.
+
+In summary, the main contributions of this work are:
+\begin{itemize}
+\item Introduction of a new technique using probabilisitic invariants to verify randomized algorithms inductively/recursively. 
+\item Verification of the orignal CVM algorithm using the new technique.
+\item Presentation of a new variant of the CVM algorithm, that is total and unbiased.
+\item Formalization of a theory of negatively associated random variables.
+\end{itemize}
+
+\todo{Classic summary of sections.}
+
+%% We set out to formalize the proof as described in the literature~\cite{chakraborty2022}, hoping to show that it would also be an easy formalization exercise.
+
+%% %TODO: a bit too long in details, we should trim this and move to later section
+%% \subparagraph*{The road not taken:}
+%% One of the difficult aspects of the proof is that it relies on an eager-lazy coin flip conversion.
+%% To understand that, we should note that none of the observable random variables, such as the presence of a stream element in the buffer or conditions on the value of $p$, are independent of the other state variables, which makes the algorithm hard to analyze and makes the application of standard techniques from probability theory, such as Hoeffding's theorem impossible.
+%% The authors resolved that problem by a simulation argument---they show that~\cref{alg:cvm} behaves stochastically identically to a different algorithm, which makes the relevant coin flips in a different order.
+%% That modified algorithm performs a column of coin flips for each stream element.
+%% An element is kept in the buffer if the first $k=\log_2(p-1)$ rows of the column are heads.
+%% At each sub-sampling step, when p is divided by two, i.e., if k increases by one.
+%% The algorithm examines the newly activated $k$-th row of the previous sequence elements to decide whether the element should be kept in the buffer.
+%% This preserves the invariant that the buffer consists of exactly those sequence elements whose associated coin flip column starts with $\log_2(p-1)$ heads.
+%% Of course, the new algorithm is not practical for actual implementation, but one can verify its correctness using standard Chernoff bounds, and on the other hand, it is possible to show that its behavior is equivalent to Algorithm 1.
+%% To summarize, while the algorithm is marvelous, the proof was still very technical.
+%% The simulation argument, in particular, is not so elegant to formalize.
+
+%% \subparagraph*{A more direct proof:}
+%% We set out to try to find a more direct proof, which also eases the formalization effort.
+%% For the following discussion, we will analyze Algorithm 1 with line 8 removed, i.e., the algorithm does not output $\bot$, nor performs a second check of $|\chi|=n$.
+%% Note that this happens if the very improbable event where none of the elements in X are removed during a subsampling step, which happens with probability at most $2^{-n/2}$.
+%% Overall, the probability of it happening during the course of the algorithm is at most $\frac{\delta}{2}$.
 (Note that removing line 8 does not affect the correctness of the algorithm, but it loses its space consumption bound.)
-It is easy to see that any probability established about the algorithm missing line 8 will be true for the original algorithm with a possible correction by, at most,  $\frac{\delta}{2}$.
-We will remember and correct this at the end of this section.
+%% It is easy to see that any probability established about the algorithm missing line 8 will be true for the original algorithm with a possible correction by, at most,  $\frac{\delta}{2}$.
+%% We will remember and correct this at the end of this section.
 
-Let us consider an imaginary situation where, somehow, $p$ is fixed at some point in the algorithm.
-For example, we could imagine a final sub-sampling loop, which is run until a fixed $p$ is reached.
-Then, the indicator random variables representing the presence of a stream element TODO
+%% Let us consider an imaginary situation where, somehow, $p$ is fixed at some point in the algorithm.
+%% For example, we could imagine a final sub-sampling loop, which is run until a fixed $p$ is reached.
+%% Then, the indicator random variables representing the presence of a stream element TODO

### 2025-01-29 paper/sec/invariants.tex
WIP: Paper   * Saving current state

\section{Probabilistic Invariants}\label{sec:invariants}
In this section, we will derive our new technique using \cref{alg:cvm} as an example. 
For this it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
The total variational distance between these two variants of the algorithms is at most $\frac{\delta}{2}$.
This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction of $\frac{\delta}{2}$.
Considering the modified version, makes the discussion easier, because we do not have to worry about $\bot$.

Let us consider the random variables $X_s$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer: $X_s := I(s \in \chi)$.\footnote{We use $I$ for the indicator of a predicate, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.} 
Before the algorithm encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far.
In particular $\chi \subseteq \{a_1,..,a_m\}$ after loop iteration $m$.

In the loop iteration, where $s$ occurs the first time, it will be inserted with a probability $p$ in lines 3--7.
This means, after line 7, we have 
\begin{equation}
  \label{eq:indicator_eq}
  \expect [p^{-1} X_s] = 1 \textrm{.}
\end{equation}
Interestingly, the equation is preserved throughout the entire algorithm.
Indeed, let us consider a subsampling step.
The probability of $\prob(X_s=1)$ is being halved, but so will $p$, which preserves the equation.

While it is easy see that this is true informally, and since we are going to build on this, let us see how we can verify \cref{eq:indicator_eq} more formally.

For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ represeting the buffer $\chi$ and a real number representing $p$.
Note that, this means $p$ and $\chi$ are random variables, with respect to distributions of the state of the algorithm.
We think of the two parts of each loop iteration lines 3--7 (resp. lines 8--10) as step 1 (resp. step 2).
The state of the algorithm can be represented as the distribution resulting from the sequential composition of alternating steps:
\[
  \mathrm{init} \, \isa{\isasymbind}\; \mathrm{step}_1\, a_1 \; \isa{\isasymbind}\; \mathrm{step}_2 \; \isa{\isasymbind}\; \mathrm{step}_1 \, a_2 \; \isa{\isasymbind}\; \cdots \isa{\isasymbind}\; \mathrm{step}_1\, a_l \; \isa{\isasymbind}\; \mathrm{step}_2 
\]
where we parameterize $\mathrm{step}_1$ with the stream element it processes.
The term $\mathrm{init}$ represents the intial state, i.e., $\mathrm{init} = \mathrm{return} (\{\},1)$.

It is not hard to show by induction, that for all the states occuring above, we will have have $0 < p \leq 1$ and $\chi \subseteq A$, so we will assume this is the case for all probability spaces representing states in the following.

Let us consider verifying that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
    \int_\Omega \int_{\Ber(p)} p^{-1} I\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
\]
Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, but it is useful to remember that these depend on $\sigma$.
To see, that the right-hand-side is equal to $1$, it is usefule to consider the cases, where $a=s$ and the converse seperately.
In the first case the right-hand-side is equal to $1$, can be seen directly, note we assume $p \in (0;1]$.
In the second case it follows from the induction hypothesis.
(In particular, the term in the inner integral is constant with respect to $\tau$.)

The same is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
Then
\[
  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
    \bigint_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
\]
Note that the true- and false-case both evaluate to the same value: $p^{-1} I(s \in \chi)$.
If $s \notin \chi$ both sides are $0$, because the subsampling operation returns a subset of $\chi$.
If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} I(s \in \chi)$.
Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.

Now, with the above it is straight-forward to derive that the expectation of the result $p^{-1} |\chi|$ of our modified algorithm (without line 11) is exactly the desired cardinality $|A|$.
Our main goal, however, is to show that the algorithm is probably approximately correct, i.e., that the probability that the relative error of the output exceeds $\varepsilon$ is less than $\delta$.

Normally, we would use Chernoff bounds for indepenent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
But they are not useful in our case, because the random variables in the case of the CVM algorithm, for example $p^{-1} X_s$, are not independent.
An alternative is the Cram\'{e}r-Chernoff method, which is a very general method to obtain tail bounds for any random variable.
It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
Note that, sometimes the moment generating function is only defined for some values of $t$ (when the corresponding integral exists).
This is typically an interval including $0$.
In which case the inequality is only applicable for those values.
Because the probability space, representing the state of the CVM algorithm is finite, in our case $M(t)$ will always be defined.
Of course, it is also possible to obtain lower tail bounds $P(X \leq a)$ which just requires estimates for the $M(t)$ for $t < 0$, instead of $t > 0$.

This means for our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
\[
  \expect [\exp( t p^{-1} |\chi| )] = \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right]
\]
for $h(x) = \exp(tx)$.
At this point, it was of course tempting to see, whether the proof for \cref{eq:indicator_eq} can also be extended to establish bounds for the above.
When we tried, we got the following result:
\[
  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|}
\]
for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
Note that the exponetial function is convex.
So this does not work directly.
However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$ instead. 
We end up with the condition that $h$ needs to be no-negative and concave only on $[0;q^{-1}]$.
This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$.
This then leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
Of course this requires, separately estimating $\prob(p < q)$, but this is fairly standard trick.
In particular the original proof by Chakraborty et al.~\cite{chakraborty2022} use the same trick.
They use $q = \frac{n}{4 |A|}$ and we could use the same cut-off.

Note that the formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof in its appendix.
The verified algorithm is actually a generalization of the above, where the subsampling probability can be some $f \in [\frac{1}{2};e^{-1/12}]$.
The algorithm presented by Chakraborty et al.~\cite{chakraborty2022}, replicated in the introduction (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are very elementary.
The main point we wanted to make in this section is that, it is possible to derive useful and general probabilistic invariants, by considering expectations of functions of the state, using recursion or induction over the algorithm itself.
As far as we know this method of establishing tail bounds for randomized algorithms is new.

A more interesting consequence of our new method is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
We think that this new variant would not be possible to verify, without the new approach presented here.

### 2025-01-29 paper/sec/negdep.tex
WIP: Paper   * Saving current state

@@
-\section{Negative Dependence}
-\label{sec:negdep}
+\section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}
 
-TODO
+Sometimes, sets of random variables possess a property called negative association.
+The concept is a generalization of independence.
+It has similar closure properties as independent random variables.
+More interestingly, the standard Chernoff-Hoeffding bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
+
+The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orphant dependence.
+
+The following preliminary definition is useful, for the definition of the concept:
+\begin{definition}
+For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, e.g., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
+\end{definition}
+
+\begin{definition}[Negative Association]
+A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$ then
+\[
+\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
+\]
+\end{definition}
+
+The following proposition summarizes some of the properties for negatively associated sets of random variables: 
+\begin{proposition}\label{pro:neg_dep_props}
+Summary of results for negatively associated random variables: 
+\begin{enumerate}
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $f_1,..,f_m : \mathbb R^n \rightarrow \mathbb R$ be all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
+\item If $X_1,..,X_n$ are independent then $X_1,...,X_n$ are negatively associated.
+\item A subset of a negatively associated set of random variables is again negatively associated.
+\end{enumerate}
+\end{proposition}
+
+These properties illustrate the trade-off between negative association and independence.
+For example property 3 would be true for independent random variables, even without the condition of monotonicty.
+Of course, on the other hand independence is a stronger property and fewer sets of random variables are indepenent.
+For our discussion the following is an important lemma about negative associated random variables.
+
+\begin{lemma}\label{le:neg_assoc_prod}
+Let $X_1,..,X_n$ be negatively associated and $f_1,..,f_n$ be all non-decreasing or all non-increasing non-negative functions, then
+\[
+  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
+\]
+\end{lemma}
+\begin{proof}
+Follows from the definition of negative associativy (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
+\end{proof}
+
+The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
+Why we are interested in the lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
+The latter is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
+\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
+However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
+Intuitively, this is interesting because it opens up an avenue to modifying the subsampling step in \cref{alg:cvm}.
+Instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, which has the disadvantage that there is a small possibility that none of the elements gets discarded, we could instead choose a random $nf$-subset, where $\frac{1}{2} \leq f < 1$, such that $nf$ is integer, in the subsampling step.
+For such a subsampling step, we can derive using \cref{le:neg_assoc_prod}:
+\begin{equation}\label{eq:subsample_with_n_subsets}
+  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(I(s \in \chi)) \leq \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(I(s \in \chi)) = \prod_{s \in S}  \expect_{\Ber(f)} g \textrm{.}
+\end{equation}
+for any non-negative $g$ and $S \subseteq \chi$.
+Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.Also, if $S$ is a singleton, the inequality becomes an equality.
+
+To understand, why the above is useful.
+When we want to establish the moment generating function, for the random variable $p^{-1} |\chi|$ --- i.e.
+the result --- inductively, we need to approximate the then-part of the following term, with the else-part.
+\[
+  \ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \prod_{s \in S} h((pf)^{-1}I(s \in \tau)) \, d \tau\right)}{\prod_{s \in S} h(p^{-1}I(s \in \tau))}
+\]
+and the inequality (\cref{eq:subsample_with_n_subsets}) is exactly, what we need.
+This means the proof for the modified algorithm in \cref{sec:invariants} works just as well with the new subsampling method.
+
+The resulting algorithm is depicted in \cref{alg:cvm_new}.
+Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
+The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
+And based on our discussion from the previous section it is unbiased.
+
+\begin{algorithm}[h!]
+	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
+  \For{$i \gets 1$ to $l$}
+    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
+    \If{$b$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
+      \State $p \gets pf$
+    \EndIf
+  \EndFor
+  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}

### 2025-01-29 paper/sec/original_proof.tex
WIP: Paper   * Saving current state

\section{Original Proof}
\todo{Explain how we formalized the original proof by Chakraborty et al. + brief section on the lazify}

### 2025-01-29 paper/sec/semra.tex
WIP: Paper   * Saving current state

\section{Semantics of randomized algorithms}
It is useful to briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad.
The key idea is to model a randomized algorithm as a probability space representing the distributions of its result.
Let us, e.g., consider \cref{alg:example}.
\begin{algorithm}[h!]
\caption{Example for sequential composition.}\label{alg:example}
\begin{algorithmic}[1]
\State $p \getsr \Ber(\frac{1}{2})$
\State $q \getsr \Ber(\frac{1}{3}+p)$
\State \Return $q$
\end{algorithmic}
\end{algorithm}%

In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise.
In the second step the algorithm flips a coin $q$ which is $1$ with probability $\frac{1}{3}$, if $p=0$ or $\frac{2}{3}$ if $q=1$.
I.e., $q$ is a compound distribution a half-half combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$,

Any algorithm can be represented using only two combinators, the bind and return combinator, as well as, primitive random operations.

\paragraph*{Primitive Random Operations}
For example a simple fair coin flip is represented using the Bernoulli distribution $\Ber(\frac{1}{2})$.

\paragraph*{Return Combinator}
Given a singleton element $x$, we can construct the singleton probabilty space, e.g., we assign the probability $1$ to $x$ and $0$ to everything else.
In notation, this is written as: $\mathrm{return}\, x$.

\paragraph*{Bind Combinator}
The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the second randomized algorithm consumes the output of the first.
In notation, this can be expressed using $m \isa{\isasymbind} f$.
Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which are then mixed using the first probability space $m$.

Let us consider an event $A$ in the probability space $m \isa{\isasymbind} f$, then its probability can be evaluated by integrating over its probabilities in $f$ with respect to $m$:
\[
  \prob_{m \isa{\isasymbind} f} (A) = \int_m \prob_{f(x)} (A) \, d x \textrm{.}
\]
Another interesting property is what we mentioned before in the introduction. If $h$ is a random variable over $m \isa{\isasymbind} f$, we can compute its expectation as:
\begin{equation}
  \label{eq:integral_bind}
  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f(x)} [h] \, d x \textrm{.}
\end{equation}

\todo{TODO: Add external references and what we have glossed over (infinite loops, measurability.)}

### 2025-01-31 paper/sec/formalization.tex
Paper   - Lots of edits   - Add formalized version of algorithm

@@
+\begin{algorithm}[h]
+\caption{Formalized Version of \cref{alg:cvm_new}}\label{alg:cvm_formalized}
+\begin{isabelle_cm}
+\isacommand{context}\isamarkupfalse%
+\isanewline
+\ \ \isakeyword{fixes}\ f\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ real\ \isakeyword{and}\ n\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ nat\isanewline
+\ \ \isakeyword{assumes}\ f{\isacharunderscore}{\kern0pt}range{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}f\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{1}}{\isacharslash}{\kern0pt}{\isadigit{2}}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}\ \isakeyword{and}\ n{\isacharunderscore}{\kern0pt}gt{\isacharunderscore}{\kern0pt}{\isadigit{0}}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}n\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+\isakeyword{begin}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ {\isacartoucheopen}initial{\isacharunderscore}{\kern0pt}state\ {\isacharequal}{\kern0pt}\ State\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}\ {\isadigit{1}}{\isacartoucheclose}\ %
+\hfill\isamarkupcmt{Setup initial state $\chi=\emptyset$ and $p=1$.\;%
+}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ subsample\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Subsampling operation: Sample random $n f$ subset.\;%
+}\isanewline
+\ \ {\isacartoucheopen}subsample\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ pmf{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}set\ {\isacharbraceleft}{\kern0pt}S{\isachardot}{\kern0pt}\ S\ {\isasymsubseteq}\ {\isasymchi}\ {\isasymand}\ card\ S\ {\isacharequal}{\kern0pt}\ n\ {\isacharasterisk}{\kern0pt}\ f{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ step\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Loop body.\;%
+}\isanewline
+\ \ {\isacartoucheopen}step\ a\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ b\ {\isasymleftarrow}\ bernoulli{\isacharunderscore}{\kern0pt}pmf\ p{\isacharsemicolon}{\kern0pt}\isanewline
+\ \ \ \ let\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ {\isacharparenleft}{\kern0pt}if\ b\ then\ {\isasymchi}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}\ else\ {\isasymchi}\ {\isacharminus}{\kern0pt}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
+\isanewline
+\ \ \ \ if\ card\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ n\ then\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ \ \ {\isasymchi}\ {\isasymleftarrow}\ subsample\ {\isasymchi}{\isacharsemicolon}{\kern0pt}\isanewline
+\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ {\isacharparenleft}{\kern0pt}p\ {\isacharasterisk}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isanewline
+\ \ \ \ {\isacharbraceright}{\kern0pt}\ else\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\isanewline
+\ \ \ \ {\isacharbraceright}{\kern0pt}\isanewline
+\ \ \ {\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ run{\isacharunderscore}{\kern0pt}steps\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Iterate loop over stream \isa{xs}.\;%
+}\isanewline
+\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}steps\ xs\ {\isacharequal}{\kern0pt}\ foldM{\isacharunderscore}{\kern0pt}pmf\ step\ xs\ initial{\isacharunderscore}{\kern0pt}state{\isacartoucheclose}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ estimate\ \isakeyword{where}\ 
+{\isacartoucheopen}estimate\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ card\ {\isasymchi}\ {\isacharslash}{\kern0pt}\ p{\isacartoucheclose}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ run{\isacharunderscore}{\kern0pt}algo\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Run algorithm and estimate.\;%
+}\isanewline
+\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}algo\ xs\ {\isacharequal}{\kern0pt}\ map{\isacharunderscore}{\kern0pt}pmf\ estimate\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}steps\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{theorem}\isamarkupfalse%
+\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{theorem}\isamarkupfalse%
+\ correctness{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}\isanewline
+\isacommand{end}
+\end{isabelle_cm}
+\end{algorithm}
 \section{Formalization}
-\todo{Overview of the two AFP entries. Include main statement of correctness for the CVM algorithm.}
+This entry is accompanied by two \todo{sessions?}.
+
+The first is a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
+The formalization of the algorithm is presented in \cref{alg:cvm_formalized}.
+We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
+The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
+Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
+\todo{Explain: card}
+\todo{Explain: map-pmf}
+\todo{Explain: pmf-of-set}
+\todo{Explain: return-pmf}

### 2025-01-31 paper/sec/invariants.tex
Paper   - Lots of edits   - Add formalized version of algorithm

@@
 In this section, we will derive our new technique using \cref{alg:cvm} as an example.
-For this it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
+
+Let us briefly, review the algorithm. The algorithm's state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
+The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
+The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
+The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely -- if the current stream element is already in the buffer -- removing them with probability $(1-p)$.
+If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
+If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returing $\bot$.
+After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
+This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
+
+For the discussion in this section, it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
 The total variational distance between these two variants of the algorithms is at most $\frac{\delta}{2}$.
@@
 \]
-Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, but it is useful to remember that these depend on $\sigma$.
-To see, that the right-hand-side is equal to $1$, it is usefule to consider the cases, where $a=s$ and the converse seperately.
+Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these depend on $\sigma$.
+To see, that the right-hand-side is equal to $1$, it is useful to consider the cases, where $a=s$ and the converse seperately.
 In the first case the right-hand-side is equal to $1$, can be seen directly, note we assume $p \in (0;1]$.
@@
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
-    \bigint_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
+    \int_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
 \]
@@
 But they are not useful in our case, because the random variables in the case of the CVM algorithm, for example $p^{-1} X_s$, are not independent.
-An alternative is the Cram\'{e}r-Chernoff method, which is a very general method to obtain tail bounds for any random variable.
+An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
 It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
@@
 So this does not work directly.
-However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$ instead. 
-We end up with the condition that $h$ needs to be no-negative and concave only on $[0;q^{-1}]$.
-This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$.
-This then leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
-Of course this requires, separately estimating $\prob(p < q)$, but this is fairly standard trick.
-In particular the original proof by Chakraborty et al.~\cite{chakraborty2022} use the same trick.
+However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$.
+We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
+This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which successfully leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
+Of course, we also have to separately estimate $\prob(p < q)$.
+We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}.
 They use $q = \frac{n}{4 |A|}$ and we could use the same cut-off.
@@
 Note that the formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof in its appendix.
-The verified algorithm is actually a generalization of the above, where the subsampling probability can be some $f \in [\frac{1}{2};e^{-1/12}]$.
+The verified algorithm is actually a generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
 The algorithm presented by Chakraborty et al.~\cite{chakraborty2022}, replicated in the introduction (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
@@
 
-A more interesting consequence of our new method is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
+When we look at the subsampling step, our proof imposes the following condition on the subsampling operation:
+\begin{equation}\label{eq:subsample_condition}
+  \int_{\mathrm{subsample}(\chi)} \left(\prod_{s \in S} g(I(s \in \tau)) \, d \tau\right) \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
+\end{equation}
+for all non-negative functions $g$.
+An interesting consequence of that, is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
 We think that this new variant would not be possible to verify, without the new approach presented here.

### 2025-01-31 paper/sec/negdep.tex
Paper   - Lots of edits   - Add formalized version of algorithm

@@
 \end{equation}
-for any non-negative $g$ and $S \subseteq \chi$.
-Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.Also, if $S$ is a singleton, the inequality becomes an equality.
+for any non-negative $g$ and $S \subseteq \chi$. Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.
+Also, if $S$ is a singleton, the inequality becomes an equality.
 
-To understand, why the above is useful.
-When we want to establish the moment generating function, for the random variable $p^{-1} |\chi|$ --- i.e.
-the result --- inductively, we need to approximate the then-part of the following term, with the else-part.
-\[
-  \ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \prod_{s \in S} h((pf)^{-1}I(s \in \tau)) \, d \tau\right)}{\prod_{s \in S} h(p^{-1}I(s \in \tau))}
-\]
-and the inequality (\cref{eq:subsample_with_n_subsets}) is exactly, what we need.
+Let us recall that this is the exact condition we needed for the subampling step (\cref{eq:subsample_condition}).
 This means the proof for the modified algorithm in \cref{sec:invariants} works just as well with the new subsampling method.
@@
 The resulting algorithm is depicted in \cref{alg:cvm_new}.
-Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
-The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
-And based on our discussion from the previous section it is unbiased.
-
-\begin{algorithm}[h!]
+\begin{algorithm}[t]
 	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
@@
   \end{algorithmic}
-\end{algorithm}
+\end{algorithm}%
+Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
+The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
+And based on our discussion from the previous section it is also an unbiased algorithm, i.e., the expectation of the result is exactly $|A|$.

### 2025-01-31 paper/sec/semra.tex
Paper   - Lots of edits   - Add formalized version of algorithm

@@
 \State $p \getsr \Ber(\frac{1}{2})$
-\State $q \getsr \Ber(\frac{1}{3}+p)$
+\State $q \getsr \Ber(\frac{1}{3}+\frac{p}{2})$
 \State \Return $q$
@@
 
-In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise.
-In the second step the algorithm flips a coin $q$ which is $1$ with probability $\frac{1}{3}$, if $p=0$ or $\frac{2}{3}$ if $q=1$.
-I.e., $q$ is a compound distribution a half-half combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$,
+In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise. (The notation $\Ber(p)$ represents the Bernoulli distribution.)
+In the second step the algorithm flips a coin $q$ which depends on $p$.
+This has the consequence, that we have to consider probability space valued functions, like: $p \mapsto Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+The resulting distribution for $q$ is a compound distribution: In this case a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
 
-Any algorithm can be represented using only two combinators, the bind and return combinator, as well as, primitive random operations.
+The above example captures the main aspects about reasoning with respect to randomized algorithms.
+Indeed, any algorithm can be represented using only two combinators, the bind and return combinator, as well as, primitive random operations.
 
@@
 \end{equation}
-
-\todo{TODO: Add external references and what we have glossed over (infinite loops, measurability.)}
+The above is just a very brief presentation of the Giry monad~\cite{giry1982}, a more thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} or Lochbihler~\cite{lochbihler2016}.

### 2025-02-04 paper/sec/formalization.tex
Paper: Section on the original proof.

@@
 \end{algorithm}
-\section{Formalization}
-This entry is accompanied by two \todo{sessions?}.
+\section{Formalization of the CVM algorithm}
+The supplementary material contains a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
 
-The first is a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
-The formalization of the algorithm is presented in \cref{alg:cvm_formalized}.
+A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
+The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
 We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
@@
 Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
-\todo{Explain: card}
-\todo{Explain: map-pmf}
-\todo{Explain: pmf-of-set}
-\todo{Explain: return-pmf}
+Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
+For readers not familiar with Isabelle, the following table contains brief descriptions of the syntactic elements used in the listing:
+
+\medskip
+\noindent\begin{tabular}{l p{9cm}}
+\toprule
+Term & Description \\
+\midrule
+\isa{card\ S} & Cardinality of a finite set $S$. \\
+\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
+\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
+\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
+\isa{nat} & Type of natural numbers (non-negative integers). \\
+\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
+\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
+\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
+\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
+\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
+\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
+\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
+\bottomrule
+\end{tabular}
+
+\medskip
+The formalization contains four theory files.
+Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
+as most of the proof is identical.
+The entry CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, wiht an abstract subsampling operation, which is required to fullfill \cref{eq:subsample_condition}.
+The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
+It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
+The total number of lines required for the verification of the original algorithm is 1004 lines.
+This is in contrast to the original proof, for which the verification required \todo{x} lines.
+
+\medskip
+\noindent\begin{tabular}{l l l l}
+\toprule
+Name & Description & Lines of Code & Comments \\
+\midrule
+CVM{\textunderscore}Preliminary & 79 \\
+CVM{\textunderscore}Abstract{\textunderscore}Algorithm & 675 \\
+CVM{\textunderscore}Original{\textunderscore}Algorithm & 250 \\
+CVM{\textunderscore}New{\textunderscore}Unbiased{\textunderscore}Algorithm & 139 \\
+\bottomrule
+\end{tabular}

### 2025-02-04 paper/sec/original_proof.tex
Paper: Section on the original proof.

@@
 \section{Original Proof}
+In this section we describe the interesting parts of the original proof by Chakraborty et al.~\cite{chakraborty2022}.
+This highlights in part, why it requires a lot of work to formalize, and why we were motivated to develop our new approach using probabilistic invariants.
+
+As we mentioned in the introduction, the main difficulty is the fact that \cref{alg:cvm}'s state variables are not independent.
+The trick Chakraborty et al.
+use to circumvent the problem is by modifying the algorithm, in a manner that obviously preserves its distribution, such that the resulting algorithm's state can be described in terms of independent coin flips.
+
+Let us consider a state, where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$. 
+Then the algorithm would normally perform a coin flip with probability $p$.
+In the modified algorithm, we perform a fixed number of coin flips for each sequence element at the beginning.
+The element is put into the sample, whenever the first $k$ coin flips associated with the sequence element are $1$.
+Note that this happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged.
+
+Moreover, during the subsampling operation those elements are kept, whose $k+1$-th associated coin flip is $1$.
+This again preserves the behaviour, that each element is discared independently with probability $1/2$.
+The operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
+For this modified algorithm, it is easy to see that the set of elements in any state are exactly, those stream elements for which the first $k$ entries of their associated coin flips are $1$.
+
+In \cref{alg:cvm_simul} we describe the new algorithm, that behaves exactly as the original one (\cref{alg:cvm}), but performs indepenent coin flips.
+Note that the function $\mathrm{last{\textunderscore}index}$ returns the index of the last-occurence of an element in the sequence, before the current loop iteration.
+It should be noted that the algorithm keeps track of the number of subsampling iterations $k$, instead of the value $p = 2^{-k}$ as the original algorithm does.
+%Like, we did in \cref{sec:invariants}, we ignore the second check, whether the subsampling operation succeeded.
+%As we explained there, the total variational difference between these two variants is $\frac{\delta}{2}$.
+%
+\begin{algorithm}[h!]
+	\caption{Modified CVM algorithm with independent coin flips.}\label{alg:cvm_simul}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
+  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ \Comment perform $l^2$ unbiased independent coin flips
+  \For{$i \gets 1$ to $l$}
+    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State $\chi \gets \{a \in \chi | b[\mathrm{last{\textunderscore}index}(a),k+1] = 1\}$
+      \State $k \gets k+1$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State \Return $\bot$
+    \EndIf
+  \EndFor
+  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}
+To roughly explain, how tail bounds can be derived for this new variant:
+It is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific value of $k$, which can be bounded by the probability that the number of stream elements with whose associated coin flips start with $k$ ones, is outside of $2^{-k} |A| (1 \pm \varepsilon)$.
+This is explained in more detail by Chakraborty et al.~\cite{chakraborty2022}.
+
+One of the key questions is, how to formalize the transformation from \cref{alg:cvm} to this new variant.
+What we discovered is that it best to solve the problem backwards, i.e., we start with the modified algorithm, which performs all the coin-flips in advance --- eagerly --- and convert it to the version, that performs the coin flips --- lazily --- at the point they are needed.
+
+We can push down the coin flips through the expression tree.
+To explain how this works.
+Let us first define the \emph{sampling} function, i.e., let $f$ be a function that depends on coin flips, more precisely $f$ takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to indepenent unbiased coin flips as.
+\begin{isabelle_cm}
+  sample\ f\ \isacharequal\ map{\isacharunderscore}pmf\ f\ {\isacharparenleft}prod{\isacharunderscore}pmf\ I\ {\isacharparenleft}\isasymlambda\isacharunderscore\isachardot\ bernoulli{\isacharunderscore}pmf \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright\isacharparenright\isacharparenright
+\end{isabelle_cm}
+%For example, we could use our modified algorithm as $f$, with the index set $I = \{0, \ldots, l-1\} \times \{0, \ldots, l-1\}$.
+The interesting fact is that we can distribute the sampling operation over composition, e.g.:
+\begin{isabelle_cm}
+  sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ f\ \isasymomega\ \isasymcirc\ g\ \isasymomega{\isacharparenright}\ \isacharequal\ sample\ g\ \isasymbind\ \isacharparenleft{\isasymlambda}x\isachardot\ sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ f\ \isasymomega\isachardot\ x\isacharparenright\isacharparenright
+\end{isabelle_cm}
+if $f$ and $g$ depend on disjoint subsets of the coin flips.
+
 \todo{Explain how we formalized the original proof by Chakraborty et al. + brief section on the lazify}

### 2025-02-05 paper/sec/formalization.tex
Paper: Save state.

@@
 The formalization contains four theory files.
+The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
 Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
 as most of the proof is identical.
-The entry CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, wiht an abstract subsampling operation, which is required to fullfill \cref{eq:subsample_condition}.
+The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, wiht an abstract subsampling operation, which is required to fullfill \cref{eq:subsample_condition}.
 The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
@@
 This is in contrast to the original proof, for which the verification required \todo{x} lines.
-
-\medskip
-\noindent\begin{tabular}{l l l l}
-\toprule
-Name & Description & Lines of Code & Comments \\
-\midrule
-CVM{\textunderscore}Preliminary & 79 \\
-CVM{\textunderscore}Abstract{\textunderscore}Algorithm & 675 \\
-CVM{\textunderscore}Original{\textunderscore}Algorithm & 250 \\
-CVM{\textunderscore}New{\textunderscore}Unbiased{\textunderscore}Algorithm & 139 \\
-\bottomrule
-\end{tabular}

### 2025-02-05 paper/sec/formalization_neg_dep.tex
Paper: Save state.

\section{Formalization of the concept of negative association}\label{sec:formalization_neg_dep}
TODO

### 2025-02-06 paper/sec/original_proof.tex
Paper: Save state.
Paper: Add missing files.   - Minor edits to original proof.

@@
     \If{$|\chi| = n$}
-      \State $\chi \gets \{a \in \chi | b[\mathrm{last{\textunderscore}index}(a),k+1] = 1\}$
+      \State $\chi \gets \{a \in \chi \;|\; b[\mathrm{last{\textunderscore}index}(a),k+1] = 1\}$
       \State $k \gets k+1$
@@
 \end{isabelle_cm}
-if $f$ and $g$ depend on disjoint subsets of the coin flips.
+if $f$ and $g$ depend on disjoint subsets of the coin flips, i.e., if $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
 
-\todo{Explain how we formalized the original proof by Chakraborty et al. + brief section on the lazify}
+By recursively applying the rule, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
+Note that, we arrive at the orignal version of the algorithm, that performs the coin flips lazily.
+
+A detail, that we simplified here is that the split of the index sets, e.g., which coin flips $f$ depends on and which coin flips $g$ depends on, maybe dynamic.
+This means the split may depend on the return value of $f$.
+For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin-flips.
+This means, we have a situation, where the previous loop iteration communicates to the next loop iteration, which coin flips it depends on using the state.
+And the next loop iteration, will indeed only read coin flips that were not read by the previous iteration.
+%For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
+%% It is possible to express the above in a point-free manner.
+%% For that, we need to introduce the reader monad.
+%% This is a deterministic monad, which provides a read operation to a global value, such as our set of coin flips.
+%% (In Haskell, the reader monad is commonly used to provide a fixed enviroment to algorithms, such as global configuration options, paths or command lines parameters~\cite{jones1995}.)
+
+%% \begin{isabelle_cm}
+%% \isacommand{datatype}\ \isacharparenleft{\isacharprime}c\isacharcomma\ {\isacharprime}a\isacharparenright\ reader{\isacharunderscore}monad\ =\ Reader\ {\isacharparenleft}run{\isacharunderscore}reader\isacharcolon\ \isacartoucheopen{\isacharprime}c\ \isasymRightarrow\ {\isacharprime}a\isacartoucheclose\isacharparenright)\isanewline
+%% \isanewline
+%% \isacommand{definition}\ bind{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}bind{\isacharunderscore}rd\ m\ f\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}r\isachardot\ run{\isacharunderscore}reader\ {\isacharparenleft}f\ {\isacharparenleft}run{\isacharunderscore}reader\ m\ r\isacharparenright\isacharparenright\ r\isacharparenright\isacartoucheclose\isanewline
+%% \isacommand{definition}\ return{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}return{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}{\isacharunderscore}\isachardot\ x\isacharparenright\isacartoucheclose\isanewline
+%% \isacommand{definition}\ get{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}get{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ id\isacartoucheclose
+%% \end{isabelle_cm}
+%% With the definition above, we can define our sampling function with respect to the reader monad, where the environment is the entire set of coin flips, which is now a monad morphism (from the reader monad to the Giry monad):
+%% \begin{isabelle_cm}
+%%   sample{\isacharunderscore}rd\ \isacharequal\ sample\ \isasymcirc\ run{\isacharunderscore}reader
+%% \end{isabelle_cm}

### 2025-02-06 paper/sec/intro.tex
Paper: Section on the original proof.
tweak the intro substantially

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
-Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its simplicity and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
+Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
 One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
-Instead, the algorithm could theoretically be implemented in a setting where the objects in the data stream only allow for equality comparisons.
-Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
+Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
+
+Another property of CVM is its simplicity, both in terms of its description---the algorithm is listed in its entirety in~\cref{alg:cvm}---and its pen-and-paper correctness proof, which only requires undergraduate-level exposure to randomized algorithms.
+This is why its authors called it ``an algorithm for the text book''.
+The motivation of this work is to explore whether CVM's simplicity transfers into a mechanization within the Isabelle proof assistant~\cite{nipkow2002}.
 
 \begin{algorithm}[h!]
-	\caption{CVM algorithm for distinct elements estimation.}\label{alg:cvm}
+	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
 	\begin{algorithmic}[1]
@@
 
+Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
+Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
+The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
+Yet, for~\cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{DBLP:journals/corr/abs-2301-10191} provides a correct proof.
+The original error serves as a side motivation for this work.}
+For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
+The aforementioned sequence of transformations bounds the error of~\cref{alg:cvm} in terms of another randomized algorithm with more desirable independence properties.
+Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.
+
+This is the point where our new analysis technique comes in---it is very similar to how deterministic algorithms are verified using a loop invariant.
+The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
+\[
+  \expect [ h ] \leq h(c)
+\]
+where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
+By first establishing such an invariant for~\cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
+For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
+We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.
+
+The main contributions of this work are:
+\begin{itemize}
+\item Introduction of a new technique using functional probabilistic invariants to verify randomized algorithms inductively/recursively.
+\item Verification of the original CVM algorithm using our new technique.
+\item Presentation and verification of a new variant of CVM that is total and unbiased.
+\item Formalization of a theory of negatively associated random variables used to analyze the new CVM variant.
+\end{itemize}
+
+Verification of CVM using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
+We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191} in Section~\todo{x}.
+
+The formalized CVM variant showcases the utility of our new proof technique.
+In this variant, the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-sized subset of $\chi$, instead of independently deciding for each element, whether to keep it.
+This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
+More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property, that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+
+To analyze the new variant, we use results from the theory of negatively dependent random variables to establish the desired functional invariant.
+The concept of negative association is a generalization of independence; negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
+It should be stressed that the theory of negatively associated RVs is orthogonal to our new technique, but the formalization of such a theory is also a contribution of this work.
+
+% TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown
+%This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.
+
+% TODO: move it into explanation of the technique
+%Note that we are able to establish tail bounds using these invariants, which has, so far, not been possible using a simple loop invariant for randomized algorithms.
+%The loop invariants are established, essentially, using the key property of the Giry monad:
+%\[
+%  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f (x)} [h] \, d x \textrm{.}
+%\]
+%Here $m \isa{\isasymbind} f$ denotes distribution of a randomized algorithm, which represents the sequential composition of $m$ with $f$.
+
 %TODO: I think this should be later, we can give a forward reference
@@
 %% This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
-The output of the algorithm is probably-approximately correct, i.e., the probability that the relative error of its output exceeds $\varepsilon$ is at most $\delta$.
-Let us assume the space usage for the elements in the stream is $b$ bits, then the algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, while the stream the algorithm processes requires far more space to store.%
-\footnote{The optimal algorithm requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+%The output of CVM is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output exceeds $\varepsilon$ is at most $\delta$.
+%Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state---far less than storing each stream element deterministically.%
+%\footnote{The optimal randomized algorithm requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
 It would not be possible to present using such elementary steps as used in \cref{alg:cvm}, and involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}
-
-The pen-and-paper proof relies on multiple conversions of the algorithm.
-The main reason for that is that the standard methods, such as Chernoff-Hoeffding bounds, usually make statements about independent random variables, such as for example, tail estimates on the deviation of their sum from their mean.
-For the original version of the algorithm, the random operations are far from being independent:
-For example in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations.
-Similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
-Indeed, it seems impossible to analyze \cref{alg:cvm} directly using known methods.
-A brief description of the reduction arguments Chakraborty et al. use to circumvent the problem is presented in Section~\todo{x}. % Maybe we should skip this?
-
-This is the point, where our new analysis technique comes in, which enables analysis, very similar to how deterministic algorithms are verified, using a loop invariant.
-Except that the loop invariant is a functional probabilistic inequality.
-For example, we consider invariants of the form:
-\[
-  \expect [ h ] \leq h(c)
-\]
-where $h$ may range over a class of functions, and the expectation is over the distribution of the state of the algorithm at each loop iteration.
-Note that we are able to establish tail bounds using these invariants, which has, so far, not been possible using a simple loop invariant for randomized algorithms.
-The loop invariants are established, essentially, using the key property of the Giry monad:
-\[
-  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f (x)} [h] \, d x \textrm{.}
-\]
-Here $m \isa{\isasymbind} f$ denotes distribution of a randomized algorithm, which represents the sequential composition of $m$ with $f$.
-
-The verification of algorithm using our new technique requires only 1044 lines, while the orignal proof required \todo{x} lines.
-To show that our new technique is more powerful, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$, instead of independently deciding for each element, whether to keep it.
-This new version has the benefit, that it is total, note that the second check in line 11 now becomes obsolete.
-Indeed, more interestingly, it can be shown that the resulting algorithm is unbiased, i.e., the expected result is exactly the cardinality of the elements in the stream.
-
-To analyze the new variant, we use results from the theory of negatively dependent random variables.
-This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.
-The concept of negative association is a generalization of independence.
-Such variables observe closure properties and fulfill Chernoff-Hoeffding bounds, similar to independent random variables.
-It should however, be stressed that the theory of negatively associated RVs is orthogonal to our new technique, but the formalization accompanying this work also contains a development of the latter.
-
-In summary, the main contributions of this work are:
-\begin{itemize}
-\item Introduction of a new technique using probabilisitic invariants to verify randomized algorithms inductively/recursively. 
-\item Verification of the orignal CVM algorithm using the new technique.
-\item Presentation of a new variant of the CVM algorithm, that is total and unbiased.
-\item Formalization of a theory of negatively associated random variables.
-\end{itemize}
+%
 

### 2025-02-06 paper/sec/intro.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
@@
 Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
-
-Another property of CVM is its simplicity, both in terms of its description---the algorithm is listed in its entirety in~\cref{alg:cvm}---and its pen-and-paper correctness proof, which only requires undergraduate-level exposure to randomized algorithms.
-This is why its authors called it ``an algorithm for the text book''.
-The motivation of this work is to explore whether CVM's simplicity transfers into a mechanization within the Isabelle proof assistant~\cite{nipkow2002}.
+Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
 
@@
 
-Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
+%Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
 Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
 The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
-Yet, for~\cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{DBLP:journals/corr/abs-2301-10191} provides a correct proof.
+Yet, for \cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
 The original error serves as a side motivation for this work.}
 For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
-The aforementioned sequence of transformations bounds the error of~\cref{alg:cvm} in terms of another randomized algorithm with more desirable independence properties.
+The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
 Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.
 
-This is the point where our new analysis technique comes in---it is very similar to how deterministic algorithms are verified using a loop invariant.
+In this paper, we present a new technique for analyzing randomized algorithms.
+The technique yields a direct and substantially more general proof of the CVM algorithm.
+It is very similar to how deterministic algorithms are verified using a loop invariant.
 The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
@@
 
-Verification of CVM using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
-We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191} in Section~\todo{x}.
+The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
+We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} in \cref{sec:transformation_based_proof}.
 
-The formalized CVM variant showcases the utility of our new proof technique.
-In this variant, the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-sized subset of $\chi$, instead of independently deciding for each element, whether to keep it.
+To show that our new technique is more general, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$, instead of independently deciding for each element, whether to keep it.
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
@@
 
+The following section provides more background information, in particular, about the semantics of randomized algorithms.
+After which, we introduce our new technique, showing how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
+\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
+\cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new library on negatively associated random variables.
+In \cref{sec:transformation_based_proof}, we present details about our alternative verification of the CVM using the transformation based proof by Chakraborty et al.
+The last sections conclude with a brief overview of related work and a summary of our results.
 % TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown
@@
 %
-
-\todo{Classic summary of sections.}
 

### 2025-02-06 paper/sec/original_proof.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.

@@
-\section{Original Proof}
-In this section we describe the interesting parts of the original proof by Chakraborty et al.~\cite{chakraborty2022}.
+\section{Transformation-based Proof}\label{sec:transformation_based_proof}
+In this section we describe the interesting parts of the original transformation-based proof by Chakraborty et al.~\cite{chakraborty2022}.
 This highlights in part, why it requires a lot of work to formalize, and why we were motivated to develop our new approach using probabilistic invariants.

### 2025-02-06 paper/main.tex
Paper   - Lots of edits   - Add formalized version of algorithm
Paper: Section on the original proof.
Minor edits   - Add classification, keywords.   - Minor typos.

@@
 
-\usepackage{bigints}
+\usepackage{booktabs}
+\usepackage{mdframed}
 \usepackage{algorithm}
@@
 %\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+
+\definecolor{shadecolor}{gray}{0.95}
+\newenvironment{isabelle_cm}{\begin{mdframed}[backgroundcolor=shadecolor,nobreak=true,linewidth=0]\begin{isabelle}}{\end{isabelle}\end{mdframed}}%
 
@@
 
-\ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\ccsdesc[500]{Theory of computation~Logic and verification}
+\ccsdesc[500]{Theory of computation~Higher order logic}
+\ccsdesc[500]{Mathematics of computing~Probabilistic algorithms}
+\ccsdesc[500]{Theory of computation~Pseudorandomness and derandomization}
 
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+\keywords{Formal Verification, Isabelle/HOL, Randomized Algorithms, Distinct Elements}
 
@@
 
+\input{sec/formalization_neg_dep}
+
 \input{sec/original_proof}
+
+\input{sec/related_work}
+
+\input{sec/conclusion}
 

### 2025-02-06 paper/sec/formalization.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.
Minor edits   - Add classification, keywords.   - Minor typos.

@@
 \end{algorithm}
-\section{Formalization of the CVM algorithm}
+\section{Formalization of the CVM algorithm}\label{sec:formalization}
 The supplementary material contains a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
@@
 as most of the proof is identical.
-The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, wiht an abstract subsampling operation, which is required to fullfill \cref{eq:subsample_condition}.
+The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
 The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).

### 2025-02-06 paper/sec/intro.tex
Minor edits   - Add classification, keywords.   - Minor typos.

@@
 %Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
-Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
+The pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
 The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
-Yet, for \cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
+Yet, for \cref{alg:cvm} the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
 The original error serves as a side motivation for this work.}
@@
 The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
-Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.
+Indeed, it seems impossible to analyze \cref{alg:cvm} directly using known methods.
 
@@
 where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
-By first establishing such an invariant for~\cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
+By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
 For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.

### 2025-02-07 paper/sec/intro.tex
minor intro edits

@@
 
-%Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
-The pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
+The pen-and-paper analysis of CVM~\cite{chakraborty2022,chakraborty2023} relies on a sequence of transformations of the algorithm.
 The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
-Yet, for \cref{alg:cvm} the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
+Yet, for \cref{alg:cvm}, the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are independent; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
 The original error serves as a side motivation for this work.}
 For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
-The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
-Indeed, it seems impossible to analyze \cref{alg:cvm} directly using known methods.
+The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the estimation results for the original algorithm can be deduced.
+To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods.
 
-In this paper, we present a new technique for analyzing randomized algorithms.
-The technique yields a direct and substantially more general proof of the CVM algorithm.
-It is very similar to how deterministic algorithms are verified using a loop invariant.
+In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
+Our approach is very similar to how deterministic algorithms are verified using a loop invariant.
 The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
@@
 By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
-For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
-We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.
+For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
+We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to ideas from interactive theorem proving.
 
@@
 
-The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
-We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} in \cref{sec:transformation_based_proof}.
+The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof (which we also mechanized) required \todo{x} lines.
+We detail some of the challenges faced when mechanizing the transformation arguments used by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} in \cref{sec:transformation_based_proof}.
 
@@
 
-To analyze the new variant, we use results from the theory of negatively dependent random variables to establish the desired functional invariant.
+To analyze the new variant, we use results from the theory of negatively dependent random variables~\cite{joagdev1983} to establish the desired functional invariant.
 The concept of negative association is a generalization of independence; negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
@@
 
-The following section provides more background information, in particular, about the semantics of randomized algorithms.
-After which, we introduce our new technique, showing how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
+\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle~\cite{nipkow2002}.
+\cref{sec:invariants} introduces our new technique, and shows how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
 \cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
@@
 The last sections conclude with a brief overview of related work and a summary of our results.
+
 % TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown
@@
 %% After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
-%% This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
-%The output of CVM is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output exceeds $\varepsilon$ is at most $\delta$.
-%Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state---far less than storing each stream element deterministically.%
-%\footnote{The optimal randomized algorithm requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
-It would not be possible to present using such elementary steps as used in \cref{alg:cvm}, and involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}
-%
-
-%% We set out to formalize the proof as described in the literature~\cite{chakraborty2022}, hoping to show that it would also be an easy formalization exercise.
 

### 2025-02-07 paper/sec/semra.tex
draft more in background

--- 2025-01-29 paper/sec/semra.tex
@@
-\section{Semantics of randomized algorithms}
-It is useful to briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad.
+\section{Background}
+\label{sec:background}
+
+\subsection{Randomized Algorithms and Distinct Elements}
+
+The CVM algorithm is a \emph{streaming} algorithm for the distinct elements problem.
+As shown in~\cref{alg:cvm}, given a data stream $a_1,\dots, a_l$, the goal of such algorithms is to return an accurate cardinality estimate for $A = \{a_1,\dots,a_l\}$.
+
+Importantly, the CVM algorithm is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output estimate $R$ satisfies $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$ for parameters $\varepsilon$ and $\delta$.
+Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}
+
+\subsection{Semantics of Randomized Algorithms}
+We briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad~\cite{giry1982}.
+A more thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
+
 The key idea is to model a randomized algorithm as a probability space representing the distributions of its result.
-Let us, e.g., consider \cref{alg:example}.
+For simplicity, consider \cref{alg:example}.
 \begin{algorithm}[h!]
@@
 
-In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise.
-In the second step the algorithm flips a coin $q$ which is $1$ with probability $\frac{1}{3}$, if $p=0$ or $\frac{2}{3}$ if $q=1$.
-I.e., $q$ is a compound distribution a half-half combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$,
+In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise. (The notation $\Ber(p)$ represents the Bernoulli distribution.)
+In the second step the algorithm flips a coin $q$ which depends on $p$.
+This has the consequence, that we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+The resulting distribution for $q$ is a compound distribution, a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
 
-Any algorithm can be represented using only two combinators, the bind and return combinator, as well as, primitive random operations.
+This example captures the main aspects of modeling randomized algorithms in the Giry monad.
+Indeed, randomized algorithms can be modeled using the following ingredients:
 
-\paragraph*{Primitive Random Operations}
-For example a simple fair coin flip is represented using the Bernoulli distribution $\Ber(\frac{1}{2})$.
+\begin{description}
+\item[Primitive Random Operations.] For example, a simple fair coin flip is represented using the Bernoulli distribution $\Ber(\frac{1}{2})$.
+\item[Return Combinator.]
+Given a singleton element $x$, we can construct the singleton probability space, assigning probability $1$ to $x$ and $0$ to everything else.
+In monad notation, this is written as: $\mathrm{\bf return}\, x$.
 
-\paragraph*{Return Combinator}
-Given a singleton element $x$, we can construct the singleton probabilty space, e.g., we assign the probability $1$ to $x$ and $0$ to everything else.
-In notation, this is written as: $\mathrm{return}\, x$.
+\item[Bind Combinator.]
+The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the second randomized algorithm consumes the output of the first; in monad notation, this is: $m \isa{\isasymbind} f$.
+Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which takes inputs from the probability space $m$.
 
-\paragraph*{Bind Combinator}
-The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the second randomized algorithm consumes the output of the first.
-In notation, this can be expressed using $m \isa{\isasymbind} f$.
-Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which are then mixed using the first probability space $m$.
-
-Let us consider an event $A$ in the probability space $m \isa{\isasymbind} f$, then its probability can be evaluated by integrating over its probabilities in $f$ with respect to $m$:
+Let us consider an event $A$ in the probability space $m \isa{\isasymbind} f$.
+Its probability can be evaluated by integrating over its probabilities in $f$ with respect to $m$:
 \[
@@
 \]
-Another interesting property is what we mentioned before in the introduction. If $h$ is a random variable over $m \isa{\isasymbind} f$, we can compute its expectation as:
+
+Another key property is the calculation of expectations;
+if $h$ is a random variable over $m \isa{\isasymbind} f$, we can compute its expectation as:
 \begin{equation}
@@
 
-\todo{TODO: Add external references and what we have glossed over (infinite loops, measurability.)}
+\cref{eq:integral_bind} is crucially used to establish invariants of the form introduced in~\cref{sec:intro}.
+\end{description}
+
+\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}

### 2025-02-07 paper/sec/formalization.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

@@
-\begin{algorithm}[h]
+\begin{algorithm}[t]
 \caption{Formalized Version of \cref{alg:cvm_new}}\label{alg:cvm_formalized}
@@
 Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
-For readers not familiar with Isabelle, the following table contains brief descriptions of the syntactic elements used in the listing:
-
-\medskip
+For readers not familiar with Isabelle, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the listing.
+%\medskip
+\begin{table}
+\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
 \noindent\begin{tabular}{l p{9cm}}
@@
 \end{tabular}
-
-\medskip
+\end{table}
+%\medskip
 The formalization contains four theory files.

### 2025-02-07 paper/sec/formalization_neg_dep.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

@@
 \section{Formalization of the concept of negative association}\label{sec:formalization_neg_dep}
-TODO
+As mentioned in the previous section, the formalization of the new unbiased variant of the CVM algorithm, requires results from the theory of negative association.
+The supplementary material contains a formalization of the concept in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
+
+The formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
+However, the authors did not correctly define the constrains for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
+In the formalization, we assume that the functions are bounded and measurable.
+On the other hand, we derive that the condition is still true, if $f, g$ are square integrable.
+Or alternatively integrable, and non-negative.
+This is derived using the monotone-convergence theorem.
+
+Another deviation from the original work is that we do not require that the random variables need to be real-valued.
+In the formalization any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
+In this case the test functions must be monotone with respect ot the respective order on that space.
+
+The formalization contains the standard Chernoff bounds for negatively associated random variables.
+This includes the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multipicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
+Although, these results are not needed for the proof of the CVM algorithm.
+
+A key issue, we faced during formalization, was the fact that there are many theorems of the form if a set of functions are either simultaneously monotone or anti-monotone.
+For this we introduce a notation, that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}.
+Which evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
+For example the FKG inequality
+\[
+  \expect [f g] \geq \expect [f] \expect [g]
+\]
+is true, if $f$ and $g$ are both montone, or both antimontone.
+On the other hand the reverse inequality is true, if $f$ is montone and $g$ is antimontone, or vice versa.
+Using our parameterized relation symbol, we could state all variants in a concise manner.

### 2025-02-07 paper/sec/negdep.tex
Paper: Section on the original proof.
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

@@
 \section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}
-
 Sometimes, sets of random variables possess a property called negative association.
@@
 
-\begin{definition}[Negative Association]
+\begin{definition}[Negative Association]\label{def:neg_assoc}
 A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$ then
@@
 The resulting algorithm is depicted in \cref{alg:cvm_new}.
-\begin{algorithm}[t]
+\begin{algorithm}[h]
 	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
@@
     \If{$|\chi| = n$}
-      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$ subset of $\chi$
       \State $p \gets pf$

### 2025-02-07 paper/sec/semra.tex
Paper:   * Add section on formalization of negatively dependend variables.   * Minor edits.

@@
 
-\cref{eq:integral_bind} is crucially used to establish invariants of the form introduced in~\cref{sec:intro}.
+\cref{eq:integral_bind} is crucially used to establish invariants of the form we will use in~\cref{sec:invariants}.
 \end{description}
-
-\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}
+%\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}

### 2025-02-07 tools/count_lines.py
Add script to count lines of the formalization.

@@ -1,0 +1,25 @@
+# Determines the number of code lines (i.e. excluding empty and comment, presentation lines)
+# of the theory files.
+
+import re
+import sys
+import glob
+
+def readfile(name):
+    f = open(name, mode='r',)
+    r = f.read()
+    f.close()
+    return r
+
+t = 0
+for file in glob.glob(sys.argv[1]+"**/*.thy", recursive=True):
+    a = readfile(file)
+    a = re.sub(r'section\s*\\<open>(.|\n)*?\\<close>','',a)
+    a = re.sub(r'text\s*\\<open>(.|\n)*?\\<close>','',a)
+    a = re.sub(r'\(\*(.|\n)*?\*\)','',a)
+    a = re.sub(r'\n\s*\n','\n',a)
+    c = a.count('\n')
+    t += c
+    print('{} {}'.format(file, c))
+
+print('Total: {}'.format(t))

### 2025-02-10 paper/sec/original_proof.tex
Paper: Added related work section. (WIP)

@@
 In \cref{alg:cvm_simul} we describe the new algorithm, that behaves exactly as the original one (\cref{alg:cvm}), but performs indepenent coin flips.
-Note that the function $\mathrm{last{\textunderscore}index}$ returns the index of the last-occurence of an element in the sequence, before the current loop iteration.
+Note that the function last{\textunderscore}index returns the index of the last-occurence of an element in the sequence, before the current loop iteration.
 It should be noted that the algorithm keeps track of the number of subsampling iterations $k$, instead of the value $p = 2^{-k}$ as the original algorithm does.
@@
     \If{$|\chi| = n$}
-      \State $\chi \gets \{a \in \chi \;|\; b[\mathrm{last{\textunderscore}index}(a),k+1] = 1\}$
+      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$
       \State $k \gets k+1$

### 2025-02-10 paper/sec/related_work.tex
Paper: Added related work section. (WIP)

\section{Related Work}\label{sec:related_work}
It is important to note that there are already good practical solutions for the distinct elements problem.
The first solution was presented by Flajolet~\cite{flajolet1985} in 1985, however he like many other authors~\cite{flajolet2007,heule2013,pettie2021}. make the assumption that a fixed hash function can be regarded as a fully random function.
Alon et al.~\cite[Section 2.3]{alon1999} presented in 1999 an easy solution, which does not require unmotivated model assumptions.
Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pair-wise independent family.
The space complexity of the algorithm is $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) b)$, where we again assume that $b$ is the number of bits required to represent the stream elements.
This is slightly better than the CVM algorithm (which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$), since we do not have a dependency on $l$.
Later Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented more and more sophisticated solutions.
The last one by B\l{}asiok is indeed optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
Karayel~\cite{karayel2023} presented a version of the latter, that preserves monotonicity and supports the merge-operation, which enables the use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
It should, however, be noted that the more recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
What makes the CVM algorithm really unique is the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms do.

The aformentioned hash based algorithms are not unbiased.
Flajolet et al.~\cite{flajolet1985} points it out and provides bounds on the bound.
Most authors do not discuss the matter.
It, however, is not hard to see that they are not.
One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
In the context of query processing unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements in the streaming model.

As far as we know probabilistic invariants have not been used to establish Gaussian tail-bounds.
It is fairly common to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}, using recursive analysis, which is easy due to the linearity of expectations and --- in the case of independent variables --- variances.
A simple example is the Morris-counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.

There is also research on the (automated) analysis of loop invariants, for probabilisitic loops, using the characteristic function of loops~\cite{batz2023, mciver2005}.
This works by establishing the limiting distribution of the state of the loop.
De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
Our approach differs from these techniques, by avoiding the computation of the distribution, which, we think, is infeasible for the CVM algorithm.
Instead, we investigate functions of the distributions, which are relevant for the analysis.

There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}. 
However, these methods, do not extend to algorithms with branches, or in general, which contain discrete operations.

Verification of randomized algorithms in general has been done by many authors~\cite{bosshard2024, eberl2020, karayel2022, karayel2023, Probabilistic_Prime_Tests-AFP, tan2024}.

### 2025-02-10 paper/sec/formalization.tex
Paper: More on the sec:formalization   + Fix formattign issue

@@
+\section{Formalization of the CVM algorithm}\label{sec:formalization}
+The supplementary material contains a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
+
+A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
+%The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
+We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
+The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
+Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
+Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
+Nevertheless, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the listing.
+%\medskip
+\begin{table}
+\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
+\noindent\begin{tabular}{l p{9cm}}
+\toprule
+Term & Description \\
+\midrule
+\isa{card\ S} & Cardinality of a finite set $S$. \\
+%\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
+%\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
+\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
+\isa{nat} & Type of natural numbers (non-negative integers). \\
+\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
+\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
+\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
+\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
+\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
+%\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
+%\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
+\bottomrule
+\end{tabular}
+\end{table}
+%\medskip
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expresed in the following snippet:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ correctness{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}
+\end{isabelle_cm}
+The first line establish the conditions on $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
+The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
+This is a bit more general than the original algorithm, where we fix $n$ to be the smallest integer larger than the same value.
+Then we define $A$ to be the cardinality of the set of elements in sequence \isa{xs}:
+More precisely \isa{set\ xs} denotes the set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
+The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$.
+
+Similarly the unbiasedness is also a theorem:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probabibiltiy space \isa{M}.
+
+The formalization contains four theory files.
+The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
+Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
+as most of the proof is identical.
+The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
+The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
+It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
+The total number of lines required for the verification of the original algorithm is 1004 lines.
+This is in contrast to the original proof, for which the verification required \todo{x} lines.
 \begin{algorithm}[t]
@@
 \ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}algo\ xs\ {\isacharequal}{\kern0pt}\ map{\isacharunderscore}{\kern0pt}pmf\ estimate\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}steps\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\isanewline
-\isacommand{theorem}\isamarkupfalse%
-\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\isanewline
-\isacommand{theorem}\isamarkupfalse%
-\ correctness{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}\isanewline
+%\isanewline
+%\isacommand{theorem}\isamarkupfalse%
+%\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\isanewline
+%\isacommand{theorem}\isamarkupfalse%
+%\ correctness{\isacharcolon}{\kern0pt}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}\isanewline
+{\normalfont [\dots]}\isanewline
 \isacommand{end}
@@
 \end{algorithm}
-\section{Formalization of the CVM algorithm}\label{sec:formalization}
-The supplementary material contains a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
-
-A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
-The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
-We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
-The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
-Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
-Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
-For readers not familiar with Isabelle, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the listing.
-%\medskip
-\begin{table}
-\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
-\noindent\begin{tabular}{l p{9cm}}
-\toprule
-Term & Description \\
-\midrule
-\isa{card\ S} & Cardinality of a finite set $S$. \\
-\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
-\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
-\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
-\isa{nat} & Type of natural numbers (non-negative integers). \\
-\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
-\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
-\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
-\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
-\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
-\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
-\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
-\bottomrule
-\end{tabular}
-\end{table}
-%\medskip
-The formalization contains four theory files.
-The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
-Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
-as most of the proof is identical.
-The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
-The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
-It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
-The total number of lines required for the verification of the original algorithm is 1004 lines.
-This is in contrast to the original proof, for which the verification required \todo{x} lines.

### 2025-02-10 paper/sec/negdep.tex
Paper: More on the sec:formalization   + Fix formattign issue

@@
 \section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}
+\begin{algorithm}[t!]
+	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
+  \For{$i \gets 1$ to $l$}
+    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
+    \If{$b$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$ subset of $\chi$
+      \State $p \gets pf$
+    \EndIf
+  \EndFor
+  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}%
 Sometimes, sets of random variables possess a property called negative association.
@@
 The resulting algorithm is depicted in \cref{alg:cvm_new}.
-\begin{algorithm}[h]
-	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
-	\begin{algorithmic}[1]
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
-  \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
-    \If{$b$}
-      \State $\chi \gets \chi \cup \{a_i\}$
-    \Else
-      \State $\chi \gets \chi - \{a_i\}$
-    \EndIf
-    \If{$|\chi| = n$}
-      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$ subset of $\chi$
-      \State $p \gets pf$
-    \EndIf
-  \EndFor
-  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
-  \end{algorithmic}
-\end{algorithm}%
 Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.

### 2025-02-13 paper/sec/intro.tex
draft more in background
add more intuitive desc on line 8 and 11
Paper:   Add a line clarifying that the new variant would not possible with the classic transformation-based route.
edit intro somewhat

@@
     \EndIf
-    \If{$|\chi| = n$}
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
       \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
@@
     \If{$|\chi| = n$}
-      \Return $\bot$ \Comment fail if $\chi$ remains too large
+      \Return $\bot$ \Comment fail if $\chi$ remains full
     \EndIf
@@
 The original error serves as a side motivation for this work.}
-For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
-The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the estimation results for the original algorithm can be deduced.
+For example, in line 3 the Bernoulli distribution is sampled for the value $p$, which itself depends on previous random operations; similarly, the subsampling step for the buffer state $\chi$ in line 9 is only applied if the buffer is full, which also depends on previous random operations.
+The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the estimation results for the original algorithm can be deduced.
 To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods.
@@
 In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
-Our approach is very similar to how deterministic algorithms are verified using a loop invariant.
+Our approach is very similar to how deterministic algorithms are verified using loop invariants.
 The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
@@
 For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
-We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to ideas from interactive theorem proving.
+We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to ideas from mechanized theorem proving.
 
@@
 
-To show that our new technique is more general, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$, instead of independently deciding for each element, whether to keep it.
+To show the generality of our new technique, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
@@
 
-To analyze the new variant, we use results from the theory of negatively dependent random variables~\cite{joagdev1983} to establish the desired functional invariant.
+The modified subsampling step leads to additional dependence for the elements in $\chi$, which cannot be readily removed using transformations following the original proof.
+Instead, we readily verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
 The concept of negative association is a generalization of independence; negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
@@
 %This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.
-
-% TODO: move it into explanation of the technique
-%Note that we are able to establish tail bounds using these invariants, which has, so far, not been possible using a simple loop invariant for randomized algorithms.
-%The loop invariants are established, essentially, using the key property of the Giry monad:
-%\[
-%  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f (x)} [h] \, d x \textrm{.}
-%\]
-%Here $m \isa{\isasymbind} f$ denotes distribution of a randomized algorithm, which represents the sequential composition of $m$ with $f$.
 

### 2025-02-13 paper/sec/semra.tex
minor changes in sec 2

@@
 
-The key idea is to model a randomized algorithm as a probability space representing the distributions of its result.
+The key idea is to model a randomized algorithm as a probability space representing the distribution of its results.
 For simplicity, consider \cref{alg:example}.
@@
 
-In the first step the algorithm flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise. (The notation $\Ber(p)$ represents the Bernoulli distribution.)
-In the second step the algorithm flips a coin $q$ which depends on $p$.
-This has the consequence, that we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
-The resulting distribution for $q$ is a compound distribution, a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
+In the first step,~\cref{alg:example} flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise; the notation $\Ber(p)$ represents the Bernoulli distribution.
+In the second step, the algorithm flips a coin $q$ which depends on $p$.
+This has the consequence that we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+The resulting distribution for $q$ is a \emph{compound distribution}, a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
 
@@
 \begin{description}
-\item[Primitive Random Operations.] For example, a simple fair coin flip is represented using the Bernoulli distribution $\Ber(\frac{1}{2})$.
+\item[Primitive Random Operations.] For example, a simple fair coin flip is represented using the Bernoulli distribution, $\Ber(\frac{1}{2})$.
 \item[Return Combinator.]

### 2025-02-13 paper/sec/invariants.tex
some editing in Sec 3

@@
-\section{Probabilistic Invariants}\label{sec:invariants}
+\section{Functional Probabilistic Invariants}\label{sec:invariants}
 In this section, we will derive our new technique using \cref{alg:cvm} as an example.
-
-Let us briefly, review the algorithm. The algorithm's state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
-The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$.
-The size is chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
-The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely -- if the current stream element is already in the buffer -- removing them with probability $(1-p)$.
+Let us start by briefly reviewing the algorithm---its state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
+The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
+The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing them with probability $(1-p)$.
 If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
 If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returing $\bot$.
-After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as an approximation of the number of distinct elements in the stream.
-This output is probably-approximately correct, i.e., the probability that the relative error of $\frac{|\chi|}{p}$ exceeds $\varepsilon$ is at most $\delta$.
+After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct value for the number of distinct elements in the stream.
 
-For the discussion in this section, it is best to consider \cref{alg:cvm} without line 11, i.e., we are going to ignore the second check whether the subsampling step succeeded, instead the algorithm continues even though the buffer may still contain $n$ elements.
-The total variational distance between these two variants of the algorithms is at most $\frac{\delta}{2}$.
-This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction of $\frac{\delta}{2}$.
-Considering the modified version, makes the discussion easier, because we do not have to worry about $\bot$.
+\subsection{A Minor Transformation}
+It is convenient to analyze \cref{alg:cvm} without line 11, i.e., we will skip the second check of $|\chi|=n$, determining whether the subsampling step succeeded.
+This modified version simplifies our subsequent analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
+This transformation is identically used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
+This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
 
-Let us consider the random variables $X_s$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer: $X_s := I(s \in \chi)$.\footnote{We use $I$ for the indicator of a predicate, i.e., $I(\mathrm{true}) = 1$ and $I(\mathrm{false}) = 0$.}
+\subsection{Deriving A Simple Probabilistic Invariant}
+Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer.\footnote{We use $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.}
 Before the algorithm encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far.
@@
 
-In the loop iteration, where $s$ occurs the first time, it will be inserted with a probability $p$ in lines 3--7.
-This means, after line 7, we have 
+In the loop iteration where $s$ occurs the first time, it will be inserted with a probability $p$ in lines 3--7.
+This means, after line 7, we have:
 \begin{equation}
@@
 \end{equation}
-Interestingly, the equation is preserved throughout the entire algorithm.
-Indeed, let us consider a subsampling step.
-The probability of $\prob(X_s=1)$ is being halved, but so will $p$, which preserves the equation.
+Interestingly, the equation is preserved for the rest of the algorithm.
+For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so will $p$, which preserves the equation.
 
-While it is easy see that this is true informally, and since we are going to build on this, let us see how we can verify \cref{eq:indicator_eq} more formally.
-
-For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ represeting the buffer $\chi$ and a real number representing $p$.
-Note that, this means $p$ and $\chi$ are random variables, with respect to distributions of the state of the algorithm.
-We think of the two parts of each loop iteration lines 3--7 (resp. lines 8--10) as step 1 (resp. step 2).
-The state of the algorithm can be represented as the distribution resulting from the sequential composition of alternating steps:
+Let us see how we can verify \cref{eq:indicator_eq} more formally.
+For that, we model the state of the algorithm as a pair $(\chi,p)$, so the algorithm computes a distribution over the random variables $\chi$ and $p$.
+%For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ representing the buffer $\chi$ and a real number representing $p$.
+%Note that, this means $p$ and $\chi$ are random variables, with respect to distributions of the state of the algorithm.
+We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for lines 3--7 (resp.~lines 8--10).
+The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps:
 \[
@@
 \]
-where we parameterize $\mathrm{step}_1$ with the stream element it processes.
-The term $\mathrm{init}$ represents the intial state, i.e., $\mathrm{init} = \mathrm{return} (\{\},1)$.
+where we parameterize $\mathrm{step}_1$ with the stream element that it processes.
+The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return} (\{\},1)$.
+It is easy to show by induction over the sequence of steps, we will have have $0 < p \leq 1$ and $\chi \subseteq A$.
 
-It is not hard to show by induction, that for all the states occuring above, we will have have $0 < p \leq 1$ and $\chi \subseteq A$, so we will assume this is the case for all probability spaces representing states in the following.
-
-Let us consider verifying that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
+Similarly, let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
+To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
 \[
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
-    \int_\Omega \int_{\Ber(p)} p^{-1} I\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
+    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
 \]
 Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these depend on $\sigma$.
-To see, that the right-hand-side is equal to $1$, it is useful to consider the cases, where $a=s$ and the converse seperately.
-In the first case the right-hand-side is equal to $1$, can be seen directly, note we assume $p \in (0;1]$.
+To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
+In the first case the right-hand-side is equal to $1$, can be seen directly (since $p \in (0;1]$).
 In the second case it follows from the induction hypothesis.
@@
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
-    \int_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{I(s \in \tau)}{p/2} \, d \tau\right)}{\frac{I(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
+    \int_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{\indicat(s \in \tau)}{p/2} \, d \tau\right)}{\frac{\indicat(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
 \]
-Note that the true- and false-case both evaluate to the same value: $p^{-1} I(s \in \chi)$.
+Note that the true- and false-case both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
 If $s \notin \chi$ both sides are $0$, because the subsampling operation returns a subset of $\chi$.
-If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} I(s \in \chi)$.
+If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
 Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
+This completes the invariance proof.
 
-Now, with the above it is straight-forward to derive that the expectation of the result $p^{-1} |\chi|$ of our modified algorithm (without line 11) is exactly the desired cardinality $|A|$.
-Our main goal, however, is to show that the algorithm is probably approximately correct, i.e., that the probability that the relative error of the output exceeds $\varepsilon$ is less than $\delta$.
+\subsection{Deriving A Functional Probabilistic Invariant}
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$.
+However, we want to show that the algorithm is probably-approximately correct.
+Normally, we would use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
+However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$, are dependent.
 
-Normally, we would use Chernoff bounds for indepenent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
-But they are not useful in our case, because the random variables in the case of the CVM algorithm, for example $p^{-1} X_s$, are not independent.
 An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
 It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
-Note that, sometimes the moment generating function is only defined for some values of $t$ (when the corresponding integral exists).
-This is typically an interval including $0$.
-In which case the inequality is only applicable for those values.
-Because the probability space, representing the state of the CVM algorithm is finite, in our case $M(t)$ will always be defined.
-Of course, it is also possible to obtain lower tail bounds $P(X \leq a)$ which just requires estimates for the $M(t)$ for $t < 0$, instead of $t > 0$.
+Note that, the moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.
+Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
+It is also possible to obtain lower tail bounds $P(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
 
-This means for our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
+In our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
 \[
@@
 for $h(x) = \exp(tx)$.
-At this point, it was of course tempting to see, whether the proof for \cref{eq:indicator_eq} can also be extended to establish bounds for the above.
-When we tried, we got the following result:
+At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be extended to establish bounds for the above.
+A possible invariant is the following:
 \[
@@
 for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
-Note that the exponetial function is convex.
+We call this a \emph{functional} probabilistic invariant because one can it for all valid choices of $h$ with a single induction over the randomized algorithm.
+
+%TODO: need to read below
+Note that the exponential function is convex.
 So this does not work directly.

### 2025-02-13 paper/sec/invariants.tex
edit sec 3

@@
 \subsection{Deriving A Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$.
-However, we want to show that the algorithm is probably-approximately correct.
-Normally, we would use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$---but we are interested the algorithm's probably-approximately correctness.
+A typical approach is to use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
 However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$, are dependent.
@@
 An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
-It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t)$ is the moment generating function of $X$: $M(t) := \expect [\exp(t X)]$.
-Note that, the moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.
+It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
+Note that the moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.
 Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
@@
 At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be extended to establish bounds for the above.
-A possible invariant is the following:
+A generalized invariant is the following:
 \[
@@
 for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
-We call this a \emph{functional} probabilistic invariant because one can it for all valid choices of $h$ with a single induction over the randomized algorithm.
+We call this a \emph{functional} probabilistic invariant because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.
 
-%TODO: need to read below
-Note that the exponential function is convex.
-So this does not work directly.
-However, when we instead try to derive tail bounds for the random variable $I(p \geq q) p^{-1} |\chi|$.
+Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
+However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
 We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
-This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which successfully leads to tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
-Of course, we also have to separately estimate $\prob(p < q)$.
-We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}.
-They use $q = \frac{n}{4 |A|}$ and we could use the same cut-off.
+This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
+To use these bounds, we also have to separately estimate $\prob(p < q)$.
+We can use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$ and separately bound the probability $\prob(p < q)$.
 
-Note that the formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof in its appendix.
-The verified algorithm is actually a generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
-The algorithm presented by Chakraborty et al.~\cite{chakraborty2022}, replicated in the introduction (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
-Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are very elementary.
-The main point we wanted to make in this section is that, it is possible to derive useful and general probabilistic invariants, by considering expectations of functions of the state, using recursion or induction over the algorithm itself.
+The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
+The algorithm we have verified is actually a slight generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
+The original CVM algorithm~\cite{chakraborty2022} (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
+Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.
+
+The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of functions of the state, proved using recursion or induction over the algorithm itself.
 As far as we know this method of establishing tail bounds for randomized algorithms is new.
 
-When we look at the subsampling step, our proof imposes the following condition on the subsampling operation:
-\begin{equation}\label{eq:subsample_condition}
-  \int_{\mathrm{subsample}(\chi)} \left(\prod_{s \in S} g(I(s \in \tau)) \, d \tau\right) \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
-\end{equation}
-for all non-negative functions $g$.
-An interesting consequence of that, is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
-We think that this new variant would not be possible to verify, without the new approach presented here.
+%Move to sec 4
+%When we look at the subsampling step, our proof imposes the following condition on the subsampling operation:
+%\begin{equation}\label{eq:subsample_condition}
+%  \int_{\mathrm{subsample}(\chi)} \left(\prod_{s \in S} g(I(s \in \tau)) \, d \tau\right) \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
+%\end{equation}
+%for all non-negative functions $g$.
+%An interesting consequence of that, is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
+%We think that this new variant would not be possible to verify, without the new approach presented here.

### 2025-02-13 paper/sec/intro.tex
edits in sec 4

@@
 \cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle~\cite{nipkow2002}.
-\cref{sec:invariants} introduces our new technique, and shows how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
+\cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
 \cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
 \cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new library on negatively associated random variables.
-In \cref{sec:transformation_based_proof}, we present details about our alternative verification of the CVM using the transformation based proof by Chakraborty et al.
-The last sections conclude with a brief overview of related work and a summary of our results.
+In \cref{sec:transformation_based_proof}, we present details about our alternative verification of the CVM using the transformation based proof by Chakraborty et al.~\cite{chakraborty2023}.
+The final sections conclude with an overview of related work and a summary of our results.
 

### 2025-02-13 paper/sec/invariants.tex
edits in sec 4

@@
 \subsection{Deriving A Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$---but we are interested the algorithm's probably-approximately correctness.
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$; but we are interested the algorithm's probably-approximately correctness.
 A typical approach is to use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
@@
 As far as we know this method of establishing tail bounds for randomized algorithms is new.
-
-%Move to sec 4
-%When we look at the subsampling step, our proof imposes the following condition on the subsampling operation:
-%\begin{equation}\label{eq:subsample_condition}
-%  \int_{\mathrm{subsample}(\chi)} \left(\prod_{s \in S} g(I(s \in \tau)) \, d \tau\right) \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
-%\end{equation}
-%for all non-negative functions $g$.
-%An interesting consequence of that, is that it allowed us to devise a refined version of the CVM algorithm, that is both total and unbiased, which we will present in the following section.
-%We think that this new variant would not be possible to verify, without the new approach presented here.

### 2025-02-13 paper/sec/negdep.tex
edits in sec 4

--- 2025-01-31 paper/sec/negdep.tex
@@
-\section{Negative Dependence and an Unbiased CVM Algorithm\label{sec:negdep}}
+\section{An Unbiased CVM Variant and Negative Dependence\label{sec:negdep}}
 
-Sometimes, sets of random variables possess a property called negative association.
-The concept is a generalization of independence.
-It has similar closure properties as independent random variables.
-More interestingly, the standard Chernoff-Hoeffding bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
+An interesting consequence of our invariant-based approach, is that it allowed us to devise and verify a refined version of the CVM algorithm, that is both total and unbiased.
 
-The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orphant dependence.
+\subsection{Unbiased CVM Variant}
 
-The following preliminary definition is useful, for the definition of the concept:
-\begin{definition}
-For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, e.g., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
-\end{definition}
-
-\begin{definition}[Negative Association]
-A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$ then
-\[
-\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
-\]
-\end{definition}
-
-The following proposition summarizes some of the properties for negatively associated sets of random variables: 
-\begin{proposition}\label{pro:neg_dep_props}
-Summary of results for negatively associated random variables: 
-\begin{enumerate}
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $f_1,..,f_m : \mathbb R^n \rightarrow \mathbb R$ be all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
-\item If $X_1,..,X_n$ are independent then $X_1,...,X_n$ are negatively associated.
-\item A subset of a negatively associated set of random variables is again negatively associated.
-\end{enumerate}
-\end{proposition}
-
-These properties illustrate the trade-off between negative association and independence.
-For example property 3 would be true for independent random variables, even without the condition of monotonicty.
-Of course, on the other hand independence is a stronger property and fewer sets of random variables are indepenent.
-For our discussion the following is an important lemma about negative associated random variables.
-
-\begin{lemma}\label{le:neg_assoc_prod}
-Let $X_1,..,X_n$ be negatively associated and $f_1,..,f_n$ be all non-decreasing or all non-increasing non-negative functions, then
-\[
-  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
-\]
-\end{lemma}
-\begin{proof}
-Follows from the definition of negative associativy (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
-\end{proof}
-
-The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
-Why we are interested in the lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
-The latter is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
-\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
-However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
-Intuitively, this is interesting because it opens up an avenue to modifying the subsampling step in \cref{alg:cvm}.
-Instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, which has the disadvantage that there is a small possibility that none of the elements gets discarded, we could instead choose a random $nf$-subset, where $\frac{1}{2} \leq f < 1$, such that $nf$ is integer, in the subsampling step.
-For such a subsampling step, we can derive using \cref{le:neg_assoc_prod}:
-\begin{equation}\label{eq:subsample_with_n_subsets}
-  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(I(s \in \chi)) \leq \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(I(s \in \chi)) = \prod_{s \in S}  \expect_{\Ber(f)} g \textrm{.}
-\end{equation}
-for any non-negative $g$ and $S \subseteq \chi$. Note that $g$ is only evaluated at two points and is thus either non-increasing or non-decreasing.
-Also, if $S$ is a singleton, the inequality becomes an equality.
-
-Let us recall that this is the exact condition we needed for the subampling step (\cref{eq:subsample_condition}).
-This means the proof for the modified algorithm in \cref{sec:invariants} works just as well with the new subsampling method.
-
-The resulting algorithm is depicted in \cref{alg:cvm_new}.
-\begin{algorithm}[t]
-	\caption{New unbiased and total CVM algorithm.}\label{alg:cvm_new}
+\begin{algorithm}
+	\caption{New unbiased and total CVM algorithm variant.}\label{alg:cvm_new}
 	\begin{algorithmic}[1]
@@
     \EndIf
-    \If{$|\chi| = n$}
-      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment Select a random $nf$ subset of $\chi$
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$-subset of $\chi$
       \State $p \gets pf$
@@
 \end{algorithm}%
-Note that it is possible to just choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
-The key advantage is that there is no line 11 and the algorithm never returns $\bot$.
-And based on our discussion from the previous section it is also an unbiased algorithm, i.e., the expectation of the result is exactly $|A|$.
+
+When we look at the subsampling step of~\cref{alg:cvm}, our invariant imposes the following condition on the subsampling operation:
+\begin{equation}\label[ineq]{i:subsample_condition}
+  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau)) \, d \tau \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
+\end{equation}
+for all non-negative functions $g$.
+
+Motivated by this, our new variant is shown in \cref{alg:cvm_new}.
+For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.\footnote{It is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.}
+Since this new subsampling step always reduces the size of $\chi$, the variant never returns $\bot$, i.e., it is \emph{total}.
+The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
+These depend crucially on establishing~\cref{i:subsample_condition} for the new subsampler, for which we need a new concept.
+
+\subsection{Background on Negative Dependence}
+Some sets of random variables possess a property called \emph{negative association}, a generalization of independence.
+The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orphant dependence.
+Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
+Negative association is defined as follows:
+\begin{definition}
+For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
+\end{definition}
+
+\begin{definition}[Negative Association]\label{def:neg_assoc}
+A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
+\[
+\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
+\]
+\end{definition}
+
+The following proposition summarizes some of the properties for negatively associated sets of random variables.
+\begin{proposition}\label{pro:neg_dep_props}
+Summary of results for negatively associated random variables.
+\begin{enumerate}
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $f_1,..,f_m : \mathbb R^n \rightarrow \mathbb R$ be all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
+\item If $X_1,..,X_n$ are independent then $X_1,...,X_n$ are negatively associated.
+\item A subset of a negatively associated set of random variables is again negatively associated.
+\end{enumerate}
+\end{proposition}
+
+These properties illustrate the trade-off between negative association and independence.
+For example, property 3 would be true for independent random variables, even without the condition of monotonicity.
+%Of course, on the other hand independence is a stronger property and fewer sets of random variables are indepenent.
+To analyze our new subsampler, the following is an important lemma about negative associated random variables.
+
+\begin{lemma}\label{le:neg_assoc_prod}
+Let $X_1,..,X_n$ be negatively associated and $f_1,..,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
+\[
+  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
+\]
+\end{lemma}
+\begin{proof}
+This follows from the definition of negative associativity (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
+\end{proof}
+
+The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
+The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
+This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
+\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
+However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
+Thus, for the new subsampling step in line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
+%TODO: double check!
+\begin{equation}\label{eq:subsample_with_n_subsets}
+  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau))\,d\tau \leq
+  \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(\indicat(s \in \tau))\,d\tau = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
+\end{equation}
+for any non-negative $g$ and $S \subseteq \chi$.
+Note that the domain of $g$ has two values, so it is either non-increasing or non-decreasing.
+Also, if $S$ is a singleton, the inequality becomes an equality.

### 2025-02-13 paper/sec/formalization.tex
Paper: Minor edits.

@@
 as most of the proof is identical.
-The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{eq:subsample_condition}.
+The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
 The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).

### 2025-02-13 paper/sec/intro.tex
Paper: Minor edits.

@@
 The original error serves as a side motivation for this work.}
-For example, in line 3 the Bernoulli distribution is sampled for the value $p$, which itself depends on previous random operations; similarly, the subsampling step for the buffer state $\chi$ in line 9 is only applied if the buffer is full, which also depends on previous random operations.
-The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the estimation results for the original algorithm can be deduced.
+For example, in Line~3 the Bernoulli distribution is sampled for the value $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
+The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
 To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods.
@@
 
-The main contributions of this work are:
-\begin{itemize}
-\item Introduction of a new technique using functional probabilistic invariants to verify randomized algorithms inductively/recursively.
-\item Verification of the original CVM algorithm using our new technique.
-\item Presentation and verification of a new variant of CVM that is total and unbiased.
-\item Formalization of a theory of negatively associated random variables used to analyze the new CVM variant.
-\end{itemize}
-
 The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof (which we also mechanized) required \todo{x} lines.
@@
 It should be stressed that the theory of negatively associated RVs is orthogonal to our new technique, but the formalization of such a theory is also a contribution of this work.
+
+% After re-reading. This summary (especially point 3/4), feels very weird because the new variant/or neg. association has not been mentioned before.
+The main contributions of this work are:
+\begin{itemize}
+\item Introduction of a new technique using functional probabilistic invariants to verify randomized algorithms inductively/recursively.
+\item Verification of the original CVM algorithm using our new technique.
+\item Presentation and verification of a new variant of CVM that is total and unbiased.
+\item Formalization of a theory of negatively associated random variables used to analyze the new CVM variant.
+\end{itemize}
 

### 2025-02-13 paper/sec/invariants.tex
Paper: Minor edits.

@@
 The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
-The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing them with probability $(1-p)$.
+The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$.
 If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
@@
 
-In the loop iteration where $s$ occurs the first time, it will be inserted with a probability $p$ in lines 3--7.
-This means, after line 7, we have:
+In the loop iteration where $s$ occurs the first time, it will be inserted with a probability $p$ in Lines 3--7.
+This means, after Line 7, we have:
 \begin{equation}
@@
 Let us see how we can verify \cref{eq:indicator_eq} more formally.
-For that, we model the state of the algorithm as a pair $(\chi,p)$, so the algorithm computes a distribution over the random variables $\chi$ and $p$.
+For that, we model the state of the algorithm as a pair $(\chi,p)$.
+Note that this means $\chi$ and $p$ are random variables on the distribution of the state of the algorithm. 
 %For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ representing the buffer $\chi$ and a real number representing $p$.
@@
 
-Similarly, let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
+Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
 To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
@@
 It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
-Note that the moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.
-Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
+%\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
+%Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
 It is also possible to obtain lower tail bounds $P(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
@@
 At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be extended to establish bounds for the above.
-A generalized invariant is the following:
+We could establish the following result:
 \[

### 2025-02-13 paper/sec/semra.tex
Paper: Minor edits.

@@
 
-Importantly, the CVM algorithm is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output estimate $R$ satisfies $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$ for parameters $\varepsilon$ and $\delta$.
+Importantly, the CVM algorithm is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output estimate $R$ exceeds $\varepsilon$ is at most $\delta$, i.e.., $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$ for parameters $\varepsilon$ and $\delta$.
+% I think here it's good to have both a text and formal description.
 Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
@@
 The key idea is to model a randomized algorithm as a probability space representing the distribution of its results.
-For simplicity, consider \cref{alg:example}.
+Let us consider \cref{alg:example}.
 \begin{algorithm}[h!]

### 2025-02-14 paper/sec/formalization.tex
fixed some typos

@@
 %\medskip
-The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expresed in the following snippet:
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expressed in the following snippet:
 \begin{isabelle_cm}
@@
 \end{isabelle_cm}
-The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probabibiltiy space \isa{M}.
+The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
 
@@
 The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
-It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
+It should be noted that only the last file, for the new unbiased algorithm, depends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
 The total number of lines required for the verification of the original algorithm is 1004 lines.

### 2025-02-14 paper/sec/formalization_neg_dep.tex
fixed some typos

@@
 The formalization contains the standard Chernoff bounds for negatively associated random variables.
-This includes the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multipicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
+This includes the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
 Although, these results are not needed for the proof of the CVM algorithm.
@@
 \]
-is true, if $f$ and $g$ are both montone, or both antimontone.
-On the other hand the reverse inequality is true, if $f$ is montone and $g$ is antimontone, or vice versa.
+is true, if $f$ and $g$ are both monotone, or both antimonotone.
+On the other hand the reverse inequality is true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
 Using our parameterized relation symbol, we could state all variants in a concise manner.

### 2025-02-14 paper/sec/negdep.tex
fixed some typos

@@
 Some sets of random variables possess a property called \emph{negative association}, a generalization of independence.
-The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orphant dependence.
+The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orthant dependence.
 Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
@@
 For example, property 3 would be true for independent random variables, even without the condition of monotonicity.
-%Of course, on the other hand independence is a stronger property and fewer sets of random variables are indepenent.
+%Of course, on the other hand independence is a stronger property and fewer sets of random variables are independent.
 To analyze our new subsampler, the following is an important lemma about negative associated random variables.

### 2025-02-14 paper/sec/original_proof.tex
fixed some typos

@@
 Moreover, during the subsampling operation those elements are kept, whose $k+1$-th associated coin flip is $1$.
-This again preserves the behaviour, that each element is discared independently with probability $1/2$.
+This again preserves the behaviour, that each element is discarded independently with probability $1/2$.
 The operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
@@
 
-In \cref{alg:cvm_simul} we describe the new algorithm, that behaves exactly as the original one (\cref{alg:cvm}), but performs indepenent coin flips.
-Note that the function last{\textunderscore}index returns the index of the last-occurence of an element in the sequence, before the current loop iteration.
+In \cref{alg:cvm_simul} we describe the new algorithm, that behaves exactly as the original one (\cref{alg:cvm}), but performs independent coin flips.
+Note that the function last{\textunderscore}index returns the index of the last-occurrence of an element in the sequence, before the current loop iteration.
 It should be noted that the algorithm keeps track of the number of subsampling iterations $k$, instead of the value $p = 2^{-k}$ as the original algorithm does.
@@
 To explain how this works.
-Let us first define the \emph{sampling} function, i.e., let $f$ be a function that depends on coin flips, more precisely $f$ takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to indepenent unbiased coin flips as.
+Let us first define the \emph{sampling} function, i.e., let $f$ be a function that depends on coin flips, more precisely $f$ takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to independent unbiased coin flips as.
 \begin{isabelle_cm}
@@
 By recursively applying the rule, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
-Note that, we arrive at the orignal version of the algorithm, that performs the coin flips lazily.
+Note that, we arrive at the original version of the algorithm, that performs the coin flips lazily.
 

### 2025-02-15 paper/sec/intro.tex
tweak intro

@@
 The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
-To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods.
+To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods. %TODO: perhaps cite a textbook or two?
 
@@
 By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
-For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
-We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to ideas from mechanized theorem proving.
-
-The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof (which we also mechanized) required \todo{x} lines.
-We detail some of the challenges faced when mechanizing the transformation arguments used by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} in \cref{sec:transformation_based_proof}.
+%For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
+We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.
 
@@
 The modified subsampling step leads to additional dependence for the elements in $\chi$, which cannot be readily removed using transformations following the original proof.
-Instead, we readily verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
+Instead, we verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
 The concept of negative association is a generalization of independence; negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
@@
 % After re-reading. This summary (especially point 3/4), feels very weird because the new variant/or neg. association has not been mentioned before.
-The main contributions of this work are:
+In summary, the main contributions of this work are:
 \begin{itemize}
@@
 
+We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
+As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
+
+The organization of this paper is as follows.
 \cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle~\cite{nipkow2002}.
@@
 \cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
-\cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new library on negatively associated random variables.
-In \cref{sec:transformation_based_proof}, we present details about our alternative verification of the CVM using the transformation based proof by Chakraborty et al.~\cite{chakraborty2023}.
+\cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new formalized library on negatively associated random variables.
+In \cref{sec:transformation_based_proof}, we present some challenges faced in our alternative verification of the CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
 The final sections conclude with an overview of related work and a summary of our results.

### 2025-02-15 paper/sec/semra.tex
edit background slightly

@@
 
-Importantly, the CVM algorithm is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output estimate $R$ exceeds $\varepsilon$ is at most $\delta$, i.e.., $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$ for parameters $\varepsilon$ and $\delta$.
-% I think here it's good to have both a text and formal description.
+Importantly, CVM is a \emph{probably-approximately correct} algorithm where its output estimate $R$ satisfies
+$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+for parameters $\varepsilon$ and $\delta$,
+i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
 Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.

### 2025-02-16 paper/sec/intro.tex
Paper: Add references to intro.

@@
 The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
-To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods. %TODO: perhaps cite a textbook or two?
+To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods~\cite{alon2000,mitzenmacher2017,motwani1995}.
 
@@
 
-To show the generality of our new technique, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
-This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
+To show the generality of our new technique, we introduce a new variant of the CVM algorithm, where the subsampling step in Line~9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
+This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
 More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property, that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
@@
 
-% After re-reading. This summary (especially point 3/4), feels very weird because the new variant/or neg. association has not been mentioned before.
 In summary, the main contributions of this work are:

### 2025-02-18 paper/sec/abstract.tex
WIP: Paper   * Saving current state
take a stab at abstract

--- 2025-01-24 isabelle/CVM_Subsample/document/root.tex
@@
-\documentclass[11pt,a4paper]{article}
-\usepackage{isabelle,isabellesym}
-\usepackage{amsfonts,amsmath, amssymb,amsthm}
-\usepackage[top=1.4in, bottom=1.4in, left=1.4in, right=1.4in]{geometry}
+\begin{abstract}
+Estimating the number of distinct elements in a data stream is a classical problem with numerous applications in computer science.
+We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (the so-called CVM algorithm).
+Their algorithm deviated considerably from the state-of-the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close to optimal logarithmic space complexity.
 
-\usepackage[only,bigsqcap]{stmaryrd}
-
-\usepackage{marginnote}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
-\let\oldsection\section
-\renewcommand\section{\clearpage\oldsection}
-
-\newcommand{\getsr}{\xleftarrow{\$}}
-
-\usepackage{mathtools}
-\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
-\DeclareMathOperator{\prob}{\mathcal P}
-\DeclareMathOperator{\expect}{\mathbb E}
-\DeclareMathOperator{\Ber}{\mathrm{Ber}}
-\DeclareMathOperator{\indicator}{\mathrm{I}}
-
-\newcommand{\eps}{\varepsilon}
-\renewcommand{\phi}{\varphi}
-\newtheorem{theorem}{Theorem}
-\theoremstyle{definition}
-\newtheorem{lemma}{Lemma}
-
-% this should be the last package used
-\usepackage{pdfsetup}
-
-% urls in roman style, theory text in math-similar italics
-\urlstyle{rm}
-\isabellestyle{it}
-
-\begin{document}
-
-\title{Verification of the CVM algorithm with a new Recursive Analysis Technique}
-\author{Emin Karayel, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan,\\and Joe Watt}
-\maketitle
-
-\begin{abstract}
-In 2022, Chakraborty et al.~\cite{chakraborty2022} published a streaming 
-algorithm (henceforth, the CVM algorithm) for the distinct
-elements problem, that deviated considerably from the state-of-the art, due to its simplicity
-and avoidance of standard derandomization techniques, while still maintaining a close to optimal
-logarithmic space complexity.
-
-In this entry, we verify the CVM algorithm's correctness using a new technique which simplifies
-the formal analysis considerably compared to the orignal proof by Chakraborty et
-al. The main idea is based on a probabilistic invariant that allows us to derive concentration using
-Cram\'{e}r Chernoff method.
-
-This new technique opens up the possible algorithm design space, and we introduce a new variant of the 
-CVM algorithm, that is total, and also has an additional property in addition 
-to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
-the desired result. The latter is also a new property, that neither the original CVM algorithm
-nor classic algorithms for the distinct elements problem posses.
+Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method.
+This simplifies the formal analysis considerably compared to the original transformation-based proof.
+Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying with a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
+In this paper, we introduce the proof technique, describe its use in Isabelle/HOL to mechanize both versions of the CVM algorithm, and present a supporting formalized library on negative association used to verify the latter variant.
 \end{abstract}
-
-\tableofcontents
-
-\parindent 0pt\parskip 0.5ex
-
-\input{session}
-
-\bibliographystyle{abbrv}
-\bibliography{root}
-
-\appendix
-\input{appendix}
-
-\end{document}

### 2025-02-18 paper/sec/abstract.tex
slight edit

@@
 Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying with a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
-In this paper, we introduce the proof technique, describe its use in Isabelle/HOL to mechanize both versions of the CVM algorithm, and present a supporting formalized library on negative association used to verify the latter variant.
+In this paper, we introduce the proof technique, describe its use in Isabelle/HOL to mechanize both versions of the CVM algorithm, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
 \end{abstract}

### 2025-02-18 paper/sec/abstract.tex
swap order of some words

--- 2025-02-18 paper/sec/abstract.tex
@@
 Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying with a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
-In this paper, we introduce the proof technique, describe its use in Isabelle/HOL to mechanize both versions of the CVM algorithm, and present a supporting formalized library on negative association used to verify the latter variant.
+In this paper, we introduce the proof technique, describe its use to mechanize both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
 \end{abstract}

### 2025-02-18 paper/sec/invariants.tex
fixed some typos
quick pass over 2-3

@@
 The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
-The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$.
-If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$.
-If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returing $\bot$.
+The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$ (Lines 3--7).
+If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$ (Lines 8--10).
+If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returning $\bot$ (Line 11).
 After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct value for the number of distinct elements in the stream.
@@
 \subsection{A Minor Transformation}
-It is convenient to analyze \cref{alg:cvm} without line 11, i.e., we will skip the second check of $|\chi|=n$, determining whether the subsampling step succeeded.
+It is convenient to analyze \cref{alg:cvm} without Line 11, i.e., we will skip the second check of $|\chi|=n$ determining whether the subsampling step succeeded.
 This modified version simplifies our subsequent analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
@@
 \end{equation}
-Interestingly, the equation is preserved for the rest of the algorithm.
-For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so will $p$, which preserves the equation.
+Interestingly, this equation is preserved for the rest of the algorithm.
+For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so will $p$ after subsampling, which preserves the equation.
 
-Let us see how we can verify \cref{eq:indicator_eq} more formally.
-For that, we model the state of the algorithm as a pair $(\chi,p)$.
-Note that this means $\chi$ and $p$ are random variables on the distribution of the state of the algorithm. 
+Let us explain how we can verify \cref{eq:indicator_eq} more formally.
+For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
 %For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ representing the buffer $\chi$ and a real number representing $p$.
 %Note that, this means $p$ and $\chi$ are random variables, with respect to distributions of the state of the algorithm.
-We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for lines 3--7 (resp.~lines 8--10).
+We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for Lines 3--7 (resp.~Lines 8--10).
 The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps:
@@
 The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return} (\{\},1)$.
-It is easy to show by induction over the sequence of steps, we will have have $0 < p \leq 1$ and $\chi \subseteq A$.
+It is easy to show by induction over the sequence of steps, we will have $0 < p \leq 1$ and $\chi \subseteq A$ for all possible states of the algorithm.
 
@@
 \]
-Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these depend on $\sigma$.
+Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
 To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
-In the first case the right-hand-side is equal to $1$, can be seen directly (since $p \in (0;1]$).
-In the second case it follows from the induction hypothesis.
+In the first case ($a = s$), the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
+In the second case ($a \not= s$), it follows from the induction hypothesis on $\Omega$.
 (In particular, the term in the inner integral is constant with respect to $\tau$.)
 
-The same is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
+The same invariant-based argument is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
 Then
@@
 \]
-Note that the true- and false-case both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
-If $s \notin \chi$ both sides are $0$, because the subsampling operation returns a subset of $\chi$.
+Note that the true- and false-cases of the inner if-then-else both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
+If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
 If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
@@
 \subsection{Deriving A Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ of the modified algorithm (without line 11) is equal to the desired cardinality $|A|$; but we are interested the algorithm's probably-approximately correctness.
-A typical approach is to use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
-However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$, are dependent.
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
+However, recall that we are interested in verifying the algorithm's PAC guarantee.
+A typical approach to establishing such a guarantee is to use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
+However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$ for $s \in A$, are dependent.
 
@@
 for $h(x) = \exp(tx)$.
-At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be extended to establish bounds for the above.
-We could establish the following result:
+At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be generalized to establish bounds for the above.
+Indeed, we managed to establish the following result:
 \[
@@
 
-The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of functions of the state, proved using recursion or induction over the algorithm itself.
+The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of classes of functions of the state, proved using recursion or induction over the algorithm itself.
 As far as we know this method of establishing tail bounds for randomized algorithms is new.

### 2025-02-18 paper/sec/semra.tex
quick pass over 2-3

--- 2025-02-13 paper/sec/semra.tex
@@
 
-Importantly, the CVM algorithm is \emph{probably-approximately correct}, i.e., the probability that the relative error of its output estimate $R$ exceeds $\varepsilon$ is at most $\delta$, i.e.., $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$ for parameters $\varepsilon$ and $\delta$.
-% I think here it's good to have both a text and formal description.
+Importantly, CVM is a \emph{probably-approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
+$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+for parameters $\varepsilon$ and $\delta$,
+i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
 Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.

### 2025-02-18 paper/sec/negdep.tex
minor edits in 3 and 4

@@
 
-\begin{algorithm}
+\begin{algorithm}[t!]
 	\caption{New unbiased and total CVM algorithm variant.}\label{alg:cvm_new}
@@
 
-When we look at the subsampling step of~\cref{alg:cvm}, our invariant imposes the following condition on the subsampling operation:
+When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation:
 \begin{equation}\label[ineq]{i:subsample_condition}
@@
 \end{equation}
-for all non-negative functions $g$.
+for all non-negative functions $g$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
 
@@
 The following proposition summarizes some of the properties for negatively associated sets of random variables.
-\begin{proposition}\label{pro:neg_dep_props}
+\begin{proposition}[\cite{joagdev1983}]\label{pro:neg_dep_props}
 Summary of results for negatively associated random variables.

### 2025-02-18 paper/sec/formalization.tex
edit & move some things from formalization section

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-The supplementary material contains a formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants}.
+In this section, we present the formalization of the CVM algorithm in Isabelle/HOL; we mainly discuss the variant from~\cref{sec:negdep} since the details are similar.
+A snippet of the formalization of~\cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
 
-A snippet of the formalization of the new version of the algorithm is presented in \cref{alg:cvm_formalized}.
-%The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds. 
-We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
-The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
-Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
-Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
-Nevertheless, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the listing.
-%\medskip
-\begin{table}
-\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
-\noindent\begin{tabular}{l p{9cm}}
-\toprule
-Term & Description \\
-\midrule
-\isa{card\ S} & Cardinality of a finite set $S$. \\
-%\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
-%\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
-\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
-\isa{nat} & Type of natural numbers (non-negative integers). \\
-\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
-\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
-\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
-\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
-\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
-%\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
-%\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
-\bottomrule
-\end{tabular}
-\end{table}
-%\medskip
-The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expressed in the following snippet:
-\begin{isabelle_cm}
-\isacommand{theorem}\isamarkupfalse%
-\ correctness{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}
-\end{isabelle_cm}
-The first line establish the conditions on $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
-The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
-This is a bit more general than the original algorithm, where we fix $n$ to be the smallest integer larger than the same value.
-Then we define $A$ to be the cardinality of the set of elements in sequence \isa{xs}:
-More precisely \isa{set\ xs} denotes the set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
-The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$.
-
-Similarly the unbiasedness is also a theorem:
-\begin{isabelle_cm}
-\isacommand{theorem}\isamarkupfalse%
-\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}
-\end{isabelle_cm}
-The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
-
-The formalization contains four theory files.
-The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
-Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
-as most of the proof is identical.
-The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
-The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
-It should be noted that only the last file, for the new unbiased algorithm, depends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
-The total number of lines required for the verification of the original algorithm is 1004 lines.
-This is in contrast to the original proof, for which the verification required \todo{x} lines.
-\begin{algorithm}[t]
-\caption{Formalized Version of \cref{alg:cvm_new}}\label{alg:cvm_formalized}
+\begin{figure}[t]
 \begin{isabelle_cm}
@@
 \end{isabelle_cm}
-\end{algorithm}
+\caption{Formalized version of \cref{alg:cvm_new}.}\label{alg:cvm_formalized}
+\end{figure}
+
+%The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds.
+We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
+The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
+Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
+Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
+Nevertheless, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the formalization.
+%\medskip
+\begin{table}
+\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
+\noindent\begin{tabular}{l p{9cm}}
+\toprule
+Term & Description \\
+\midrule
+\isa{card\ S} & Cardinality of a finite set $S$. \\
+%\isa{set\ xs} & For a sequence \isa{xs} the set of elements in the sequence. \\
+%\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
+\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
+\isa{nat} & Type of natural numbers (non-negative integers). \\
+\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
+\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
+\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
+\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
+\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
+%\isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} & Probability of predicate $P$ in the probability space $M$. \\
+%\isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} & Expectation of the random variable $f$ over the probability space $M$. \\
+\bottomrule
+\end{tabular}
+\end{table}
+%\medskip
+
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expressed in the following snippet:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ correctness{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}
+\end{isabelle_cm}
+The first line gives conditions on parameters $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
+The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$; this is slightly more general than the original algorithm, where we fix $n$ to be the smallest integer larger than the same value.
+%Then, let $A$ to be the cardinality of the set of elements in sequence \isa{xs}.
+%More precisely, \isa{set\ xs} denotes the finite set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
+The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for \isa{run-algo}.
+
+Similarly, we have also formalized unbiasedness of the algorithm:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
+
+% TODO: I think we should give more of the details in the formalization, as opposed to giving a glossary of it
+% For example, here are some of the intricate details I would suggest:
+
+% 1) showing a high-level view of the way the invariant is established by induction?
+% 2) describing that we actually used locales and instances to give a general shared argument for both algorithms?
+% 3) we can write that we omit intricate details of the calculations, but some high-level (minor) details can be in one paragraph, e.g., any numerical analysis you needed were dealt with using existing libraries XXX and YYY (e.g., automating derivatives, approximation, etc.)?
+
+%The formalization contains four theory files.
+The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
+%Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
+%as most of the proof is identical.
+%The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
+%The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
+%It should be noted that only the last file, for the new unbiased algorithm, depends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
+%The total number of lines required for the verification of the original algorithm is 1004 lines.
+%This is in contrast to the original proof, for which the verification required \todo{x} lines.
+%

### 2025-02-18 paper/sec/formalization_neg_dep.tex
edit & move some things from formalization section

@@
-\section{Formalization of the concept of negative association}\label{sec:formalization_neg_dep}
+\section{Formalization of a Library for Negative Association}\label{sec:formalization_neg_dep}
 As mentioned in the previous section, the formalization of the new unbiased variant of the CVM algorithm, requires results from the theory of negative association.
-The supplementary material contains a formalization of the concept in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
 

### 2025-02-18 paper/sec/intro.tex
edit & move some things from formalization section

@@
 
+The supplementary material contains:
+\begin{itemize}
+\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants};
+\item formalization of a library for negative association (\cref{sec:negdep}); and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
+\item formalization of CVM following the original proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} (\cref{sec:transformation_based_proof}).
+\end{itemize}
+
 % TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown

### 2025-02-18 paper/sec/related_work.tex
fixed some typos
add some sectioning to related work

@@
 \section{Related Work}\label{sec:related_work}
-It is important to note that there are already good practical solutions for the distinct elements problem.
-The first solution was presented by Flajolet~\cite{flajolet1985} in 1985, however he like many other authors~\cite{flajolet2007,heule2013,pettie2021}.
-make the assumption that a fixed hash function can be regarded as a fully random function.
-Alon et al.~\cite[Section 2.3]{alon1999} presented in 1999 an easy solution, which does not require unmotivated model assumptions.
-Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pair-wise independent family.
-The space complexity of the algorithm is $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) b)$, where we again assume that $b$ is the number of bits required to represent the stream elements.
-This is slightly better than the CVM algorithm (which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$), since we do not have a dependency on $l$.
-Later Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented more and more sophisticated solutions.
-The last one by B\l{}asiok is indeed optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
-Karayel~\cite{karayel2023} presented a version of the latter, that preserves monotonicity and supports the merge-operation, which enables the use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
-It should, however, be noted that the more recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
-What makes the CVM algorithm really unique is the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms do.
+\subsection{Algorithms for the Distinct Elements Problem}
+It is important to note that there are several practical solutions for the distinct elements problem.
+The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013,pettie2021}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
+Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
+Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family; the space complexity of this algorithm is $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) b)$, where we assume that $b$ is the number of bits required to represent the stream elements.
+This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$ since there is no logarithmic dependency on $l$.
 
-The aformentioned hash based algorithms are not unbiased.
+Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented increasingly sophisticated solutions.
+The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
+Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables the use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
+It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
+What makes the CVM algorithm unique is the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
+
+The aforementioned hash based algorithms are not unbiased.
+%TODO: is there a typo? does this mean bounds on the deviation from the mean?
 Flajolet et al.~\cite{flajolet1985} points it out and provides bounds on the bound.
@@
 
+\subsection{Probabilistic Invariants and Formalization}
 As far as we know probabilistic invariants have not been used to establish Gaussian tail-bounds.
@@
 
-There is also research on the (automated) analysis of loop invariants, for probabilisitic loops, using the characteristic function of loops~\cite{batz2023, mciver2005}.
+There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using the characteristic function of loops~\cite{batz2023, mciver2005}.
 This works by establishing the limiting distribution of the state of the loop.

### 2025-02-18 paper/sec/intro.tex
minor edits

--- 2025-02-16 paper/sec/intro.tex
@@
 Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
-Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.
+Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is displayed in its entirety in~\cref{alg:cvm}.
 
@@
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
-More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property, that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 
-The modified subsampling step leads to additional dependence for the elements in $\chi$, which cannot be readily removed using transformations following the original proof.
+The modified subsampling step leads to additional dependence for the elements in $\chi$ which cannot be readily removed using transformations following the original proof.
 Instead, we verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
-The concept of negative association is a generalization of independence; negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
-It should be stressed that the theory of negatively associated RVs is orthogonal to our new technique, but the formalization of such a theory is also a contribution of this work.
+The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
+It should be stressed that the theory of negatively association is orthogonal to our new technique, but its formalization is also a contribution of this work.
 
-In summary, the main contributions of this work are:
+In summary, our main contributions are:
 \begin{itemize}
@@
 \cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
-\cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new formalized library on negatively associated random variables.
-In \cref{sec:transformation_based_proof}, we present some challenges faced in our alternative verification of the CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
+\cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
+In \cref{sec:transformation_based_proof}, we discuss some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
 The final sections conclude with an overview of related work and a summary of our results.
+
+The supplementary material contains:
+\begin{itemize}
+\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants};
+\item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
+\item formalization of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
+\end{itemize}
 

### 2025-02-18 paper/sec/semra.tex
minor edits

@@
 We briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad~\cite{giry1982}.
-A more thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
+A thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
 

### 2025-02-18 paper/sec/invariants.tex
try to explain how inv is used

@@
 Indeed, we managed to establish the following result:
-\[
-  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|}
-\]
-for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
-We call this a \emph{functional} probabilistic invariant because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.
+\begin{align}
+  \expect \left[ \prod_{s \in S} h(p^{-1} X_s) \right] \leq h(1)^{|S|} \label[ineq]{i:func_invariant}
+\end{align}
+for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$ and any subset $S \subseteq A$ of the elements in the stream.
+We call this a \emph{functional probabilistic invariant} because one can establish it for all valid choices of $h$ in a single induction over the randomized algorithm.
 
-Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
-However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
-We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
-This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
+\subsection{Using the Functional Probabilistic Invariant}
+After establishing~\cref{i:func_invariant}, it remains to carry out the tail bounds analysis using the Cram\'er--Chernoff method.
+The formalization accompanying this work contains a detailed step-by-step informal proof in its appendix, so we give brief illustrative details here.
+
+Firstly, note that choosing $S = A$ in~\cref{i:func_invariant} gives a bound of the form needed for the moment generating function.
+However, the exponential function in $M(t)$ is convex, so just establishing the invariant is not the full story.
+Instead, we can derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, which leads to a weaker condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
+This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, giving tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
 To use these bounds, we also have to separately estimate $\prob(p < q)$.
-We can use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$ and separately bound the probability $\prob(p < q)$.
+We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$ and separately bound the probability $\prob(p < q)$.
 
-The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
-The algorithm we have verified is actually a slight generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
-The original CVM algorithm~\cite{chakraborty2022} (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
-Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.
-
+As a simpler concrete example, consider choosing $S = \{s\}$ for any element $s \in A$.
+If we pick $h_1(x) = x$ and $h_2(x) = M+1-x$ (for sufficiently large $M$), we obtain from~\cref{i:func_invariant}:
+\begin{align*}
+\expect \left[ p^{-1} X_s \right] &= \expect \left[ \prod_{s \in S} h_1(p^{-1} X_s) \right] \leq h_1(1)^{|S|} = 1\\
+\expect [M + 1 - p^{-1} X_s] &= \expect \left[ \prod_{s \in S} h_2(p^{-1} X_s) \right] \leq h_2(1)^{|S|} = M
+\end{align*}
+This implies the simple invariant $\expect \left[ p^{-1} X_s \right] = 1$ (and thus, unbiasedness) without needing a separate induction over the algorithm.
 The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of classes of functions of the state, proved using recursion or induction over the algorithm itself.

### 2025-02-18 paper/sec/invariants.tex
slight edit

@@
 \end{align*}
-This implies the simple invariant $\expect \left[ p^{-1} X_s \right] = 1$ (and thus, unbiasedness) without needing a separate induction over the algorithm.
+This implies the simple invariant $\expect \left[ p^{-1} X_s \right] = 1$ without needing a separate induction over the algorithm.
 The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of classes of functions of the state, proved using recursion or induction over the algorithm itself.

### 2025-02-18 paper/sec/formalization.tex
some suggestions

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-In this section, we present the formalization of the CVM algorithm in Isabelle/HOL; we mainly discuss the variant from~\cref{sec:negdep} since the details are similar.
-A snippet of the formalization of~\cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
+In this section, we present the formalization of the CVM algorithm in Isabelle/HOL; we mainly discuss the~\cref{alg:cvm_new} variant from~\cref{sec:negdep} since the details are similar.
+Note that using our approach, it was also straightforward to verify a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
+
+\subsection{Formalized Results}
 
@@
 %The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds.
+A snippet of the formalization of~\cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
 We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
@@
 Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
-Nevertheless, \cref{tab:isabelle_syntax} contains brief descriptions of the syntactic elements used in the formalization.
+Nevertheless, \cref{tab:isabelle_syntax} contains a brief glossary of the syntactic elements used in the formalization.
 %\medskip
@@
 
-% TODO: I think we should give more of the details in the formalization, as opposed to giving a glossary of it
-% For example, here are some of the intricate details I would suggest:
+\subsection{Formalization Details}
 
-% 1) showing a high-level view of the way the invariant is established by induction?
-% 2) describing that we actually used locales and instances to give a general shared argument for both algorithms?
-% 3) we can write that we omit intricate details of the calculations, but some high-level (minor) details can be in one paragraph, e.g., any numerical analysis you needed were dealt with using existing libraries XXX and YYY (e.g., automating derivatives, approximation, etc.)?
+\todo{I think we can afford to go into some high-level details of the formalization. Here's what I (YK) suggest}
+
+\begin{itemize}
+\item Show a structural outline of the Isar induction that looks intuitively similar to the bind sequence in Section 3.2
+\item Briefly say that we used locales to abstract out the main details (essentially, we use the aforementioned intuition that \isa{subsample} can be replaced
+\item Maybe Mention some arithmetic intricacies that were easily solved using Isabelle's automation
+\end{itemize}
 

### 2025-02-18 paper/sec/related_work.tex
Paper: Fix inaccuracy in related work section.

@@
 The aforementioned hash based algorithms are not unbiased.
-%TODO: is there a typo? does this mean bounds on the deviation from the mean?
-Flajolet et al.~\cite{flajolet1985} points it out and provides bounds on the bound.
+Flajolet et al.~\cite{flajolet1985} points it out and provides bounds on the distance between the expected result and the cardinality of the stream.
 Most authors do not discuss the matter.
@@
 One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
-In the context of query processing unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements in the streaming model.
+In the context of query processing unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements problem in the streaming model.
 

### 2025-02-19 paper/sec/abstract.tex
Paper: Add more details on how the invariant is used in Sec 3.   Minor edits: Avoid ``transformation-based`` in abstract (which requires additional context)   Section 4 (clarify range of S)

@@
 Estimating the number of distinct elements in a data stream is a classical problem with numerous applications in computer science.
-We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (the so-called CVM algorithm).
+We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called CVM algorithm).
 Their algorithm deviated considerably from the state-of-the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close to optimal logarithmic space complexity.
@@
 Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method.
-This simplifies the formal analysis considerably compared to the original transformation-based proof.
+This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
 Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying with a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.

### 2025-02-19 paper/sec/invariants.tex
Paper: Revert commit cd47c60
This is not a good change:
- First we explain how to establish unbiasedness (for the alg without Line 11).
- Then we generalize the proof to establish tail bounds.
- Then we show how to derive unbiasedness using the generalized result.
  - Note that at this stage unbiasedness is not even a motivated result (because we are not talking about an actually relevant algorithm.)

For readers who are reading the text the first time, they will get confused -- like, I thought we knew this already, why are we deriving it
agian. Or they will ask, why was the first proof presen

--- 2025-02-18 paper/sec/invariants.tex
@@
 Indeed, we managed to establish the following result:
-\[
-  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|}
-\]
+\begin{align}
+  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant}
+\end{align}
 for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
@@
 We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
-This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $\prob( |p^{-1} |\chi| - |A|| \geq \varepsilon |A| \wedge p \geq q)$.
+This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
+As an example, the upper tail bound can be derived as follows:
+\begin{align*}
+  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq 
+  \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
+  & \leq \exp(-t(1+\varepsilon) |A|) M(t) \\ % = exp(-t(1+\varepsilon) |A|) \expect \left[ \prod_{s \in A} \exp( t p^{-1} X_s) \right]
+  & \leq \exp(-t(1+\varepsilon)|A|) \expect \left[ I(p \geq q) \prod_{s \in A} h(p^{-1} X_s) \right] \\
+  & \leq \exp(-t(1+\varepsilon)|A|) h(1)^{|A|} \leq \exp(-\frac{n}{12} \varepsilon^2)
+\end{align*}
+where we choose $h(x) = 1+qx (\exp(t/q)-1)$.
+For the last step, we have to find the $t$ that produces the required bound.%
+\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.} 
+Note that $h$ is affine and it can be be easily checked%
+\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
+that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
 To use these bounds, we also have to separately estimate $\prob(p < q)$.
-We can use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$ and separately bound the probability $\prob(p < q)$.
+We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$.
 

### 2025-02-19 paper/sec/negdep.tex
Paper: Add more details on how the invariant is used in Sec 3.   Minor edits: Avoid ``transformation-based`` in abstract (which requires additional context)   Section 4 (clarify range of S)

@@
 \end{equation}
-for all non-negative functions $g$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
+for all non-negative functions $g$ and any $S \subset \chi$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
 

### 2025-02-20 paper/sec/invariants.tex
Remove comment about the formalization generalizing f from Section 3.
minor edits

@@
 \footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.} 
-Note that $h$ is affine and it can be be easily checked%
+Note that $h$ is affine and it can be easily checked%
 \footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
@@
 The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
-The algorithm we have verified is actually a slight generalization of the above, where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$.
-The original CVM algorithm~\cite{chakraborty2022} (\cref{alg:cvm}) is the special case $f=\frac{1}{2}$.
 Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

### 2025-02-20 paper/sec/negdep.tex
minor edits

--- 2025-02-18 paper/sec/negdep.tex
@@
 \end{equation}
-for all non-negative functions $g$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
+for all non-negative functions $g$ and any $S \subseteq \chi$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
 
@@
 However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
-Thus, for the new subsampling step in line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
+Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
 %TODO: double check!

### 2025-02-20 paper/sec/original_proof.tex
minor edits

@@
 As we mentioned in the introduction, the main difficulty is the fact that \cref{alg:cvm}'s state variables are not independent.
-The trick Chakraborty et al.
-use to circumvent the problem is by modifying the algorithm, in a manner that obviously preserves its distribution, such that the resulting algorithm's state can be described in terms of independent coin flips.
+The trick Chakraborty et al.\ use to circumvent the problem is by modifying the algorithm, in a manner that obviously preserves its distribution, such that the resulting algorithm's state can be described in terms of independent coin flips.
 

### 2025-02-20 paper/sec/formalization_neg_dep.tex
some suggestions

@@
 The formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
-However, the authors did not correctly define the constrains for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
+However, the authors did not correctly define the constraints for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
 In the formalization, we assume that the functions are bounded and measurable.
@@
 In the formalization any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
-In this case the test functions must be monotone with respect ot the respective order on that space.
+In this case the test functions must be monotone with respect to the respective order on that space.
 
@@
 Using our parameterized relation symbol, we could state all variants in a concise manner.
+
+\todo{I think we can afford to go into some high-level details of the formalization. Here's what I (YK) suggest}
+
+\begin{itemize}
+\item It would be good to show the definition of NA and compare it to Definition 2
+\item It would be good to show formalized versions of Prop 3, Lemma 4, and the NA of permutation distributions
+\end{itemize}

### 2025-02-20 paper/sec/abstract.tex
abstract minor edits

@@
 Estimating the number of distinct elements in a data stream is a classical problem with numerous applications in computer science.
-We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called CVM algorithm).
-Their algorithm deviated considerably from the state-of-the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close to optimal logarithmic space complexity.
+We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
+Their algorithm deviated considerably from the state of the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close-to-optimal logarithmic space complexity.
 
@@
 This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
-Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying with a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
-In this paper, we introduce the proof technique, describe its use to mechanize both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
+Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
+In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
 \end{abstract}

### 2025-02-21 paper/sec/invariants.tex
Paper: More details on the derivation of the upper tail bound.

@@
 However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
-We end up with the condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
+This leads to a similar inequality:
+\begin{align}
+  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
+\end{align}
+with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
 This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
 As an example, the upper tail bound can be derived as follows:
-\begin{align*}
-  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq 
+\begin{eqnarray*}
+  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
   \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
-  & \leq \exp(-t(1+\varepsilon) |A|) M(t) \\ % = exp(-t(1+\varepsilon) |A|) \expect \left[ \prod_{s \in A} \exp( t p^{-1} X_s) \right]
-  & \leq \exp(-t(1+\varepsilon)|A|) \expect \left[ I(p \geq q) \prod_{s \in A} h(p^{-1} X_s) \right] \\
-  & \leq \exp(-t(1+\varepsilon)|A|) h(1)^{|A|} \leq \exp(-\frac{n}{12} \varepsilon^2)
-\end{align*}
-where we choose $h(x) = 1+qx (\exp(t/q)-1)$.
-For the last step, we have to find the $t$ that produces the required bound.%
-\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.} 
+  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) \exp( t p^{-1} X_s) \right] \\
+  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \\
+  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
+  & \underset{\textrm{By calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
+\end{eqnarray*}
+where we choose $h(x) = 1+qx (e^{t/q}-1)$.
 Note that $h$ is affine and it can be easily checked%
@@
 that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
+For the last step, we have to find the $t$ that produces the required bound.%
+\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
 To use these bounds, we also have to separately estimate $\prob(p < q)$.

### 2025-02-23 paper/sec/formalization.tex
Paper: Add more on formalization of neg dep.

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-In this section, we present the formalization of the CVM algorithm in Isabelle/HOL; we mainly discuss the~\cref{alg:cvm_new} variant from~\cref{sec:negdep} since the details are similar.
-Note that using our approach, it was also straightforward to verify a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
+In this section, we present the formalization of the CVM algorithm in Isabelle/HOL using our new functional probabilistic invariant based approach.
+We verified both the total variant, we introduced in the previous section and the original variant (\cref{alg:cvm}) from the introduction.
+The invariant-based verficiation approach is in the directory CVM{\textunderscore}Invariant from the supplementary material.
 
-\subsection{Formalized Results}
+The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
+The specialization happens in the following theories, where CVM{\textunderscore}Original{\textunderscore}Algorithm is verficiation of the original algorithm, and CVM{\textunderscore}New{\textunderscore}Algorithm verifies the new total and unbiased variant.
+It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables, which is included in a separate directory, which we describe in more detail in the \cref{sec:formalization_neg_dep}.
+The total number of lines required for the verification of the original algorithm is 1004 lines.
+
+%; we mainly discuss the~\cref{alg:cvm_new} variant from~\cref{sec:negdep} since the details are similar.
+We verifed a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
+
+Another slight deviation is that we have better constant factors compared to the original proof by Chakraborty et al.~\cite{chakraborty2023}.
+The buffer size in \cref{alg:cvm} is $12 \varepsilon^{-2} \ln(6 l \delta^{-1})$ while the buffer size used by Chakraborty et al.
+is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
+
+This is mainly due to the fact that the invariant-based proof, does not require conditioning on particular values of $p$ --- the subsampling probability at the end of the algorithm --- in contrast the transformation-based proof applies Chernoff bounds for each possible value of $p$ (upto some lower bound $q$) and establishes the final error probability as a union bound.
+
+However, we would like to note that, neither Chakraborty et al. seem to, nor we, have tried to optimize constants.
+It would be possible to find improvements using a more careful analysis in both cases.
 
@@
 \ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}algo\ xs\ {\isacharequal}{\kern0pt}\ map{\isacharunderscore}{\kern0pt}pmf\ estimate\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}steps\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-%\isanewline
-%\isacommand{theorem}\isamarkupfalse%
-%\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-%\isanewline
-%\isacommand{theorem}\isamarkupfalse%
-%\ correctness{\isacharcolon}{\kern0pt}\isanewline
-%\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
-%\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-%\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-%\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}\isanewline
 {\normalfont [\dots]}\isanewline
@@
 %The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds.
-A snippet of the formalization of~\cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
+A snippet of the formalization of \cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
 We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
@@
 
-The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability $1-\delta$ is expressed in the following snippet:
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability at least $1-\delta$ is expressed in the following snippet:
 \begin{isabelle_cm}
@@
 The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$; this is slightly more general than the original algorithm, where we fix $n$ to be the smallest integer larger than the same value.
-%Then, let $A$ to be the cardinality of the set of elements in sequence \isa{xs}.
+Then, we introduce the abrreviation \isa{A} for the cardinality of the set of elements in the sequence \isa{xs}.
 %More precisely, \isa{set\ xs} denotes the finite set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
@@
 
-\subsection{Formalization Details}
+The informal proof of the algorithm, which is included in the appendix of the formal proof document, is useful to follow the formalized proof.
+There are references from the formalized lemmas to the informal ones for convenience.
 
-\todo{I think we can afford to go into some high-level details of the formalization. Here's what I (YK) suggest}
-
-\begin{itemize}
-\item Show a structural outline of the Isar induction that looks intuitively similar to the bind sequence in Section 3.2
-\item Briefly say that we used locales to abstract out the main details (essentially, we use the aforementioned intuition that \isa{subsample} can be replaced
-\item Maybe Mention some arithmetic intricacies that were easily solved using Isabelle's automation
-\end{itemize}
-
-%The formalization contains four theory files.
-The file CVM{\textunderscore}Preliminary definitions and lemmas, such as properties about the \isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} operation we introduced.
-%Of course, we do not verify \cref{alg:cvm} and \cref{alg:cvm_new} separately,
-%as most of the proof is identical.
-%The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
-%The specialization happens in the following theories CVM{\textunderscore}Original{\textunderscore}Algorithm (resp. CVM{\textunderscore}New{\textunderscore}Algorithm).
-%It should be noted that only the last file, for the new unbiased algorithm, depends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables.
-%The total number of lines required for the verification of the original algorithm is 1004 lines.
-%This is in contrast to the original proof, for which the verification required \todo{x} lines.
-%
+Originally, we wrote down the informal proof first before starting the formalization effort.
+We could express the same arguments and same general strategy in Isabelle with little effort; we did not encounter any suprises during mechanization.
+This is in contrast to the mechanization of the original transformation-based proof by Chakraborty et al., which we detail in \cref{sec:transformation_based_proof}.

### 2025-02-23 paper/sec/formalization_neg_dep.tex
Paper: Add more on formalization of neg dep.

@@
 
-The formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
+The formalization, which we intend to submit to the AFP as a separate entry is included in the supplementary material in the directory Neg{\textunderscore}Assoc.
+It follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
 However, the authors did not correctly define the constraints for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
@@
 Which evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
-For example the FKG inequality
+For example the FKG~\cite[Ch. 6]{alon2000},\cite{fortuin1971} inequality
 \[
@@
 \]
-is true, if $f$ and $g$ are both monotone, or both antimonotone.
+is true, if $f$ and $g$ are both monotone, or both antimonotone (on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function).
 On the other hand the reverse inequality is true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
-Using our parameterized relation symbol, we could state all variants in a concise manner.
+Using our parameterized relation symbol, we could state all variants in a concise manner:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ fkg{\isacharunderscore}{\kern0pt}inequality{\isacharunderscore}{\kern0pt}pmf{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{fixes}\ M\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}a\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ finite{\isacharunderscore}{\kern0pt}distrib{\isacharunderscore}{\kern0pt}lattice{\isacharparenright}{\kern0pt}\ pmf{\isacartoucheclose}\isanewline
+\ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharprime}{\kern0pt}a\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}x\ y{\isachardot}{\kern0pt}\ pmf\ M\ x\ {\isacharasterisk}{\kern0pt}\ pmf\ M\ y\ {\isasymle}\ pmf\ M\ {\isacharparenleft}{\kern0pt}x\ {\isasymsqunion}\ y{\isacharparenright}{\kern0pt}\ {\isacharasterisk}{\kern0pt}\ pmf\ M\ {\isacharparenleft}{\kern0pt}x\ {\isasymsqinter}\ y{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymtau}\isactrlesub {\isacharparenright}{\kern0pt}\ f{\isacartoucheclose}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymsigma}\isactrlesub {\isacharparenright}{\kern0pt}\ g{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ f\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}\ {\isacharasterisk}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ g\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}\ {\isasymle}{\isasymge}\isactrlbsub {\isasymtau}\ {\isacharasterisk}{\kern0pt}\ {\isasymsigma}\isactrlesub \ {\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ f\ x\ {\isacharasterisk}{\kern0pt}\ g\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions.
+It is possible to multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
+The first assumption denotes the log-supermodularilty of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
+The FKG inequality is a key result which enables verification of negative association of random variables.
+Such as the one indicator variables for the new subsampling operation, we introduced in \cref{sec:negdep}.
 
-\todo{I think we can afford to go into some high-level details of the formalization. Here's what I (YK) suggest}
+Let us look into a few example results about negatively associated random variables:
+\begin{isabelle_cm}
+\isacommand{lemma}\isamarkupfalse%
+\ neg{\isacharunderscore}{\kern0pt}assoc{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}mult{\isacharunderscore}{\kern0pt}mono{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}i\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}c{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}linorder{\isacharunderscore}{\kern0pt}topology{\isacharparenright}{\kern0pt}\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}J\ {\isasymsubseteq}\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}f\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}g\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ f{\isacartoucheclose}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ g{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ f\ J{\isacartoucheclose}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ g\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}f\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ J\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ \isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}g\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}{\isasymomega}{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasympartial}M{\isacharparenright}{\kern0pt}{\isasymle}{\isacharparenleft}{\kern0pt}{\isasymintegral}\isasymomega{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}\isasymomega{\isachardot}{\kern0pt}\ g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+This corresponds to Property~\ref{it:neg_dep_props:mult_mono} of \cref{pro:neg_dep_props}:
+If $X_i$ for $i \in I$ form a set of negatively associated variables, and $f$, $g$ are simultaneously monotone (or simultaneously antimonotone) real valued square integrable measurable functions, depending on disjoint components of $I$ then the expectations of the product are smaller than then product of the expectations of the composition of the functions $f,g$ with random variables $X_i$.
+Note that, if $X_i$ where independent the above would be true, without the monotonicity assumption.
 
-\begin{itemize}
-\item It would be good to show the definition of NA and compare it to Definition 2
-\item It would be good to show formalized versions of Prop 3, Lemma 4, and the NA of permutation distributions
-\end{itemize}
+The following is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables:
+\begin{isabelle_cm}
+\isacommand{lemma}\isamarkupfalse%
+\ hoeffding{\isacharunderscore}{\kern0pt}bound{\isacharunderscore}{\kern0pt}two{\isacharunderscore}{\kern0pt}sided{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}finite\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ a\ i\ {\isasymle}\ b\ i{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ AE\ {\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}a\ i{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}b\ i{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}I\ {\isasymnoteq}\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}n\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ I{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}{\isasymmu}\ {\isasymequiv}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ expectation\ {\isacharparenleft}{\kern0pt}X\ i{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymge}\ {\isadigit{0}}{\isacartoucheclose}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\end{isabelle_cm}

### 2025-02-23 paper/sec/negdep.tex
Paper: Add more on formalization of neg dep.

@@
 \begin{enumerate}
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
 \item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.

### 2025-02-23 paper/sec/conclusion.tex
draft a conclusion

\section{Conclusion}\label{sec:conclusion}
We have presented the first formalization of the CVM algorithm using Isabelle/HOL.
Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative approach to analyzing the CVM algorithm via the Cram\'{e}r--Chernoff method.
Interestingly, the technique readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this version using the same invariant, together with a new library of results for negative association.
In future work, it would be interesting to formalize other variations subsampling variants of CVM.
More generally, it would be interesting to explore whether the technique could be applied towards proving other randomized (streaming) algorithms.

### 2025-02-23 paper/sec/conclusion.tex
conclusion edit

@@
 \section{Conclusion}\label{sec:conclusion}
-We have presented the first formalization of the CVM algorithm using Isabelle/HOL.
-Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative approach to analyzing the CVM algorithm via the Cram\'{e}r--Chernoff method.
-Interestingly, the technique readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this version using the same invariant, together with a new library of results for negative association.
-In future work, it would be interesting to formalize other variations subsampling variants of CVM.
-More generally, it would be interesting to explore whether the technique could be applied towards proving other randomized (streaming) algorithms.
+We presented the first formalization of the CVM algorithm using Isabelle/HOL.
+Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
+Interestingly, the technique readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+In future work, it would be interesting to formalize other variations of subsampling for CVM.
+More generally, one could further explore whether the technique we introduced here could be applied towards proofs of other randomized (streaming) algorithms.

### 2025-02-23 paper/sec/intro.tex
minor fixes (mostly macros or uniformity)

@@
 We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
-As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
+As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \todo{x} lines.
 
 The organization of this paper is as follows.
-\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle~\cite{nipkow2002}.
+\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL~\cite{nipkow2002}.
 \cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.

### 2025-02-23 paper/sec/invariants.tex
minor fixes (mostly macros or uniformity)

@@
 An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
-It can be stated simply as $P(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
+It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
 %\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
 %Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
-It is also possible to obtain lower tail bounds $P(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
+It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
 
@@
 This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
-As an example, the upper tail bound can be derived as follows:
+As an example, the upper tail bound can be derived as follows: %TODO: should we use exp(...) everywhere instead of mixing with e^ ?
 \begin{eqnarray*}
   \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
-  \prob( I(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
-  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) \exp( t p^{-1} X_s) \right] \\
-  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \\
+  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
+  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
+  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
   & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
@@
 To use these bounds, we also have to separately estimate $\prob(p < q)$.
-We use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, where they use $q = \frac{n}{4 |A|}$.
+For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, with $q = \frac{n}{4 |A|}$.
 

### 2025-02-23 paper/sec/negdep.tex
minor fixes (mostly macros or uniformity)

@@
 \end{equation}
-for all non-negative functions $g$ and any $S \subseteq \chi$---any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
+for all non-negative functions $g$ and any $S \subseteq \chi$.
+Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
 

### 2025-02-23 paper/sec/semra.tex
minor fixes (mostly macros or uniformity)

@@
 \subsection{Semantics of Randomized Algorithms}
-We briefly review how reasoning about randomized algorithms works in Isabelle using the Giry monad~\cite{giry1982}.
-A thorough discussion of the concept in the context of Isabelle has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
+We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
+A thorough discussion of the concept in the context of Isabelle/HOL has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
 

### 2025-02-23 paper/sec/conclusion.tex
Paper: Add note on comparison to conclusion.
minor fixes (mostly macros or uniformity)
edit concl slightly

@@
 Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
-Interestingly, the technique readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with 1004 vs.~2646 lines).
+Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+
 In future work, it would be interesting to formalize other variations of subsampling for CVM.

### 2025-02-24 paper/sec/original_proof.tex
update original proof section

--- 2025-02-10 paper/sec/original_proof.tex
@@
 \section{Transformation-based Proof}\label{sec:transformation_based_proof}
-In this section we describe the interesting parts of the original transformation-based proof by Chakraborty et al.~\cite{chakraborty2022}.
-This highlights in part, why it requires a lot of work to formalize, and why we were motivated to develop our new approach using probabilistic invariants.
+In this section, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023},
+focusing on the challenging parts in its formalization.
+An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
 
-As we mentioned in the introduction, the main difficulty is the fact that \cref{alg:cvm}'s state variables are not independent.
-The trick Chakraborty et al.
-use to circumvent the problem is by modifying the algorithm, in a manner that obviously preserves its distribution, such that the resulting algorithm's state can be described in terms of independent coin flips.
+As mentioned in~\cref{sec:intro}, the main difficulty in directly analyzing \cref{alg:cvm} is the lack of independence in its state variables.
+The technique Chakraborty et al.\ use to circumvent this issue is by progressively modifying the algorithm (\circled{A}--\circled{D}), in a manner that obviously bounds (or preserves) its distribution, and such that the final algorithm \circled{D} can be described using a simple random process with independent coin flips.
 
-Let us consider a state, where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$. 
-Then the algorithm would normally perform a coin flip with probability $p$.
-In the modified algorithm, we perform a fixed number of coin flips for each sequence element at the beginning.
-The element is put into the sample, whenever the first $k$ coin flips associated with the sequence element are $1$.
-Note that this happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged.
+For interested readers, \circled{A} corresponds to~\cite[Algorithm 1]{chakraborty2023}, \circled{B} is~\cite[Algorithm 2]{chakraborty2023}, and \circled{D} is~\cite[Algorithm 3]{chakraborty2023}.
+Whereas Chakraborty et al.\ move directly from \circled{B} to \circled{D} with an informal argument, \circled{C} is a transformation we added in the formalization to bridge this gap.
 
-Moreover, during the subsampling operation those elements are kept, whose $k+1$-th associated coin flip is $1$.
-This again preserves the behaviour, that each element is discared independently with probability $1/2$.
-The operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
-For this modified algorithm, it is easy to see that the set of elements in any state are exactly, those stream elements for which the first $k$ entries of their associated coin flips are $1$.
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[node distance=2.7cm, auto]
 
-In \cref{alg:cvm_simul} we describe the new algorithm, that behaves exactly as the original one (\cref{alg:cvm}), but performs indepenent coin flips.
-Note that the function last{\textunderscore}index returns the index of the last-occurence of an element in the sequence, before the current loop iteration.
-It should be noted that the algorithm keeps track of the number of subsampling iterations $k$, instead of the value $p = 2^{-k}$ as the original algorithm does.
+    % Define nodes
+    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center] {\circled{A} CVM \\ (Alg.~\ref{alg:cvm})};
+    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=A] {\circled{B} CVM w/o\\ Line 11};
+    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=B] {\circled{C} Eager Ver.\\(Alg.~\ref{alg:cvm_simul})};
+    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=C] {\circled{D} Random \\ Process};
+    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=D] {Chernoff \\ Bounds};
+
+    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, below of=B, node distance = 1.8cm] {Functional\\Invariant};
+
+    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, below of=E, node distance = 1.8cm] {CVM PAC\\ Guarantee};
+
+    \draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(E.south east) + (0.3,-0.6)$);
+    \node[anchor=south] at ($(D.south) + (0.0,-0.6)$) {Transformation-based Proof};
+
+    \draw[dashed,thick] ($(A.north west) + (-0.5,0.4)$) rectangle ($(C'.south east) + (0.2,-0.2)$);
+    \node[anchor=south,align=center] at ($(C'.west) + (-1.5,-0.45)$) {Probabilistic\\Invariant Proof};
+
+    % Draw arrows
+    \draw[->] (A) -- (B);
+    \draw[->] (B) -- (C);
+    \draw[->] (B) -- (C');
+    \draw[->] (C) -- (D);
+    \draw[->] (D) -- (E);
+    \draw[->] (E) -- (pac);
+    \draw[->] (C') -- (pac);
+
+\end{tikzpicture}
+\end{center}
+\caption{An overview of two proof approaches for the CVM algorithm.}\label{fig:proof_overview}
+\end{figure}
+
+\subsection{A Bridging Transformation}
+Let us consider algorithm \circled{B} in a state where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$.
+The algorithm would perform a coin flip lazily with probability $p$ when it encounters the next stream element.
+
+The transformation \circled{C} is shown in \cref{alg:cvm_simul}, and we prove that it precisely computes the same distribution as \circled{B}.
+In \circled{C}, we eagerly perform a fixed number of coin flips for each sequence element at the beginning.
+Now, each element is put into the state $\chi$, whenever the first $k$ coin flips associated with the sequence element are all $1$.
+Note that this happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged from \circled{B}.
+Similarly, in the subsampling operation, only those elements whose $k+1$-th associated coin flip is $1$ are kept; the operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
+This again preserves the behaviour of \circled{B} that each element is discarded independently with probability $1/2$.
+
+It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
+The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream for each fixed $K$; one relates \circled{C} to \circled{D} by:
+\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
+for fixed values of $K$ and $X$.
+To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific values $k = K$.
+The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ heads is outside $2^{-K} |A| (1 \pm \varepsilon)$.
+This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.
+
 %Like, we did in \cref{sec:invariants}, we ignore the second check, whether the subsampling operation succeeded.
@@
 \begin{algorithm}[h!]
-	\caption{Modified CVM algorithm with independent coin flips.}\label{alg:cvm_simul}
+	\caption{Modified CVM algorithm with independent coin flips.
+The function last{\textunderscore}index returns the index of the last occurrence of an element in the sequence, before the current iteration.}\label{alg:cvm_simul}
 	\begin{algorithmic}[1]
@@
     \EndIf
-    \If{$|\chi| = n$}
-      \State \Return $\bot$
-    \EndIf
+    %\If{$|\chi| = n$}
+    %  \State \Return $\bot$
+    %\EndIf
   \EndFor
@@
 \end{algorithm}
-To roughly explain, how tail bounds can be derived for this new variant:
-It is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific value of $k$, which can be bounded by the probability that the number of stream elements with whose associated coin flips start with $k$ ones, is outside of $2^{-k} |A| (1 \pm \varepsilon)$.
-This is explained in more detail by Chakraborty et al.~\cite{chakraborty2022}.
 
-One of the key questions is, how to formalize the transformation from \cref{alg:cvm} to this new variant.
-What we discovered is that it best to solve the problem backwards, i.e., we start with the modified algorithm, which performs all the coin-flips in advance --- eagerly --- and convert it to the version, that performs the coin flips --- lazily --- at the point they are needed.
+\subsection{Eager to Lazy Coin Flips}
+A remaining question is how to formalize the from \circled{B} to \circled{C}.
+Our insight is that it is best to solve the problem backwards, i.e., we start with the modified algorithm \circled{C}, which performs all the coin-flips in advance \emph{eagerly} and convert it back to \circled{B} which implicitly performs the coin flips \emph{lazily} at the point they are needed.
 
-We can push down the coin flips through the expression tree.
-To explain how this works.
-Let us first define the \emph{sampling} function, i.e., let $f$ be a function that depends on coin flips, more precisely $f$ takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to indepenent unbiased coin flips as.
+The main idea is to automatically push down the coin flips through the expression tree of \cref{alg:cvm_simul}.
+To explain how this works, let us first define the \emph{sampling} function, i.e., let $f$ be a function that takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to independent unbiased coin flips as.
 \begin{isabelle_cm}
@@
 %For example, we could use our modified algorithm as $f$, with the index set $I = \{0, \ldots, l-1\} \times \{0, \ldots, l-1\}$.
-The interesting fact is that we can distribute the sampling operation over composition, e.g.:
+The interesting fact is that we can distribute the sampling operation over composition:
 \begin{isabelle_cm}
@@
 if $f$ and $g$ depend on disjoint subsets of the coin flips, i.e., if $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+By recursively applying the rule, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
+This lets us readily transform \circled{C} to \circled{B} and prove their distributions equivalent.
 
-By recursively applying the rule, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
-Note that, we arrive at the orignal version of the algorithm, that performs the coin flips lazily.
-
-A detail, that we simplified here is that the split of the index sets, e.g., which coin flips $f$ depends on and which coin flips $g$ depends on, maybe dynamic.
-This means the split may depend on the return value of $f$.
+A detail that we have simplified here is that the split of the index sets, e.g., which coin flips $f$ depends on and which coin flips $g$ depends on, may be dynamic.
 For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin-flips.
@@
 And the next loop iteration, will indeed only read coin flips that were not read by the previous iteration.
-%For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
-%% It is possible to express the above in a point-free manner.
-%% For that, we need to introduce the reader monad.
-%% This is a deterministic monad, which provides a read operation to a global value, such as our set of coin flips.
-%% (In Haskell, the reader monad is commonly used to provide a fixed enviroment to algorithms, such as global configuration options, paths or command lines parameters~\cite{jones1995}.)
+For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
 
-%% \begin{isabelle_cm}
-%% \isacommand{datatype}\ \isacharparenleft{\isacharprime}c\isacharcomma\ {\isacharprime}a\isacharparenright\ reader{\isacharunderscore}monad\ =\ Reader\ {\isacharparenleft}run{\isacharunderscore}reader\isacharcolon\ \isacartoucheopen{\isacharprime}c\ \isasymRightarrow\ {\isacharprime}a\isacartoucheclose\isacharparenright)\isanewline
-%% \isanewline
-%% \isacommand{definition}\ bind{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}bind{\isacharunderscore}rd\ m\ f\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}r\isachardot\ run{\isacharunderscore}reader\ {\isacharparenleft}f\ {\isacharparenleft}run{\isacharunderscore}reader\ m\ r\isacharparenright\isacharparenright\ r\isacharparenright\isacartoucheclose\isanewline
-%% \isacommand{definition}\ return{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}return{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}{\isacharunderscore}\isachardot\ x\isacharparenright\isacartoucheclose\isanewline
-%% \isacommand{definition}\ get{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}get{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ id\isacartoucheclose
-%% \end{isabelle_cm}
-%% With the definition above, we can define our sampling function with respect to the reader monad, where the environment is the entire set of coin flips, which is now a monad morphism (from the reader monad to the Giry monad):
-%% \begin{isabelle_cm}
-%%   sample{\isacharunderscore}rd\ \isacharequal\ sample\ \isasymcirc\ run{\isacharunderscore}reader
-%% \end{isabelle_cm}
+It is possible to express the above in a point-free manner using the reader monad.
+This is a deterministic monad, which provides a read operation to a global value, such as our set of coin flips.
+(In Haskell, the reader monad is commonly used to provide a fixed environment to algorithms, such as global configuration options, paths or command lines parameters~\cite{jones1995}.)
+
+\begin{isabelle_cm}
+\isacommand{datatype}\ \isacharparenleft{\isacharprime}c\isacharcomma\ {\isacharprime}a\isacharparenright\ reader{\isacharunderscore}monad\ =\ Reader\ {\isacharparenleft}run{\isacharunderscore}reader\isacharcolon\ \isacartoucheopen{\isacharprime}c\ \isasymRightarrow\ {\isacharprime}a\isacartoucheclose\isacharparenright)\isanewline
+\isanewline
+\isacommand{definition}\ bind{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}bind{\isacharunderscore}rd\ m\ f\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}r\isachardot\ run{\isacharunderscore}reader\ {\isacharparenleft}f\ {\isacharparenleft}run{\isacharunderscore}reader\ m\ r\isacharparenright\isacharparenright\ r\isacharparenright\isacartoucheclose\isanewline
+\isacommand{definition}\ return{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}return{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}{\isacharunderscore}\isachardot\ x\isacharparenright\isacartoucheclose\isanewline
+\isacommand{definition}\ get{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}get{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ id\isacartoucheclose
+\end{isabelle_cm}
+With the definition above, we can define our sampling function with respect to the reader monad, where the environment is the entire set of coin flips, which is now a monad morphism (from the reader monad to the Giry monad):
+\begin{isabelle_cm}
+  sample{\isacharunderscore}rd\ \isacharequal\ sample\ \isasymcirc\ run{\isacharunderscore}reader
+\end{isabelle_cm}

### 2025-02-24 paper/sec/related_work.tex
minor fix

@@
 
-The aforementioned hash based algorithms are not unbiased.
-Flajolet et al.~\cite{flajolet1985} points it out and provides bounds on the distance between the expected result and the cardinality of the stream.
-Most authors do not discuss the matter.
-It, however, is not hard to see that they are not.
+The aforementioned hash-based algorithms are biased; Flajolet et al.~\cite{flajolet1985} points this out and also provides bounds on the distance between the expected result and the cardinality of the stream.
+Most authors do not discuss the matter of unbiasedness but it is not hard to see that they are not.
 One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
-In the context of query processing unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements problem in the streaming model.
+In the context of query processing, unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements problem in the streaming model.
 
@@
 De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
-Our approach differs from these techniques, by avoiding the computation of the distribution, which, we think, is infeasible for the CVM algorithm.
-Instead, we investigate functions of the distributions, which are relevant for the analysis.
+Our approach differs from these techniques, by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
+Instead, we investigate classes of functions of the distributions, which are relevant for the analysis.
 
 There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
-However, these methods, do not extend to algorithms with branches, or in general, which contain discrete operations.
+However, these methods, do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.
 

### 2025-02-24 paper/main.tex
some editing in Sec 3
edits in sec 4
test newline
minor edits
update original proof section
some space saving

@@
 \DeclareMathOperator{\expect}{\mathrm{E}}
+\DeclareMathOperator{\indicat}{\mathrm{I}}
 \DeclareMathOperator{\bigo}{\mathcal O}
@@
 
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{shapes.geometric}
+\usetikzlibrary{decorations.pathreplacing,calligraphy}
+\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
+            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0pt] (char) {#1};}}
+
+\crefname{ineq}{inequality}{inequalities}
+\creflabelformat{ineq}{#2{\upshape(#1)}#3}
+
 \bibliographystyle{plainurl}% the mandatory bibstyle
 
-\title{Verification of the CVM algorithm with a New Recursive Analysis Technique} %TODO Please add
+%\title{Verification of the CVM algorithm with a New Recursive Analysis Technique} %TODO Please add
+\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant} %TODO maybe?
 
@@
 
-\author{Joan R.
+%\author{Joan R.
 Public\footnote{Optional footnote, e.g.
 to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\author{Joan R.
+%
+%\author{Joan R.
 Public\footnote{Optional footnote, e.g.
 to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\author{Joan R.
+%
+%\author{Joan R.
 Public\footnote{Optional footnote, e.g.
 to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\author{Joan R.
+%
+%\author{Joan R.
 Public\footnote{Optional footnote, e.g.
@@
 
-\keywords{Formal Verification, Isabelle/HOL, Randomized Algorithms, Distinct Elements}
+\keywords{Verification, Isabelle/HOL, Randomized Algorithms, Distinct Elements}
 

### 2025-02-24 paper/sec/invariants.tex
some space saving

@@
 Let us start by briefly reviewing the algorithm---its state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
-The buffer contains a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
+The buffer tracks a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
 The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$ (Lines 3--7).
@@
 If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returning $\bot$ (Line 11).
-After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct value for the number of distinct elements in the stream.
+After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct estimate for the number of distinct elements in the stream.
 
-\subsection{A Minor Transformation}
-It is convenient to analyze \cref{alg:cvm} without Line 11, i.e., we will skip the second check of $|\chi|=n$ determining whether the subsampling step succeeded.
-This modified version simplifies our subsequent analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
-This transformation is identically used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
+\begin{remark}
+For our discussion below, it is convenient to analyze \cref{alg:cvm} without Line 11, i.e., we will skip the second check of $|\chi|=n$ determining whether the subsampling step succeeded.
+This modified version simplifies our analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
+This transformation is also used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
 This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
+\end{remark}
 
 \subsection{Deriving A Simple Probabilistic Invariant}
-Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer.\footnote{We use $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.}
-Before the algorithm encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far.
-In particular $\chi \subseteq \{a_1,..,a_m\}$ after loop iteration $m$.
+Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
+Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,..,a_m\}$ after loop iteration $m$.
 
-In the loop iteration where $s$ occurs the first time, it will be inserted with a probability $p$ in Lines 3--7.
+In the loop iteration where element $s$ occurs the first time, it will be inserted with a probability $p$ in Lines 3--7.
 This means, after Line 7, we have:
@@
 Interestingly, this equation is preserved for the rest of the algorithm.
-For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so will $p$ after subsampling, which preserves the equation.
+For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so is $p$ after subsampling, which preserves the equation.
 
@@
 To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
-In the first case ($a = s$), the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
-In the second case ($a \not= s$), it follows from the induction hypothesis on $\Omega$.
-(In particular, the term in the inner integral is constant with respect to $\tau$.)
+When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
+When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.
 
@@
 Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
-This completes the invariance proof.
+This completes the invariance proof for~\cref{eq:indicator_eq}.
 
 \subsection{Deriving A Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expectation of the output $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
 However, recall that we are interested in verifying the algorithm's PAC guarantee.
-A typical approach to establishing such a guarantee is to use Chernoff bounds for independent random variables, which provide exponential tail bounds for the deviation of their sums from their mean.
+A typical approach to establishing such a guarantee is to use Chernoff bounds which provide exponential tail bounds for the deviation of sums of independent random variables from their mean.
 However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$ for $s \in A$, are dependent.
@@
 for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
-We call this a \emph{functional} probabilistic invariant because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.
+We call this a \emph{functional probabilistic invariant} because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.
 
@@
 \begin{eqnarray*}
-  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq &
+  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & = &
   \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
@@
   & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
-  & \underset{\textrm{By calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
+  & \underset{\textrm{Calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
 \end{eqnarray*}
@@
 For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, with $q = \frac{n}{4 |A|}$.
-
-The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
+%The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
+The formalization i nthe supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
 Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

### 2025-02-24 paper/sec/original_proof.tex
shorten sec 7

@@
 As mentioned in~\cref{sec:intro}, the main difficulty in directly analyzing \cref{alg:cvm} is the lack of independence in its state variables.
-The technique Chakraborty et al.\ use to circumvent this issue is by progressively modifying the algorithm (\circled{A}--\circled{D}), in a manner that obviously bounds (or preserves) its distribution, and such that the final algorithm \circled{D} can be described using a simple random process with independent coin flips.
+The technique Chakraborty et al.\ use to circumvent this issue is by progressively modifying the algorithm (\circled{A}--\circled{D} in~\cref{fig:proof_overview}), in a manner that obviously bounds (or preserves) its distribution, and such that the final algorithm \circled{D} can be described using a simple random process with independent coin flips.
 
@@
 
-The transformation \circled{C} is shown in \cref{alg:cvm_simul}, and we prove that it precisely computes the same distribution as \circled{B}.
-In \circled{C}, we eagerly perform a fixed number of coin flips for each sequence element at the beginning.
-Now, each element is put into the state $\chi$, whenever the first $k$ coin flips associated with the sequence element are all $1$.
-Note that this happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged from \circled{B}.
-Similarly, in the subsampling operation, only those elements whose $k+1$-th associated coin flip is $1$ are kept; the operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
-This again preserves the behaviour of \circled{B} that each element is discarded independently with probability $1/2$.
-
-It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
-The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream for each fixed $K$; one relates \circled{C} to \circled{D} by:
-\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
-for fixed values of $K$ and $X$.
-To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific values $k = K$.
-The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ heads is outside $2^{-K} |A| (1 \pm \varepsilon)$.
-This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.
-
-%Like, we did in \cref{sec:invariants}, we ignore the second check, whether the subsampling operation succeeded.
-%As we explained there, the total variational difference between these two variants is $\frac{\delta}{2}$.
-%
-\begin{algorithm}[h!]
+\begin{algorithm}[t]
 	\caption{Modified CVM algorithm with independent coin flips.
@@
 
+The transformation \circled{C} is shown in \cref{alg:cvm_simul}, and we prove that it computes precisely the same distribution as \circled{B}.
+In \circled{C}, we eagerly perform a fixed number of coin flips for each sequence element at the beginning.
+Now, each element is put into the state $\chi$, whenever the first $k$ coin flips associated with the sequence element are all $1$s.
+This happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged from \circled{B}.
+Similarly, in the subsampling operation, only those elements whose $k+1$-th associated coin flip is $1$ are kept; the operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
+This again preserves the behaviour of \circled{B} that each element is discarded independently with probability $1/2$.
+
+It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
+The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream, taking $K$ as a fixed parameter; one relates \circled{C} to \circled{D} by:
+\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
+for fixed values of $K$ and $X$.
+To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific values $k = K$.
+The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
+This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.
+
+%Like, we did in \cref{sec:invariants}, we ignore the second check, whether the subsampling operation succeeded.
+%As we explained there, the total variational difference between these two variants is $\frac{\delta}{2}$.
+%
 \subsection{Eager to Lazy Coin Flips}
-A remaining question is how to formalize the from \circled{B} to \circled{C}.
+A remaining question is how to formalize the transformation from \circled{B} to \circled{C}.
 Our insight is that it is best to solve the problem backwards, i.e., we start with the modified algorithm \circled{C}, which performs all the coin-flips in advance \emph{eagerly} and convert it back to \circled{B} which implicitly performs the coin flips \emph{lazily} at the point they are needed.
@@
 For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
+In our formalization, we abstract these bookkeeping details using the reader monad.
 
-It is possible to express the above in a point-free manner using the reader monad.
-This is a deterministic monad, which provides a read operation to a global value, such as our set of coin flips.
-(In Haskell, the reader monad is commonly used to provide a fixed environment to algorithms, such as global configuration options, paths or command lines parameters~\cite{jones1995}.)
-
-\begin{isabelle_cm}
-\isacommand{datatype}\ \isacharparenleft{\isacharprime}c\isacharcomma\ {\isacharprime}a\isacharparenright\ reader{\isacharunderscore}monad\ =\ Reader\ {\isacharparenleft}run{\isacharunderscore}reader\isacharcolon\ \isacartoucheopen{\isacharprime}c\ \isasymRightarrow\ {\isacharprime}a\isacartoucheclose\isacharparenright)\isanewline
-\isanewline
-\isacommand{definition}\ bind{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}bind{\isacharunderscore}rd\ m\ f\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}r\isachardot\ run{\isacharunderscore}reader\ {\isacharparenleft}f\ {\isacharparenleft}run{\isacharunderscore}reader\ m\ r\isacharparenright\isacharparenright\ r\isacharparenright\isacartoucheclose\isanewline
-\isacommand{definition}\ return{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}return{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}{\isacharunderscore}\isachardot\ x\isacharparenright\isacartoucheclose\isanewline
-\isacommand{definition}\ get{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}get{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ id\isacartoucheclose
-\end{isabelle_cm}
-With the definition above, we can define our sampling function with respect to the reader monad, where the environment is the entire set of coin flips, which is now a monad morphism (from the reader monad to the Giry monad):
-\begin{isabelle_cm}
-  sample{\isacharunderscore}rd\ \isacharequal\ sample\ \isasymcirc\ run{\isacharunderscore}reader
-\end{isabelle_cm}
+%It is possible to express the above in a point-free manner using the reader monad.
+%This is a deterministic monad, which provides a read operation to a global value, such as our set of coin flips.
+%(In Haskell, the reader monad is commonly used to provide a fixed environment to algorithms, such as global configuration options, paths or command lines parameters~\cite{jones1995}.)
+%
+%\begin{isabelle_cm}
+%\isacommand{datatype}\ \isacharparenleft{\isacharprime}c\isacharcomma\ {\isacharprime}a\isacharparenright\ reader{\isacharunderscore}monad\ =\ Reader\ {\isacharparenleft}run{\isacharunderscore}reader\isacharcolon\ \isacartoucheopen{\isacharprime}c\ \isasymRightarrow\ {\isacharprime}a\isacartoucheclose\isacharparenright)\isanewline
+%\isanewline
+%\isacommand{definition}\ bind{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}bind{\isacharunderscore}rd\ m\ f\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}r\isachardot\ run{\isacharunderscore}reader\ {\isacharparenleft}f\ {\isacharparenleft}run{\isacharunderscore}reader\ m\ r\isacharparenright\isacharparenright\ r\isacharparenright\isacartoucheclose\isanewline
+%\isacommand{definition}\ return{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}return{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ \isacharparenleft{\isasymlambda}{\isacharunderscore}\isachardot\ x\isacharparenright\isacartoucheclose\isanewline
+%\isacommand{definition}\ get{\isacharunderscore}rd\ \isacommand{where}\ {\isacartoucheopen}get{\isacharunderscore}rd\ x\ \isacharequal\ Reader\ id\isacartoucheclose
+%\end{isabelle_cm}
+%With the definition above, we can define our sampling function with respect to the reader monad, where the environment is the entire set of coin flips, which is now a monad morphism (from the reader monad to the Giry monad):
+%\begin{isabelle_cm}
+%  sample{\isacharunderscore}rd\ \isacharequal\ sample\ \isasymcirc\ run{\isacharunderscore}reader
+%\end{isabelle_cm}

### 2025-02-24 paper/main.tex
minor edit
Paper: Minor pedantic change.   Use plural for anonymous :)

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 %This is a template for producing LIPIcs articles.
@@
 \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
-            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0pt] (char) {#1};}}
+            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0.5pt] (char) {#1};}}
 
@@
 
-\author{Jane {Open Access}}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional.
-Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.
+\author{\textcolor{red}{Anounymous authors}}{\textcolor{red}{Anonymous affiliations}}{}{}{}
 
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-%
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-%
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-%
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+\authorrunning{\textcolor{red}{Anonymous}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{\textcolor{red}{Anonymous}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 

### 2025-02-24 paper/sec/invariants.tex
Paper: Reply - minor edit to inequality chain for upper tail bound.

@@
 This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
-As an example, the upper tail bound can be derived as follows: %TODO: should we use exp(...) everywhere instead of mixing with e^ ?
+As an example, the upper tail bound can be derived as follows:
+%TODO: should we use exp(...) everywhere instead of mixing with e^ ?
+EK: That was the previous verison, main issue is the chain looks very lopsided, with almost all of the terms being on the right edge.
+(It looked too ugly.)
+% The convention is now, if exp is applied to a random variable we use the long notation and otherwise the power notation, which might actually be (a bit) helpful.
 \begin{eqnarray*}
-  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & = &
+  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq & % Is not equal when |A|=0.
   \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
@@
   & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
-  & \underset{\textrm{Calculus}}{\leq} & \exp\left(-\frac{n}{12} \varepsilon^2\right)
+  & \underset{\textrm{Calculus}}{\leq} & e^{-n\varepsilon^2 / 12}
 \end{eqnarray*}

### 2025-02-24 paper/sec/abstract.tex
Paper: Minor edits.   * Use central place for LOCs.

@@
 \begin{abstract}
-Estimating the number of distinct elements in a data stream is a classical problem with numerous applications in computer science.
+Estimating the number of distinct elements in a data stream is a classic problem with numerous applications in computer science.
 We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
@@
 
-Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method.
+Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
 This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.

### 2025-02-24 paper/sec/conclusion.tex
Paper: Minor edits.   * Use central place for LOCs.

@@
 Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
-Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with 1004 vs.~2646 lines).
+Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with \locnew vs.
+\locold~lines).
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-02-24 paper/sec/intro.tex
Paper: Minor edits.   * Use central place for LOCs.

@@
 We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
-As a rough point of comparison, verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \todo{x} lines.
+As a rough point of comparison, verification of the CVM algorithm using our new technique requires only \locnew~lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \locold~lines.
 
@@
 \begin{itemize}
-\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) with the approach presented in \cref{sec:invariants};
+\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using probabilistic functional invariants;
 \item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.

### 2025-02-25 paper/sec/invariants.tex
Paper: Fix typo in Sec. 3.
minor edits (.. -> ...)

@@
 Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
-Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,..,a_m\}$ after loop iteration $m$.
+Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,\dots,a_m\}$ after loop iteration $m$.
 
@@
 Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
-To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, i.e., we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
+To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
 \[
@@
 \]
-Note that, we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
+Note that we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
 To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
@@
 EK: That was the previous verison, main issue is the chain looks very lopsided, with almost all of the terms being on the right edge.
-(It looked too ugly.)
+(It looked too ugly.)--- OK
 % The convention is now, if exp is applied to a random variable we use the long notation and otherwise the power notation, which might actually be (a bit) helpful.
@@
 To use these bounds, we also have to separately estimate $\prob(p < q)$.
-For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2022}, with $q = \frac{n}{4 |A|}$.
+For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2023}, with $q = \frac{n}{4 |A|}$.
 %The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
-The formalization i nthe supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
+The formalization in the supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
 Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.

### 2025-02-25 paper/sec/negdep.tex
some space saving
minor edits (.. -> ...)

@@
 The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orthant dependence.
+%TODO: perhaps citations for these two?
 Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
@@
 \begin{definition}[Negative Association]\label{def:neg_assoc}
-A set of random variables $X_1,..,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
+A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
 \[
@@
 \begin{enumerate}
-\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ also for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector valued random variables $X$ and $Y$ are independent, then the union $X_1,..,X_n,Y_1,..,Y_m$ is a set of negatively associated random variables.
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $f_1,..,f_m : \mathbb R^n \rightarrow \mathbb R$ be all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
-\item If $X_1,..,X_n$ are independent then $X_1,...,X_n$ are negatively associated.
+\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector-valued random variables $X$ and $Y$ are independent, then the union $X_1,\dots,X_n,Y_1,\dots,Y_m$ is a set of negatively associated random variables.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated and $f_1, \dots ,f_m : \mathbb R^n \rightarrow \mathbb R$ are all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
+\item If $X_1,\dots,X_n$ are independent then $X_1,\dots,X_n$ are negatively associated.
 \item A subset of a negatively associated set of random variables is again negatively associated.
@@
 These properties illustrate the trade-off between negative association and independence.
-For example, property 3 would be true for independent random variables, even without the condition of monotonicity.
+For example, Property 3 would be true for independent random variables, even without the condition of monotonicity.
 %Of course, on the other hand independence is a stronger property and fewer sets of random variables are independent.
@@
 \begin{lemma}\label{le:neg_assoc_prod}
-Let $X_1,..,X_n$ be negatively associated and $f_1,..,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
+Let $X_1,\dots,X_n$ be negatively associated and $f_1,\dots,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
 \[

### 2025-02-25 paper/sec/semra.tex
some space saving
minor edits (.. -> ...)

@@
 i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
-Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.\footnote{The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
-It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.}
+Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
+
+\begin{remark}
+The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
+\end{remark}
 
@@
 
-\cref{eq:integral_bind} is crucially used to establish invariants of the form we will use in~\cref{sec:invariants}.
+\cref{eq:integral_bind} is crucially used to establish the invariants we introduce next in~\cref{sec:invariants}.
 \end{description}
-%\todo{I think we should briefly say that a streaming algorithm is modeled as folding over the monad?}

### 2025-02-25 paper/sec/conclusion.tex
minor fixes

--- 2025-02-23 paper/sec/conclusion.tex
@@
 We presented the first formalization of the CVM algorithm using Isabelle/HOL.
-Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
-Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our new invariant-based technique yields a considerably shorter formalization (with 1004 vs.~2646 lines).
+Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-02-25 paper/sec/invariants.tex
minor fixes

@@
 This transformation is also used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
-This means probability bounds derived for the modified version, can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
-\end{remark}
+Thus, probability bounds derived for the modified version can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
+\lipicsEnd\end{remark}
 

### 2025-02-25 paper/sec/formalization.tex
Paper: Minor edits.   * Use central place for LOCs.
more uniform structure in formalization secs

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-In this section, we present the formalization of the CVM algorithm in Isabelle/HOL using our new functional probabilistic invariant based approach.
+Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach.
 We verified both the total variant, we introduced in the previous section and the original variant (\cref{alg:cvm}) from the introduction.
-The invariant-based verficiation approach is in the directory CVM{\textunderscore}Invariant from the supplementary material.
 
-The file CVM{\textunderscore}Abstract{\textunderscore}Algorithm verifies a generalized version of the CVM algorithm, with an abstract subsampling operation, which is required to fulfill \cref{i:subsample_condition}.
-The specialization happens in the following theories, where CVM{\textunderscore}Original{\textunderscore}Algorithm is verficiation of the original algorithm, and CVM{\textunderscore}New{\textunderscore}Algorithm verifies the new total and unbiased variant.
-It should be noted that only the last file, for the new unbiased algorithm, dependends on the new library (\cref{sec:formalization_neg_dep}) on negatively associated random variables, which is included in a separate directory, which we describe in more detail in the \cref{sec:formalization_neg_dep}.
-The total number of lines required for the verification of the original algorithm is 1004 lines.
+%TODO: make sure these names are correct!
+\begin{note}
+In our supplementary material, the invariant-based verification approach is in the directory \verb|CVM_Invariant|.
+The file \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
+The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Algorithm| verifies the new total and unbiased variant.
+It should be noted that only \verb|CVM_New_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
+The total number of lines required for the verification of the original algorithm is \locnew~lines.
+\lipicsEnd\end{note}
 
-%; we mainly discuss the~\cref{alg:cvm_new} variant from~\cref{sec:negdep} since the details are similar.
-We verifed a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
-
-Another slight deviation is that we have better constant factors compared to the original proof by Chakraborty et al.~\cite{chakraborty2023}.
-The buffer size in \cref{alg:cvm} is $12 \varepsilon^{-2} \ln(6 l \delta^{-1})$ while the buffer size used by Chakraborty et al.
+We verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$; and we allow $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
+Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-
-This is mainly due to the fact that the invariant-based proof, does not require conditioning on particular values of $p$ --- the subsampling probability at the end of the algorithm --- in contrast the transformation-based proof applies Chernoff bounds for each possible value of $p$ (upto some lower bound $q$) and establishes the final error probability as a union bound.
-
-However, we would like to note that, neither Chakraborty et al. seem to, nor we, have tried to optimize constants.
-It would be possible to find improvements using a more careful analysis in both cases.
+This is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+However, we would like to note that, neither Chakraborty et al., nor us, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
 
@@
 The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
-Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability, that the stream elements are in the buffer.
-Isabelle provides notation closely related to informal pseudocode, such that it is usually feasible to read a formalization without expert knowledge.
-Nevertheless, \cref{tab:isabelle_syntax} contains a brief glossary of the syntactic elements used in the formalization.
+Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability that the stream elements are in the buffer, respectively.
+Isabelle/HOL provides notation closely related to informal pseudocode, so it is usually feasible to read a formal statement without expert knowledge.
+Nevertheless, \cref{tab:isabelle_syntax} contains a brief glossary of the syntax used in the formalization.
 %\medskip
 \begin{table}
-\caption{Isabelle syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
+\caption{Isabelle/HOL syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
 \noindent\begin{tabular}{l p{9cm}}
@@
 \isa{nat} & Type of natural numbers (non-negative integers). \\
-\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
+\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the Boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
 \isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
@@
 The first line gives conditions on parameters $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
-The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$; this is slightly more general than the original algorithm, where we fix $n$ to be the smallest integer larger than the same value.
-Then, we introduce the abrreviation \isa{A} for the cardinality of the set of elements in the sequence \isa{xs}.
+The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
+Then, we introduce the abbreviation \isa{A} for the cardinality of the set of elements in the sequence \isa{xs}.
 %More precisely, \isa{set\ xs} denotes the finite set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
-The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for \isa{run-algo}.
+The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for \isa{run{\isacharunderscore}{\kern0pt}algo}.
 
-Similarly, we have also formalized unbiasedness of the algorithm:
+Similarly, we have also formalized unbiasedness of \cref{alg:cvm_new}:
 \begin{isabelle_cm}
@@
 
-The informal proof of the algorithm, which is included in the appendix of the formal proof document, is useful to follow the formalized proof.
-There are references from the formalized lemmas to the informal ones for convenience.
+Our proofs are available in the supplement, both in mechanized form in Isabelle/HOL and as a pen-and-paper proof included in the proof document.
+In practice, we developed the latter proof first and then mechanized it in Isabelle/HOL without much surprises.
+Most of the lemmas can be one-to-one identified between both proofs; Isabelle's existing libraries, automation capabilities, and structured proof format were used extensively in our proofs.
 
-Originally, we wrote down the informal proof first before starting the formalization effort.
-We could express the same arguments and same general strategy in Isabelle with little effort; we did not encounter any suprises during mechanization.
-This is in contrast to the mechanization of the original transformation-based proof by Chakraborty et al., which we detail in \cref{sec:transformation_based_proof}.
+%For our proof, we first sketched out our invariant-based approach in an informal proof before transferring this to Isabelle/HOL.
+%Originally, we wrote down the informal proof first before starting the formalization effort.
+%We could express the same arguments and same general strategy in Isabelle/HOL with little effort; we did not encounter any surprises during mechanization.
+%This is in contrast to the mechanization of the original transformation-based proof by Chakraborty et al., which we detail in \cref{sec:transformation_based_proof}.
+
+%The informal proof of the algorithm, which is included in the appendix of the formal proof document, is useful to follow the formalized proof.
+%There are references from the formalized lemmas to the informal ones for convenience.

### 2025-02-25 paper/sec/formalization_neg_dep.tex
more uniform structure in formalization secs

@@
 \section{Formalization of a Library for Negative Association}\label{sec:formalization_neg_dep}
-As mentioned in the previous section, the formalization of the new unbiased variant of the CVM algorithm, requires results from the theory of negative association.
+As mentioned in~\cref{sec:negdep}, formalizing the total and unbiased variant of the CVM algorithm requires results from the theory of negative association.
 
-The formalization, which we intend to submit to the AFP as a separate entry is included in the supplementary material in the directory Neg{\textunderscore}Assoc.
-It follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
+\begin{note}
+In our supplementary material, the formalization of negative association is in the directory \verb|Neg_Assoc|.
+This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distribution|).
+Although not needed for CVM, we have also mechanized the standard Chernoff bounds (\verb|Neg_Assoc_Chernoff_Bounds|), including the additive bounds by Hoeffding~\cite[Th.
+1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th.
+4.1, 4.2]{motwani1995}.
+Another example application of the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
+\lipicsEnd\end{note}
+
+Our formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
+% TODO: YK: I am struggling to understand the paragraph below.
+% In reference to 2.1 in Joag-Dev/Proschan, do you mean they are missing the integrability conditions for f,g?
 However, the authors did not correctly define the constraints for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
 In the formalization, we assume that the functions are bounded and measurable.
-On the other hand, we derive that the condition is still true, if $f, g$ are square integrable.
+On the other hand, we derive that the condition is still true, if $f, g$ are square-integrable.
 Or alternatively integrable, and non-negative.
@@
 
-Another deviation from the original work is that we do not require that the random variables need to be real-valued.
-In the formalization any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
-In this case the test functions must be monotone with respect to the respective order on that space.
-
-The formalization contains the standard Chernoff bounds for negatively associated random variables.
-This includes the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
-Although, these results are not needed for the proof of the CVM algorithm.
+Another deviation from the original work is that we do not require that the random variables are real-valued.
+In the formalization, any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
+In this case, the test functions must be monotone with respect to the respective order on the range space.
 
@@
 Which evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
-For example the FKG~\cite[Ch. 6]{alon2000},\cite{fortuin1971} inequality
+For example the FKG inequality~\cite[Ch. 6]{alon2000},\cite{fortuin1971}
 \[
@@
 is true, if $f$ and $g$ are both monotone, or both antimonotone (on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function).
-On the other hand the reverse inequality is true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
-Using our parameterized relation symbol, we could state all variants in a concise manner:
+The reverse inequality is also true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
+Using our parameterized relation symbol, we can state all variants in a concise manner:
 \begin{isabelle_cm}
@@
 \end{isabelle_cm}
-Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions.
-It is possible to multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
+Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, where we can multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
 The first assumption denotes the log-supermodularilty of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
-The FKG inequality is a key result which enables verification of negative association of random variables.
-Such as the one indicator variables for the new subsampling operation, we introduced in \cref{sec:negdep}.
+The FKG inequality is a key result which enables verification of negative association for random variables.
+This includes the indicator variables for the new subsampling operation we introduced in \cref{sec:negdep}.
 
@@
 This corresponds to Property~\ref{it:neg_dep_props:mult_mono} of \cref{pro:neg_dep_props}:
-If $X_i$ for $i \in I$ form a set of negatively associated variables, and $f$, $g$ are simultaneously monotone (or simultaneously antimonotone) real valued square integrable measurable functions, depending on disjoint components of $I$ then the expectations of the product are smaller than then product of the expectations of the composition of the functions $f,g$ with random variables $X_i$.
+If $X_i$ for $i \in I$ form a set of negatively associated variables, and $f$, $g$ are simultaneously monotone (or simultaneously antimonotone), real-valued, square-integrable, measurable functions, depending on disjoint components of $I$ then the expectations of the product are smaller than then product of the expectations of the composition of the functions $f,g$ with random variables $X_i$.
 Note that, if $X_i$ where independent the above would be true, without the monotonicity assumption.
 
+%TODO: I suggest to put n_subsets_distribution_neg_assoc instead here instead, since the Hoeffding inequality is never used in our CVM proof.
 The following is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables:

### 2025-02-25 paper/sec/original_proof.tex
minor edit
more uniform structure in formalization secs

@@
 \section{Transformation-based Proof}\label{sec:transformation_based_proof}
-In this section, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023},
-focusing on the challenging parts in its formalization.
+Here, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}, focusing on the challenging parts in its formalization.
 An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
@@
 Whereas Chakraborty et al.\ move directly from \circled{B} to \circled{D} with an informal argument, \circled{C} is a transformation we added in the formalization to bridge this gap.
+
+\begin{note}
+In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_TODO|.
+The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
+\lipicsEnd\end{note}
 
@@
 \end{center}
-\caption{An overview of two proof approaches for the CVM algorithm.}\label{fig:proof_overview}
+\caption{An overview of the two formalized proof approaches for the CVM algorithm.}\label{fig:proof_overview}
 \end{figure}

### 2025-02-25 paper/sec/formalization_neg_dep.tex
slight edit

@@
 
-A key issue, we faced during formalization, was the fact that there are many theorems of the form if a set of functions are either simultaneously monotone or anti-monotone.
-For this we introduce a notation, that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}.
-Which evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
+A key issue we faced during formalization, was that there are many theorems that condition on a set of functions being either simultaneously monotone or simultaneously anti-monotone.
+To reduce duplication, we introduce a notation, that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}; it evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
 For example the FKG inequality~\cite[Ch. 6]{alon2000},\cite{fortuin1971}
@@
 \]
-is true, if $f$ and $g$ are both monotone, or both antimonotone (on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function).
+is true, if $f$ and $g$ are both monotone, or both antimonotone on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function.
 The reverse inequality is also true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
@@
 \end{isabelle_cm}
-Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, where we can multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
+Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, where we multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
 The first assumption denotes the log-supermodularilty of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
@@
 
-Let us look into a few example results about negatively associated random variables:
+Let us summarize a few key example results about negatively associated random variables used in our proof:
 \begin{isabelle_cm}

### 2025-02-25 paper/sec/related_work.tex
slight cleanup related work

@@
 The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
-Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables the use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
+Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
 It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
-What makes the CVM algorithm unique is the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
+What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
 
 The aforementioned hash-based algorithms are biased; Flajolet et al.~\cite{flajolet1985} points this out and also provides bounds on the distance between the expected result and the cardinality of the stream.
-Most authors do not discuss the matter of unbiasedness but it is not hard to see that they are not.
+Most authors do not discuss the matter of biasedness but it is not hard to show.
 One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
@@
 \subsection{Probabilistic Invariants and Formalization}
-As far as we know probabilistic invariants have not been used to establish Gaussian tail-bounds.
-It is fairly common to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}, using recursive analysis, which is easy due to the linearity of expectations and --- in the case of independent variables --- variances.
+As far as we know, probabilistic invariants have not been used to establish Gaussian tail-bounds.
+However, it is fairly common to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}, using recursive analysis techniques.
+This is easy due to the linearity of expectations and---for independent random variables---variances.
 A simple example is the Morris-counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.
 
-There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using the characteristic function of loops~\cite{batz2023, mciver2005}.
-This works by establishing the limiting distribution of the state of the loop.
+There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using their characteristic functions~\cite{batz2023, mciver2005}.
+This approach works by establishing the limiting distribution of the state of the loop.
 De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
@@
 Instead, we investigate classes of functions of the distributions, which are relevant for the analysis.
-
 There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
@@
 
-Verification of randomized algorithms in general has been done by many authors~\cite{bosshard2024, eberl2020, karayel2022, karayel2023, Probabilistic_Prime_Tests-AFP, tan2024}.
+Finally, verification of randomized algorithms has been tackled by authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
+The most closely related efforts are the mechanizations of frequency moments algorithms by Karayel~\cite{karayel2022, karayel2023}.
+The functional invariant proof technique we introduce here should be applicable in any higher-order setting.

### 2025-02-25 paper/sec/original_proof.tex
more comments in Alg 4
minor edits

@@
     % Define nodes
-    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center] {\circled{A} CVM \\ (Alg.~\ref{alg:cvm})};
-    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=A] {\circled{B} CVM w/o\\ Line 11};
-    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=B] {\circled{C} Eager Ver.\\(Alg.~\ref{alg:cvm_simul})};
-    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=C] {\circled{D} Random \\ Process};
-    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, right of=D] {Chernoff \\ Bounds};
+    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center] {\circled{A} CVM \\ (Alg.~\ref{alg:cvm})};
+    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=A] {\circled{B} CVM w/o\\ Line 11};
+    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=B] {\circled{C} Eager Ver.\\(Alg.~\ref{alg:cvm_simul})};
+    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=C] {\circled{D} Random \\ Process};
+    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=D] {Chernoff \\ Bounds};
 
-    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, below of=B, node distance = 1.8cm] {Functional\\Invariant};
+    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=B, node distance = 2.0cm] {Functional\\Invariant};
 
-    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=0.8cm, align=center, below of=E, node distance = 1.8cm] {CVM PAC\\ Guarantee};
+    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=E, node distance = 2.0cm] {CVM PAC\\ Guarantee};
 
@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
@@
   \For{$i \gets 1$ to $l$}
-    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$}
+    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} \Comment insert $a_i$ if first $k$ flips are $1$s.
       \State $\chi \gets \chi \cup \{a_i\}$
@@
     \EndIf
-    \If{$|\chi| = n$}
-      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ \Comment keep elems.~whose $k+1$-th flip is $1$
       \State $k \gets k+1$

### 2025-02-25 paper/sec/intro.tex
minor edits
Editing pass in 1

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
@@
 The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
-Yet, for \cref{alg:cvm}, the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are independent; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
+However, for \cref{alg:cvm}, the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are independent; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
 The original error serves as a side motivation for this work.}
 For example, in Line~3 the Bernoulli distribution is sampled for the value $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
-The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
-To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known methods~\cite{alon2000,mitzenmacher2017,motwani1995}.
+The aforementioned sequence of transformations by Chakraborty et al.~results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
+To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known textbook methods~\cite{alon2000,mitzenmacher2017,motwani1995}.
 
@@
 
-To show the generality of our new technique, we introduce a new variant of the CVM algorithm, where the subsampling step in Line~9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
+To show the generality of our technique, we introduce a new variant of the CVM algorithm, where the subsampling step in Line~9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
@@
 
-The organization of this paper is as follows.
-\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL~\cite{nipkow2002}.
+The rest of this paper is organized as follows.
+\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL.
 \cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
@@
 \cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
-In \cref{sec:transformation_based_proof}, we discuss some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}.
+\cref{sec:transformation_based_proof} discusses some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.
 The final sections conclude with an overview of related work and a summary of our results.

### 2025-02-25 paper/sec/semra.tex
minor fixes
sec 2 minor pass

@@
 The CVM algorithm is a \emph{streaming} algorithm for the distinct elements problem.
-As shown in~\cref{alg:cvm}, given a data stream $a_1,\dots, a_l$, the goal of such algorithms is to return an accurate cardinality estimate for $A = \{a_1,\dots,a_l\}$.
+As shown in~\cref{alg:cvm}, given a data stream $a_1,\dots, a_l$, the goal of such algorithms is to return an accurate cardinality estimate for the set $A = \{a_1,\dots,a_l\}$.
 
@@
 for parameters $\varepsilon$ and $\delta$,
-i.e., the probability that the relative error of $R$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
+i.e., the probability that the relative error of $R$ with respect to $|A|$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
 Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
@@
 It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
-\end{remark}
+\lipicsEnd\end{remark}
 
@@
 In the second step, the algorithm flips a coin $q$ which depends on $p$.
-This has the consequence that we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
-The resulting distribution for $q$ is a \emph{compound distribution}, a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
+This has the consequence that, to semantically model $q$, we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
 
@@
 \item[Bind Combinator.]
-The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the second randomized algorithm consumes the output of the first; in monad notation, this is: $m \isa{\isasymbind} f$.
+The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the latter randomized algorithm consumes the output of the former; in monad notation, this is: $m \isa{\isasymbind} f$.
 Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which takes inputs from the probability space $m$.

### 2025-02-25 paper/sec/invariants.tex
sec 3 minor edits

@@
 
-In the loop iteration where element $s$ occurs the first time, it will be inserted with a probability $p$ in Lines 3--7.
+In the loop iteration where element $s$ occurs for the first time, it will be inserted with probability $p$ in Lines 3--7.
 This means, after Line 7, we have:
@@
 
-Let us explain how we can verify \cref{eq:indicator_eq} more formally.
+Let us see how we can verify \cref{eq:indicator_eq} more formally.
 For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
@@
 We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for Lines 3--7 (resp.~Lines 8--10).
-The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps:
+The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps over the stream:
 \[
@@
 where we parameterize $\mathrm{step}_1$ with the stream element that it processes.
-The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return} (\{\},1)$.
-It is easy to show by induction over the sequence of steps, we will have $0 < p \leq 1$ and $\chi \subseteq A$ for all possible states of the algorithm.
+The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return}\,(\{\},1)$.
+It is easy to show by induction over the sequence of steps, we have $0 < p \leq 1$ and $\chi \subseteq A$ for all possible states of the algorithm.
 
@@
 To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
+By~\cref{eq:integral_bind},
 \[
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
-    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma
+    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
 \]
@@
 The same invariant-based argument is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
-Then
+Then by~\cref{eq:integral_bind},
 \[
@@
 \subsection{Deriving A Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
-However, recall that we are interested in verifying the algorithm's PAC guarantee.
-A typical approach to establishing such a guarantee is to use Chernoff bounds which provide exponential tail bounds for the deviation of sums of independent random variables from their mean.
+With \cref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
+However, recall that we are interested in verifying the estimate's PAC guarantee.
+A typical approach to establishing such a guarantee is to use Chernoff bounds which provide exponential tail bounds (i.e., concentration bounds) for the deviation of sums of independent random variables from their mean.
 However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$ for $s \in A$, are dependent.
@@
 An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
-It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of $X$.
+It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of the random variable $X$.
 %\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
 %Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
-It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
+It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using the Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
 
@@
 At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be generalized to establish bounds for the above.
-Indeed, we managed to establish the following result:
+Indeed, we managed to establish the following generalized result:
 \begin{align}
@@
 However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
-This leads to a similar inequality:
+This leads to a similar invariant inequality:
 \begin{align}

### 2025-02-25 paper/sec/formalization.tex
minor edits till 7

@@
 Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach.
-We verified both the total variant, we introduced in the previous section and the original variant (\cref{alg:cvm}) from the introduction.
+We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.
 
@@
 
-We verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$; and we allow $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
+We verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
+Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
 Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-This is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+This different is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
 However, we would like to note that, neither Chakraborty et al., nor us, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
@@
 %The snippet also includes the relevant correctness statements, both unbiasedness and tail bounds.
-A snippet of the formalization of \cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized}.
+A snippet of the formalization of \cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized} (the formalization of \cref{alg:cvm} is very similar).
 We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
 The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
-Instead of representing the state using pairs, as we did in the informal proof, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability that the stream elements are in the buffer, respectively.
+Instead of representing the state using pairs, as we did in the informal discussion, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability that the stream elements are in the buffer, respectively.
 Isabelle/HOL provides notation closely related to informal pseudocode, so it is usually feasible to read a formal statement without expert knowledge.
@@
 %More precisely, \isa{set\ xs} denotes the finite set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
-The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for \isa{run{\isacharunderscore}{\kern0pt}algo}.
+The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for the output estimate \isa{R} from \isa{run{\isacharunderscore}{\kern0pt}algo}.
 
@@
 \end{isabelle_cm}
-The expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
+where the expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
 
-Our proofs are available in the supplement, both in mechanized form in Isabelle/HOL and as a pen-and-paper proof included in the proof document.
+Our proofs are available both in mechanized form in Isabelle/HOL and as a pen-and-paper proof included in the associated proof document.
 In practice, we developed the latter proof first and then mechanized it in Isabelle/HOL without much surprises.

### 2025-02-25 paper/sec/formalization_neg_dep.tex
minor edits till 7

@@
 4.1, 4.2]{motwani1995}.
-Another example application of the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
+Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
 \lipicsEnd\end{note}
@@
 \]
-is true, if $f$ and $g$ are both monotone, or both antimonotone on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function.
+is true, if $f$ and $g$ are both monotone, or both antimonotone, on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function.
 The reverse inequality is also true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
@@
 Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, where we multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
-The first assumption denotes the log-supermodularilty of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
+The first assumption denotes the log-supermodularity of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
 The FKG inequality is a key result which enables verification of negative association for random variables.
@@
 
-Let us summarize a few key example results about negatively associated random variables used in our proof:
+Let us summarize a few key formalized results for negatively associated random variables used in our proof:
 \begin{isabelle_cm}
@@
 \ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\end{isabelle_cm}
+
+%TODO: I think the following footnote should be placed here, if we put negative association for permutation distribution here.
+%\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
+%However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-02-25 paper/sec/invariants.tex
minor edits till 7

@@
 
-\subsection{Deriving A Simple Probabilistic Invariant}
+\subsection{Deriving a Simple Probabilistic Invariant}
 Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
@@
 
-\subsection{Deriving A Functional Probabilistic Invariant}
+\subsection{Deriving a Functional Probabilistic Invariant}
 With \cref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.

### 2025-02-25 paper/sec/negdep.tex
minor edits till 7

@@
 
-The following proposition summarizes some of the properties for negatively associated sets of random variables.
+The following proposition summarizes some important properties of negatively associated sets of random variables.
 \begin{proposition}[\cite{joagdev1983}]\label{pro:neg_dep_props}
@@
 This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
-\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
-However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
 Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:

### 2025-02-25 paper/sec/original_proof.tex
minor edits till 7

@@
-\section{Transformation-based Proof}\label{sec:transformation_based_proof}
+\section{Transformation-Based Proof}\label{sec:transformation_based_proof}
 Here, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}, focusing on the challenging parts in its formalization.
 An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
+\begin{note}
+In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_TODO|.
+The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
+\lipicsEnd\end{note}
 
@@
 Whereas Chakraborty et al.\ move directly from \circled{B} to \circled{D} with an informal argument, \circled{C} is a transformation we added in the formalization to bridge this gap.
-
-\begin{note}
-In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_TODO|.
-The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
-\lipicsEnd\end{note}
 
@@
 \subsection{A Bridging Transformation}
-Let us consider algorithm \circled{B} in a state where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$.
-The algorithm would perform a coin flip lazily with probability $p$ when it encounters the next stream element.
-
 \begin{algorithm}[t]
@@
 
+Let us consider algorithm \circled{B} in a state where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$.
+The algorithm would perform a coin flip lazily with probability $p$ when it encounters the next stream element.
 The transformation \circled{C} is shown in \cref{alg:cvm_simul}, and we prove that it computes precisely the same distribution as \circled{B}.

### 2025-02-25 paper/sec/related_work.tex
minor update related work

@@
 As far as we know, probabilistic invariants have not been used to establish Gaussian tail-bounds.
-However, it is fairly common to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}, using recursive analysis techniques.
-This is easy due to the linearity of expectations and---for independent random variables---variances.
+However, it is fairly common to use recursive analysis techniques to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}.
+This is easy due to the linearity of expectations and---for independent random variables---linearity of variances.
 A simple example is the Morris-counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.
@@
 Our approach differs from these techniques, by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
-Instead, we investigate classes of functions of the distributions, which are relevant for the analysis.
+Instead, we investigate invariants of classes of functions of the distributions, which are relevant for the analysis.
 There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.

### 2025-02-25 paper/sec/formalization_neg_dep.tex
change explanation around NA formal def

@@
 Our formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
-% TODO: YK: I am struggling to understand the paragraph below.
-% In reference to 2.1 in Joag-Dev/Proschan, do you mean they are missing the integrability conditions for f,g?
-However, the authors did not correctly define the constraints for the possible set of test functions $f$ and $g$ in \cref{def:neg_assoc}, which does not make sense formally.
-In the formalization, we assume that the functions are bounded and measurable.
-On the other hand, we derive that the condition is still true, if $f, g$ are square-integrable.
-Or alternatively integrable, and non-negative.
+However, their definition leaves the class of test functions $f$ and $g$ (in \cref{def:neg_assoc}) imprecise.
+In particular, for the formalization, we limit these functions to those that are bounded and measurable.
+Additionally, we provide elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is still true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.
 This is derived using the monotone-convergence theorem.
@@
 
-Let us summarize a few key formalized results for negatively associated random variables used in our proof:
+Let us summarize a few key formalized results for negatively associated random variables used in our CVM proof:
 \begin{isabelle_cm}

### 2025-02-25 paper/sec/negdep.tex
Paper: Replace example for neg. assoc.  + Improve end of original proof.

@@
 The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
-This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.%
+This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.
 Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
-%TODO: double check!
 \begin{equation}\label{eq:subsample_with_n_subsets}
@@
 Also, if $S$ is a singleton, the inequality becomes an equality.
+With this ingredient, we can conclude that our results about the original algorithm we derived in the previous section also hold for our new alternative algorithm (\cref{alg:cvm_new}).
+% Note: After reading this section. It stops too abrubtly at the previous sentence, without establishing what we have actually shown.

### 2025-02-25 paper/sec/original_proof.tex
Paper: Replace example for neg. assoc.  + Improve end of original proof.

@@
 The interesting fact is that we can distribute the sampling operation over composition:
+\begin{observation}\label{o:sample_distrib} Let $f,g$ be functions consuming a set of coin flips (indexed by $I$), where $g$ also consumes the output of $f$, such that,
+$f$ depends only on the coin-flips indexed by $J \subseteq I$ and $g$ depends on the complement $J - I$, then:
 \begin{isabelle_cm}
-  sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ f\ \isasymomega\ \isasymcirc\ g\ \isasymomega{\isacharparenright}\ \isacharequal\ sample\ g\ \isasymbind\ \isacharparenleft{\isasymlambda}x\isachardot\ sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ f\ \isasymomega\isachardot\ x\isacharparenright\isacharparenright
+  sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ g\ \isasymomega\ \isasymcirc\ f\ \isasymomega{\isacharparenright}\ \isacharequal\ sample\ f\ \isasymbind\ \isacharparenleft{\isasymlambda}x\isachardot\ sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ g\ \isasymomega\ x\isacharparenright\isacharparenright
 \end{isabelle_cm}
-if $f$ and $g$ depend on disjoint subsets of the coin flips, i.e., if $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
-By recursively applying the rule, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
+\end{observation}
+By recursively applying the observation, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
 This lets us readily transform \circled{C} to \circled{B} and prove their distributions equivalent.
@@
 And the next loop iteration, will indeed only read coin flips that were not read by the previous iteration.
-For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
-In our formalization, we abstract these bookkeeping details using the reader monad.
+%For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
 
+To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the cut between the set of coin flips $f$ and $g$ depend on, may depend on the result of $f$.
 %It is possible to express the above in a point-free manner using the reader monad.

### 2025-02-25 paper/sec/formalization.tex

@@
 We verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
-Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
+%Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
 Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-This different is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
-However, we would like to note that, neither Chakraborty et al., nor us, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
+The difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+However, we would like to note that, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
+% It feels weird to imply that Chakraborty et al. did not try to optimize. Added qualifier to make sure "we think" that they didn't.
 

### 2025-02-25 paper/sec/formalization_neg_dep.tex

@@
 \ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}i\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}c{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}linorder{\isacharunderscore}{\kern0pt}topology{\isacharparenright}{\kern0pt}\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}J\ {\isasymsubseteq}\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}J\ {\isasymsubseteq}\ I{\isacartoucheclose}\isanewline
 \ \ \isakeyword{assumes}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}f\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}g\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
@@
 
-%TODO: I suggest to put n_subsets_distribution_neg_assoc instead here instead, since the Hoeffding inequality is never used in our CVM proof.
-The following is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables:
+The following is the key result we use for the verification of the new CVM algorithm:
+The negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than $k$).
 \begin{isabelle_cm}
 \isacommand{lemma}\isamarkupfalse%
-\ hoeffding{\isacharunderscore}{\kern0pt}bound{\isacharunderscore}{\kern0pt}two{\isacharunderscore}{\kern0pt}sided{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}finite\ I{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ a\ i\ {\isasymle}\ b\ i{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ AE\ {\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}a\ i{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}b\ i{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}I\ {\isasymnoteq}\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{defines}\ {\isacartoucheopen}n\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ I{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{defines}\ {\isacartoucheopen}{\isasymmu}\ {\isasymequiv}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ expectation\ {\isacharparenleft}{\kern0pt}X\ i{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymge}\ {\isadigit{0}}{\isacartoucheclose}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\end{isabelle_cm}
+\ n{\isacharunderscore}{\kern0pt}subsets{\isacharunderscore}{\kern0pt}distribution{\isacharunderscore}{\kern0pt}neg{\isacharunderscore}{\kern0pt}assoc{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}finite\ S{\isacartoucheclose}\ {\isacartoucheopen}k\ {\isasymle}\ card\ S{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}p\ {\isasymequiv}\ pmf{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}set\ {\isacharbraceleft}{\kern0pt}T{\isachardot}{\kern0pt}\ T\ {\isasymsubseteq}\ S\ {\isasymand}\ card\ T\ {\isacharequal}{\kern0pt}\ k{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}neg{\isacharunderscore}{\kern0pt}assoc\ p\ {\isacharparenleft}{\kern0pt}{\isasymin}{\isacharparenright}{\kern0pt}\ S{\isacartoucheclose}
+\end{isabelle_cm}
+
+%TODO: I suggest to put n_subsets_distribution_neg_assoc instead here instead, since the Hoeffding inequality is never used in our CVM proof.
+%The following is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables:
+%\begin{isabelle_cm}
+%\isacommand{lemma}\isamarkupfalse%
+%\ hoeffding{\isacharunderscore}{\kern0pt}bound{\isacharunderscore}{\kern0pt}two{\isacharunderscore}{\kern0pt}sided{\isacharcolon}{\kern0pt}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}finite\ I{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ a\ i\ {\isasymle}\ b\ i{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ AE\ {\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}a\ i{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}b\ i{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}I\ {\isasymnoteq}\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{defines}\ {\isacartoucheopen}n\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ I{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{defines}\ {\isacartoucheopen}{\isasymmu}\ {\isasymequiv}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ expectation\ {\isacharparenleft}{\kern0pt}X\ i{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymge}\ {\isadigit{0}}{\isacartoucheclose}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\end{isabelle_cm}
 

### 2025-02-25 paper/sec/formalization_neg_dep.tex
Paper: Further edits on neg assoc section   * Add footnote about the proof of perm distributions.   * Due to space reasons: Replace example snippet (Property 1) with Hoeffding ineq for neg assoc vars, which requires less explanation.

@@
 
-Let us summarize a few key formalized results for negatively associated random variables used in our CVM proof:
+Let us summarize a few key formalized results for negatively associated random variables:
 \begin{isabelle_cm}
 \isacommand{lemma}\isamarkupfalse%
-\ neg{\isacharunderscore}{\kern0pt}assoc{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}mult{\isacharunderscore}{\kern0pt}mono{\isacharcolon}{\kern0pt}\isanewline
-\ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}i\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}c{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}linorder{\isacharunderscore}{\kern0pt}topology{\isacharparenright}{\kern0pt}\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}J\ {\isasymsubseteq}\ I{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}f\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}g\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ f{\isacartoucheclose}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ g{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ f\ J{\isacartoucheclose}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ g\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}f\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ J\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ \isanewline
-\ \ \isakeyword{assumes}\ {\isacartoucheopen}g\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
-\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}{\isasymomega}{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasympartial}M{\isacharparenright}{\kern0pt}{\isasymle}{\isacharparenleft}{\kern0pt}{\isasymintegral}\isasymomega{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}\isasymomega{\isachardot}{\kern0pt}\ g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\ hoeffding{\isacharunderscore}{\kern0pt}bound{\isacharunderscore}{\kern0pt}two{\isacharunderscore}{\kern0pt}sided{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}finite\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ a\ i\ {\isasymle}\ b\ i{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ AE\ {\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}a\ i{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}b\ i{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}I\ {\isasymnoteq}\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}n\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ I{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}{\isasymmu}\ {\isasymequiv}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ expectation\ {\isacharparenleft}{\kern0pt}X\ i{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymge}\ {\isadigit{0}}{\isacartoucheclose}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}
 \end{isabelle_cm}
-This corresponds to Property~\ref{it:neg_dep_props:mult_mono} of \cref{pro:neg_dep_props}:
-If $X_i$ for $i \in I$ form a set of negatively associated variables, and $f$, $g$ are simultaneously monotone (or simultaneously antimonotone), real-valued, square-integrable, measurable functions, depending on disjoint components of $I$ then the expectations of the product are smaller than then product of the expectations of the composition of the functions $f,g$ with random variables $X_i$.
-Note that, if $X_i$ where independent the above would be true, without the monotonicity assumption.
-
+The above is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables.
+%\begin{isabelle_cm}
+%\isacommand{lemma}\isamarkupfalse%
+%\ neg{\isacharunderscore}{\kern0pt}assoc{\isacharunderscore}{\kern0pt}imp{\isacharunderscore}{\kern0pt}mult{\isacharunderscore}{\kern0pt}mono{\isacharcolon}{\kern0pt}\isanewline
+%\ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}i\ {\isasymRightarrow}\ {\isacharprime}{\kern0pt}c{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}linorder{\isacharunderscore}{\kern0pt}topology{\isacharparenright}{\kern0pt}\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}J\ {\isasymsubseteq}\ I{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}f\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}square{\isacharunderscore}{\kern0pt}integrable\ M\ {\isacharparenleft}{\kern0pt}g\ {\isasymcirc}\ flip\ X{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ f{\isacartoucheclose}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymeta}\isactrlesub {\isacharparenright}{\kern0pt}\ g{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ f\ J{\isacartoucheclose}\ {\isacartoucheopen}depends{\isacharunderscore}{\kern0pt}on\ g\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}f\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ J\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\ \isanewline
+%\ \ \isakeyword{assumes}\ {\isacartoucheopen}g\ {\isasymin}\ borel{\isacharunderscore}{\kern0pt}measurable\ {\isacharparenleft}{\kern0pt}Pi\isactrlsub M\ {\isacharparenleft}{\kern0pt}I{\isacharminus}{\kern0pt}J{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymlambda}{\isacharunderscore}{\kern0pt}{\isachardot}{\kern0pt}\ borel{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+%\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}{\isasymomega}{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasympartial}M{\isacharparenright}{\kern0pt}{\isasymle}{\isacharparenleft}{\kern0pt}{\isasymintegral}\isasymomega{\isachardot}{\kern0pt}\ f{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}{\isasymintegral}{\kern0pt}\isasymomega{\isachardot}{\kern0pt}\ g{\isacharparenleft}{\kern0pt}{\isasymlambda}i{\isachardot}{\kern0pt}\ X\ i\ \isasymomega{\isacharparenright}{\kern0pt}{\isasympartial}M{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+%\end{isabelle_cm}
+%This corresponds to Property~\ref{it:neg_dep_props:mult_mono} of \cref{pro:neg_dep_props}:
+%If $X_i$ for $i \in I$ form a set of negatively associated variables, and $f$, $g$ are simultaneously monotone (or simultaneously antimonotone), real-valued, square-integrable, measurable functions, depending on disjoint components of $I$ then the expectations of the product are smaller than then product of the expectations of the composition of the functions $f,g$ with random variables $X_i$.
+%Note that, if $X_i$ where independent the above would be true, without the monotonicity assumption.
 The following is the key result we use for the verification of the new CVM algorithm:
@@
 \end{isabelle_cm}
-
+It is a consequence of a more general result, which we have also shown, that permuation distributions are negatively associated.
+We relied on the proof by Dubashi using the FKG inequality~\cite[Th. 10]{dubhashi1996}.%
+\footnote{There is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which seemed to be incorrect.}
+%However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
 %TODO: I suggest to put n_subsets_distribution_neg_assoc instead here instead, since the Hoeffding inequality is never used in our CVM proof.
@@
 %TODO: I think the following footnote should be placed here, if we put negative association for permutation distribution here.
-%\footnote{We could not verify the original proof by Joag-Dev and Proschan, which seems to be incorrect.
-%However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-02-26 paper/sec/intro.tex
add more line count to show benefit of invariant

@@
 
-We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
-As a rough point of comparison, verification of the CVM algorithm using our new technique requires only \locnew~lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \locold~lines.
+Our mechanization is carried out in Isabelle/HOL~\cite{nipkow2002}, where we generically formalized the invariant-based argument underlying both CVM variants in 753 lines.
+Specializing to the original CVM algorithm took an additional 250 lines, while the new variant took 191 more lines (with 2974 lines for the supporting negative association library).
+We also mechanized the transformation-based proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
+As a rough point of comparison showing the simplicity of our new proof, verification of the CVM algorithm using our new technique required only \locnew~(753 + 250)~lines, while the original proof required \locold~lines.
 
@@
 \cref{sec:transformation_based_proof} discusses some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.
-The final sections conclude with an overview of related work and a summary of our results.
+The final sections present related work and a summary of our results.
 

### 2025-02-26 paper/sec/intro.tex
Edit intro slightly.   - Move line count for neg assoc library to section 6.

--- 2025-02-25 paper/sec/intro.tex
@@
 
-We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
-As a rough point of comparison, verification of the CVM algorithm using our new technique requires only \locnew~lines in Isabelle/HOL~\cite{nipkow2002}, while the original proof required \locold~lines.
+We carried out the mechanizations using Isabelle/HOL~\cite{nipkow2002}, which comes with a large repository of foundational libraries~\cite{afp} for the verification of randomized algorithms.
+We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}, which provides a rough point of comparison:
+Indeed, the verification of the CVM algorithm using our new technique requires only \locnew~lines, while the original proof required \locold~lines.
 

### 2025-02-27 isabelle/CVM_Transforms/document/root.tex
wip docs

\documentclass[11pt, a4paper]{report}
\usepackage[T1]{fontenc}
\usepackage{isabelle, isabellesym, amsmath}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

\begin{document}

\title{Submission Example}
\author{placeholder}
\maketitle

\begin{abstract}
  This is an example submission to the Archive of Formal Proofs. It
  shows submission requirements and explains the structure of a simple
  typical submission.
\end{abstract}

\tableofcontents

% include generated text of all theories
\input{session}

% \bibliographystyle{abbrv}
% \bibliography{root}

\end{document}

### 2025-02-28 isabelle/CVM_Invariant/document/root.tex
reorganise, rename, improve readme

--- 2025-01-25 isabelle/CVM_Subsample/document/root.tex
@@
 \documentclass[11pt,a4paper]{article}
-\usepackage{isabelle,isabellesym}
-\usepackage{amsfonts,amsmath, amssymb,amsthm}
+
+\usepackage[T1]{fontenc}
+\usepackage{isabelle, isabellesym}
+\usepackage{amsfonts, amsmath, amssymb, amsthm}
 \usepackage[top=1.4in, bottom=1.4in, left=1.4in, right=1.4in]{geometry}
@@
 \title{Verification of the CVM algorithm with a New Recursive Analysis Technique}
-\author{Emin Karayel, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan,\\and Joe Watt}
+\author{Emin Karayel, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan,\\and Seng Joe Watt}
 \maketitle
@@
 In this entry, we verify the CVM algorithm's correctness using a new technique which simplifies
-the formal analysis considerably compared to the orignal proof by Chakraborty et
-al. The main idea is based on a probabilistic invariant that allows us to derive concentration bounds
-using the Cram\'{e}r--Chernoff method.
+the analysis considerably compared to the orignal proof by Chakraborty et al. The main idea is
+based on a probabilistic invariant that allows us to derive concentration bounds using the
+Cram\'{e}r--Chernoff method.
 
-This new technique opens up the possible algorithm design space, and we introduce a new variant of the
-CVM algorithm, that is total, and also has an additional property in addition
-to concentration: unbiasedness. This means the expected result of the algorithm is exactly equal to
-the desired result. The latter is also a new property, that neither the original CVM algorithm
-nor classic algorithms for the distinct elements problem possess.
+This new technique opens up the possible algorithm design space, and we introduce a new variant of
+the CVM algorithm, that is total, and also has an additional property in addition to concentration:
+unbiasedness. This means the expected result of the algorithm is exactly equal to the desired
+result. The latter is also a new property, that neither the original CVM algorithm nor classic
+algorithms for the distinct elements problem possess.
 \end{abstract}

### 2025-02-28 isabelle/Neg_Assoc/document/root.tex
reorganise, rename, improve readme

\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage{isabelle, isabellesym}
\usepackage{amsfonts, amsmath, amssymb}

\usepackage[only,bigsqcap]{stmaryrd}

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

\begin{document}

\title{Negatively Associated Random Variables}
\author{Emin Karayel}
\maketitle

\begin{abstract}
Negative Association is a generalization of independence for random variables, that retains some
of the key properties of independent random variables. In particular closure properties, such as
composition with monotone functions, as well as, the well-known Chernoff-Hoeffding bounds.

This entry introduces the concept and verifies the most important closure properties, as well as,
the concentration inequalities. It also verifies the FKG inequality, which is a generalization of
Chebyshev's sum inequality for distributive lattices and a key tool for establishing negative
association, but has also many applications beyond the context of negative association, in 
particular, statistical physics and graph theory.

As an example, permutation distributions are shown to be negatively associated, from which
many more sets of negatively random variables can be derived, such as, e.g., n-subsets, or the 
the balls-into-bins process.

Finally, the entry derives a correct false-positive rate for Bloom filters using the library.
\end{abstract}

\tableofcontents

\parindent 0pt\parskip 0.5ex

\input{session}

\bibliographystyle{abbrv}
\bibliography{root}

\end{document}

### 2025-02-28 isabelle/build.sh

@@ -1,2 +1,0 @@
-# isabelle build -v -b -D .
-isabelle build -v -b -o browser_info -o "document=pdf" -o "document_variants=document:outline=/proof,/ML" -D .

### 2025-02-28 cvm_unbiased/counter_example/src/Main.hs
Add haskell tool to verify counterexample pointed out in cvm_unbiased/main.tex (last section.)

@@ -1,0 +1,63 @@
+module Main where
+
+import Data.List (delete)
+import SimulationMonad
+
+threshold :: Int
+threshold = 10
+
+stream :: [Int]
+stream = [0..15]
+
+-- Remove the nth element of a list e.g. remove 1 [a,b,c] = [a,c]
+remove :: Int -> [a] -> [a]
+remove _ [] = error "unexpected"
+remove 0 (_:xs) = xs
+remove n (x:xs) = x:(remove (n-1) xs)
+
+-- Initial state
+f0_init :: ([a],Rational)
+f0_init = ([],1)
+
+-- Subsampling operation remove randomly (treshold/2 elements)
+subsample :: [Int] -> Dist [Int]
+subsample xs
+  | length xs * 2 == threshold = return xs
+  | otherwise = 
+      do index <- sample (length xs)
+         subsample (remove index xs)
+
+f0_loop :: ([Int], Rational) -> [Int] -> Dist ([Int],Rational)
+f0_loop (chi, p) [] = return (chi, p)
+f0_loop (chi, p) (x:xs) =
+  do let chi' = delete x chi
+     coin <- bernoulli p
+     let chi'' = (if coin then x:chi' else chi')
+     if length chi'' == threshold then -- buffer full 
+        do chi''' <- dedupe $ subsample chi''
+           f0_loop (chi''', p / 2) xs
+     else
+        f0_loop (chi'', p) xs
+
+f0_alg :: [Int] -> Dist ([Int], Rational)
+f0_alg xs = f0_loop f0_init xs
+
+test_f1 :: ([Int], Rational) -> Rational
+test_f1 (chi, _) = if (length $ filter (\x -> x < 10) chi) >= 5 then 1 else 0
+
+test_f2 :: ([Int], Rational) -> Rational
+test_f2 (chi, _) = fromIntegral $ length $ filter (\x -> x >= 10) chi
+
+-- Note: Both test_f1, test_f2 are monotone increasing and depend on disjoint indicator variables of chi.
+-- If the indicator variables of chi were n.a., then we would have E test_f1 * test_f2 <= E test_f1 * E test_f2
+-- I.e. the last inequality must be true. (But when we run the code, it will return False.) 
+main :: IO ()
+main = 
+  do let state = f0_alg stream
+     let x = expectation (fmap (test_f1) state)
+     let y = expectation (fmap (test_f2) state)
[…]

### 2025-02-28 cvm_unbiased/counter_example/src/SimulationMonad.hs

@@ -1,0 +1,42 @@
+-- This module introduces a monad with which it is possible to exactly evaluate the distribution
+-- of randomized algorithms. This works by keeping track of the possible states with their respective
+-- probabilities (using rational number arithmetic with big integers.)
+module SimulationMonad where
+
+import Data.Ratio
+import Data.List.Extra (groupSort)
+
+data Dist a = Dist { unDist :: [(a,Rational)] }
+
+dedupe :: (Ord a) => Dist a -> Dist a
+dedupe (Dist xs) = Dist $ map (\(x,vs) -> (x, sum vs)) $ groupSort xs
+
+instance Monad Dist where
+  return x = Dist [(x,1)]
+  m >>= f = Dist $
+    do (v,p) <- unDist m
+       (w,q) <- unDist (f v)
+       return (w,p*q)
+
+instance Applicative Dist where
+    pure = return
+    a <*> b = do
+        f <- a
+        x <- b
+        return (f x)
+        
+instance Functor Dist where
+  fmap f m = m >>= (return . f)
+
+expectation :: Dist Rational -> Rational
+expectation (Dist d) = sum $ map (\(x,p) -> x * p) d
+
+bernoulli :: Rational -> Dist Bool
+bernoulli p = Dist [(True,p),(False,1-p)]
+
+-- Sample uniformly from {0,..,n-1}
+sample :: Int -> Dist Int
+sample n = Dist $ map (\x -> (x,1 / fromIntegral n)) [0..n-1]
+
+
+

### 2025-02-28 paper/sec/formalization.tex
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 In our supplementary material, the invariant-based verification approach is in the directory \verb|CVM_Invariant|.
-The file \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
-The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Algorithm| verifies the new total and unbiased variant.
-It should be noted that only \verb|CVM_New_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
+The theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
+The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.
+Note that only \verb|CVM_New_Unbiased_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
 The total number of lines required for the verification of the original algorithm is \locnew~lines.
@@
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-The difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
-However, we would like to note that, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
+This difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+However, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
 % It feels weird to imply that Chakraborty et al. did not try to optimize. Added qualifier to make sure "we think" that they didn't.
@@
 %\isa{length\ xs} & For a sequence \isa{xs} the length of the sequence. \\
-\isa{real} & Type of real numbers and conversion between natural numbers into real numbers. \\
+\isa{real} & Type of real numbers and conversion from natural numbers into real numbers. \\
 \isa{nat} & Type of natural numbers (non-negative integers). \\
@@
 
-The theorem that establishes the correctness of the algorithm, i.e., that the relative error will be smaller than $\varepsilon$ with probability at least $1-\delta$ is expressed in the following snippet:
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will exceed $\varepsilon$ with probability at most $\delta$ is expressed in the following snippet:
 \begin{isabelle_cm}

### 2025-02-28 paper/sec/formalization_neg_dep.tex
Edit intro slightly.   - Move line count for neg assoc library to section 6.
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 In our supplementary material, the formalization of negative association is in the directory \verb|Neg_Assoc|.
-This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distribution|).
+This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distributions|).
 Although not needed for CVM, we have also mechanized the standard Chernoff bounds (\verb|Neg_Assoc_Chernoff_Bounds|), including the additive bounds by Hoeffding~\cite[Th.
@@
 Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
+In total the library contains 2974~lines of Isabelle code.
 \lipicsEnd\end{note}
@@
 A key issue we faced during formalization, was that there are many theorems that condition on a set of functions being either simultaneously monotone or simultaneously anti-monotone.
-To reduce duplication, we introduce a notation, that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}; it evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
+To reduce duplication, we introduce a notation that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}; it evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
 For example the FKG inequality~\cite[Ch. 6]{alon2000},\cite{fortuin1971}
@@
 \end{isabelle_cm}
-Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, where we multiply relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction, if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the reverse direction, otherwise.
+Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} multiplies relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and it is the reverse direction otherwise.
 The first assumption denotes the log-supermodularity of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
@@
 
-Let us summarize a few key formalized results for negatively associated random variables:
+Let us summarize a few key formalized results for negatively associated random variables in our library.
+The following is the well-known Hoeffding inequality~\cite{hoeffding1963} generalized for negatively associated random variables.
 \begin{isabelle_cm}
@@
 \end{isabelle_cm}
-The above is the well-known Hoeffding inequality~\cite{hoeffding1963} for negatively associated random variables.
 %\begin{isabelle_cm}
@@
 %Note that, if $X_i$ where independent the above would be true, without the monotonicity assumption.
-The following is the key result we use for the verification of the new CVM algorithm:
-The negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than $k$).
+
+Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than $k$).
 \begin{isabelle_cm}
@@
 \end{isabelle_cm}
-It is a consequence of a more general result, which we have also shown, that permuation distributions are negatively associated.
-We relied on the proof by Dubashi using the FKG inequality~\cite[Th. 10]{dubhashi1996}.%
-\footnote{There is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which seemed to be incorrect.}
+This is a consequence of a more general result, which we have also shown, that permutation distributions are negatively associated.
+We relied on the proof by Dubashi using the FKG inequality~\cite[Th. 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which seemed to be incorrect.
 %However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-02-28 paper/sec/intro.tex
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}, which provides a rough point of comparison:
-Indeed, the verification of the CVM algorithm using our new technique requires only \locnew~lines, while the original proof required \locold~lines.
+verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.
 
@@
 \begin{itemize}
-\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using probabilistic functional invariants;
+\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
 \item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.

### 2025-02-28 paper/sec/invariants.tex
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 \subsection{Deriving a Simple Probabilistic Invariant}
-Let us consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, i.e., $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
+Consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, so $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
 Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,\dots,a_m\}$ after loop iteration $m$.
@@
 
-The same invariant-based argument is possible for $\mathrm{step}_2$, i.e., let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
+The same invariant-based argument is possible for $\mathrm{step}_2$.
+Let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
 Then by~\cref{eq:integral_bind},
@@
 Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
-However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$.
+However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
 This leads to a similar invariant inequality:

### 2025-02-28 paper/sec/original_proof.tex
rename dirs
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 \begin{note}
-In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_TODO|.
+In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_Transforms|.
+To formalize probabilistic transformations (relating two distributions), we built on an existing relational program logic in Isabelle/HOL~\cite{lochbihler2016}.
 The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
@@
 A remaining question is how to formalize the transformation from \circled{B} to \circled{C}.
-Our insight is that it is best to solve the problem backwards, i.e., we start with the modified algorithm \circled{C}, which performs all the coin-flips in advance \emph{eagerly} and convert it back to \circled{B} which implicitly performs the coin flips \emph{lazily} at the point they are needed.
+Our insight is that it is best to solve the problem backwards, i.e., we start with the modified algorithm \circled{C}, which performs all the coin flips in advance \emph{eagerly} and convert it back to \circled{B} which implicitly performs the coin flips \emph{lazily} at the point they are needed.
 
@@
 \begin{observation}\label{o:sample_distrib} Let $f,g$ be functions consuming a set of coin flips (indexed by $I$), where $g$ also consumes the output of $f$, such that,
-$f$ depends only on the coin-flips indexed by $J \subseteq I$ and $g$ depends on the complement $J - I$, then:
+$f$ depends only on the coin flips indexed by $J \subseteq I$ and $g$ depends on the complement $I - J$, then:
 \begin{isabelle_cm}
@@
 A detail that we have simplified here is that the split of the index sets, e.g., which coin flips $f$ depends on and which coin flips $g$ depends on, may be dynamic.
-For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin-flips.
+For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin flips.
 This means, we have a situation, where the previous loop iteration communicates to the next loop iteration, which coin flips it depends on using the state.

### 2025-02-28 paper/sec/semra.tex
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
 The key idea is to model a randomized algorithm as a probability space representing the distribution of its results.
-Let us consider \cref{alg:example}.
+As an example, let us consider \cref{alg:example}.
 \begin{algorithm}[h!]
@@
 This has the consequence that, to semantically model $q$, we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
-The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ and $\Ber(\frac{2}{3})$.
+The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ (when $p = 0$) and $\Ber(\frac{5}{6})$ (when $p = 1$).
 

### 2025-02-28 paper/sec/formalization.tex
Paper: Remove comment about constants.   - I was intending to include a similar comment for the original proof as well.   - But requires too much space. (Better to drop it completely.)

--- 2025-02-25 paper/sec/formalization.tex
@@
 %Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
-Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
+%Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-The difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
-However, we would like to note that, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
+%This difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+%However, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
 % It feels weird to imply that Chakraborty et al. did not try to optimize. Added qualifier to make sure "we think" that they didn't.

### 2025-02-28 paper/sec/formalization.tex
Paper: Remove comment about constants.   - I was intending to include a similar comment for the original proof as well.   - But requires too much space. (Better to drop it completely.)

--- 2025-02-28 paper/sec/formalization.tex
@@
 %Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.
-Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
+%Another deviation is that we have better constant factors compared to the original presentation---the buffer size used by Chakraborty et al.
 is $12 \varepsilon^{-2} \log_2(8 l \delta^{-1}) \approx 17.3 \varepsilon^{-2} \ln(8 l \delta^{-1})$.
-This difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
-However, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
+%This difference is mainly due to the fact that the invariant-based proof does not require conditioning on particular values of $p$ (the subsampling probability at the end of the algorithm) in contrast to the transformation-based proof which applies Chernoff bounds for each possible value of $p$ (down to some lower bound $q$) and establishes the final error probability as a union bound.
+%However, we think neither Chakraborty et al., nor we, have tried to optimize constants and it should be possible to find constant factor improvements using a more careful analysis in both proofs.
 % It feels weird to imply that Chakraborty et al. did not try to optimize. Added qualifier to make sure "we think" that they didn't.

### 2025-03-01 paper/sec/formalization.tex
move (now) sentence into note

@@
 The total number of lines required for the verification of the original algorithm is \locnew~lines.
+In addition, we actually verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
 \lipicsEnd\end{note}
 
-We verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
 %Our formalization also allows $n \geq 12 \varepsilon^{-2} \ln(6 l \delta^{-1})$, i.e., $n$ can be any integer value greater than the bound.

### 2025-03-03 paper/sec/invariants.tex
delete extra word

@@
 with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
-This then allows us to use approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
+This then allows us to approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
 As an example, the upper tail bound can be derived as follows:

### 2025-03-03 isabelle/CVM_Transforms/document/root.tex
wip docs
reorganise, rename, improve readme
wip latex, trim whitespaces
condense theories, add docs
tweak formatting and text
fix citations
anonymize files
draft abstract

@@
 \documentclass[11pt, a4paper]{report}
+
 \usepackage[T1]{fontenc}
-\usepackage{isabelle, isabellesym, amsmath}
+\usepackage{isabelle, isabellesym}
+\usepackage{amsfonts, amsmath, amssymb}
+\usepackage[htt]{hyphenat}
 
@@
 
-\title{Submission Example}
-\author{placeholder}
+\title{Verification of the CVM algorithm}
+
+\author{
+  Seng Joe Watt,
+  Derek Khu,
+  Emin Karayel,
+  Kuldeep S. Meel,
+  \\
+  and Yong Kiam Tan
+}
+
 \maketitle
@@
 \begin{abstract}
-  This is an example submission to the Archive of Formal Proofs. It
-  shows submission requirements and explains the structure of a simple
-  typical submission.
+  Published in 2022, the CVM algorithm \cite{cvm_2022} is a remarkably simple
+  one for the distinct elements problem, avoiding intricate derandomization
+  techniques, while maintaining a close to optimal logarithmic space complexity.
+  Later, it was discovered that the proof was incorrect, with the authors
+  presenting a new proof in \cite{cvm_2023}.
+
+  This project formalises the new, corrected proof of the CVM algorithm, while
+  strengthening the correctness result of Theorem 2 in \cite{cvm_2023} to provide
+  the same correctness guarantee using a smaller threshold
+  (and hence space usage) of
+  $\lceil \frac{12}{\varepsilon^2} \text{log}_2 \frac{3m}{\delta} \rceil$,
+  where $m$ is the length of the input stream.
+
+  This corrected proof performs a sequence of transformations of the CVM
+  algorithm, akin to cryptographic game-hopping.
+  The reason for these transformations is that standard techniques for analysing
+  randomised algorithms such as the Chernoff bounds usually require independence
+  of state random variables, which is unfortunately absent from the CVM
+  algorithm, but was incorrecty claimed and used in the original proof.
+  The transformations address this by successively simplifying the
+  algorithm into one where the state variables are independent, enabling
+  the use of the Chernoff bounds.
+
+  As part of the project, we also formalise various useful results, in particular:
+  \begin{enumerate}
+    \item Unary and relational Hoare rules for reasoning about monadic folds
+    (ie. loops) over the \texttt{reader}, \texttt{pmf} and \texttt{spmf} monads,
+    which we use to justify the various program transformations.
+
+    \item Strong forms of the multiplicative Chernoff bounds for the binomial
+    distribution, derived from the Bennet-Bernstein inequalities.
+  \end{enumerate}
 \end{abstract}
@@
 
-% include generated text of all theories
-\input{session}
+\parindent 0pt
+\parskip 0.5ex
 
-% \bibliographystyle{abbrv}
-% \bibliography{root}
+\chapter{CVM algorithm formalisation}
+
+\input{CVM_Algo}
+\input{CVM_Algo_No_Fail}
+\input{CVM_Algo_Lazy_Eager}
+\input{CVM_Algo_Nondet_Binomial}
+
+\input{CVM_Error_Bounds}
+\input{CVM_Correctness}
+\input{CVM_Correctness_Instance}
+
+\chapter{Utility definitions and results}
+
+\input{Utils_Basic}
+\input{Utils_Reader_Monad}
+\input{Utils_Approx_Algo}
+\input{Utils_Concentration_Ineqs}
+
+\input{Utils_PMF_Basic}
+\input{Utils_PMF_Lazify}
+
+\input{Utils_SPMF_Basic}
+\input{Utils_SPMF_Rel_Hoare}
+
+\bibliographystyle{abbrv}
+\bibliography{root}
 

### 2025-03-04 paper/sec/abstract.tex
minor rewording of abstract

@@
 This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
-Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
+Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is both total and unbiased---neither of which are properties of the original algorithm.
 In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.

### 2025-03-04 paper/sec/original_proof.tex
minor edits to original_proof section

@@
 for fixed values of $K$ and $X$.
-To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with a specific values $k = K$.
+To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with specific values $k = K$.
 The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
@@
 The main idea is to automatically push down the coin flips through the expression tree of \cref{alg:cvm_simul}.
-To explain how this works, let us first define the \emph{sampling} function, i.e., let $f$ be a function that takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to independent unbiased coin flips as.
+To explain how this works, let us first define the \emph{sampling} function, i.e., let $f$ be a function that takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to independent unbiased coin flips as:
 \begin{isabelle_cm}
@@
 For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin flips.
-This means, we have a situation, where the previous loop iteration communicates to the next loop iteration, which coin flips it depends on using the state.
-And the next loop iteration, will indeed only read coin flips that were not read by the previous iteration.
+This means we have a situation where the previous loop iteration communicates to the next loop iteration which coin flips it depends on using the state.
+And the next loop iteration will indeed only read coin flips that were not read by the previous iteration.
 %For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.

### 2025-03-04 paper/sec/conclusion.tex
minor changes to related_work and conclusion

@@
 We presented the first formalization of the CVM algorithm using Isabelle/HOL.
-Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
+Central to our formalization is a novel invariant-based proof technique for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-03-04 paper/sec/related_work.tex
minor changes to related_work and conclusion

@@
 Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
-It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity is rather large.
+It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
 What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
@@
 There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
-However, these methods, do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.
+However, these methods do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.
 
-Finally, verification of randomized algorithms has been tackled by authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
+Finally, verification of randomized algorithms has been tackled by various authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
 The most closely related efforts are the mechanizations of frequency moments algorithms by Karayel~\cite{karayel2022, karayel2023}.

### 2025-03-04 isabelle/CVM_Transforms/document/root.tex
wording
slight fixes

@@
   strengthening the correctness result of Theorem 2 in \cite{cvm_2023} to provide
-  the same correctness guarantee using a smaller threshold
+  the same correctness guarantee using a slightly smaller threshold
   (and hence space usage) of
@@
 
-  This corrected proof performs a sequence of transformations of the CVM
-  algorithm, akin to cryptographic game-hopping.
-  The reason for these transformations is that standard techniques for analysing
-  randomised algorithms such as the Chernoff bounds usually require independence
-  of state random variables, which is unfortunately absent from the CVM
-  algorithm, but was incorrecty claimed and used in the original proof.
-  The transformations address this by successively simplifying the
-  algorithm into one where the state variables are independent, enabling
-  the use of the Chernoff bounds.
+  This proof performs a sequence of transformations from the CVM algorithm,
+  akin to cryptographic game-hopping.  The reason for these transformations is
+  that standard techniques for analysing randomised algorithms such as the
+  Chernoff bounds usually require independence of state random variables, which
+  is unfortunately absent from the CVM algorithm.  The transformations address
+  this by successively simplifying the algorithm into one where the state
+  variables are independent, enabling the use of the Chernoff bounds.
 
@@
     \item Strong forms of the multiplicative Chernoff bounds for the binomial
-    distribution, derived from the Bennet-Bernstein inequalities.
+    distribution, derived from the Bennett-Bernstein inequalities.
   \end{enumerate}

### 2025-03-04 paper/main.tex
Paper: Minor edits.   * Use central place for LOCs.
minor edits
fix bib except math
add more line count to show benefit of invariant
Edit intro slightly.   - Move line count for neg assoc library to section 6.
assorted minor fixes
main - latest line count, bib url
intro - minor
sec 2 - fix probabilities
sec 3 - minor on role of q
sec 4 - minor edits
sec 5 - minor edits
sec 6 - minor edits
sec 7 - mention relational reasoning, fix small bug

@@
+\RequirePackage[hyphens]{url}
 \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
@@
 \usepackage{tikz}
+\usetikzlibrary{arrows.meta}
 \usetikzlibrary{calc}
@@
 
+\newcommand\locnew{1003}
+\newcommand\locold{2630}
+
 \bibliographystyle{plainurl}% the mandatory bibstyle
@@
 
-\author{\textcolor{red}{Anounymous authors}}{\textcolor{red}{Anonymous affiliations}}{}{}{}
+\author{\textcolor{red}{Anonymous author(s)}}{\textcolor{red}{Anonymous affiliation(s)}}{}{}{}
 
-\authorrunning{\textcolor{red}{Anonymous}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{\textcolor{red}{Anonymous author(s)}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\Copyright{\textcolor{red}{Anonymous}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{\textcolor{red}{Anonymous author(s)}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 

### 2025-03-04 paper/sec/formalization_neg_dep.tex

@@
 
-Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than $k$).
+Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than or equal to $k$).
 \begin{isabelle_cm}

### 2025-03-04 paper/sec/negdep.tex

@@
 
-An interesting consequence of our invariant-based approach, is that it allowed us to devise and verify a refined version of the CVM algorithm, that is both total and unbiased.
+An interesting consequence of our invariant-based approach is that it allowed us to devise and verify a refined version of the CVM algorithm that is both total and unbiased.
 
@@
 \begin{algorithm}[t!]
-	\caption{New unbiased and total CVM algorithm variant.}\label{alg:cvm_new}
+	\caption{New total and unbiased CVM algorithm variant.}\label{alg:cvm_new}
 	\begin{algorithmic}[1]
@@
 
-Motivated by this, our new variant is shown in \cref{alg:cvm_new}.
-For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.\footnote{It is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.}
-Since this new subsampling step always reduces the size of $\chi$, the variant never returns $\bot$, i.e., it is \emph{total}.
+Motivated by this observation, our new variant is shown in \cref{alg:cvm_new}.
+For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.
+For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
+Since this new subsampling step always reduces the size of $\chi$, the variant is \emph{total} (never returns $\bot$).
 The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
@@
 \begin{definition}[Negative Association]\label{def:neg_assoc}
-A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds.
+A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds:
 \[
@@
 Also, if $S$ is a singleton, the inequality becomes an equality.
-With this ingredient, we can conclude that our results about the original algorithm we derived in the previous section also hold for our new alternative algorithm (\cref{alg:cvm_new}).
+With this ingredient, we can conclude that our results about the original algorithm derived in the previous section also hold for our new variant (\cref{alg:cvm_new}).
 % Note: After reading this section. It stops too abrubtly at the previous sentence, without establishing what we have actually shown.

### 2025-03-04 paper/sec/original_proof.tex

@@
 for fixed values of $K$ and $X$.
-To see how tail bounds can be derived from this inequality, note that it is possible to union bound the probability that the estimate exceeds the desired interval, with the probability of the event in conjunction with specific values $k = K$.
-The latter can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
+To see how tail bounds can be derived from this inequality, let us first consider the failure event where the algorithm \circled{C}'s estimate exceeds the desired estimation interval and it ends with some fixed value $k = K$.
+Using \circled{D}, this can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
+Now, we can take a union bound over all the possible values $K$ to establish a global bound for the failure event in \circled{C}.
 This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.

### 2025-03-28 paper/sec/intro.tex
fix a typo (too late now, but for later)

@@
 The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
-It should be stressed that the theory of negatively association is orthogonal to our new technique, but its formalization is also a contribution of this work.
+It should be stressed that the theory of negative association is orthogonal to our new technique, but its formalization is also a contribution of this work.
 

### 2025-06-18 paper/sec/formalization.tex
wip

@@
 The first line gives conditions on parameters $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
-The next line requires the buffer size $n$ to be larger than $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
+The next line requires the buffer size $n$ to be larger than or equal to $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
 Then, we introduce the abbreviation \isa{A} for the cardinality of the set of elements in the sequence \isa{xs}.
 %More precisely, \isa{set\ xs} denotes the finite set of distinct elements in a given sequence, and \isa{card} returns the cardinality of a finite set.
-The notation \isa{\isasymP{\isacharparenleft}x\ in\ M{\isachardot}\ P\ x\isacharparenright} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for the output estimate \isa{R} from \isa{run{\isacharunderscore}{\kern0pt}algo}.
+The notation \isaprob{x}{M}{P x} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for the output estimate \isa{R} from \isa{run{\isacharunderscore}{\kern0pt}algo}.
 
@@
 Our proofs are available both in mechanized form in Isabelle/HOL and as a pen-and-paper proof included in the associated proof document.
-In practice, we developed the latter proof first and then mechanized it in Isabelle/HOL without much surprises.
-Most of the lemmas can be one-to-one identified between both proofs; Isabelle's existing libraries, automation capabilities, and structured proof format were used extensively in our proofs.
+In practice, we developed the latter proof first and then mechanized it in Isabelle/HOL without much surprise.
+Most of the lemmas can be identified one-to-one between both proofs; Isabelle's existing libraries, automation capabilities, and structured proof format were used extensively in our proofs.
 

### 2025-06-18 paper/sec/formalization_neg_dep.tex
wip

@@
 Additionally, we provide elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is still true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.
-This is derived using the monotone-convergence theorem.
+This is derived using the monotone convergence theorem.
 
@@
 
-A key issue we faced during formalization, was that there are many theorems that condition on a set of functions being either simultaneously monotone or simultaneously anti-monotone.
+A key issue we faced during formalization was that there are many theorems that condition on a set of functions being either simultaneously monotone or simultaneously anti-monotone.
 To reduce duplication, we introduce a notation that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}; it evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.

### 2025-06-18 paper/sec/intro.tex
wip

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
@@
 Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
-Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is displayed in its entirety in~\cref{alg:cvm}.
+Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
+The algorithm is displayed in its entirety in~\cref{alg:cvm}.
+% Here, $\cdot \getsr \cdot$ denotes random sampling, and $\Ber(p)$ denotes a Bernoulli distribution with parameter $p$.
 
@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
@@
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
-More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 
-The modified subsampling step leads to additional dependence for the elements in $\chi$ which cannot be readily removed using transformations following the original proof.
+The modified subsampling step leads to additional dependence for the elements in $\chi$ which cannot be readily removed using transformations as was done in the original proof.
 Instead, we verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
-The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similar to independent random variables.
+The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similarly to independent random variables.
 It should be stressed that the theory of negative association is orthogonal to our new technique, but its formalization is also a contribution of this work.
@@
 verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.
+\footnote{
+  We count the total number of lines of Isabelle code in the whole project, excluding empty, comment and presentation related lines.
+}
 

### 2025-06-18 paper/sec/invariants.tex
wip

@@
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
-    \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
+    \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
+    % \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
 \]
-Note that we write $p$ or $\chi$ even though, we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
-To see that the right-hand-side is equal to $1$, it is useful to consider the cases where $a=s$ and the converse separately.
+Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
+To see that the right-hand side is equal to $1$, it is useful to consider cases on whether $a=s$.
 When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
@@
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
-    \int_{\Omega} \left(\ift{|\chi|=n}{\left(\int_{\mathrm{subsample}(\chi)} \frac{\indicat(s \in \tau)}{p/2} \, d \tau\right)}{\frac{\indicat(s \in \chi)}{p}} \right) \, d \sigma \textrm{.}
+    \integral{\Omega}
+      {\left(\ift{|\chi|=n}{\left(\integral{\mathrm{subsample}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}\right)}{\frac{\indicat(s \in \chi)}{p}} \right)}
+      {\sigma}
+    \textrm{.}
 \]
-Note that the true- and false-cases of the inner if-then-else both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
+Note that the \emph{true} and \emph{false} cases of the inner if-then-else both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
 If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
@@
 
-Of course, the exponential function in $M(t)$ is convex, so this is not yet the full story.
-However, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
+While the exponential function in $M(t)$ is convex, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
 This leads to a similar invariant inequality:

### 2025-06-18 paper/sec/semra.tex
wip

@@
 
-Importantly, CVM is a \emph{probably-approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
+Importantly, CVM is a \emph{probably approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
 $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
@@
 i.e., the probability that the relative error of $R$ with respect to $|A|$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
-Moreover, let us assume the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
+Moreover, let us assume the space needed to store each element in the stream is $b$ bits, so that the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
 
 \begin{remark}
-The optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+The theoretically optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
 It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
@@
 In the second step, the algorithm flips a coin $q$ which depends on $p$.
-This has the consequence that, to semantically model $q$, we have to consider probability space-valued functions, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+This has the consequence that, to semantically model $q$, we have to consider functions returning probability spaces, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
 The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ (when $p = 0$) and $\Ber(\frac{5}{6})$ (when $p = 1$).
@@
 \item[Return Combinator.]
-Given a singleton element $x$, we can construct the singleton probability space, assigning probability $1$ to $x$ and $0$ to everything else.
+Given an element $x$, we can construct the singleton probability space, assigning probability $1$ to $x$ and $0$ to everything else.
 In monad notation, this is written as: $\mathrm{\bf return}\, x$.

### 2025-06-18 paper/sec/intro.tex
elan on vars, bold if-then-else

@@
 \]
-where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
+where $c$ is a non-negative real constant, $h : \Rnonneg \rightarrow \Rnonneg$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
 By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.

### 2025-06-18 paper/sec/intro.tex
marvelous -> remarkable, rephrase

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} unexpectedly published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).

### 2025-06-19 paper/main.tex
wip
bold font for PL exprs
elan on vars, bold if-then-else

@@
 \usepackage[noend]{algpseudocode}
+
 \renewcommand{\algorithmicrequire}{\textbf{Input:}}
@@
 \newcommand{\todo}[1]{\textcolor{red}{#1}}
-\newcommand{\ift}[3]{\mathrm{if} \; #1 \; \mathrm{then} \; #2 \; \mathrm{else} \; #3}
+\newcommand{\ift}[3]{\mathbf{if} \; #1 \; \mathbf{then} \; #2 \; \mathbf{else} \; #3}
+\newcommand{\integral}[3]{\int_{#1} \! #2 \, \mathrm{d} #3}
+\newcommand{\isaprob}[3]{\isa{\isasymP{\isacharparenleft}#1 \isatext{in} #2{\isachardot} #3\isacharparenright}}
 
@@
 \DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
+\DeclareMathOperator{\Rnonneg}{\mathbb R_{\geq 0}}
 \DeclareMathOperator{\Ber}{\mathrm{Ber}}
@@
 
+% TODO: put in everyone's names here
+
+\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
+% \author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
+
 \author{\textcolor{red}{Anonymous author(s)}}{\textcolor{red}{Anonymous affiliation(s)}}{}{}{}
@@
 
-%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%\supplementdetails[linktext={opt.
+\supplement{TODO}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+% AFP entry
+\supplementdetails[linktext={opt.
+text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
+Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
+% Zenodo code repo
+\supplementdetails[linktext={opt.
 text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.

### 2025-06-20 paper/sec/intro.tex
address PR comments

--- 2025-06-18 paper/sec/intro.tex
@@
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
-More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value returned by the algorithm is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 

### 2025-06-20 paper/sec/semra.tex
address PR comments

@@
 \begin{remark}
-The theoretically optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+The asymptotically optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
 It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.

### 2025-06-20 paper/sec/intro.tex

@@
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
-More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value returned by the algorithm is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value of the algorithm's output is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 

### 2025-06-20 paper/sec/invariants.tex

@@
 \]
-Note that the \emph{true} and \emph{false} cases of the inner if-then-else both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
+Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
 If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
@@
 \end{align}
-for every non-negative concave function $h : \mathbb R_{\geq 0} \rightarrow \mathbb R_{\geq 0}$.
+for every non-negative concave function $h : \Rnonneg \rightarrow \Rnonneg$.
 We call this a \emph{functional probabilistic invariant} because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.

### 2025-06-20 paper/sec/negdep.tex

@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, s.t., $nf$ integer
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, such that $nf$ integer
   \For{$i \gets 1$ to $l$}
@@
 \begin{equation}\label[ineq]{i:subsample_condition}
-  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau)) \, d \tau \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
+  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))}{\tau} \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
 \end{equation}
@@
 Motivated by this observation, our new variant is shown in \cref{alg:cvm_new}.
-For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we instead pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and such that $nf$ is an integer.
+For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and $nf$ is an integer.
 For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
@@
 \begin{definition}
-For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, s.t., $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
+For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, such that $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
 \end{definition}
@@
 The following proposition summarizes some important properties of negatively associated sets of random variables.
-\begin{proposition}[\cite{joagdev1983}]\label{pro:neg_dep_props}
-Summary of results for negatively associated random variables.
+\begin{proposition}[Summary of results for negatively associated random variables \cite{joagdev1983}]\label{pro:neg_dep_props}
 \begin{enumerate}
@@
 \item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector-valued random variables $X$ and $Y$ are independent, then the union $X_1,\dots,X_n,Y_1,\dots,Y_m$ is a set of negatively associated random variables.
-\item If $X=(X_1,\ldots,X_n)$ are negatively associated and $f_1, \dots ,f_m : \mathbb R^n \rightarrow \mathbb R$ are all non-increasing or all non-decreasing functions, s.t., $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated and $f_1, \dots ,f_m : \mathbb R^n \rightarrow \mathbb R$ are all non-increasing or all non-decreasing functions, such that $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
 \item If $X_1,\dots,X_n$ are independent then $X_1,\dots,X_n$ are negatively associated.
@@
 %Of course, on the other hand independence is a stronger property and fewer sets of random variables are independent.
-To analyze our new subsampler, the following is an important lemma about negative associated random variables.
+To analyze our new subsampler, the following is an important lemma about negatively associated random variables.
 
@@
 \begin{proof}
-This follows from the definition of negative associativity (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
+This follows from the definition of negative association (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
 \end{proof}
@@
 The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
-This is a consequence of the fact that permutation distributions are negatively-associated~\cite[Th. 2.11]{joagdev1983}.
+This is a consequence of the fact that permutation distributions are negatively associated~\cite[Th. 2.11]{joagdev1983}.
 Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
 \begin{equation}\label{eq:subsample_with_n_subsets}
-  \int_{\mathrm{subsample}(\chi)} \prod_{s \in S} g(\indicat(s \in \tau))\,d\tau \leq
-  \prod_{s \in S} \int_{\mathrm{subsample}(\chi)} g(\indicat(s \in \tau))\,d\tau = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
+  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))} {\tau} \leq
+  \prod_{s \in S} \integral{\mathrm{subsample}(\chi)}{g(\indicat(s \in \tau))}{\tau} = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
 \end{equation}

### 2025-06-20 paper/sec/related_work.tex

@@
 The aforementioned hash-based algorithms are biased; Flajolet et al.~\cite{flajolet1985} points this out and also provides bounds on the distance between the expected result and the cardinality of the stream.
-Most authors do not discuss the matter of biasedness but it is not hard to show.
+Most authors do not discuss the matter of bias but it is not hard to show.
 One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
@@
 This is easy due to the linearity of expectations and---for independent random variables---linearity of variances.
-A simple example is the Morris-counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.
+A simple example is the Morris counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.
 
@@
 De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
-Our approach differs from these techniques, by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
+Our approach differs from these techniques by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
 Instead, we investigate invariants of classes of functions of the distributions, which are relevant for the analysis.
-There is research on automated evaluation of moments for restricted classes of loops, which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
+There is research on automated evaluation of moments for restricted classes of loops which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
 However, these methods do not extend to algorithms with branches or, more generally, algorithms which contain discrete operations.

### 2025-06-20 paper/main.tex
Fill in author information

@@
 % TODO: put in everyone's names here
+\author{Emin Karayel}{Technical University of Munich, Department of Computer Science, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
+\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{TODO}
+\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
+\author{Kuldeep S. Meel}{Georgia Institute of Technology \and University of Toronto}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{TODO}
+\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{TODO}
 
-\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
-% \author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
+\authorrunning{E. Karayel, S. J. Watt, Y. K. Tan, K. S. Meel, D. Khu} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\author{\textcolor{red}{Anonymous author(s)}}{\textcolor{red}{Anonymous affiliation(s)}}{}{}{}
-
-\authorrunning{\textcolor{red}{Anonymous author(s)}} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{\textcolor{red}{Anonymous author(s)}} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Emin Karayel, Seng Joe Watt, Yong Kiam Tan, Kuldeep S. Meel, Derek Khu} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 

### 2025-06-20 paper/main.tex
Add AFP entries as supplementary material.
- Note: We also have to reference the transform based proof on github.

@@
 % TODO: put in everyone's names here
-\author{Emin Karayel}{Technical University of Munich, Department of Computer Science, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
+\author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
 \author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{TODO}
@@
 
-\supplement{TODO}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-% AFP entry
-\supplementdetails[linktext={opt.
+\supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+%Template
+%\supplementdetails[linktext={opt.
 text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
 Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-% Zenodo code repo
-\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
+
+% TODO Add Reference to transformation based proof.
+\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
+\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}
 

### 2025-06-20 paper/sec/conclusion.tex
Clarify that the novel part of the new technique is that we are establishing exponential tail bounds.

--- 2025-02-25 paper/sec/conclusion.tex
@@
 We presented the first formalization of the CVM algorithm using Isabelle/HOL.
-Central to our formalization is a novel invariant-based proof technique for randomized algorithms which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
+Central to our formalization is a novel invariant-based proof technique to establish exponentially decreasing tail-bounds for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.

### 2025-06-20 paper/sec/related_work.tex
Clarify that the novel part of the new technique is that we are establishing exponential tail bounds.

@@
 \subsection{Probabilistic Invariants and Formalization}
-As far as we know, probabilistic invariants have not been used to establish Gaussian tail-bounds.
+As far as we know, probabilistic invariants have not been used to establish exponentially decreasing tail-bounds.
 However, it is fairly common to use recursive analysis techniques to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}.

### 2025-06-20 paper/sec/related_work.tex
Remove third person reference.

@@
 The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
-Karayel~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
+We~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
 It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
@@
 Finally, verification of randomized algorithms has been tackled by various authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
-The most closely related efforts are the mechanizations of frequency moments algorithms by Karayel~\cite{karayel2022, karayel2023}.
+The most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
 The functional invariant proof technique we introduce here should be applicable in any higher-order setting.

### 2025-06-20 paper/sec/intro.tex
Clarify that the novel part of the new technique is that we are establishing exponential tail bounds.
Add tangent in intro to explain where the term functional probabilistic invariant'' comes from.

@@
 %For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
+The term ``functional probabilistic invariant'' is new.
+It is borrowed loosely from the theory of the calculus of variations, where scalar valued maps---called functionals---from the problem space are used to solve optimization problems.
+In our case we are using scalar valued maps---i.e., functionals---from the distribution of the state of the algorithm, which lead us to the name for our new technique.
 We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.
@@
 \begin{itemize}
-\item Introduction of a new technique using functional probabilistic invariants to verify randomized algorithms inductively/recursively.
+\item Introduction of a new technique using functional probabilistic invariants to verify tail-bounds for randomized algorithms inductively/recursively.
 \item Verification of the original CVM algorithm using our new technique.

### 2025-06-21 paper/main.tex
add my funding source

@@
 \author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
-\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{TODO}
+\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
 \author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}

### 2025-06-21 paper/main.tex
add Kuldeep's funding source

@@
 \author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
-\author{Kuldeep S. Meel}{Georgia Institute of Technology \and University of Toronto}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{TODO}
-\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{TODO}
+\author{Kuldeep S.
+Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
+\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
 

### 2025-06-21 paper/sec/intro.tex
add comment in alg 1

@@
 Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
-The algorithm is displayed in its entirety in~\cref{alg:cvm}.
+The algorithm is displayed in its entirety in~\cref{alg:cvm}; h.
 % Here, $\cdot \getsr \cdot$ denotes random sampling, and $\Ber(p)$ denotes a Bernoulli distribution with parameter $p$.
@@
   \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
-    \If{$b$}
+    \State $b \getsr \Ber(p)$ \Comment randomly sample a bit $b$ from the Bernoulli distribution
+    \If{$b$} \Comment insert $a_i$ ($b$ is true with prob. $p$)
       \State $\chi \gets \chi \cup \{a_i\}$
-    \Else
+    \Else \Comment remove $a_i$ otherwise
       \State $\chi \gets \chi - \{a_i\}$

### 2025-06-21 paper/sec/conclusion.tex
Improve definition of the method and explain that it requires further examples to refine.

@@
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+In future work, it would be interesting to formalize other variations of subsampling for CVM.
 
-In future work, it would be interesting to formalize other variations of subsampling for CVM.
-More generally, one could further explore whether the technique we introduced here could be applied towards proofs of other randomized (streaming) algorithms.
+Our novel technique can be summarized by the following two steps:
+\begin{enumerate}
+\item Find functions of the state of the algorithm, for which it is possible to establish upper bounds of their expectation recursively. 
+\item Use those bounds to establish tail bounds on the result of the algorithm.
+\end{enumerate}
+However, we would like to note that, we do not yet have good further examples, where our approach can be applied to. (It is easy to construct artificial examples.)
+This also means that our definition of the method may require refinement.
+Another issue is that we do not have general strategies to identify the functionals to consider.
+Identification of realistic applications for this new method is an interesting avenue for future work.

### 2025-06-21 paper/sec/intro.tex
minor fixes in intro

@@
 
-In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} unexpectedly published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
 Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
@@
     \State $b \getsr \Ber(p)$ \Comment randomly sample a bit $b$ from the Bernoulli distribution
-    \If{$b$} \Comment insert $a_i$ ($b$ is true with prob. $p$)
+    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
       \State $\chi \gets \chi \cup \{a_i\}$

### 2025-06-21 paper/sec/intro.tex
some edits to intro

@@
 \]
-where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
+where the expectation is taken over the distribution of the state of the algorithm; $h$ is allowed to range over a class of functions (mapping states to real values); and $c$ is a fixed state (possibly chosen differently at each loop iteration).
 By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
 %For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a proof of the CVM algorithm.
-The term ``functional probabilistic invariant'' is new.
-It is borrowed loosely from the theory of the calculus of variations, where scalar valued maps---called functionals---from the problem space are used to solve optimization problems.
-In our case we are using scalar valued maps---i.e., functionals---from the distribution of the state of the algorithm, which lead us to the name for our new technique.
+We coined the term ``functional probabilistic invariant'', borrowing loosely from the theory of the calculus of variations, where scalar-valued maps---called functionals---from the problem space are used to solve optimization problems.
+In our case we are using scalar-valued maps---i.e., functionals---from the distribution of the state of the algorithm, which lead us to the name for our new technique.
 We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.
@@
 This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
-More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected value of the algorithm's output is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
+More interestingly, the variant is \emph{unbiased}, i.e., the expected value of the algorithm's output is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 

### 2025-06-21 paper/sec/conclusion.tex
reorder conclusion

--- 2025-03-04 paper/sec/conclusion.tex
@@
 We presented the first formalization of the CVM algorithm using Isabelle/HOL.
-Central to our formalization is a novel invariant-based proof technique for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method; comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
+Central to our formalization is a novel invariant-based proof technique to establish exponentially decreasing tail-bounds for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
+Our technique can be summarized by the following two steps:
+\begin{enumerate}
+\item Find functionals over the state distribution of the algorithm, for which it is possible to establish bounds on their expectation inductively/recursively.
+\item Use those bounds to establish tail bounds on the result of the algorithm.
+\end{enumerate}
+Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
 Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+In future work, it would be interesting to formalize other variations of subsampling for CVM.
 
-In future work, it would be interesting to formalize other variations of subsampling for CVM.
-More generally, one could further explore whether the technique we introduced here could be applied towards proofs of other randomized (streaming) algorithms.
+Note that we have yet to apply our proof technique to examples beyond CVM. (It is easy to construct artificial examples.)
+Identifying realistic applications for our new method is an interesting avenue for future work---this could lead to further refinements of the method, and a better understanding of how to identify suitable functionals for the proofs.

### 2025-06-21 paper/sec/invariants.tex
Explain where inequality i:subsample_condition comes from.
Note: This is due to reviewer A comment:
> l261: What is "f"? I generally have some trouble understanding where (5) comes from

@@
 By~\cref{eq:integral_bind},
-\[
+\begin{equation}\label{eq:step_1_exp}
   \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
@@
     % \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
-\]
+\end{equation}
 Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
@@
 Let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
-Then by~\cref{eq:integral_bind},
-\[
-  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right] =
+Then by~\cref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
+\begin{equation}\label{eq:step_2_exp}
+%  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right]
     \integral{\Omega}
@@
     \textrm{.}
-\]
+\end{equation}
 Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.

### 2025-06-21 paper/sec/negdep.tex
Explain where inequality i:subsample_condition comes from.
Note: This is due to reviewer A comment:
> l261: What is "f"? I generally have some trouble understanding where (5) comes from

@@
 
-When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation:
+When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation.
+It should be noted that the condition becomes apparent while establishing \cref{i:func_invariant} using similar but more general steps as described in Equations~\ref{eq:step_1_exp} and \ref{eq:step_2_exp}.%
+\footnote{A step-by-step proof of the derivation is also available in the appendix of the formalization~\cite[Appendix~A]{CVM_Distinct_Elements-AFP}.}
 \begin{equation}\label[ineq]{i:subsample_condition}

### 2025-06-23 paper/sec/abstract.tex
make abstract slightly more grammatically correct?

--- 2025-02-24 paper/sec/abstract.tex
@@
 
-Central to our formalization is a new proof technique based on \emph{functional probabilistic invariants}, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
+Central to our formalization is a new proof technique based on functional probabilistic invariants, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
 This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
-Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is total and unbiased---neither property is possessed by the original algorithm.
+Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is both total and unbiased---neither of which is a property of the original algorithm.
 In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.

### 2025-06-23 paper/sec/intro.tex
remove stray characters
fix too much whitespace

@@
 Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
-The algorithm is displayed in its entirety in~\cref{alg:cvm}; h.
+The algorithm is displayed in its entirety in~\cref{alg:cvm}.
 % Here, $\cdot \getsr \cdot$ denotes random sampling, and $\Ber(p)$ denotes a Bernoulli distribution with parameter $p$.
@@
 We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}, which provides a rough point of comparison:
-verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.
-\footnote{
-  We count the total number of lines of Isabelle code in the whole project, excluding empty, comment and presentation related lines.
+verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.\footnote{
+We count the total number of lines of Isabelle code in the whole project, excluding empty, comment, and presentation-related lines.
 }

### 2025-06-26 paper/sec/formalization_neg_dep.tex
Explain incompleteness of Joag-devs proof regarding neg. association of permutation distributions.

@@
 This is a consequence of a more general result, which we have also shown, that permutation distributions are negatively associated.
-We relied on the proof by Dubashi using the FKG inequality~\cite[Th. 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which seemed to be incorrect.
+We relied on the proof by Dubashi et al.\ using the FKG inequality~\cite[Th.
+10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th.
+2.11]{joagdev1983}, which is incomplete\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$.
+That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality.
+We think Dubashi et al.\ developed their proof to complete it.}.
 %However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-06-28 paper/main.tex
fix author order
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 \author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
-\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
+\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
 \author{Kuldeep S.
 Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
-\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
+\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
 
-\authorrunning{E. Karayel, S. J. Watt, Y. K. Tan, K. S. Meel, D. Khu} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\Copyright{Emin Karayel, Seng Joe Watt, Yong Kiam Tan, Kuldeep S. Meel, Derek Khu} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 \supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-%Template
-%\supplementdetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g.
-Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
-
-% TODO Add Reference to transformation based proof.
-\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
-\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}
+\supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
+\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:0d79e265cf5d4d19a0dd1d7e79ed0119cf39ae22]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
+\supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}
 

### 2025-06-28 paper/sec/formalization.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach.
+Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach~\cite{CVM_Distinct_Elements}.
 We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.
@@
 \begin{note}
-In our supplementary material, the invariant-based verification approach is in the directory \verb|CVM_Invariant|.
-The theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
+In our supplementary material~\cite{CVM_Distinct_Elements-AFP}, the theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
 The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.

### 2025-06-28 paper/sec/formalization_neg_dep.tex
fix spelling of Dubhashi
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 \begin{note}
-In our supplementary material, the formalization of negative association is in the directory \verb|Neg_Assoc|.
+The formalization of the theory of negative association is included in a separate AFP entry~\cite{Negative_Association-AFP}.
 This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distributions|).
@@
 This is a consequence of a more general result, which we have also shown, that permutation distributions are negatively associated.
-We relied on the proof by Dubashi et al.\ using the FKG inequality~\cite[Th.
+We relied on the proof by Dubhashi et al.\ using the FKG inequality~\cite[Th.
 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th.
@@
 That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality.
-We think Dubashi et al.\ developed their proof to complete it.}.
-%However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
+We think Dubhashi et al.\ developed their proof to complete it.}.
+%However Dubhashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}
 %TODO: I suggest to put n_subsets_distribution_neg_assoc instead here instead, since the Hoeffding inequality is never used in our CVM proof.

### 2025-06-28 paper/sec/intro.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 \begin{itemize}
-\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
-\item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
-\item formalization of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
+\item formalization~\cite{CVM_Distinct_Elements-AFP} of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
+\item formalization~\cite{Negative_Association-AFP} of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
+\item formalization~\cite{CVM_Transforms-Github} of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
 \end{itemize}

### 2025-06-28 paper/sec/original_proof.tex
wip
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 \begin{note}
-In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_Transforms|.
+The transformation-based formalization of the CVM algorthm is included in the supplementary material~\cite{CVM_Transforms-Github}.
 To formalize probabilistic transformations (relating two distributions), we built on an existing relational program logic in Isabelle/HOL~\cite{lochbihler2016}.
@@
   \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$

### 2025-06-28 paper/sec/related_work.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

@@
 Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
-Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family; the space complexity of this algorithm is $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) b)$, where we assume that $b$ is the number of bits required to represent the stream elements.
-This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$ since there is no logarithmic dependency on $l$.
+Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family.
 
-Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented increasingly sophisticated solutions.
-The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
+Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} improved on the solution by Alon et al.
+For example, Bar-Yossef et al.\ present a solution (Algorithm~3 in their work) with a space-complexity of $\bigo(\ln (\delta^{-1}) (\varepsilon^{-2}(\ln(\varepsilon^{-1})+\ln b) + b)))$, which can be implemented in practice.
+This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$. In particular, there is no dependency on the length of the stream $l$.
+The more recent and more sophisticated solution by B\l{}asiok is space-optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
 We~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.

### 2025-06-28 paper/main.tex
Paper: Update links to supplementary material

@@
 \supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
-\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:0d79e265cf5d4d19a0dd1d7e79ed0119cf39ae22]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
+\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:c677f1e3ced670373c05ac835f1cd342a2777e20]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
 \supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

### 2025-06-30 paper/sec/formalization.tex
minor fixes to intro

@@
 \section{Formalization of the CVM algorithm}\label{sec:formalization}
-Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach~\cite{CVM_Distinct_Elements}.
+Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach~\cite{CVM_Distinct_Elements-AFP}.
 We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.

### 2025-06-30 paper/sec/intro.tex
minor fixes to intro

@@
 The original error serves as a side motivation for this work.}
-For example, in Line~3 the Bernoulli distribution is sampled for the value $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
+For example, in Line~3 the Bernoulli distribution is sampled with the parameter $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
 The aforementioned sequence of transformations by Chakraborty et al.~results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.

### 2025-06-30 paper/sec/formalization_neg_dep.tex
minor fixes

@@
 Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
-In total the library contains 2974~lines of Isabelle code.
+In total, the library contains 2974~lines of Isabelle code.
 \lipicsEnd\end{note}
@@
 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th.
-2.11]{joagdev1983}, which is incomplete\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$.
+2.11]{joagdev1983}, which is incomplete.\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$.
 That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality.
-We think Dubhashi et al.\ developed their proof to complete it.}.
+We think Dubhashi et al.\ developed their proof to complete it.}
 %However Dubhashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-06-30 paper/sec/negdep.tex
Explain the parameter f.
This related to the reviewer A comment:
l261: What is "f"?
minor fixes

@@
 \end{equation}
-for all non-negative functions $g$ and any $S \subseteq \chi$.
+for all non-negative functions $g$ and any $S \subseteq \chi$, where $f$ is the probability of retaining each element in the subsampling step of the algorithm.
 Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
@@
 The following proposition summarizes some important properties of negatively associated sets of random variables.
-\begin{proposition}[Summary of results for negatively associated random variables \cite{joagdev1983}]\label{pro:neg_dep_props}
+\begin{proposition}[Summary of results for negatively associated random variables \cite{joagdev1983}]\label{pro:neg_dep_props}~
 \begin{enumerate}

### 2025-06-30 paper/sec/original_proof.tex
minor fixes

@@
 For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin flips.
-This means we have a situation where the previous loop iteration communicates to the next loop iteration which coin flips it depends on using the state.
-And the next loop iteration will indeed only read coin flips that were not read by the previous iteration.
+This means we have a situation where the previous loop iteration communicates to the next loop iteration which coin flips it depends on using the state; and the next loop iteration will indeed only read coin flips that were not read by the previous iteration.
 %For the column index of the coin flips, this is based on the stream index, but for the row index of the coin flips, it is based on the state variable $k$.
 
-To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the cut between the set of coin flips $f$ and $g$ depend on, may depend on the result of $f$.
+To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the index $J$ splitting the set of coin flips $f$ and $g$ depend on, may itself depend on the result of $f$.
 %It is possible to express the above in a point-free manner using the reader monad.

### 2025-06-30 paper/sec/semra.tex
minor fixes

@@
 i.e., the probability that the relative error of $R$ with respect to $|A|$ exceeds $\varepsilon$ is at most $\delta$. % I think here it's good to have both a text and formal description.
-Moreover, let us assume the space needed to store each element in the stream is $b$ bits, so that the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
+Moreover, let us assume that the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
 

### 2025-06-30 paper/sec/related_work.tex
Paper: Additional related work and discussion of probabilistic hoare logic.

@@
 It is important to note that there are several practical solutions for the distinct elements problem.
-The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013,pettie2021}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
+The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
 Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
@@
 It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
+A comprehensive review of distinct elements algorithms has be compiled by Pettie and Wang~\cite[Table~1]{pettie2021}.
 What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
@@
 
-Finally, verification of randomized algorithms has been tackled by various authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
+Finally, verification of randomized algorithms expressed in functional programming languages has been tackled by various authors using various proof assistants~\cite{audebaud2006,bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024, tassarotti2021}.
+Moreover, for classical imperatively expressed randomized algorithms, there are approaches based on probabilistic hoare logic~\cite{denhartog1999}.
+Interesting is the work by Haselwarter et al.~\cite{haselwarter2025} in this context, which discusses some of the issues we encountered in the transformation-based proof (Section~\ref{sec:transformation_based_proof}), such (approximate) equivalence between related algorithms.
+However, because this work emphasizes the representation of randomized algorithms using the Giry monad, we did not deeply explore probabilistic logic based methods.
+
 The most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
-The functional invariant proof technique we introduce here should be applicable in any higher-order setting.

### 2025-06-30 paper/sec/semra.tex
Paper: Additional related work and discussion of probabilistic hoare logic.

@@
 We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
-A thorough discussion of the concept in the context of Isabelle/HOL has been written, for example, by Eberl et al.~\cite{eberl2020} and Lochbihler~\cite{lochbihler2016}.
+Multiple authors provide more thorough discussions of the concept in the context of Isabelle and other proof assistants~\cite{audebaud2006,eberl2020, lochbihler2016}.
 

### 2025-06-30 paper/sec/formalization_neg_dep.tex
Paper: Improve explanation regarding test functions for negative association.

@@
 However, their definition leaves the class of test functions $f$ and $g$ (in \cref{def:neg_assoc}) imprecise.
-In particular, for the formalization, we limit these functions to those that are bounded and measurable.
-Additionally, we provide elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is still true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.
+In the formalization, we use different conditions for introduction vs. elimination rules.
+In paritcular, for introduction rules, the test functions are bounded and measurable.
+However, we provide stronger elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.
 This is derived using the monotone convergence theorem.

### 2025-06-30 paper/sec/original_proof.tex
minor edit

@@
 
-To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the index $J$ splitting the set of coin flips $f$ and $g$ depend on, may itself depend on the result of $f$.
+To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the set of indices $J$ splitting the set of coin flips $f$ and $g$ depend on, may itself depend on the result of $f$.
 %It is possible to express the above in a point-free manner using the reader monad.

### 2025-06-30 paper/sec/formalization_neg_dep.tex
fix bibs and minor typos

@@
 However, their definition leaves the class of test functions $f$ and $g$ (in \cref{def:neg_assoc}) imprecise.
-In the formalization, we use different conditions for introduction vs. elimination rules.
-In paritcular, for introduction rules, the test functions are bounded and measurable.
+In the formalization, we use different conditions for introduction and elimination rules.
+In particular, for introduction rules, the test functions are bounded and measurable.
 However, we provide stronger elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.

### 2025-06-30 paper/sec/related_work.tex
fix bibs and minor typos

--- 2025-06-28 paper/sec/related_work.tex
@@
 It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
+A comprehensive review of distinct elements algorithms has been compiled by Pettie and Wang~\cite[Table~1]{pettie2021}.
 What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
@@
 
-Finally, verification of randomized algorithms has been tackled by various authors using various proof assistants~\cite{bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024}.
-The most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
-The functional invariant proof technique we introduce here should be applicable in any higher-order setting.
+Finally, verification of randomized algorithms expressed in functional programming languages has been tackled by various authors using various proof assistants~\cite{audebaud2009,bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024, tassarotti2021}; the most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
+Moreover, for classical imperative randomized algorithms, there are approaches based on probabilistic Hoare logic~\cite{denhartog2002}.
+An interesting related work by Haselwarter et al.~\cite{haselwarter2025} discusses some of the issues we encountered in the transformation-based proof (Section~\ref{sec:transformation_based_proof}), such as (approximate) equivalence between related randomized algorithms.
+However, because our work reasons directly over the semantics of randomized algorithms expressed in the Giry monad, we did not deeply explore probabilistic logic-based methods.

### 2025-06-30 paper/sec/semra.tex
fix bibs and minor typos

@@
 We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
-Multiple authors provide more thorough discussions of the concept in the context of Isabelle and other proof assistants~\cite{audebaud2006,eberl2020, lochbihler2016}.
+Multiple authors provide more thorough discussions of the concept in the context of Isabelle and other proof assistants~\cite{audebaud2009,eberl2020, lochbihler2016}.
 

### 2025-06-30 archive/cvm_unbiased/main.tex
archive dir

--- 2025-01-06 cvm_unbiased/main.tex
@@
   & \underset{\textrm{Markov}}{\leq} & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ I(p \geq q) \exp ( t p^{-1} |\chi| ) \right] \\
- & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
+ & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) \exp ( t p^{-1} I(s \in \chi) ) \right] \\
  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} I(p \geq q) h( p^{-1} I(s \in \chi) ) \right] \\
@@
 \[
-  f(\eps) := -q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
+  f(\eps) := q \ln(1-\eps) (1-\eps) -\ln(1 - q \eps) - \frac{q}{2}\eps^2 = t (1-\eps) - \ln(h(1)) - \frac{q}{2} \eps^2
 \]

### 2025-06-30 paper/sec/formalization.tex
Paper: Additional related work and discussion of probabilistic hoare logic.
Paper: Update equation references.

@@
 
-%TODO: make sure these names are correct!
 \begin{note}
-In our supplementary material~\cite{CVM_Distinct_Elements-AFP}, the theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
+In our supplementary material~\cite{CVM_Distinct_Elements-AFP}, the theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill Inequality~\ref{i:subsample_condition}.
 The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.

### 2025-06-30 paper/sec/invariants.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.
Paper: Update equation references.

@@
 
-Let us see how we can verify \cref{eq:indicator_eq} more formally.
+Let us see how we can verify Equation~\ref{eq:indicator_eq} more formally.
 For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
@@
 
-Let us verify that~\cref{eq:indicator_eq} is preserved as an invariant over all steps.
-To verify that $\mathrm{step}_1\, a$ preserves \cref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills \cref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
-By~\cref{eq:integral_bind},
+Let us verify that Equation~\ref{eq:indicator_eq} is preserved as an invariant over all steps.
+To verify that $\mathrm{step}_1\, a$ preserves Equation~\ref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills Equation~\ref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
+By Equation~\ref{eq:integral_bind},
 \begin{equation}\label{eq:step_1_exp}
@@
 The same invariant-based argument is possible for $\mathrm{step}_2$.
-Let us assume $\Omega$ is a probability space of states fulfilling \cref{eq:indicator_eq}.
-Then by~\cref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
+Let us assume $\Omega$ is a probability space of states fulfilling Equation~\ref{eq:indicator_eq}.
+Then by Equation~\ref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
 \begin{equation}\label{eq:step_2_exp}
@@
 Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
-This completes the invariance proof for~\cref{eq:indicator_eq}.
+This completes the invariance proof for Equation~\ref{eq:indicator_eq}.
 
 \subsection{Deriving a Functional Probabilistic Invariant}
-With \cref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
+With Equation~\ref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
 However, recall that we are interested in verifying the estimate's PAC guarantee.
@@
 for $h(x) = \exp(tx)$.
-At this point, it is tempting to see whether the proof for \cref{eq:indicator_eq} can be generalized to establish bounds for the above.
+At this point, it is tempting to see whether the proof for Equation~\ref{eq:indicator_eq} can be generalized to establish bounds for the above.
 Indeed, we managed to establish the following generalized result:
@@
 %The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
-The formalization in the supplementary material contains a detailed informal step-by-step proof using our approach in its appendix.
-Besides the use of \cref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.
+The formalization in the supplementary material~\cite{CVM_Distinct_Elements-AFP} contains a detailed informal step-by-step proof using our approach in its appendix.
+Besides the use of Equation~\ref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.
 

### 2025-06-30 paper/sec/negdep.tex
Paper: Update equation references.

@@
 
-When we look at the subsampling step of~\cref{alg:cvm}, our invariant~\cref{i:func_invariant} imposes the following condition on the subsampling operation.
-It should be noted that the condition becomes apparent while establishing \cref{i:func_invariant} using similar but more general steps as described in Equations~\ref{eq:step_1_exp} and \ref{eq:step_2_exp}.%
+When we look at the subsampling step of~\cref{alg:cvm}, our invariant (Inequality~\ref{i:func_invariant}) imposes the following condition on the subsampling operation.
+It should be noted that the condition becomes apparent while establishing Inequality~\ref{i:func_invariant} using similar but more general steps as described in Equations~\ref{eq:step_1_exp} and \ref{eq:step_2_exp}.%
 \footnote{A step-by-step proof of the derivation is also available in the appendix of the formalization~\cite[Appendix~A]{CVM_Distinct_Elements-AFP}.}
@@
 for all non-negative functions $g$ and any $S \subseteq \chi$, where $f$ is the probability of retaining each element in the subsampling step of the algorithm.
-Any subsampling step that satisfies this functional inequality can be used while still preserving~\cref{i:func_invariant} for the algorithm.
+Any subsampling step that satisfies this functional inequality can be used while still preserving Inequality~\ref{i:func_invariant} for the algorithm.
 
@@
 The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
-These depend crucially on establishing~\cref{i:subsample_condition} for the new subsampler, for which we need a new concept.
+These depend crucially on establishing Inequality~\ref{i:subsample_condition} for the new subsampler, for which we need a new concept.
 

### 2025-06-30 paper/sec/related_work.tex
bib fix
Paper: Add reference to paper by Haslbeck,Nipkow 2016 (mentioned by reviewers).

@@
 This is easy due to the linearity of expectations and---for independent random variables---linearity of variances.
-A simple example is the Morris counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2005}.
+A simple example is the Morris counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2017}.
 
@@
 
-Finally, verification of randomized algorithms expressed in functional programming languages has been tackled by various authors using various proof assistants~\cite{audebaud2009,bosshard2024,demedeiros2024, eberl2020,gopinathan20,hurd03, Probabilistic_Prime_Tests-AFP, tan2024, tassarotti2021}; the most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
+Finally, verification of randomized algorithms expressed in functional programming languages has been tackled by various authors using various proof assistants~\cite{audebaud2009,bosshard2024,demedeiros2024, eberl2020,gopinathan20,haslbeck2016,hurd03,Probabilistic_Prime_Tests-AFP,tan2024,tassarotti2021}; the most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
 Moreover, for classical imperative randomized algorithms, there are approaches based on probabilistic Hoare logic~\cite{denhartog2002}.

### 2025-07-01 paper/sec/invariants.tex
Paper: Clarify the definition of functional probabilistic invariants in Section 3.2

@@
 For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
-%For that, let us describe the states of the algorithm using pairs composed of a subset of $A$ representing the buffer $\chi$ and a real number representing $p$.
-%Note that, this means $p$ and $\chi$ are random variables, with respect to distributions of the state of the algorithm.
 We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for Lines 3--7 (resp.~Lines 8--10).
@@
     \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
-    % \int_\Omega \int_{\Ber(p)} p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right) \, d \tau d \sigma \textrm{.}
 \end{equation}
@@
 \begin{equation}\label{eq:step_2_exp}
-%  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} \left[\frac{X_s}{p}\right]
     \integral{\Omega}
@@
 It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of the random variable $X$.
-%\footnote{The moment generating function is sometimes only defined for some values of $t$ (when the corresponding integral exists), and this is typically an interval including $0$; the inequality is only applicable wherever $M(t)$ is defined.}
-%Since the probability space representing the state of the CVM algorithm is finite, $M(t)$ will always be defined in our case.
 It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using the Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
@@
 for every non-negative concave function $h : \Rnonneg \rightarrow \Rnonneg$.
-We call this a \emph{functional probabilistic invariant} because one can establish it for all valid choices of $h$ with a single induction over steps of the randomized algorithm.
-
-While the exponential function in $M(t)$ is convex, we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
+However, the exponential function in $M(t)$ is convex, but we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
 This leads to a similar invariant inequality:
@@
 As an example, the upper tail bound can be derived as follows:
-%TODO: should we use exp(...) everywhere instead of mixing with e^ ?
-EK: That was the previous verison, main issue is the chain looks very lopsided, with almost all of the terms being on the right edge.
-(It looked too ugly.)--- OK
-% The convention is now, if exp is applied to a random variable we use the long notation and otherwise the power notation, which might actually be (a bit) helpful.
 \begin{eqnarray*}
@@
 For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2023}, with $q = \frac{n}{4 |A|}$.
-%The formalization accompanying this work~\todo{[cite]} contains a detailed informal step-by-step proof using our approach in its appendix.
 The formalization in the supplementary material~\cite{CVM_Distinct_Elements-AFP} contains a detailed informal step-by-step proof using our approach in its appendix.
@@
 
-The main takeaway here is that it is possible to derive useful and general probabilistic invariants by considering expectations of classes of functions of the state, proved using recursion or induction over the algorithm itself.
-As far as we know this method of establishing tail bounds for randomized algorithms is new.
+We call inequalities like Inequality~\ref{i:func_invariant} and \ref{i:func_invariant_capped}: \emph{functional probabilistic invariants}.
+The inequalities can be established using induction over the steps of the randomized algorithm.
+Even, though, the actual distribution of the states themselves are not known, nor do we think it is possible to find useful closed form descriptions for them, the invariant establishes valuable information about the distribution of the state.
+In this case, enough information, to establish exponential tail bounds for the algorithm.

### 2025-07-02 paper/sec/negdep.tex
Paper: Clarify discrete operations.

@@
 for all non-negative functions $g$ and any $S \subseteq \chi$, where $f$ is the probability of retaining each element in the subsampling step of the algorithm.
-Any subsampling step that satisfies this functional inequality can be used while still preserving Inequality~\ref{i:func_invariant} for the algorithm.
+(This parameter $f$ was fixed to $\frac{1}{2}$ in the original presentation of the algorithm for simplicity.)
+Any subsampling step that satisfies Inequality~\ref{i:subsample_condition} can be used while still preserving Inequalities~\ref{i:func_invariant} and \ref{i:func_invariant_capped} for the algorithm.
 
@@
 For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and $nf$ is an integer.
-For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding a random element from $\chi$ in the subsampling step.
+For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding just one random element from $\chi$ in the subsampling step.
 Since this new subsampling step always reduces the size of $\chi$, the variant is \emph{total} (never returns $\bot$).

### 2025-07-04 paper/main.tex
minor fixes to intro
Preparations for submission to LIPIcs.   - Inline included files.   - Remove commented blocks.   - Update volume macros from portal.

@@
 \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-%This is a template for producing LIPIcs articles.
-%See lipics-v2021-authors-guidelines.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling autoref support, use "autoref"
-%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-%for enabling thm-restate support, use "thm-restate"
-%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
-%for producing a PDF according the PDF/A standard, add "pdfa"
-
-%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
 
@@
 \DeclareMathOperator{\bigo}{\mathcal O}
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
 
@@
 \newcommand\locnew{1003}
-\newcommand\locold{2630}
+\newcommand\locold{2634}
 
@@
 
-%\title{Verification of the CVM algorithm with a New Recursive Analysis Technique} %TODO Please add
-\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant} %TODO maybe?
+\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant}
 
-%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line
-
-% TODO: put in everyone's names here
 \author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
@@
 
-\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan}
 
-\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan}
 
@@
 
-\category{} %optional, e.g. invited paper
+\category{}
+\relatedversion{}
 
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversiondetails[linktext={opt.
-text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g.
-Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional
+\supplement{Isabelle/HOL Formalization}
 
-\supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
 \supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
@@
 
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-%\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
+\EventEditors{Yannick Forster and Chantal Keller}
 \EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
+\EventLongTitle{16th International Conference on Interactive Theorem Proving (ITP 2025)}
+\EventShortTitle{ITP 2025}
+\EventAcronym{ITP}
+\EventYear{2025}
+\EventDate{September 29--October 2, 2025}
+\EventLocation{Reykjavik, Iceland}
 \EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+\SeriesVolume{352}
+\ArticleNo{34}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%NOTE: we have to inline all of these into one file for the final version if accepted.
+\begin{abstract}
+Estimating the number of distinct elements in a data stream is a classic problem with numerous applications in computer science.
+We formalize a recent, remarkably simple, randomized algorithm for this problem due to Chakraborty, Vinodchandran, and Meel (called the CVM algorithm).
+Their algorithm deviated considerably from the state of the art, due to its avoidance of intricate derandomization techniques, while still maintaining a close-to-optimal logarithmic space complexity.
 
-\input{sec/abstract}
+Central to our formalization is a new proof technique based on functional probabilistic invariants, which allows us to derive concentration bounds using the Cram\'{e}r--Chernoff method without relying on independence.
+This simplifies the formal analysis considerably compared to the original proof by Chakraborty et al.
+Moreover, our technique opens up the possible algorithm design space; we demonstrate this by introducing and verifying a new variant of the CVM algorithm that is both total and unbiased---neither of which is a property of the original algorithm.
+In this paper, we introduce the proof technique, describe its use in mechanizing both versions of the CVM algorithm in Isabelle/HOL, and present a supporting formalized library on negatively associated random variables used to verify the latter variant.
+\end{abstract}
 
-\input{sec/intro}
+\section{Introduction}
+\label{sec:intro}
 
-\input{sec/semra}
+In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a remarkable streaming algorithm for the distinct elements problem~\cite{quanta}.
+Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
+One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
+Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
+Another property is its simplicity, which is why the authors called it ``an algorithm for the textbook''.
+The algorithm is displayed in its entirety in~\cref{alg:cvm}.
 
-\input{sec/invariants}
+\begin{algorithm}[h!]
+	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
+  \For{$i \gets 1$ to $l$}
+    \State $b \getsr \Ber(p)$ \Comment randomly sample a bit $b$ from the Bernoulli distribution
+    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else \Comment remove $a_i$ otherwise
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment discard elements of $\chi$ independently with prob. $\frac{1}{2}$
+      \State $p \gets \frac{p}{2}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \Return $\bot$ \Comment fail if $\chi$ remains full
+    \EndIf
+  \EndFor
+  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}
 
-\input{sec/negdep}
+The pen-and-paper analysis of CVM~\cite{chakraborty2022,chakraborty2023} relies on a sequence of transformations of the algorithm.
+The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
+However, for \cref{alg:cvm}, the state variables are far from being independent.\footnote{There is an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022} that the indicator functions for elements in $\chi$ are independent; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
+The original error serves as a side motivation for this work.}
+For example, in Line~3 the Bernoulli distribution is sampled with the parameter $p$, which itself depends on previous random operations; similarly, the subsampling step in Line~9 is only applied if the buffer $\chi$ is full, which also depends on previous random operations.
+The aforementioned sequence of transformations by Chakraborty et al.~results in another randomized algorithm which can be analyzed using standard methods, and from which the desired results for the original algorithm can be deduced.
+To our knowledge, it seems impossible to analyze \cref{alg:cvm} more directly using known textbook methods~\cite{alon2000,mitzenmacher2017,motwani1995}.
 
-\input{sec/formalization}
+In this paper, we present a new technique for analyzing randomized algorithms which yields a direct and substantially more general proof of the CVM algorithm.
+Our approach is very similar to how deterministic algorithms are verified using loop invariants.
+The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
+\[
+  \expect [ h ] \leq h(c)
+\]
+where the expectation is taken over the distribution of the state of the algorithm; $h$ is allowed to range over a class of functions (mapping states to real values); and $c$ is a fixed state (possibly chosen differently at each loop iteration).
+By first establishing such an invariant for \cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
+We coined the term ``functional probabilistic invariant'', borrowing loosely from the theory of the calculus of variations, where scalar-valued maps---called functionals---from the problem space are used to solve optimization problems.
+In our case we are using scalar-valued maps---i.e., functionals---from the distribution of the state of the algorithm, which lead us to the name for our new technique.
+We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to mechanized theorem proving.
 
-\input{sec/formalization_neg_dep}
+To show the generality of our technique, we introduce a new variant of the CVM algorithm, where the subsampling step in Line~9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$ instead of independently discarding each element with some probability.
+This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in Line~11 becomes obsolete.
+More interestingly, the variant is \emph{unbiased}, i.e., the expected value of the algorithm's output is exactly the cardinality of the elements in the stream; this is a new property that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.
 
-\input{sec/original_proof}
+The modified subsampling step leads to additional dependence for the elements in $\chi$ which cannot be readily removed using transformations as was done in the original proof.
+Instead, we verify the new variant with our probabilistic invariant-based approach, using results from the theory of negatively associated random variables~\cite{joagdev1983} to establish the desired functional invariant.
+The concept of negative association is a generalization of independence; importantly, negatively associated variables observe closure properties and fulfill Chernoff--Hoeffding bounds similarly to independent random variables.
+It should be stressed that the theory of negative association is orthogonal to our new technique, but its formalization is also a contribution of this work.
 
-\input{sec/related_work}
+In summary, our main contributions are:
+\begin{itemize}
+\item Introduction of a new technique using functional probabilistic invariants to verify tail-bounds for randomized algorithms inductively/recursively.
+\item Verification of the original CVM algorithm using our new technique.
+\item Presentation and verification of a new variant of CVM that is total and unbiased.
+\item Formalization of a theory of negatively associated random variables used to analyze the new CVM variant.
+\end{itemize}
 
-\input{sec/conclusion}
+We carried out the mechanizations using Isabelle/HOL~\cite{nipkow2002}, which comes with a large repository of foundational libraries~\cite{afp} for the verification of randomized algorithms.
+We have also mechanized the transformation-based CVM proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}, which provides a rough point of comparison:
+verification of the CVM algorithm using our new technique required only \locnew~lines, while the original proof required \locold~lines.\footnote{
+We count the total number of lines of Isabelle code in the whole project, excluding empty, comment, and presentation-related lines.
+}
 
-%%
-%% Bibliography
-%%
+The rest of this paper is organized as follows.
+\cref{sec:background} provides background information on randomized algorithms, in particular on their semantics in Isabelle/HOL.
+\cref{sec:invariants} introduces our new technique and explains how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
+\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
+\cref{sec:formalization} presents the formalization of both variants of the algorithm, and \cref{sec:formalization_neg_dep} describes our new formalized library on negatively associated random variables.
+\cref{sec:transformation_based_proof} discusses some challenges faced in our alternative verification of CVM using the transformation-based proof by Chakraborty et al.
+The final sections present related work and a summary of our results.
 
-%% Please use bibtex,
+The supplementary material contains:
+\begin{itemize}
+\item formalization~\cite{CVM_Distinct_Elements-AFP} of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
+\item formalization~\cite{Negative_Association-AFP} of a library for negative association; and
+\item formalization~\cite{CVM_Transforms-Github} of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
+\end{itemize}
+
+\section{Background}
+\label{sec:background}
+
+\subsection{Randomized Algorithms and Distinct Elements}
+
+The CVM algorithm is a \emph{streaming} algorithm for the distinct elements problem.
+As shown in~\cref{alg:cvm}, given a data stream $a_1,\dots, a_l$, the goal of such algorithms is to return an accurate cardinality estimate for the set $A = \{a_1,\dots,a_l\}$.
+
+Importantly, CVM is a \emph{probably approximately correct} (PAC) algorithm where its output estimate $R$ satisfies
+$\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+for parameters $\varepsilon$ and $\delta$,
+i.e., the probability that the relative error of $R$ with respect to $|A|$ exceeds $\varepsilon$ is at most $\delta$.
+Moreover, let us assume that the space needed to store each element in the stream is $b$ bits, then the CVM algorithm requires only $\bigo(\varepsilon^{-2} b \ln(\delta^{-1} l))$ bits of mutable state, which is far less than storing each stream element deterministically.
+
+\begin{remark}
+The asymptotically optimal randomized algorithm for distinct elements requires $\bigo( \varepsilon^{-2} \ln \delta + b)$ bits, but it requires more advanced algorithmic techniques.
+It would not be possible to present using such elementary steps as in \cref{alg:cvm} as it involves computations in finite fields and random walks in expander graphs~\cite{blasiok2020, karayel2023}.
+\lipicsEnd\end{remark}
+
+\subsection{Semantics of Randomized Algorithms}
+We briefly review how reasoning about randomized algorithms works in Isabelle/HOL using the Giry monad~\cite{giry1982}.
+Multiple authors provide more thorough discussions of the concept in the context of Isabelle and other proof assistants~\cite{audebaud2009,eberl2020, lochbihler2016}.
+
+The key idea is to model a randomized algorithm as a probability space representing the distribution of its results.
+As an example, let us consider \cref{alg:example}.
+\begin{algorithm}[h!]
+\caption{Example for sequential composition.}\label{alg:example}
+\begin{algorithmic}[1]
+\State $p \getsr \Ber(\frac{1}{2})$
+\State $q \getsr \Ber(\frac{1}{3}+\frac{p}{2})$
+\State \Return $q$
+\end{algorithmic}
+\end{algorithm}%
+
+In the first step,~\cref{alg:example} flips a fair coin, such that $p$ is $1$ with probability $\frac{1}{2}$ and $0$ otherwise; the notation $\Ber(p)$ represents the Bernoulli distribution.
+In the second step, the algorithm flips a coin $q$ which depends on $p$.
+This has the consequence that, to semantically model $q$, we have to consider functions returning probability spaces, like: $p \mapsto \Ber(\frac{1}{3}+\frac{p}{2})$, which is being \emph{bound} to the distribution of $p$.
+The resulting distribution for $q$ is a \emph{compound distribution} resulting from a combination of $\Ber(\frac{1}{3})$ (when $p = 0$) and $\Ber(\frac{5}{6})$ (when $p = 1$).
+
+This example captures the main aspects of modeling randomized algorithms in the Giry monad.
+Indeed, randomized algorithms can be modeled using the following ingredients:
+
+\begin{description}
+\item[Primitive Random Operations.] For example, a simple fair coin flip is represented using the Bernoulli distribution, $\Ber(\frac{1}{2})$.
+\item[Return Combinator.]
+Given an element $x$, we can construct the singleton probability space, assigning probability $1$ to $x$ and $0$ to everything else.
+In monad notation, this is written as: $\mathrm{\bf return}\, x$.
+
+\item[Bind Combinator.]
+The bind combinator represents sequential composition of two randomized algorithms $m$ and $f$, where the latter randomized algorithm consumes the output of the former; in monad notation, this is: $m \isa{\isasymbind} f$.
+Mathematically, this is the most involved operation, because $f$ is a function returning probability spaces, which takes inputs from the probability space $m$.
+
+Let us consider an event $A$ in the probability space $m \isa{\isasymbind} f$.
+Its probability can be evaluated by integrating over its probabilities in $f$ with respect to $m$:
+\[
+  \prob_{m \isa{\isasymbind} f} (A) = \int_m \prob_{f(x)} (A) \, d x \textrm{.}
+\]
+
+Another key property is the calculation of expectations;
+if $h$ is a random variable over $m \isa{\isasymbind} f$, we can compute its expectation as:
+\begin{equation}
+  \label{eq:integral_bind}
+  \expect_{m \isa{\isasymbind} f} [h] = \int_m \expect_{f(x)} [h] \, d x \textrm{.}
+\end{equation}
+
+Equation~\ref{eq:integral_bind} is crucially used to establish the invariants we introduce next in~\cref{sec:invariants}.
+\end{description}
+
+\section{Functional Probabilistic Invariants}\label{sec:invariants}
+In this section, we will derive our new technique using \cref{alg:cvm} as an example.
+Let us start by briefly reviewing the algorithm---its state is a buffer $\chi$ (initially empty) and a fraction $p > 0$ (initially set to $1$).
+The buffer tracks a subset of the elements of the stream encountered so far, with maximal size $n$ chosen according to the desired accuracy parameters $\varepsilon$, $\delta$, and the stream size $l$.
+The algorithm iterates over the stream elements, adding each one to the buffer with probability $p$ or conversely---if the current stream element is already in the buffer---removing it with probability $(1-p)$ (Lines 3--7).
+If the number of elements in the buffer reaches the maximal size $n$, the subsampling operation is executed, which discards each element in $\chi$ independently with probability $\frac{1}{2}$; then, $p$ is adjusted to reflect the fact that the buffer now contains each element with probability $p_\text{new} = \frac{p_\text{old}}{2}$ (Lines 8--10).
+If the subsampling operation fails, i.e., if no elements get discarded, then the algorithm fails returning $\bot$ (Line 11).
+After processing the stream, the algorithm returns $\frac{|\chi|}{p}$ as a probably-approximately correct estimate for the number of distinct elements in the stream.
+
+\begin{remark}
+For our discussion below, it is convenient to analyze \cref{alg:cvm} without Line 11, i.e., we will skip the second check of $|\chi|=n$ determining whether the subsampling step succeeded.
+This modified version simplifies our analysis as we do not have to worry about the possibility of the algorithm failing (returning $\bot$).
+This transformation is also used in the original CVM proof~\cite{chakraborty2023}, where the total variational distance between these two variants of the algorithms is shown to be at most $\frac{\delta}{2}$.
+Thus, probability bounds derived for the modified version can be transferred to the original algorithm, with a correction term of $\frac{\delta}{2}$.
+\lipicsEnd\end{remark}
+
+\subsection{Deriving a Simple Probabilistic Invariant}
+Consider the random variables $X_s := \indicat(s \in \chi)$ indicating the presence of a stream element $s \in A = \{a_1,\ldots,a_l\}$ in the buffer, where we write $\indicat$ for the indicator of a predicate, so $\indicat(\mathrm{true}) = 1$ and $\indicat(\mathrm{false}) = 0$.
+Before the algorithm first encounters the stream element $s$, $X_s$ will be $0$ unconditionally, because the buffer $\chi$ is always a subset of the stream elements processed so far, i.e., $\chi \subseteq \{a_1,\dots,a_m\}$ after loop iteration $m$.
+
+In the loop iteration where element $s$ occurs for the first time, it will be inserted with probability $p$ in Lines 3--7.
+This means, after Line 7, we have:
+\begin{equation}
+  \label{eq:indicator_eq}
+  \expect [p^{-1} X_s] = 1 \textrm{.}
+\end{equation}
+Interestingly, this equation is preserved for the rest of the algorithm.
+For example, let us consider a subsampling step: each $s$ is independently discarded with probability $\frac{1}{2}$ so $\prob(X_s=1)$ is halved, but so is $p$ after subsampling, which preserves the equation.
+
+Let us see how we can verify Equation~\ref{eq:indicator_eq} more formally.
+For that, we model the state of the randomized algorithm as a pair $(\chi,p)$ and we write $\chi$ and $p$ for the random variables projecting their respective components from the distribution of the state of the algorithm.
+We will refer to parts of each loop iteration in \cref{alg:cvm} as $\mathrm{step}_1$ (resp.~$\mathrm{step}_2$) for Lines 3--7 (resp.~Lines 8--10).
+The final distribution of the algorithm is the distribution resulting from the sequential composition of alternating steps over the stream:
+\[
+  \mathrm{init} \, \isa{\isasymbind}\; \mathrm{step}_1\, a_1 \; \isa{\isasymbind}\; \mathrm{step}_2 \; \isa{\isasymbind}\; \mathrm{step}_1 \, a_2 \; \isa{\isasymbind}\; \cdots \isa{\isasymbind}\; \mathrm{step}_1\, a_l \; \isa{\isasymbind}\; \mathrm{step}_2
+\]
+where we parameterize $\mathrm{step}_1$ with the stream element that it processes.
+The term $\mathrm{init}$ represents the initial state, i.e., $\mathrm{init} = \mathrm{\bf return}\,(\{\},1)$.
+It is easy to show by induction over the sequence of steps, we have $0 < p \leq 1$ and $\chi \subseteq A$ for all possible states of the algorithm.
+
+Let us verify that Equation~\ref{eq:indicator_eq} is preserved as an invariant over all steps.
+To verify that $\mathrm{step}_1\, a$ preserves Equation~\ref{eq:indicator_eq}, we assume some probability space of states $\Omega$ fulfills Equation~\ref{eq:indicator_eq} and we would like to show that it is still true for $\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a$.
+By Equation~\ref{eq:integral_bind},
+\begin{equation}\label{eq:step_1_exp}
+  \expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step}_1\, a} [ p^{-1} X_s ] =
+    \integral{\Omega}{\integral{\Ber(p)}{p^{-1} \indicat\left(s \in (\ift{\tau}{\chi \cup \{a\}}{\chi-\{a\}} )\right)}{\tau}}{\sigma} \textrm{.}
+\end{equation}
+Note that we write $p$ or $\chi$ even though we should actually write $p(\sigma)$ or $\chi(\sigma)$, i.e., we remember that these implicitly depend on $\sigma$.
+To see that the right-hand side is equal to $1$, it is useful to consider cases on whether $a=s$.
+When $a = s$, the right-hand-side is equal to $1$ by definition of the Bernoulli distribution (since $p \in (0;1]$).
+When $a \not= s$, it follows from the induction hypothesis on $\Omega$; in particular, the term in the inner integral is constant with respect to $\tau$.
+
+The same invariant-based argument is possible for $\mathrm{step}_2$.
+Let us assume $\Omega$ is a probability space of states fulfilling Equation~\ref{eq:indicator_eq}.
+Then by Equation~\ref{eq:integral_bind}, $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1}X_s]$ equals
+\begin{equation}\label{eq:step_2_exp}
+    \integral{\Omega}
+      {\left(\ift{|\chi|=n}{\left(\integral{\mathrm{subsample}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}\right)}{\frac{\indicat(s \in \chi)}{p}} \right)}
+      {\sigma}
+    \textrm{.}
+\end{equation}
+Note that the \emph{true} and \emph{false} cases of the inner \textbf{if-then-else} both evaluate to the same value: $p^{-1} \indicat(s \in \chi)$.
+If $s \notin \chi$ both sides of the equation are $0$, because the subsampling operation returns a subset of $\chi$.
+If $s \in \chi$ the probability that the element gets subsampled is $1/2$, so we arrive again at $\frac{1/2}{p/2} = p^{-1} \indicat(s \in \chi)$.
+Hence: $\expect_{\Omega \; \isa{\isasymbind}\; \mathrm{step_2}} [p^{-1} X_s] = \expect_\Omega [p^{-1} X_s] = 1$.
+This completes the invariance proof for Equation~\ref{eq:indicator_eq}.
+
+\subsection{Deriving a Functional Probabilistic Invariant}
+With Equation~\ref{eq:indicator_eq} established, it is straightforward to show that the expected value of the output estimate $p^{-1} |\chi|$ for the modified algorithm (without Line 11) is equal to the desired cardinality $|A|$.
+However, recall that we are interested in verifying the estimate's PAC guarantee.
+A typical approach to establishing such a guarantee is to use Chernoff bounds which provide exponential tail bounds (i.e., concentration bounds) for the deviation of sums of independent random variables from their mean.
+However, these are not directly useful in the CVM algorithm because the key random variables, e.g., $p^{-1} X_s$ for $s \in A$, are dependent.
+
+An alternative is the Cram\'{e}r--Chernoff method, which is a general method to obtain tail bounds for any random variable.
+It can be stated simply as $\prob(X \geq a) \leq M(t) e^{-ta}$ for all $t > 0$, where $M(t) := \expect [\exp(t X)]$ is the moment generating function of the random variable $X$.
+It is also possible to obtain lower tail bounds $\prob(X \leq a)$ using the Cram\'{e}r--Chernoff method, which just requires estimates for $M(t)$ for $t < 0$, instead of $t > 0$.
+
+In our case, we are interested in estimating the moment generating function of the random variable $p^{-1} |\chi|$ for the CVM algorithm:
+\[
+  \expect [\exp( t p^{-1} |\chi| )] = \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right]
+\]
+for $h(x) = \exp(tx)$.
+At this point, it is tempting to see whether the proof for Equation~\ref{eq:indicator_eq} can be generalized to establish bounds for the above.
+Indeed, we managed to establish the following generalized result:
+\begin{align}
+  \expect \left[ \prod_{s \in A} h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant}
+\end{align}
+for every non-negative concave function $h : \Rnonneg \rightarrow \Rnonneg$.
+However, the exponential function in $M(t)$ is convex, but we can instead try to derive tail bounds for the random variable $\indicat(p \geq q) p^{-1} |\chi|$, for some fixed constant $q > 0$.
+This leads to a similar invariant inequality:
+\begin{align}
+  \expect \left[ \prod_{s \in A} I(p \geq q) h(p^{-1} X_s) \right] \leq h(1)^{|A|} \label[ineq]{i:func_invariant_capped}
+\end{align}
+with the new condition that $h$ needs to be non-negative and concave only on $[0;q^{-1}]$.
+This then allows us to approximate the exponential function from above with an affine function $h$ on the range $[0;q^{-1}]$, which yields tail bounds for $p^{-1} |\chi|$ under the condition $p \geq q$.
+As an example, the upper tail bound can be derived as follows:
+\begin{eqnarray*}
+  \prob( p^{-1} |\chi| \geq (1+\varepsilon) |A| \wedge p \geq q) & \leq & % Is not equal when |A|=0.
+  \prob( \indicat(p \geq q) p^{-1} |\chi| \geq (1+\varepsilon) |A|) \\
+  & \underset{\textrm{Markov}}{\leq} & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) \exp( t p^{-1} X_s) \right] \\
+  & \leq & e^{-t(1+\varepsilon) |A|} \expect \left[ \prod_{s \in A} \indicat(p \geq q) h(p^{-1} X_s) \right] \\
+  & \underset{\textrm{Ineq.~\ref{i:func_invariant_capped}}}{\leq} & e^{-t(1+\varepsilon) |A|} h(1)^{|A|} \\
+  & \underset{\textrm{Calculus}}{\leq} & e^{-n\varepsilon^2 / 12}
+\end{eqnarray*}
+where we choose $h(x) = 1+qx (e^{t/q}-1)$.
+Note that $h$ is affine and it can be easily checked%
+\footnote{Because the exponential function is convex and $h$ is affine, we only have to check the end points: $0, q^{-1}$.}
+that it is an upper approximation of $\exp(tx)$ for $x \in [0;q^{-1}]$.
+For the last step, we have to find the $t$ that produces the required bound.%
+\footnote{We use $t = q \ln(1+\varepsilon)$ which is not the real optimum, but better for algebraic evaluation.}
+To use these bounds, we also have to separately estimate $\prob(p < q)$.
+For that, we use a similar strategy, as in the original proof by Chakraborty et al.~\cite{chakraborty2023}, with $q = \frac{n}{4 |A|}$.
+The formalization in the supplementary material~\cite{CVM_Distinct_Elements-AFP} contains a detailed informal step-by-step proof using our approach in its appendix.
+Besides the use of Equation~\ref{eq:integral_bind} and the Cram\'er--Chernoff method, the steps are elementary.
+
+We call inequalities like Inequality~\ref{i:func_invariant} and \ref{i:func_invariant_capped}: \emph{functional probabilistic invariants}.
+The inequalities can be established using induction over the steps of the randomized algorithm.
+Even, though, the actual distribution of the states themselves are not known, nor do we think it is possible to find useful closed form descriptions for them, the invariant establishes valuable information about the distribution of the state.
+In this case, enough information, to establish exponential tail bounds for the algorithm.
+
+\section{An Unbiased CVM Variant and Negative Dependence\label{sec:negdep}}
+An interesting consequence of our invariant-based approach is that it allowed us to devise and verify a refined version of the CVM algorithm that is both total and unbiased.
+
+\subsection{Unbiased CVM Variant}
+\begin{algorithm}[t!]
+	\caption{New total and unbiased CVM algorithm variant.}\label{alg:cvm_new}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{3l}{\delta})} }$, $\frac{1}{2} \leq f < 1$, such that $nf$ integer
+  \For{$i \gets 1$ to $l$}
+    \State $b \getsr \Ber(p)$ \Comment insert $a_i$ with probability $p$ (and remove it otherwise)
+    \If{$b$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+      \State $\chi \getsr \mathrm{subsample}(\chi)$ \Comment select a random $nf$-subset of $\chi$
+      \State $p \gets pf$
+    \EndIf
+  \EndFor
+  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}%
+
+When we look at the subsampling step of~\cref{alg:cvm}, our invariant (Inequality~\ref{i:func_invariant}) imposes the following condition on the subsampling operation.
+It should be noted that the condition becomes apparent while establishing Inequality~\ref{i:func_invariant} using similar but more general steps as described in Equations~\ref{eq:step_1_exp} and \ref{eq:step_2_exp}.%
+\footnote{A step-by-step proof of the derivation is also available in the appendix of the formalization~\cite[Appendix~A]{CVM_Distinct_Elements-AFP}.}
+\begin{equation}\label[ineq]{i:subsample_condition}
+  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))}{\tau} \leq \prod_{s \in S} \expect_{\Ber(f)} [g]
+\end{equation}
+for all non-negative functions $g$ and any $S \subseteq \chi$, where $f$ is the probability of retaining each element in the subsampling step of the algorithm.
+(This parameter $f$ was fixed to $\frac{1}{2}$ in the original presentation of the algorithm for simplicity.)
+Any subsampling step that satisfies Inequality~\ref{i:subsample_condition} can be used while still preserving Inequalities~\ref{i:func_invariant} and \ref{i:func_invariant_capped} for the algorithm.
+
+Motivated by this observation, our new variant is shown in \cref{alg:cvm_new}.
+For the subsampling step, instead of keeping each element of $\chi$ with probability $\frac{1}{2}$, we pick a uniform random $nf$-subset of $\chi$, where $\frac{1}{2} \leq f < 1$ and $nf$ is an integer.
+For example, it is possible to choose $f = \frac{n-1}{n}$, i.e., discarding just one random element from $\chi$ in the subsampling step.
+Since this new subsampling step always reduces the size of $\chi$, the variant is \emph{total} (never returns $\bot$).
+The invariant-based approach allows us to show that the algorithm is probably-approximately correct and also \emph{unbiased}, i.e., the expectation of the result is exactly $|A|$.
+These depend crucially on establishing Inequality~\ref{i:subsample_condition} for the new subsampler, for which we need a new concept.
+
+\subsection{Background on Negative Dependence}
+Some sets of random variables possess a property called \emph{negative association}, a generalization of independence.
+The concept was introduced by Joag-Dev and Proschan~\cite{joagdev1983}, who showed that it has many useful closure properties compared to other previously introduced notions of negative dependence, such as negative correlation or negative orthant dependence.
+%TODO: perhaps citations for these two?
+Importantly, standard Chernoff--Hoeffding type bounds still apply to negatively associated random variables~\cite[Prop. 7]{dubhashi1998}.
+Negative association is defined as follows:
+\begin{definition}
+For a function defined on $n$-tuples $f: V^n \rightarrow W$, we will denote by $\mathrm{dep}(f)$ the set of coordinates the function depends on, i.e., $dep(f) \subseteq \{1,\ldots,n\}$ is minimal, such that $f(x) = f(y)$ for all $x, y \in V^n$ with $x_i = y_i$ for all $i \in dep(f)$.
+\end{definition}
+
+\begin{definition}[Negative Association]\label{def:neg_assoc}
+A set of random variables $X_1,\dots,X_n: \Omega \rightarrow \mathbb R$ is negatively associated if, for all non-decreasing functions $f,g: \mathbb R^n \rightarrow \mathbb R$, which depend on disjoint sets of the variables, i.e., $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$, the following inequality holds:
+\[
+\expect [f(X_1,\ldots,X_n) g(X_1,\ldots,X_n)] \leq \expect [f(X_1,\ldots,X_n)] \expect [g(X_1,\ldots,X_n)] \textrm{.}
+\]
+\end{definition}
+
+The following proposition summarizes some important properties of negatively associated sets of random variables.
+\begin{proposition}[Summary of results for negatively associated random variables \cite{joagdev1983}]\label{pro:neg_dep_props}~
+\begin{enumerate}
+\item \label{it:neg_dep_props:mult_mono} If $X=(X_1,\ldots,X_n)$ are negatively associated then $\expect [f(X) g(X)] \leq \expect [f(X)] \expect [g(X)]$ for non-increasing functions $f,g$ with $\mathrm{dep}(f) \cap \mathrm{dep}(g) = \emptyset$.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated, $Y=(Y_1,\ldots,Y_m)$ are negatively associated, and the pair of vector-valued random variables $X$ and $Y$ are independent, then the union $X_1,\dots,X_n,Y_1,\dots,Y_m$ is a set of negatively associated random variables.
+\item If $X=(X_1,\ldots,X_n)$ are negatively associated and $f_1, \dots ,f_m : \mathbb R^n \rightarrow \mathbb R$ are all non-increasing or all non-decreasing functions, such that $\mathrm{dep}(f_i) \cap \mathrm{dep}(f_j) = \emptyset$ for $i \neq j$, then $f_1(X),\ldots,f_m(X)$ form a set of negatively dependent random variables of size $m$.
+\item If $X_1,\dots,X_n$ are independent then $X_1,\dots,X_n$ are negatively associated.
+\item A subset of a negatively associated set of random variables is again negatively associated.
+\end{enumerate}
+\end{proposition}
+
+These properties illustrate the trade-off between negative association and independence.
+For example, Property 3 would be true for independent random variables, even without the condition of monotonicity.
+To analyze our new subsampler, the following is an important lemma about negatively associated random variables.
+
+\begin{lemma}\label{le:neg_assoc_prod}
+Let $X_1,\dots,X_n$ be negatively associated and $f_1,\dots,f_n$ be all non-decreasing (or all non-increasing), non-negative functions, then
+\[
+  \expect \left[\prod_{i=1}^{n} f_i(X_i)\right] \leq \prod_{i=1}^{n} f_i(\expect [X_i]) \textrm{.}
+\]
+\end{lemma}
+\begin{proof}
+This follows from the definition of negative association (or Property~1 of \cref{pro:neg_dep_props}, if the $f_i$ are non-increasing) using induction.
+\end{proof}
+
+The case for non-decreasing functions of the above lemma is pointed out by Joag-Dev and Proschan~\cite[P.2]{joagdev1983}.
+The reason for our interest in this lemma stems from the fact that indicator variables of random $m$-subsets are negatively associated.
+This is a consequence of the fact that permutation distributions are negatively associated~\cite[Th. 2.11]{joagdev1983}.
+Thus, for the new subsampling step in Line 9 of~\cref{alg:cvm_new}, we can derive using \cref{le:neg_assoc_prod}:
+\begin{equation}\label{eq:subsample_with_n_subsets}
+  \integral{\mathrm{subsample}(\chi)}{\prod_{s \in S} g(\indicat(s \in \tau))} {\tau} \leq
+  \prod_{s \in S} \integral{\mathrm{subsample}(\chi)}{g(\indicat(s \in \tau))}{\tau} = \prod_{s \in S}  \expect_{\Ber(f)}[g] \textrm{.}
+\end{equation}
+for any non-negative $g$ and $S \subseteq \chi$.
+Note that the domain of $g$ has two values, so it is either non-increasing or non-decreasing.
+Also, if $S$ is a singleton, the inequality becomes an equality.
+With this ingredient, we can conclude that our results about the original algorithm derived in the previous section also hold for our new variant (\cref{alg:cvm_new}).
+
+\section{Formalization of the CVM algorithm}\label{sec:formalization}
+Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach~\cite{CVM_Distinct_Elements-AFP}.
+We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.
+
+\begin{note}
+In our supplementary material~\cite{CVM_Distinct_Elements-AFP}, the theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill Inequality~\ref{i:subsample_condition}.
+The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.
+Note that only \verb|CVM_New_Unbiased_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
+The total number of lines required for the verification of the original algorithm is \locnew~lines.
+In addition, we actually verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
+\lipicsEnd\end{note}
+
+\begin{figure}[t]
+\begin{isabelle_cm}
+\isacommand{context}\isamarkupfalse%
+\isanewline
+\ \ \isakeyword{fixes}\ f\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ real\ \isakeyword{and}\ n\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ nat\isanewline
+\ \ \isakeyword{assumes}\ f{\isacharunderscore}{\kern0pt}range{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}f\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{1}}{\isacharslash}{\kern0pt}{\isadigit{2}}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}\ \isakeyword{and}\ n{\isacharunderscore}{\kern0pt}gt{\isacharunderscore}{\kern0pt}{\isadigit{0}}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}n\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+\isakeyword{begin}\isanewline
+\isanewline
+\isacommand{definition}\isamarkupfalse%
+\ {\isacartoucheopen}initial{\isacharunderscore}{\kern0pt}state\ {\isacharequal}{\kern0pt}\ State\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}\ {\isadigit{1}}{\isacartoucheclose}\ %
+\hfill\isamarkupcmt{Setup initial state $\chi=\emptyset$ and $p=1$.\;%
+}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ subsample\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Subsampling operation: Sample random $n f$ subset.\;%
+}\isanewline
+\ \ {\isacartoucheopen}subsample\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ pmf{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}set\ {\isacharbraceleft}{\kern0pt}S{\isachardot}{\kern0pt}\ S\ {\isasymsubseteq}\ {\isasymchi}\ {\isasymand}\ card\ S\ {\isacharequal}{\kern0pt}\ n\ {\isacharasterisk}{\kern0pt}\ f{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ step\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Loop body.\;%
+}\isanewline
+\ \ {\isacartoucheopen}step\ a\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ b\ {\isasymleftarrow}\ bernoulli{\isacharunderscore}{\kern0pt}pmf\ p{\isacharsemicolon}{\kern0pt}\isanewline
+\ \ \ \ let\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ {\isacharparenleft}{\kern0pt}if\ b\ then\ {\isasymchi}\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}\ else\ {\isasymchi}\ {\isacharminus}{\kern0pt}\ {\isacharbraceleft}{\kern0pt}a{\isacharbraceright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacharsemicolon}{\kern0pt}\isanewline
+\isanewline
+\ \ \ \ if\ card\ {\isasymchi}\ {\isacharequal}{\kern0pt}\ n\ then\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ \ \ {\isasymchi}\ {\isasymleftarrow}\ subsample\ {\isasymchi}{\isacharsemicolon}{\kern0pt}\isanewline
+\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ {\isacharparenleft}{\kern0pt}p\ {\isacharasterisk}{\kern0pt}\ f{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\isanewline
+\ \ \ \ {\isacharbraceright}{\kern0pt}\ else\ do\ {\isacharbraceleft}{\kern0pt}\isanewline
+\ \ \ \ \ \ return{\isacharunderscore}{\kern0pt}pmf\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\isanewline
+\ \ \ \ {\isacharbraceright}{\kern0pt}\isanewline
+\ \ \ {\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ run{\isacharunderscore}{\kern0pt}steps\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Iterate loop over stream \isa{xs}.\;%
+}\isanewline
+\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}steps\ xs\ {\isacharequal}{\kern0pt}\ foldM{\isacharunderscore}{\kern0pt}pmf\ step\ xs\ initial{\isacharunderscore}{\kern0pt}state{\isacartoucheclose}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ estimate\ \isakeyword{where}\
+{\isacartoucheopen}estimate\ {\isacharparenleft}{\kern0pt}State\ {\isasymchi}\ p{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ card\ {\isasymchi}\ {\isacharslash}{\kern0pt}\ p{\isacartoucheclose}\isanewline
+\isacommand{fun}\isamarkupfalse%
+\ run{\isacharunderscore}{\kern0pt}algo\ \isakeyword{where}\ %
+\hfill\isamarkupcmt{Run algorithm and estimate.\;%
+}\isanewline
+\ \ {\isacartoucheopen}run{\isacharunderscore}{\kern0pt}algo\ xs\ {\isacharequal}{\kern0pt}\ map{\isacharunderscore}{\kern0pt}pmf\ estimate\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}steps\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+{\normalfont [\dots]}\isanewline
+\isacommand{end}
+\end{isabelle_cm}
+\caption{Formalized version of \cref{alg:cvm_new}.}\label{alg:cvm_formalized}
+\end{figure}
+
+A snippet of the formalization of \cref{alg:cvm_new} is presented in \cref{alg:cvm_formalized} (the formalization of \cref{alg:cvm} is very similar).
+We use the same variables as in the informal presentation: $n$ for the maximal size of the buffer, $f$ for the fraction of elements to keep in the buffer when subsampling.
+The condition \isa{{\isacartoucheopen}n\ {\isacharasterisk}{\kern0pt}\ f\ {\isasymin}\ {\isasymnat}{\isacartoucheclose}} expresses the requirement that the $nf$ must be integer.
+Instead of representing the state using pairs, as we did in the informal discussion, we use a datatype with the single constructor \isa{State}, which has two arguments $\chi$ and $p$, the buffer and the probability that the stream elements are in the buffer, respectively.
+Isabelle/HOL provides notation closely related to informal pseudocode, so it is usually feasible to read a formal statement without expert knowledge.
+Nevertheless, \cref{tab:isabelle_syntax} contains a brief glossary of the syntax used in the formalization.
+\begin{table}
+\caption{Isabelle/HOL syntax used in \cref{alg:cvm_formalized}.}\label{tab:isabelle_syntax}
+\noindent\begin{tabular}{l p{9cm}}
+\toprule
+Term & Description \\
+\midrule
+\isa{card\ S} & Cardinality of a finite set $S$. \\
+\isa{real} & Type of real numbers and conversion from natural numbers into real numbers. \\
+\isa{nat} & Type of natural numbers (non-negative integers). \\
+\isa{bernoulli{\isacharunderscore}pmf\ p} & The probability space over the Boolean values, where the probability of \isa{True} is $p$. (Bernoulli distribution.) \\
+\isa{pmf{\isacharunderscore}of{\isacharunderscore}set\ S} & For a finite set $S$, the uniform probability space on $S$. (Every element of $S$ is equiprobable.) \\
+\isa{map{\isacharunderscore}pmf\ f\ A} & The probability space representing the distribution of the random variable $f$ over the probability space $A$. \\
+\isa{return{\isacharunderscore}pmf\ x} & The probability space of the singleton $\{x\}$. \\
+\isa{foldM{\isacharunderscore}pmf\ f\ xs\ a} & Iterate randomized algorithm $f$ over the sequence $xs$ using the initial state $a$. \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+The theorem that establishes the correctness of the algorithm, i.e., that the relative error will exceed $\varepsilon$ with probability at most $\delta$ is expressed in the following snippet:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ correctness{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymepsilon}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}{\isadigit{0}}{\isacharless}{\kern0pt}{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}{\isacharless}{\kern0pt}{\isadigit{1}}{\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}real{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}real\ n\ {\isasymge}\ {\isadigit{1}}{\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isasymepsilon}\isactrlsup {\isadigit{2}}\ {\isacharasterisk}{\kern0pt}\ ln\ {\isacharparenleft}{\kern0pt}{\isadigit{3}}\ {\isacharasterisk}{\kern0pt}\ real\ {\isacharparenleft}{\kern0pt}length\ xs{\isacharparenright}{\kern0pt}\ {\isacharslash}{\kern0pt}\ {\isasymdelta}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}A\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}R\ in\ run{\isacharunderscore}{\kern0pt}algo\ xs{\isachardot}{\kern0pt}\ {\isasymbar}R\ {\isacharminus}{\kern0pt}\ A{\isasymbar}\ {\isachargreater}{\kern0pt}\ {\isasymepsilon}\ {\isacharasterisk}{\kern0pt}\ A{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isasymdelta}{\isacartoucheclose}
+\end{isabelle_cm}
+The first line gives conditions on parameters $\varepsilon$ and $\delta$, which must be strictly between $0$ and $1$.
+The next line requires the buffer size $n$ to be larger than or equal to $12 \varepsilon^{-2} \ln ( 3 \delta^{-1} l)$.
+Then, we introduce the abbreviation \isa{A} for the cardinality of the set of elements in the sequence \isa{xs}.
+The notation \isaprob{x}{M}{P x} denotes the probability of a predicate $P$ in the probability space $M$, so the conclusion gives the PAC guarantee for the output estimate \isa{R} from \isa{run{\isacharunderscore}{\kern0pt}algo}.
+
+Similarly, we have also formalized unbiasedness of \cref{alg:cvm_new}:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ unbiasedness{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}expectation\ {\isacharparenleft}{\kern0pt}run{\isacharunderscore}{\kern0pt}algo\ xs{\isacharparenright}{\kern0pt}\ id\ {\isacharequal}{\kern0pt}\ card\ {\isacharparenleft}{\kern0pt}set\ xs{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+where the expression \isa{measure{\isacharunderscore}pmf{\isachardot}expectation\ M\ f} denotes the expectation of the random variable \isa{f} on the probability space \isa{M}.
+
+Our proofs are available both in mechanized form in Isabelle/HOL and as a pen-and-paper proof included in the associated proof document.
+In practice, we developed the latter proof first and then mechanized it in Isabelle/HOL without much surprise.
+Most of the lemmas can be identified one-to-one between both proofs; Isabelle's existing libraries, automation capabilities, and structured proof format were used extensively in our proofs.
+
+\section{Formalization of a Library for Negative Association}\label{sec:formalization_neg_dep}
+As mentioned in~\cref{sec:negdep}, formalizing the total and unbiased variant of the CVM algorithm requires results from the theory of negative association.
+
+\begin{note}
+The formalization of the theory of negative association is included in a separate AFP entry~\cite{Negative_Association-AFP}.
+This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distributions|).
+Although not needed for CVM, we have also mechanized the standard Chernoff bounds (\verb|Neg_Assoc_Chernoff_Bounds|), including the additive bounds by Hoeffding~\cite[Th.
+1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th.
+4.1, 4.2]{motwani1995}.
+Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
+In total, the library contains 2974~lines of Isabelle code.
+\lipicsEnd\end{note}
+
+Our formalization follows the definitions by Joag-Dev and Proschan~\cite{joagdev1983} closely.
+However, their definition leaves the class of test functions $f$ and $g$ (in \cref{def:neg_assoc}) imprecise.
+In the formalization, we use different conditions for introduction and elimination rules.
+In particular, for introduction rules, the test functions are bounded and measurable.
+However, we provide stronger elimination rules, showing that if $X_1,\dots,X_n$ are negatively associated, then the inequality on expectations is true even if $f, g$ are only square-integrable; or, alternatively, integrable and non-negative.
+This is derived using the monotone convergence theorem.
+
+Another deviation from the original work is that we do not require that the random variables are real-valued.
+In the formalization, any linearly ordered topological space with the Borel $\sigma$-algebra is allowed as the range space.
+In this case, the test functions must be monotone with respect to the respective order on the range space.
+
+A key issue we faced during formalization was that there are many theorems that condition on a set of functions being either simultaneously monotone or simultaneously anti-monotone.
+To reduce duplication, we introduce a notation that allows us to abstract over the direction of relations: \isa{\isasymle\isasymge\isactrlbsub\isasymeta\isactrlesub}; it evaluates to the forward version of the relation $\leq$ if \isa{\isasymeta\ \isacharequal\ Fwd} and the converse: $\geq$ if \isa{\isasymeta\ \isacharequal\ Rev}.
+For example the FKG inequality~\cite[Ch. 6]{alon2000},\cite{fortuin1971}
+\[
+  \expect [f g] \geq \expect [f] \expect [g]
+\]
+is true, if $f$ and $g$ are both monotone, or both antimonotone, on a probability space whose domain is a finite distributive lattice with a log-supermodular probability mass function.
+The reverse inequality is also true, if $f$ is monotone and $g$ is antimonotone, or vice versa.
+Using our parameterized relation symbol, we can state all variants in a concise manner:
+\begin{isabelle_cm}
+\isacommand{theorem}\isamarkupfalse%
+\ fkg{\isacharunderscore}{\kern0pt}inequality{\isacharunderscore}{\kern0pt}pmf{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{fixes}\ M\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isacharprime}{\kern0pt}a\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ finite{\isacharunderscore}{\kern0pt}distrib{\isacharunderscore}{\kern0pt}lattice{\isacharparenright}{\kern0pt}\ pmf{\isacartoucheclose}\isanewline
+\ \ \isakeyword{fixes}\ f\ g\ {\isacharcolon}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isacartoucheopen}{\isacharprime}{\kern0pt}a\ {\isasymRightarrow}\ real{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}x\ y{\isachardot}{\kern0pt}\ pmf\ M\ x\ {\isacharasterisk}{\kern0pt}\ pmf\ M\ y\ {\isasymle}\ pmf\ M\ {\isacharparenleft}{\kern0pt}x\ {\isasymsqunion}\ y{\isacharparenright}{\kern0pt}\ {\isacharasterisk}{\kern0pt}\ pmf\ M\ {\isacharparenleft}{\kern0pt}x\ {\isasymsqinter}\ y{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymtau}\isactrlesub {\isacharparenright}{\kern0pt}\ f{\isacartoucheclose}\ {\isacartoucheopen}monotone\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isacharparenright}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymle}{\isasymge}\isactrlbsub {\isasymsigma}\isactrlesub {\isacharparenright}{\kern0pt}\ g{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ f\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}\ {\isacharasterisk}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ g\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}\ {\isasymle}{\isasymge}\isactrlbsub {\isasymtau}\ {\isacharasterisk}{\kern0pt}\ {\isasymsigma}\isactrlesub \ {\isacharparenleft}{\kern0pt}{\isasymintegral}x{\isachardot}{\kern0pt}\ f\ x\ {\isacharasterisk}{\kern0pt}\ g\ x\ {\isasympartial}M{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+Here, \isa{\isasymsigma} and \isa{\isasymtau} are relation directions, and \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} multiplies relation directions, i.e., \isa{\isasymsigma\ \isacharasterisk\ \isasymtau} is the forward direction if \isa{\isasymsigma} and \isa{\isasymtau} have the same direction, and it is the reverse direction otherwise.
+The first assumption denotes the log-supermodularity of the probability mass function, while the second assumptions are the parametric monotonicity conditions.
+The FKG inequality is a key result which enables verification of negative association for random variables.
+This includes the indicator variables for the new subsampling operation we introduced in \cref{sec:negdep}.
+
+Let us summarize a few key formalized results for negatively associated random variables in our library.
+The following is the well-known Hoeffding inequality~\cite{hoeffding1963} generalized for negatively associated random variables.
+\begin{isabelle_cm}
+\isacommand{lemma}\isamarkupfalse%
+\ hoeffding{\isacharunderscore}{\kern0pt}bound{\isacharunderscore}{\kern0pt}two{\isacharunderscore}{\kern0pt}sided{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}neg{\isacharunderscore}{\kern0pt}assoc\ X\ I{\isacartoucheclose}\ {\isacartoucheopen}finite\ I{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ a\ i\ {\isasymle}\ b\ i{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymAnd}i{\isachardot}{\kern0pt}\ i{\isasymin}I\ {\isasymLongrightarrow}\ AE\ {\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}\ {\isasymin}\ {\isacharbraceleft}{\kern0pt}a\ i{\isachardot}{\kern0pt}{\isachardot}{\kern0pt}b\ i{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\ {\isacartoucheopen}I\ {\isasymnoteq}\ {\isacharbraceleft}{\kern0pt}{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}n\ {\isasymequiv}\ real\ {\isacharparenleft}{\kern0pt}card\ I{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}{\isasymmu}\ {\isasymequiv}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ expectation\ {\isacharparenleft}{\kern0pt}X\ i{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}{\isasymdelta}\ {\isasymge}\ {\isadigit{0}}{\isacartoucheclose}\ {\isacartoucheopen}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}\ {\isachargreater}{\kern0pt}\ {\isadigit{0}}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}{\isasymP}{\isacharparenleft}{\kern0pt}{\isasymomega}\ in\ M{\isachardot}{\kern0pt}\ {\isasymbar}{\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ X\ i\ {\isasymomega}{\isacharparenright}{\kern0pt}{\isacharminus}{\kern0pt}{\isasymmu}{\isasymbar}\ {\isasymge}\ {\isasymdelta}{\isacharasterisk}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isasymle}\ {\isadigit{2}}{\isacharasterisk}{\kern0pt}exp\ {\isacharparenleft}{\kern0pt}{\isacharminus}{\kern0pt}{\isadigit{2}}{\isacharasterisk}{\kern0pt}{\isacharparenleft}{\kern0pt}n{\isacharasterisk}{\kern0pt}{\isasymdelta}{\isacharparenright}{\kern0pt}\isactrlsup {\isadigit{2}}\ {\isacharslash}{\kern0pt}\ {\isacharparenleft}{\kern0pt}{\isasymSum}i{\isasymin}I{\isachardot}{\kern0pt}\ {\isacharparenleft}{\kern0pt}b\ i{\isacharminus}{\kern0pt}a\ i{\isacharparenright}\isactrlsup {\isadigit{2}}{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}{\isacartoucheclose}
+\end{isabelle_cm}
+
+Another key result (shown below) used for the verification of our CVM variant is the negative-associativity of the indicator functions of random $k$-subsets of a finite set $S$ (with cardinality greater than or equal to $k$).
+\begin{isabelle_cm}
+\isacommand{lemma}\isamarkupfalse%
+\ n{\isacharunderscore}{\kern0pt}subsets{\isacharunderscore}{\kern0pt}distribution{\isacharunderscore}{\kern0pt}neg{\isacharunderscore}{\kern0pt}assoc{\isacharcolon}{\kern0pt}\isanewline
+\ \ \isakeyword{assumes}\ {\isacartoucheopen}finite\ S{\isacartoucheclose}\ {\isacartoucheopen}k\ {\isasymle}\ card\ S{\isacartoucheclose}\isanewline
+\ \ \isakeyword{defines}\ {\isacartoucheopen}p\ {\isasymequiv}\ pmf{\isacharunderscore}{\kern0pt}of{\isacharunderscore}{\kern0pt}set\ {\isacharbraceleft}{\kern0pt}T{\isachardot}{\kern0pt}\ T\ {\isasymsubseteq}\ S\ {\isasymand}\ card\ T\ {\isacharequal}{\kern0pt}\ k{\isacharbraceright}{\kern0pt}{\isacartoucheclose}\isanewline
+\ \ \isakeyword{shows}\ {\isacartoucheopen}measure{\isacharunderscore}{\kern0pt}pmf{\isachardot}{\kern0pt}neg{\isacharunderscore}{\kern0pt}assoc\ p\ {\isacharparenleft}{\kern0pt}{\isasymin}{\isacharparenright}{\kern0pt}\ S{\isacartoucheclose}
+\end{isabelle_cm}
+This is a consequence of a more general result, which we have also shown, that permutation distributions are negatively associated.
+We relied on the proof by Dubhashi et al.\ using the FKG inequality~\cite[Th.
+10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th.
+2.11]{joagdev1983}, which is incomplete.\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$.
+That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality.
+We think Dubhashi et al.\ developed their proof to complete it.}
+
+\section{Transformation-Based Proof}\label{sec:transformation_based_proof}
+Here, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}, focusing on the challenging parts in its formalization.
+An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
+\begin{note}
+The transformation-based formalization of the CVM algorthm is included in the supplementary material~\cite{CVM_Transforms-Github}.
+To formalize probabilistic transformations (relating two distributions), we built on an existing relational program logic in Isabelle/HOL~\cite{lochbihler2016}.
+The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
+\lipicsEnd\end{note}
+
+As mentioned in~\cref{sec:intro}, the main difficulty in directly analyzing \cref{alg:cvm} is the lack of independence in its state variables.
+The technique Chakraborty et al.\ use to circumvent this issue is by progressively modifying the algorithm (\circled{A}--\circled{D} in~\cref{fig:proof_overview}), in a manner that obviously bounds (or preserves) its distribution, and such that the final algorithm \circled{D} can be described using a simple random process with independent coin flips.
+
+For interested readers, \circled{A} corresponds to~\cite[Algorithm 1]{chakraborty2023}, \circled{B} is~\cite[Algorithm 2]{chakraborty2023}, and \circled{D} is~\cite[Algorithm 3]{chakraborty2023}.
+Whereas Chakraborty et al.\ move directly from \circled{B} to \circled{D} with an informal argument, \circled{C} is a transformation we added in the formalization to bridge this gap.
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[node distance=2.7cm, auto]
+
+    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center] {\circled{A} CVM \\ (Alg.~\ref{alg:cvm})};
+    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=A] {\circled{B} CVM w/o\\ Line 11};
+    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=B] {\circled{C} Eager Ver.\\(Alg.~\ref{alg:cvm_simul})};
+    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=C] {\circled{D} Random \\ Process};
+    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=D] {Chernoff \\ Bounds};
+
+    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=B, node distance = 2.0cm] {Functional\\Invariant};
+
+    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=E, node distance = 2.0cm] {CVM PAC\\ Guarantee};
+
+    \draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(E.south east) + (0.3,-0.6)$);
+    \node[anchor=south] at ($(D.south) + (0.0,-0.6)$) {Transformation-based Proof};
+
+    \draw[dashed,thick] ($(A.north west) + (-0.5,0.4)$) rectangle ($(C'.south east) + (0.2,-0.2)$);
+    \node[anchor=south,align=center] at ($(C'.west) + (-1.5,-0.45)$) {Probabilistic\\Invariant Proof};
+
+    \draw[->] (A) -- (B);
+    \draw[->] (B) -- (C);
+    \draw[->] (B) -- (C');
+    \draw[->] (C) -- (D);
+    \draw[->] (D) -- (E);
+    \draw[->] (E) -- (pac);
+    \draw[->] (C') -- (pac);
+
+\end{tikzpicture}
+\end{center}
+\caption{An overview of the two formalized proof approaches for the CVM algorithm.}\label{fig:proof_overview}
+\end{figure}
+
+\subsection{A Bridging Transformation}
+\begin{algorithm}[t]
+	\caption{Modified CVM algorithm with independent coin flips.
+The function last{\textunderscore}index returns the index of the last occurrence of an element in the sequence, before the current iteration.}\label{alg:cvm_simul}
+	\begin{algorithmic}[1]
+  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
+  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
+  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
+  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ \Comment perform $l^2$ unbiased independent coin flips
+  \For{$i \gets 1$ to $l$}
+    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} \Comment insert $a_i$ if first $k$ flips are $1$s.
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ \Comment keep elems.~whose $k+1$-th flip is $1$
+      \State $k \gets k+1$
+    \EndIf
+  \EndFor
+  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
+  \end{algorithmic}
+\end{algorithm}
+
+Let us consider algorithm \circled{B} in a state where $k$ subsampling steps have been performed, i.e., $p = 2^{-k}$.
+The algorithm would perform a coin flip lazily with probability $p$ when it encounters the next stream element.
+The transformation \circled{C} is shown in \cref{alg:cvm_simul}, and we prove that it computes precisely the same distribution as \circled{B}.
+In \circled{C}, we eagerly perform a fixed number of coin flips for each sequence element at the beginning.
+Now, each element is put into the state $\chi$, whenever the first $k$ coin flips associated with the sequence element are all $1$s.
+This happens exactly with probability $2^{-k}$, which means the behaviour of the algorithm is unchanged from \circled{B}.
+Similarly, in the subsampling operation, only those elements whose $k+1$-th associated coin flip is $1$ are kept; the operation $p \mapsto \frac{p}{2}$ is replaced with $k \mapsto k+1$.
+This again preserves the behaviour of \circled{B} that each element is discarded independently with probability $1/2$.
+
+It is easy to show for \circled{C} that the coin flips are independent, and that the set of elements in $\chi$ in any state are exactly those stream elements for which the first $k$ entries of their associated coin flips are $1$.
+The final random process $\circled{D}$ directly computes the final set of elements in $\chi$ after the stream, taking $K$ as a fixed parameter; one relates \circled{C} to \circled{D} by:
+\[ \prob_{\circled{C}}(k = K \land \chi = X)  \leq \prob_{\circled{D}_K}(\chi = X)\]
+for fixed values of $K$ and $X$.
+To see how tail bounds can be derived from this inequality, let us first consider the failure event where the algorithm \circled{C}'s estimate exceeds the desired estimation interval and it ends with some fixed value $k = K$.
+Using \circled{D}, this can be bounded using a Chernoff bound for the probability that the number of stream elements whose associated coin flips start with $K$ $1$s is outside $2^{-K} |A| (1 \pm \varepsilon)$.
+Now, we can take a union bound over all the possible values $K$ to establish a global bound for the failure event in \circled{C}.
+This is explained in more detail by Chakraborty et al.~\cite{chakraborty2023}.
+
+\subsection{Eager to Lazy Coin Flips}
+A remaining question is how to formalize the transformation from \circled{B} to \circled{C}.
+Our insight is that it is best to solve the problem backwards, i.e., we start with the modified algorithm \circled{C}, which performs all the coin flips in advance \emph{eagerly} and convert it back to \circled{B} which implicitly performs the coin flips \emph{lazily} at the point they are needed.
+
+The main idea is to automatically push down the coin flips through the expression tree of \cref{alg:cvm_simul}.
+To explain how this works, let us first define the \emph{sampling} function, i.e., let $f$ be a function that takes as argument a vector of coin flips indexed by $I$, then we can express the distribution of $f$ with respect to independent unbiased coin flips as:
+\begin{isabelle_cm}
+  sample\ f\ \isacharequal\ map{\isacharunderscore}pmf\ f\ {\isacharparenleft}prod{\isacharunderscore}pmf\ I\ {\isacharparenleft}\isasymlambda\isacharunderscore\isachardot\ bernoulli{\isacharunderscore}pmf \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright\isacharparenright\isacharparenright
+\end{isabelle_cm}
+The interesting fact is that we can distribute the sampling operation over composition:
+\begin{observation}\label{o:sample_distrib} Let $f,g$ be functions consuming a set of coin flips (indexed by $I$), where $g$ also consumes the output of $f$, such that,
+$f$ depends only on the coin flips indexed by $J \subseteq I$ and $g$ depends on the complement $I - J$, then:
+\begin{isabelle_cm}
+  sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ g\ \isasymomega\ \isasymcirc\ f\ \isasymomega{\isacharparenright}\ \isacharequal\ sample\ f\ \isasymbind\ \isacharparenleft{\isasymlambda}x\isachardot\ sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ g\ \isasymomega\ x\isacharparenright\isacharparenright
+\end{isabelle_cm}
+\end{observation}
+By recursively applying the observation, we end up with elementary lookup operations, e.g., \isa{sample\ \isacharparenleft\isasymlambda\isasymomega\isachardot\ \isasymomega\ i\isacharparenright}, for which it is easy to see that it is just a coin flip, i.e., equal to \isa{bernoulli{\isacharunderscore}pmf\ \isacharparenleft\isadigit{1}/\isadigit{2}\isacharparenright}.
+This lets us readily transform \circled{C} to \circled{B} and prove their distributions equivalent.
+
+A detail that we have simplified here is that the split of the index sets, e.g., which coin flips $f$ depends on and which coin flips $g$ depends on, may be dynamic.
+For example, when the algorithm increases the subsampling counter $k$, it will have read the corresponding row of coin flips.
+This means we have a situation where the previous loop iteration communicates to the next loop iteration which coin flips it depends on using the state; and the next loop iteration will indeed only read coin flips that were not read by the previous iteration.
+
+To handle these situations we generalized \cref{o:sample_distrib} to allow for the case where the set of indices $J$ splitting the set of coin flips $f$ and $g$ depend on, may itself depend on the result of $f$.
+
+\section{Related Work}\label{sec:related_work}
+\subsection{Algorithms for the Distinct Elements Problem}
+It is important to note that there are several practical solutions for the distinct elements problem.
+The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
+Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
+Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family.
+
+Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} improved on the solution by Alon et al.
+For example, Bar-Yossef et al.\ present a solution (Algorithm~3 in their work) with a space-complexity of $\bigo(\ln (\delta^{-1}) (\varepsilon^{-2}(\ln(\varepsilon^{-1})+\ln b) + b)))$, which can be implemented in practice.
+This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$. In particular, there is no dependency on the length of the stream $l$.
+The more recent and more sophisticated solution by B\l{}asiok is space-optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
+We~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
+It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
+A comprehensive review of distinct elements algorithms has been compiled by Pettie and Wang~\cite[Table~1]{pettie2021}.
+What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
+
+The aforementioned hash-based algorithms are biased; Flajolet et al.~\cite{flajolet1985} points this out and also provides bounds on the distance between the expected result and the cardinality of the stream.
+Most authors do not discuss the matter of bias but it is not hard to show.
+One issue, for example, is that the usual method to amplify the accuracy of these algorithms is using the median, which does not preserve expectations.
+In the context of query processing, unbiasedness has been discussed~\cite[Section 2.1]{haas1995}, but we could not find any similar discussion for the distinct elements problem in the streaming model.
+
+\subsection{Probabilistic Invariants and Formalization}
+As far as we know, probabilistic invariants have not been used to establish exponentially decreasing tail-bounds.
+However, it is fairly common to use recursive analysis techniques to establish results about expectations or variance of random variables, such as their run-time~\cite[Section 1.4]{motwani1995}.
+This is easy due to the linearity of expectations and---for independent random variables---linearity of variances.
+A simple example is the Morris counter~\cite{morris1978} or the expected run-time of the quick-sort algorithm~\cite[Section 2.5]{mitzenmacher2017}.
+
+There is also research on the (automated) analysis of loop invariants, for probabilistic loops, using their characteristic functions~\cite{batz2023, mciver2005}.
+This approach works by establishing the limiting distribution of the state of the loop.
+De Medeiros et al.~\cite[Section 3.2]{demedeiros2024} also establish methods to derive limiting distributions of probabilistic loops.
+Our approach differs from these techniques by avoiding computation of the distribution, which, we think, is infeasible for the CVM algorithm.
+Instead, we investigate invariants of classes of functions of the distributions, which are relevant for the analysis.
+There is research on automated evaluation of moments for restricted classes of loops which contain only polynomial assignments and no branches~\cite{bartocci2019,kofnov2022}.
+However, these methods do not extend to algorithms with non-continuous operations, or control flow that depends on non-deterministic state variables.
+
+Finally, verification of randomized algorithms expressed in functional programming languages has been tackled by various authors using various proof assistants~\cite{audebaud2009,bosshard2024,demedeiros2024, eberl2020,gopinathan20,haslbeck2016,hurd03,Probabilistic_Prime_Tests-AFP,tan2024,tassarotti2021}; the most closely related efforts are our mechanizations of frequency moments algorithms~\cite{karayel2022, karayel2023}.
+Moreover, for classical imperative randomized algorithms, there are approaches based on probabilistic Hoare logic~\cite{denhartog2002}.
+An interesting related work by Haselwarter et al.~\cite{haselwarter2025} discusses some of the issues we encountered in the transformation-based proof (Section~\ref{sec:transformation_based_proof}), such as (approximate) equivalence between related randomized algorithms.
+However, because our work reasons directly over the semantics of randomized algorithms expressed in the Giry monad, we did not deeply explore probabilistic logic-based methods.
+
+\section{Conclusion}\label{sec:conclusion}
+We presented the first formalization of the CVM algorithm using Isabelle/HOL.
+Central to our formalization is a novel invariant-based proof technique to establish exponentially decreasing tail-bounds for randomized algorithms, which is inspired by our alternative analysis of the CVM algorithm via the Cram\'{e}r--Chernoff method.
+Our technique can be summarized by the following two steps:
+\begin{enumerate}
+\item Find functionals over the state distribution of the algorithm, for which it is possible to establish bounds on their expectation inductively/recursively.
+\item Use those bounds to establish tail bounds on the result of the algorithm.
+\end{enumerate}
+Comparing our approach against the original proof by Chakraborty et al.~\cite{chakraborty2023} shows that our technique yields a considerably shorter formalization (with \locnew~vs.~\locold~lines).
+Interestingly, our technique also readily generalized to a new CVM variant with stronger properties (totality and unbiasedness)---we formalized this latter version using the same invariant, together with a new library of results for negative association.
+In future work, it would be interesting to formalize other variations of subsampling for CVM.
+
+Note that we have yet to apply our proof technique to examples beyond CVM. (It is easy to construct artificial examples.)
+Identifying realistic applications for our new method is an interesting avenue for future work---this could lead to further refinements of the method, and a better understanding of how to identify suitable functionals for the proofs.
 

### 2025-07-04 paper/main.tex
Paper: Update SWHID references.

@@
 \supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
-\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:c677f1e3ced670373c05ac835f1cd342a2777e20]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
+\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:7df92f0347e7bd150efef42e3edbbde0037498cc]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
 \supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

### 2025-07-04 paper/main.tex
Paper: Fixes suggested by Yong Kiam.

@@
 \supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}
+
+\nolinenumbers
 
@@
 The inequalities can be established using induction over the steps of the randomized algorithm.
-Even, though, the actual distribution of the states themselves are not known, nor do we think it is possible to find useful closed form descriptions for them, the invariant establishes valuable information about the distribution of the state.
+Even though the actual distribution of the states themselves are not known, nor do we think it is possible to find useful closed form descriptions for them, the invariant establishes valuable information about the distribution of the state.
 In this case, enough information, to establish exponential tail bounds for the algorithm.

### 2025-09-19 presentation/algorithm_animation.tex
Presentation Draft

%\newcommand{streamcolor}{#5D8AA8}
\definecolor{streamcolor1}{rgb}{1.0, 0.94, 0.0}
%\definecolor{streamcolor1}{rgb}{0.36, 0.54, 0.66}
\definecolor{streamcolor2}{rgb}{0.9, 0.17, 0.31}
\definecolor{streamcolor3}{rgb}{0.0, 0.5, 0.0}
\definecolor{streamcolor4}{rgb}{0.0, 0.5, 1.0}
\definecolor{streamcolor5}{rgb}{0.74, 0.2, 0.64}
\definecolor{tailcoinbordercolor}{rgb}{0.13, 0.67, 0.8}
\definecolor{tailcoinfillcolor}{rgb}{0.54, 0.81, 0.94}
\definecolor{headcoinbordercolor}{rgb}{1.0, 0.44, 0.37}
\definecolor{headcoinfillcolor}{rgb}{0.98, 0.91, 0.71}

\newcommand{\illustratestate}[7]{
  \begin{tikzpicture}[
    streamcell/.style={circle, draw=black, fill=black!15, thick, text centered,inner sep=3pt,minimum width=0.6cm},
    arrow/.style={thick,-{Latex[length=.32cm, width=.32cm,fill=white]}},
    coin/.style={circle, draw=orange, fill=yellow, thick},
  ]
  \node[streamcell,fill=streamcolor3] at (0cm,0 cm) {};
  \node[streamcell,fill=streamcolor1] at (1cm,0 cm) {};
  \node[streamcell,fill=streamcolor3] at (2cm,0 cm) {};
  \node[streamcell,fill=streamcolor2] at (3cm,0 cm) {};
  \node[streamcell,fill=streamcolor4] at (4cm,0 cm) {};
  \node[streamcell,fill=streamcolor5] at (5cm,0 cm) {};
  \node[streamcell,fill=streamcolor1] at (6cm,0 cm) {};
  \node[streamcell,fill=streamcolor2] at (7cm,0 cm) {};
  \node[streamcell,fill=streamcolor3] at (8cm,0 cm) {};
  \node at (9cm,0cm) {$\cdots$};
  \draw[arrow] (1cm * #1, -1.5cm) to (1cm * #1, -0.5cm);
  \node (robot) at (0,-3.5cm)
    {\includegraphics[width=.25\textwidth]{contrib/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545_655/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545}};

  \node[anchor=west] at (1.4cm, -3.1cm) {$\chi = \{$};
  \ifthenelse{\equal{#3}{empty}}{}{\node[streamcell,fill=#3] at (3cm, -3.1cm) {};}
  \ifthenelse{\equal{#4}{empty}}{}{\node[streamcell,fill=#4] at (3.8cm, -3.1cm) {};}
  \ifthenelse{\equal{#5}{empty}}{}{\node[streamcell,fill=#5] at (4.6cm, -3.1cm) {};}
  \ifthenelse{\equal{#6}{empty}}{}{\node[streamcell,fill=#6] at (5.4cm, -3.1cm) {};}
  \node at (6cm, -3.1cm) {$\}$};

  \ifthenelse{\equal{#7}{head}}{
    \node[coin,fill=headcoinfillcolor,draw=headcoinbordercolor] at (-1cm,-3.5cm) {H}; 
  }{}
  \ifthenelse{\equal{#7}{tail}}{
    \node[coin,fill=tailcoinfillcolor,draw=tailcoinbordercolor] at (-1cm,-3.5cm) {T}; 
  }{}

  \node[anchor=west] at (1.4cm, -3.8cm) {$p = #2, n = 4$};

  \end{tikzpicture}
}

\begin{frame}
\frametitle{Illustration of the Algorithm}

\only<1>{\illustratestate{0}{1}{empty}{empty}{empty}{empty}{none}}
\only<2>{\illustratestate{0}{1}{empty}{empty}{empty}{empty}{head}}
\only<3>{\illustratestate{0}{1}{streamcolor3}{empty}{empty}{empty}{head}}
\only<4>{\illustratestate{1}{1}{streamcolor3}{empty}{empty}{empty}{head}}
\only<5>{\illustratestate{2}{1}{streamcolor3}{streamcolor1}{empty}{empty}{head}}
\only<6>{\illustratestate{3}{1}{streamcolor3}{streamcolor1}{empty}{empty}{head}}
\only<7>{\illustratestate{4}{1}{streamcolor3}{streamcolor1}{streamcolor2}{empty}{head}}
\only<8>{\illustratestate{4}{1}{streamcolor3}{streamcolor1}{streamcolor2}{streamcolor4}{head}}
\only<9>{\illustratestate{4}{1}{empty}{streamcolor1}{streamcolor2}{empty}{none}}
\only<10>{\illustratestate{4}{\frac{1}{2}}{empty}{streamcolor1}{streamcolor2}{empty}{none}}
\only<11>{\illustratestate{5}{\frac{1}{2}}{empty}{streamcolor1}{streamcolor2}{empty}{none}}
\only<12>{\illustratestate{5}{\frac{1}{2}}{empty}{streamcolor1}{streamcolor2}{empty}{head}}
\only<13>{\illustratestate{5}{\frac{1}{2}}{streamcolor5}{streamcolor1}{streamcolor2}{empty}{head}}
\only<14>{\illustratestate{6}{\frac{1}{2}}{streamcolor5}{streamcolor1}{streamcolor2}{empty}{none}}
\only<15>{\illustratestate{6}{\frac{1}{2}}{streamcolor5}{streamcolor1}{streamcolor2}{empty}{tail}}
\only<16>{\illustratestate{6}{\frac{1}{2}}{streamcolor5}{empty}{streamcolor2}{empty}{taiil}}
\only<17>{\illustratestate{7}{\frac{1}{2}}{streamcolor5}{empty}{streamcolor2}{empty}{taiil}}
\end{frame}

### 2025-09-19 presentation/main.tex
Presentation Draft

\documentclass{beamer}
\usetheme{Rochester}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{calc}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw, minimum size=3.5mm,inner sep=0.5pt] (char) {#1};}}

\usepackage{ifthen}
\usepackage[noend]{algpseudocode}
\usepackage{soul}
\usepackage{isabelle}
\usepackage{isabellesym}
\isabellestyle{it}

\DeclareMathOperator{\mbind}{\text{\isa{\isasymbind}}}
\DeclareMathOperator{\expect}{\mathrm{E}}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand{\ift}[3]{\mathbf{if} \; #1 \; \mathbf{then} \; #2 \; \mathbf{else} \; #3}
\newcommand{\integral}[3]{\int_{#1} \! #2 \, \mathrm{d} #3}

\newcommand{\getsr}{\xleftarrow{\$}}

\title{Verification of the CVM algorithm with a Functional Probabilistic Invariant}
\author{E. Karayel\inst{1}, J. Watt\inst{2}, D. Khu\inst{2}, K. Meel\inst{3}, Y. K. Tan\inst{2}}
\institute[VFU]{
  \inst{1}%
    Technical University of Munich
  \and
  \inst{2}%
    Institute for Infocomm Research (I$^2$R), A*STAR, Singapore
  \and
  \inst{3}%
    University of Toronto
}

\usepackage{mathtools}
\DeclareMathOperator{\prob}{\mathcal P}
\DeclareMathOperator{\bigo}{\mathcal O}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclareMathOperator{\Ber}{\mathrm{Ber}}
\DeclareMathOperator{\indicat}{\mathrm{I}}

\begin{document}
\AtBeginSection[]
{
    \begin{frame}
        \frametitle{Table of Contents}
        \tableofcontents[currentsection]
    \end{frame}
}

\frame{\titlepage}
\section{Introduction to the CVM algorithm}
\begin{frame}
\frametitle{What is the CVM algorithm?}
\begin{itemize}
\item Name after initals of S. Chakraborty, N. V. Vinodchandran, K. S. Meel
\item Streaming Algorithm (one-pass)
\[
  a_1, \ldots, a_l
\]
\item Algorithm has only limited mutable state
\[
  \bigo( \varepsilon^{-2} \ln(\delta^{-1} l ) b ) 
\]
\item Approximation of the count of distinct elements
\item Randomized
\item Small error with high probability
\[
  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
\]
for choosable $\varepsilon > 0$, $\delta > 0$.
\end{itemize}
\end{frame}
\begin{frame}
\frametitle{Illustration of Streaming Algorithms}
\begin{figure}
\begin{center}
\begin{tikzpicture}[
  streamcell/.style={rectangle, draw=black, fill=black!15, thick, text centered,inner sep=3pt,minimum width=0.6cm},
]
\node[streamcell] (cell1) at (-3,2) {$a_1$};
\node[streamcell] (cell1) at (-2.4,2) {$a_1$};
\node[streamcell] (cell1) at (-1.8,2) {$a_2$};
\node at (0,2) {$\cdots$};
\node[streamcell] (cell1) at (2,2) {$a_{l}$};
\node (robot) at (0,0)
    {\includegraphics[width=.25\textwidth]{contrib/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545_655/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545}};
\end{tikzpicture}
\end{center}
\end{figure}
\end{frame}
%\section{Comparison with Previous Algorithms}
\begin{frame}[label=cvm_algorithm]
\frametitle{The CVM Algorithm}
%\begin{algorithm}[h!]
%	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
\begin{algorithmic}
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment random bit $b$ from the Bernoulli distribution
    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else \Comment remove $a_i$ otherwise
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$ \Comment fail if $\chi$ remains full
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
\end{algorithmic}
%\end{algorithm}
\end{frame}

\input{algorithm_animation}

\againframe{cvm_algorithm}

\begin{frame}
\frametitle{Comparison with Previous Algorithms}

\begin{table}
\begin{tabular}{l | l | l | l}
\hline
Year & Author(s) & Space & Remarks \\
\hline
2001 & Bar-Yossef et al.~\cite{baryossef2002} & $\tilde{O}( \ln(\delta^{-1}) (\varepsilon^{-2}  + b ))$ & M. H. \\
2007 & HLL~\cite{flajolet2007} & $O(\ln(\delta^{-1} \varepsilon^{-2} \ln b)$ & O. M. \\
2010 & Kane et al.~\cite{kane2010} & $O( \ln(\delta^-1) (\varepsilon^{-2} + b) )$ & H. \\
2020 & B\l{}asiok~\cite{blasiok2020} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. \\
2022 & Karayel~\cite{karayel2022} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. M. \\
2023 & Chakraborty et al.~\cite{chakraborty2023} & $O( \varepsilon^{-2} \ln( \delta^{-1} l) b )$ & - \\
\end{tabular}
\end{table}
Abbreviations: 
\begin{itemize}
\item M: Supports merging of sketches
\item H: Relies on hash functions.
\item E: Relies on expander graphs.
\item O: Random oracle model.
\end{itemize}
\end{frame}

\section{Summary of Results}
\begin{frame}
\frametitle{Summary of Results}
\begin{itemize}
\item <2-> Verification of Correctness \only<6->{using the original and new method}
\item <3-> Discovery of a new more succint proof (1003 lines vs. 2634 lines.)\only<4->{, which is based on a \emph{new} proof technique for randomized algorithms.}
\item <5-> We call it: \emph{Functional Probabilistic Invariants}
\item <7-> Discovery of an unbiased total variant of the algorithm. (Application of our new technique.)
\end{itemize}
\end{frame}

\section{Verification using Functional Probabilistic Invariants}
\begin{frame}[label=algsplit]
\frametitle{Another slide with the algorithm}
\begin{algorithmic}
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$
\end{algorithmic}

\tikz[overlay, shift=(current page.north west)]{
\begin{scope}[x={(current page.north east)},y={(current page.south west)}]

\only<2->{
        \fill[blue!50, opacity=0.5] (0.07,0.4) rectangle (0.7,0.67);
        \node [anchor=west] at (0.45,0.535) {Step 1};
}
\only<3->{
        \fill[red!50, opacity=0.5] (0.07,0.675) rectangle (0.7,0.845);
        \node [anchor=west] at (0.45,0.76) {Step 2};
}
\only<4>{
        \fill[yellow!50,opacity=0.5] (0.07,0.846) rectangle (0.7,0.895);
        \node [anchor=west] at (0.45,0.87) {Step 3};
}
\only<5->{
        \fill[white] (0.07,0.846) rectangle (0.7,0.895);
}
\only<6->{
  \fill[green!50, opacity=0.5] (0.03, 0.895) rectangle (0.7, 0.97);
   \node [anchor=west] at (0.45,0.932) {Estimate};
}
\end{scope}
}
\end{frame}

\begin{frame}
\frametitle{Giry Monad}
We can represent randomized algorithms using the Giry monad:
\begin{itemize}
\item Primitive random operations, e.g., $\Ber(p)$
\item Return operation $\mathrm{return} \, x$
\item Sequential compositon $m \mbind f$
\end{itemize}
\pause
\begin{block}{Fact}
For an event $E$: 
\begin{eqnarray*}
\prob_{m \mbind f}( E )  & = & \int_{m} P_{f(x)} (E) \, dx \onslide<3->{= \sum_x P_{f{x}} (E) P_m({x})} \\
\onslide<4->{\expect_{m \mbind f}[ g ]  & = & \int_{m} \expect_{f(x)} [g] \, dx}
\end{eqnarray*}
\end{block}
%\end{frame}
%\begin{frame}
%\frametitle{Giry Monad II}
\end{frame}
\againframe<5>{algsplit}
\begin{frame}
\frametitle{Our algorithm in monadic notation}
\begin{eqnarray*}
  \mathrm{init} & \mbind & \mathrm{step}_1 a_1 \mbind \mathrm{step}_2 \mbind \mathrm{step}_1 a_2 \mbind \mathrm{step}_2 \cdots \\
  & \mbind & \mathrm{step_1} a_l \mbind \mathrm{step}_2 \mbind \mathrm{estimate} \\
  \mathrm{init} & = & \mathrm{return} \, (1, \emptyset)
\end{eqnarray*}
\pause
Key invariant:
\[
\expect \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1
\]
for all $x$ that are present in the stream.
%\end{block}
\end{frame}

\begin{frame}
\frametitle{Verifying the invariant: Induction step for Step 2}
\begin{block}{Assumption}
$\expect_m \left[ \frac{\indicat(x \in \chi)}{p} \right] = 1$
\end{block}
\begin{block}{Goal}
$ L := \expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1$
\end{block}
\[
L =
\only<1>{\expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]}
\only<2>{\integral{m}
      {\left(\ift{|\chi|=n}{\integral{\mathrm{subs.}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}}
\only<3>{\integral{m}
      {\left(\ift{|\chi|=n}{\frac{p}{2}\integral{\mathrm{subs.}(\chi)}{\indicat(s \in \tau)}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}}
\only<4>{\integral{m}
      {\left(\ift{|\chi|=n}{\frac{p}{2}\frac{\indicat(s \in \chi)}{2}}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}}
\only<5>{\integral{m}
      {\left(\ift{|\chi|=n}{\frac{\indicat(s \in \chi)}{p}}{\frac{\indicat(s \in \chi)}{p}} \right)}
      {\sigma}}
\only<6>{\integral{m}
      {\left(\frac{\indicat(s \in \chi)}{p} \right)}
      {\sigma}}
\only<7>{1}
\]
\end{frame}

\begin{frame}<1>[label=observations]
\frametitle{Observations}
\begin{itemize}
\item Expected result is the count of distinct elements.
\only<2->{\item Can extend the above recursive analysis to more complex expressions: }
\only<3>{
\[
  \expect \left[ h \left( \frac{\indicat(s \in \chi)}{p} \right)\right] \leq h(1)
\]
for every non-negative concave function $h$ and stream elements $s$.}
\only<4->{
\[
  \expect \left[ \prod_{s \in S}  h \left( \frac{\indicat(s \in \chi)}{p} \right)\right] \leq h(1)^{|S|}
\]
for every non-negative concave function $h$ and subset of stream elements $S$.}
\only<5->{\item $\Rightarrow$ Approximation Guarantee
\[
  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
\]
}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Concentration}
\begin{tikzpicture}[scale=0.5]
\begin{axis}[
    xmin=0, xmax = 6.5,
    ymin=0, ymax=0.7,
    axis lines = left,
    xlabel = {},
    xtick={2.75,3.75,4.75},
    xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
    ytick=\empty,
    ylabel = {pdf},
    height = \textwidth,
    width = \textwidth,
]
\addplot [
    domain=0:6, 
    samples=1000, 
]{max(0,35/2*(x/6)^4*(1-x/6)^2)};
\addplot [
    domain=4.75:6, 
    samples=1000, 
    fill=gray,
]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
\addplot [
    domain=0:2.75, 
    samples=1000, 
    fill=gray,
]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
\draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}[scale=0.5]
\begin{axis}[
    xmin=0, xmax = 6.5,
    ymin=0, ymax=0.7,
    axis lines = left,
    xlabel = {},
    xtick={2.75,3.75,4.75},
    xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
    ytick=\empty,
    ylabel = {pdf},
    height = \textwidth,
    width = \textwidth,
]
\addplot [
    domain=0:6, 
    samples=1000, 
]{0.65*exp(-(2*(x-3.75))^2)};
\addplot [
    domain=4.75:6, 
    samples=1000, 
    fill=gray,
]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
\addplot [
    domain=0:2.75, 
    samples=1000, 
    fill=gray,
]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
\draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
\end{axis}
\end{tikzpicture}
\end{frame}

\againframe<2->{observations}

\begin{frame}
\frametitle{Observations II}
\begin{itemize}
\item The new proof is much shorter than the original proof (1003 lines vs. 2634 lines.)
\only<2->{\item \emph{As far as we can tell,} this technique is new.}
\only<3->{\item The new proof opens up the design-space of the algorithm.}
\only<4>{\item For example we can select the subsampling ratio dynamically.}
\only<6->{\item More interestingly: It is possible to use a different subsampling operation.}
\only<7->{\item Select a random $nf$-subset (where $\frac{1}{2} \leq f < 1$, $nf \in \mathbb Z$).}
\only<8->{\item $\Rightarrow$ Unbiased algorithm}
\end{itemize}
\end{frame}
\againframe<4>{algsplit}

\section{Sketch of the Original Proof}
\begin{frame}
\frametitle{Original Proof}
\begin{tikzpicture}[node distance=2.7cm, auto,scale=0.8,every node/.style={scale=0.8}]

    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center] {CVM};
    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=A] {CVM w/o\\ Step 3};
    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=B] {Eager Alg.};
    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=C] {Random \\ Process};
    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=D] {Chernoff \\ Bounds};

    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=B, node distance = 2.0cm] {Functional\\Invariant};

    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=E, node distance = 2.0cm] {CVM PAC\\ Guarantee};

    \draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(E.south east) + (0.3,-0.6)$);
    \node[anchor=south] at ($(D.south) + (0.0,-0.6)$) {Transformation-based Proof};

    \draw[dashed,thick] ($(A.north west) + (-0.5,0.4)$) rectangle ($(C'.south east) + (0.2,-0.2)$);
    \node[anchor=south,align=center] at ($(C'.west) + (-1.5,-0.45)$) {Probabilistic\\Invariant Proof};

    \draw[->] (A) -- (B);
    \draw[->] (B) -- (C);
    \draw[->] (B) -- (C');
    \draw[->] (C) -- (D);
    \draw[->] (D) -- (E);
    \draw[->] (E) -- (pac);
    \draw[->] (C') -- (pac);

\end{tikzpicture}

\end{frame}

\againframe{cvm_algorithm}

\begin{frame}
\frametitle{CVM Algorithm II}
\begin{algorithmic}
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(2^{-k})$ \Comment random bit $b$ from the Bernoulli distribution
    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else \Comment remove $a_i$ otherwise
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $k \gets k + 1$
    \EndIf
  \EndFor
  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
\end{algorithmic}
\end{frame}

\begin{frame}
\frametitle{Eager Algorithm}
\begin{algorithmic}
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ %such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
  \For{$i \gets 1$ to $l$}
    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} % \Comment insert $a_i$ if first $k$ flips are $1$s.
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ % \Comment keep elems.~whose $k+1$-th flip is $1$
      \State $k \gets k+1$
    \EndIf
  \EndFor
  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{frame}

\begin{frame}
\frametitle{Random Process}
Previous algorithm preserves the invariant:
\[
  s \in \chi \leftrightarrow \textrm{b[ last{\textunderscore}index}(s), 0] = \cdots = \textrm{b[ last{\textunderscore}index}(s), k-1] = 1
\]
Establish Chernoff bounds using independence of the coin flips of $b$ for all possible outcomes of $k$.
\end{frame}

\begin{frame}
\frametitle{Negative Association}
\begin{itemize}
\item Generalization of the concept of Independence
\pause
\item Obey rules similar to independent random variables
\pause
\item Ober Chernoff-Bounds similar to independent random variables
\pause
\item Formalized in a separate AFP entry: Negative{\textunderscore}Association
\pause
\item Example: The sequence of bits in an idealized Bloom Filter form an negative dependent set
\end{itemize}

\section*{References}
\begin{frame}[allowframebreaks]
        \frametitle{References}
        \bibliographystyle{plain}
        \bibliography{main.bib}
\end{frame}

\end{document}

### 2025-09-27 presentation/main.tex
Edits after dry-run

@@
-\documentclass{beamer}
+\documentclass[xcolor=table]{beamer}
 \usetheme{Rochester}
@@
 \usetikzlibrary{calc}
+\usetikzlibrary{decorations.pathmorphing}
 \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
@@
 \title{Verification of the CVM algorithm with a Functional Probabilistic Invariant}
-\author{E. Karayel\inst{1}, J. Watt\inst{2}, D. Khu\inst{2}, K. Meel\inst{3}, Y. K. Tan\inst{2}}
+\author{E. Karayel\inst{1}, J. Watt\inst{2}, D. Khu\inst{2}, K. Meel\inst{4,5}, Y. K. Tan\inst{2,3}}
 \institute[VFU]{
   \inst{1}%
-    Technical University of Munich
+    Technical University of Munich, Germany
   \and
@@
   \inst{3}%
-    University of Toronto
+    Nanyang Technological University, Singapore
+  \and
+  \inst{4}%
+    Georgia Institute of Technology, Atlanta, GA, USA
+  \and
+  \inst{5}%
+    University of Toronto, Canada
 }
@@
 \frame{\titlepage}
+\begin{frame}
+\frametitle{Table of Contents}
+\tableofcontents
+\end{frame}
+
 \section{Introduction to the CVM algorithm}
@@
 \item Name after initals of S. Chakraborty, N. V. Vinodchandran, K. S. Meel
-\item Streaming Algorithm (one-pass)
-\[
-  a_1, \ldots, a_l
-\]
-\item Algorithm has only limited mutable state
-\[
-  \bigo( \varepsilon^{-2} \ln(\delta^{-1} l ) b ) 
-\]
+\pause
+\item Streaming Algorithm (one sequential pass): $a_1, \ldots, a_l$
+\pause
+\item Algorithm has only limited mutable state: $\bigo( \varepsilon^{-2} \ln(\delta^{-1} l ) b )$
+\pause
 \item Approximation of the count of distinct elements
+\pause
 \item Randomized
+\pause
 \item Small error with high probability
 \[
-  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
+  \prob\left(|X - |A|| \geq \varepsilon |A| \right) \leq \delta 
 \]
-for choosable $\varepsilon > 0$, $\delta > 0$.
+for choosable $\varepsilon > 0$, $\delta > 0$
+\pause
+\item Suprisingly simple compared to previous algorithms for the same problem \pause (see e.g. article in Quanta Magazine\cite{quanta})
+\pause
+\item No hashing
 \end{itemize}
@@
 \end{frame}
-%\section{Comparison with Previous Algorithms}
+
 \begin{frame}[label=cvm_algorithm]
 \frametitle{The CVM Algorithm}
-%\begin{algorithm}[h!]
-%	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
 \begin{algorithmic}
@@
 \end{algorithmic}
-%\end{algorithm}
 \end{frame}
@@
 \frametitle{Comparison with Previous Algorithms}
-
+\alt<2>{\definecolor{colorrowa}{rgb}{0,0.8,0.6}}{\definecolor{colorrowa}{rgb}{1,1,1}}
+\alt<3>{\definecolor{colorrowb}{rgb}{0,0.8,0.6}}{\definecolor{colorrowb}{rgb}{1,1,1}}
+\alt<4>{\definecolor{colorrowc}{rgb}{0,0.8,0.6}}{\definecolor{colorrowc}{rgb}{1,1,1}}
+\alt<5>{\definecolor{colorrowd}{rgb}{0,0.8,0.6}}{\definecolor{colorrowd}{rgb}{1,1,1}}
+\alt<6>{\definecolor{colorrowe}{rgb}{0,0.8,0.6}}{\definecolor{colorrowe}{rgb}{1,1,1}}
+\alt<7>{\definecolor{colorrowf}{rgb}{0,0.8,0.6}}{\definecolor{colorrowf}{rgb}{1,1,1}}
 \begin{table}
@@
 \hline
-2001 & Bar-Yossef et al.~\cite{baryossef2002} & $\tilde{O}( \ln(\delta^{-1}) (\varepsilon^{-2}  + b ))$ & M. H. \\
-2007 & HLL~\cite{flajolet2007} & $O(\ln(\delta^{-1} \varepsilon^{-2} \ln b)$ & O. M. \\
-2010 & Kane et al.~\cite{kane2010} & $O( \ln(\delta^-1) (\varepsilon^{-2} + b) )$ & H. \\
-2020 & B\l{}asiok~\cite{blasiok2020} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. \\
-2022 & Karayel~\cite{karayel2022} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. M. \\
-2023 & Chakraborty et al.~\cite{chakraborty2023} & $O( \varepsilon^{-2} \ln( \delta^{-1} l) b )$ & - \\
+\rowcolor{colorrowa} 2001 & Bar-Yossef et al.~\cite{baryossef2002} & $\tilde{O}( \ln(\delta^{-1}) (\varepsilon^{-2}  + b ))$ &  M. H. \\
+\rowcolor{colorrowb} 2007 & HLL~\cite{flajolet2007} &  $O(\ln(\delta^{-1} \varepsilon^{-2} \ln b)$ & O. M. H. \\
+\rowcolor{colorrowc} 2010 & Kane et al.~\cite{kane2010} & $O( \ln(\delta^-1) (\varepsilon^{-2} + b) )$ & H. \\
+\rowcolor{colorrowd} 2020 & B\l{}asiok~\cite{blasiok2020} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. \\
+\rowcolor{colorrowe} 2022 & Karayel~\cite{karayel2022} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. M. \\
+\rowcolor{colorrowf} 2023 & Chakraborty et al.~\cite{chakraborty2023} & $O( \varepsilon^{-2} \ln( \delta^{-1} l) b )$ & - \\
 \end{tabular}
@@
 \begin{itemize}
-\item M: Supports merging of sketches
+\item M: Supports merging of sketches.
 \item H: Relies on hash functions.
@@
 \item <2-> Verification of Correctness \only<6->{using the original and new method}
-\item <3-> Discovery of a new more succint proof (1003 lines vs. 2634 lines.)\only<4->{, which is based on a \emph{new} proof technique for randomized algorithms.}
+\item <3-> Discovery of a new more succinct proof\only<6->{ (1003 lines vs.\ 2634 lines)}\only<4->{, which is based on a \emph{new} proof technique for randomized algorithms.}
 \item <5-> We call it: \emph{Functional Probabilistic Invariants}
@@
 \section{Verification using Functional Probabilistic Invariants}
-\begin{frame}[label=algsplit]
+\begin{frame}<1-5>[label=algsplit]
 \frametitle{Another slide with the algorithm}
 \begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{\only<6>{3}\only<1-5>{6}l}{\delta})} }$
   \For{$i \gets 1$ to $l$}
@@
     \If{$|\chi| = n$}
-      \State $\chi \getsr \mathrm{subsample}(\chi)$
+      \State $\chi \getsr \only<1-5>{\mathrm{subsample}}\only<6>{\mathrm{subsample'}}(\chi)$
       \State $p \gets \frac{p}{2}$
@@
 }
-\only<4>{
-        \fill[yellow!50,opacity=0.5] (0.07,0.846) rectangle (0.7,0.895);
-        \node [anchor=west] at (0.45,0.87) {Step 3};
+\only<4->{
+        \draw[thick, decorate, decoration={snake,amplitude=0.6pt}] (0.072,0.87) -- (0.42, 0.87);
+}
+\only<6>{
+       \fill[white] (0.07,0.846) rectangle (0.7,0.895);
+%        \node [anchor=west] at (0.45,0.87) {Step 3};
 }
 \only<5->{
-        \fill[white] (0.07,0.846) rectangle (0.7,0.895);
-}
-\only<6->{
   \fill[green!50, opacity=0.5] (0.03, 0.895) rectangle (0.7, 0.97);
@@
 \begin{eqnarray*}
-\prob_{m \mbind f}( E )  & = & \int_{m} P_{f(x)} (E) \, dx \onslide<3->{= \sum_x P_{f{x}} (E) P_m({x})} \\
+\prob_{m \mbind f}( E )  & = & \int_{m} P_{f(x)} (E) \, dx \onslide<3->{= \sum_x P_{f(x)} (E) P_m({x})} \\
 \onslide<4->{\expect_{m \mbind f}[ g ]  & = & \int_{m} \expect_{f(x)} [g] \, dx}
@@
 \end{block}
-%\end{frame}
-%\begin{frame}
-%\frametitle{Giry Monad II}
 \end{frame}
+
 \againframe<5>{algsplit}
+
 \begin{frame}
 \frametitle{Our algorithm in monadic notation}
+\newcommand{\stepone}{\textcolor{blue}{\mathrm{step}_1}}
+\newcommand{\steptwo}{\textcolor{red}{\mathrm{step}_2}}
+\newcommand{\estimate}{\textcolor{green}{\mathrm{estimate}}}
 \begin{eqnarray*}
-  \mathrm{init} & \mbind & \mathrm{step}_1 a_1 \mbind \mathrm{step}_2 \mbind \mathrm{step}_1 a_2 \mbind \mathrm{step}_2 \cdots \\
-  & \mbind & \mathrm{step_1} a_l \mbind \mathrm{step}_2 \mbind \mathrm{estimate} \\
+  \mathrm{init} & \mbind & \stepone a_1 \mbind \steptwo \mbind \stepone a_2 \mbind \steptwo \cdots \\
+  & \mbind & \stepone a_l \mbind \steptwo \mbind \estimate \\
   \mathrm{init} & = & \mathrm{return} \, (1, \emptyset)
@@
 \pause
-Key invariant:
+Example invariant:
 \[
-\expect \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1
+\expect \left[ \frac{\indicat(s \in \chi)}{p} \right]= 1
 \]
-for all $x$ that are present in the stream.
-%\end{block}
+for all $s$ that are present in the stream.
 \end{frame}
@@
 \begin{block}{Assumption}
-$\expect_m \left[ \frac{\indicat(x \in \chi)}{p} \right] = 1$
+$\expect_m \left[ \frac{\indicat(s \in \chi)}{p} \right] = 1$
 \end{block}
 \begin{block}{Goal}
-$ L := \expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1$
+$ L := \expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(s \in \chi)}{p} \right]= 1$
 \end{block}
@@
 L =
-\only<1>{\expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]}
+\only<1>{\expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(s \in \chi)}{p} \right]}
 \only<2>{\integral{m}
@@
 \only<3>{\integral{m}
-      {\left(\ift{|\chi|=n}{\frac{p}{2}\integral{\mathrm{subs.}(\chi)}{\indicat(s \in \tau)}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
+      {\left(\ift{|\chi|=n}{\frac{2}{p}\integral{\mathrm{subs.}(\chi)}{\indicat(s \in \tau)}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
       {\sigma}}
 \only<4>{\integral{m}
-      {\left(\ift{|\chi|=n}{\frac{p}{2}\frac{\indicat(s \in \chi)}{2}}{\frac{\indicat(s \in \chi)}{p}} \right)}
+      {\left(\ift{|\chi|=n}{\frac{2}{p}\frac{\indicat(s \in \chi)}{2}}{\frac{\indicat(s \in \chi)}{p}} \right)}
       {\sigma}}
@@
 \only<6>{\integral{m}
-      {\left(\frac{\indicat(s \in \chi)}{p} \right)}
+      {\frac{\indicat(s \in \chi)}{p} }
       {\sigma}}
@@
 
-\begin{frame}<1>[label=observations]
+\begin{frame}[label=observations]
 \frametitle{Observations}
@@
 \item Expected result is the count of distinct elements.
-\only<2->{\item Can extend the above recursive analysis to more complex expressions: }
-\only<3>{
+\only<2->{\item We want to establish concentration}
+\only<4->{\item Can extend the above recursive analysis to more complex expressions: }
+\only<5>{
 \[
@@
 for every non-negative concave function $h$ and stream elements $s$.}
-\only<4->{
+\only<6->{
 \[
@@
 for every non-negative concave function $h$ and subset of stream elements $S$.}
-\only<5->{\item $\Rightarrow$ Approximation Guarantee
+\only<7->{\item $\Rightarrow$\only<8->{*} Approximation Guarantee
 \[
-  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
+  \prob\left(|X - |A|| \geq \varepsilon |A| \right) \leq \delta 
 \]
+\only<8->{\item More details in the paper.}
 }
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Concentration}
-\begin{tikzpicture}[scale=0.5]
+\only<3>{
+\begin{figure}
+\centering
+\begin{tikzpicture}[scale=0.4]
 \begin{axis}[
@@
 \end{tikzpicture}
-\begin{tikzpicture}[scale=0.5]
+\begin{tikzpicture}[scale=0.4]
 \begin{axis}[
@@
 \end{tikzpicture}
+\end{figure}
+}
 \end{frame}
 
-\againframe<2->{observations}
+%% \begin{frame}
+%% \frametitle{Concentration}
+%% \begin{tikzpicture}[scale=0.5]
+%% \begin{axis}[
+%%     xmin=0, xmax = 6.5,
+%%     ymin=0, ymax=0.7,
+%%     axis lines = left,
+%%     xlabel = {},
+%%     xtick={2.75,3.75,4.75},
+%%     xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
+%%     ytick=\empty,
+%%     ylabel = {pdf},
+%%     height = \textwidth,
+%%     width = \textwidth,
+%% ]
+%% \addplot [
+%%     domain=0:6, 
+%%     samples=1000, 
+%% ]{max(0,35/2*(x/6)^4*(1-x/6)^2)};
+%% \addplot [
+%%     domain=4.75:6, 
+%%     samples=1000, 
+%%     fill=gray,
+%% ]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
+%% \addplot [
+%%     domain=0:2.75, 
+%%     samples=1000, 
+%%     fill=gray,
+%% ]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
+%% \draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
+%% \end{axis}
+%% \end{tikzpicture}
+%% \begin{tikzpicture}[scale=0.5]
+%% \begin{axis}[
+%%     xmin=0, xmax = 6.5,
+%%     ymin=0, ymax=0.7,
+%%     axis lines = left,
+%%     xlabel = {},
+%%     xtick={2.75,3.75,4.75},
+%%     xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
+%%     ytick=\empty,
+%%     ylabel = {pdf},
+%%     height = \textwidth,
+%%     width = \textwidth,
+%% ]
+%% \addplot [
+%%     domain=0:6, 
+%%     samples=1000, 
+%% ]{0.65*exp(-(2*(x-3.75))^2)};
+%% \addplot [
+%%     domain=4.75:6, 
+%%     samples=1000, 
+%%     fill=gray,
+%% ]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
+%% \addplot [
+%%     domain=0:2.75, 
+%%     samples=1000, 
+%%     fill=gray,
+%% ]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
+%% \draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
+%% \end{axis}
+%% \end{tikzpicture}
+%% \end{frame}
+
+%% \againframe<2->{observations}
 
@@
 \end{frame}
-\againframe<4>{algsplit}
+\againframe<5-6>{algsplit}
 
@@
 
-\againframe{cvm_algorithm}
+\newcommand{\overlaytransform}[1]{
+\begin{tikzpicture}[node distance=2.7cm, auto,scale=0.8,every node/.style={scale=0.8}]
+
+\ifthenelse{#1=1}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
+\node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor] {CVM};
+\ifthenelse{#1=2}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
+\node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=A] {CVM w/o\\ Step 3};
+\ifthenelse{#1=3}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
+\node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=B] {Eager Alg.};
+\ifthenelse{#1=4}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
+\node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=C] {Random \\ Process};
+\draw[->] (A) -- (B);
+\draw[->] (B) -- (C);
+\draw[->] (C) -- (D);
+\draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(D.south east) + (0.3,-0.3)$);
+\end{tikzpicture}
+}
+
+\begin{frame}
+\frametitle{CVM}
+\overlaytransform{1}
+\begin{algorithmic}
+  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6 l}{\delta})} }$
+  \For{$i \gets 1$ to $l$}
+    \State $b \getsr \Ber(p)$
+    \If{$b$}
+      \State $\chi \gets \chi \cup \{a_i\}$
+    \Else
+      \State $\chi \gets \chi - \{a_i\}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \State $\chi \getsr \mathrm{subsample}(\chi)$
+      \State $p \gets \frac{p}{2}$
+    \EndIf
+    \If{$|\chi| = n$}
+      \Return $\bot$
+    \EndIf
+  \EndFor
+  \State \Return $\frac{|\chi|}{p}$
+\end{algorithmic}
+\end{frame}
 
@@
 \frametitle{CVM Algorithm II}
+\overlaytransform{2}
 \begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
   \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(2^{-k})$ \Comment random bit $b$ from the Bernoulli distribution
-    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
+    \State $b \getsr \Ber(2^{-k})$ 
+    \If{$b$} 
       \State $\chi \gets \chi \cup \{a_i\}$
-    \Else \Comment remove $a_i$ otherwise
+    \Else 
       \State $\chi \gets \chi - \{a_i\}$
     \EndIf
-    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
+    \If{$|\chi| = n$} 
       \State $\chi \getsr \mathrm{subsample}(\chi)$
@@
   \EndFor
-  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
+  \State \Return $2^k |\chi|$ 
 \end{algorithmic}
@@
 \frametitle{Eager Algorithm}
+\overlaytransform{3}
 \begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ %such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
   \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
+  \only<2->{\color{red}}
   \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
+  \only<2->{\color{black}}
   \For{$i \gets 1$ to $l$}
-    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} % \Comment insert $a_i$ if first $k$ flips are $1$s.
+    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$}
       \State $\chi \gets \chi \cup \{a_i\}$
@@
     \EndIf
-    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
+    \If{$|\chi| = n$} % \Comment if buffer $\chi$ is full
       \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ % \Comment keep elems.~whose $k+1$-th flip is $1$
@@
   \EndFor
-  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
+  \State \Return $2^k |\chi|$ 
   \end{algorithmic}
@@
 \frametitle{Random Process}
-Previous algorithm preserves the invariant:
+\overlaytransform{4}
+
+\begin{block}{Observation}
+The eager algorithm preserves the invariant:
 \[
-  s \in \chi \leftrightarrow \textrm{b[ last{\textunderscore}index}(s), 0] = \cdots = \textrm{b[ last{\textunderscore}index}(s), k-1] = 1
+  s \in \chi \leftrightarrow \text{For all $j < k$:} b[ \textrm{last{\textunderscore}index}(s), j] = 1
 \]
-Establish Chernoff bounds using independence of the coin flips of $b$ for all possible outcomes of $k$.
+\end{block}
+\pause
+Non-deterministic algorithm:
+\begin{algorithmic}
+  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
+  \State $k \gets \{0,\dots,k_\mathrm{max}\}$ \Comment Non-deterministc step.
+  \State $\chi \gets \{ s \in A \mid b[ \textrm{last{\textunderscore}index}(s), j] = 1 \textrm { for all } j < k \}$
+  \State \Return $2^k |\chi|$ 
+\end{algorithmic}
+%\pause
+%Establish Chernoff bounds using union-bound  for all possible outcomes of $k$\pause, based on independence of the coin flips of $b$.
 \end{frame}
@@
 \begin{frame}
-\frametitle{Negative Association}
+\frametitle{Conclusion/Summary}
 \begin{itemize}
-\item Generalization of the concept of Independence
+\item Verification of the CVM Algorithm
 \pause
-\item Obey rules similar to independent random variables
+\item New Technique for the Verfication of Randomized Algorithms
 \pause
-\item Ober Chernoff-Bounds similar to independent random variables
+\item New unbiased/total variant of the CVM Algorithm
 \pause
-\item Formalized in a separate AFP entry: Negative{\textunderscore}Association
+\item Development of a Library for Negative Association (a generaliztion of the concept of independence)
 \pause
-\item Example: The sequence of bits in an idealized Bloom Filter form an negative dependent set
+\begin{itemize}
+\item Example: Indicator random variables of an idealized Bloom-Filter
 \end{itemize}
+\pause
+\item $\Rightarrow$ Everything in the AFP
+\pause
+\item Thank You!
+\end{itemize}
+\end{frame}
 

### 2025-09-27 presentation/main.tex
Minor edits and split into sections.

--- 2025-09-19 presentation/main.tex
@@
 \newcommand{\getsr}{\xleftarrow{\$}}
+\newcommand{\robot}[1]{\includegraphics[width=#1]{contrib/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545_655/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545}}
 
 \title{Verification of the CVM algorithm with a Functional Probabilistic Invariant}
-\author{E. Karayel\inst{1}, J. Watt\inst{2}, D. Khu\inst{2}, K. Meel\inst{3}, Y. K. Tan\inst{2}}
+\author{E. Karayel\inst{1}, J. Watt\inst{2}, D. Khu\inst{2}, K. Meel\inst{4,5}, Y. K. Tan\inst{2,3}}
 \institute[VFU]{
   \inst{1}%
-    Technical University of Munich
+    Technical University of Munich, Germany
   \and
@@
 \frame{\titlepage}
-\section{Introduction to the CVM algorithm}
 \begin{frame}
-\frametitle{What is the CVM algorithm?}
-\begin{itemize}
-\item Name after initals of S. Chakraborty, N. V. Vinodchandran, K. S. Meel
-\item Streaming Algorithm (one-pass)
-\[
-  a_1, \ldots, a_l
-\]
-\item Algorithm has only limited mutable state
-\[
-  \bigo( \varepsilon^{-2} \ln(\delta^{-1} l ) b ) 
-\]
-\item Approximation of the count of distinct elements
-\item Randomized
-\item Small error with high probability
-\[
-  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
-\]
-for choosable $\varepsilon > 0$, $\delta > 0$.
-\end{itemize}
-\end{frame}
-\begin{frame}
-\frametitle{Illustration of Streaming Algorithms}
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[
-  streamcell/.style={rectangle, draw=black, fill=black!15, thick, text centered,inner sep=3pt,minimum width=0.6cm},
-]
-\node[streamcell] (cell1) at (-3,2) {$a_1$};
-\node[streamcell] (cell1) at (-2.4,2) {$a_1$};
-\node[streamcell] (cell1) at (-1.8,2) {$a_2$};
-\node at (0,2) {$\cdots$};
-\node[streamcell] (cell1) at (2,2) {$a_{l}$};
-\node (robot) at (0,0)
-    {\includegraphics[width=.25\textwidth]{contrib/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545_655/vecteezy_cute-robot-holding-clipboard-cartoon-icon-illustration_53779545}};
-\end{tikzpicture}
-\end{center}
-\end{figure}
-\end{frame}
-%\section{Comparison with Previous Algorithms}
-\begin{frame}[label=cvm_algorithm]
-\frametitle{The CVM Algorithm}
-%\begin{algorithm}[h!]
-%	\caption{CVM algorithm for distinct elements estimation~\cite{chakraborty2022}.}\label{alg:cvm}
-\begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
-  \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(p)$ \Comment random bit $b$ from the Bernoulli distribution
-    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
-      \State $\chi \gets \chi \cup \{a_i\}$
-    \Else \Comment remove $a_i$ otherwise
-      \State $\chi \gets \chi - \{a_i\}$
-    \EndIf
-    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
-      \State $\chi \getsr \mathrm{subsample}(\chi)$
-      \State $p \gets \frac{p}{2}$
-    \EndIf
-    \If{$|\chi| = n$}
-      \Return $\bot$ \Comment fail if $\chi$ remains full
-    \EndIf
-  \EndFor
-  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
-\end{algorithmic}
-%\end{algorithm}
+\frametitle{Table of Contents}
+\tableofcontents
 \end{frame}
 
-\input{algorithm_animation}
+\input{sections/intro.tex}
 
-\againframe{cvm_algorithm}
+\input{sections/prob_fun_invariants.tex}
 
-\begin{frame}
-\frametitle{Comparison with Previous Algorithms}
+\input{sections/original_proof.tex}
 
-\begin{table}
-\begin{tabular}{l | l | l | l}
-\hline
-Year & Author(s) & Space & Remarks \\
-\hline
-2001 & Bar-Yossef et al.~\cite{baryossef2002} & $\tilde{O}( \ln(\delta^{-1}) (\varepsilon^{-2}  + b ))$ & M. H. \\
-2007 & HLL~\cite{flajolet2007} & $O(\ln(\delta^{-1} \varepsilon^{-2} \ln b)$ & O. M. \\
-2010 & Kane et al.~\cite{kane2010} & $O( \ln(\delta^-1) (\varepsilon^{-2} + b) )$ & H. \\
-2020 & B\l{}asiok~\cite{blasiok2020} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. \\
-2022 & Karayel~\cite{karayel2022} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. M. \\
-2023 & Chakraborty et al.~\cite{chakraborty2023} & $O( \varepsilon^{-2} \ln( \delta^{-1} l) b )$ & - \\
-\end{tabular}
-\end{table}
-Abbreviations: 
-\begin{itemize}
-\item M: Supports merging of sketches
-\item H: Relies on hash functions.
-\item E: Relies on expander graphs.
-\item O: Random oracle model.
-\end{itemize}
-\end{frame}
-
-\section{Summary of Results}
-\begin{frame}
-\frametitle{Summary of Results}
-\begin{itemize}
-\item <2-> Verification of Correctness \only<6->{using the original and new method}
-\item <3-> Discovery of a new more succint proof (1003 lines vs. 2634 lines.)\only<4->{, which is based on a \emph{new} proof technique for randomized algorithms.}
-\item <5-> We call it: \emph{Functional Probabilistic Invariants}
-\item <7-> Discovery of an unbiased total variant of the algorithm. (Application of our new technique.)
-\end{itemize}
-\end{frame}
-
-\section{Verification using Functional Probabilistic Invariants}
-\begin{frame}[label=algsplit]
-\frametitle{Another slide with the algorithm}
-\begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
-  \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(p)$
-    \If{$b$}
-      \State $\chi \gets \chi \cup \{a_i\}$
-    \Else
-      \State $\chi \gets \chi - \{a_i\}$
-    \EndIf
-    \If{$|\chi| = n$}
-      \State $\chi \getsr \mathrm{subsample}(\chi)$
-      \State $p \gets \frac{p}{2}$
-    \EndIf
-    \If{$|\chi| = n$}
-      \Return $\bot$
-    \EndIf
-  \EndFor
-  \State \Return $\frac{|\chi|}{p}$
-\end{algorithmic}
-
-\tikz[overlay, shift=(current page.north west)]{
-\begin{scope}[x={(current page.north east)},y={(current page.south west)}]
-
-\only<2->{
-        \fill[blue!50, opacity=0.5] (0.07,0.4) rectangle (0.7,0.67);
-        \node [anchor=west] at (0.45,0.535) {Step 1};
-}
-\only<3->{
-        \fill[red!50, opacity=0.5] (0.07,0.675) rectangle (0.7,0.845);
-        \node [anchor=west] at (0.45,0.76) {Step 2};
-}
-\only<4>{
-        \fill[yellow!50,opacity=0.5] (0.07,0.846) rectangle (0.7,0.895);
-        \node [anchor=west] at (0.45,0.87) {Step 3};
-}
-\only<5->{
-        \fill[white] (0.07,0.846) rectangle (0.7,0.895);
-}
-\only<6->{
-  \fill[green!50, opacity=0.5] (0.03, 0.895) rectangle (0.7, 0.97);
-   \node [anchor=west] at (0.45,0.932) {Estimate};
-}
-\end{scope}
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Giry Monad}
-We can represent randomized algorithms using the Giry monad:
-\begin{itemize}
-\item Primitive random operations, e.g., $\Ber(p)$
-\item Return operation $\mathrm{return} \, x$
-\item Sequential compositon $m \mbind f$
-\end{itemize}
-\pause
-\begin{block}{Fact}
-For an event $E$: 
-\begin{eqnarray*}
-\prob_{m \mbind f}( E )  & = & \int_{m} P_{f(x)} (E) \, dx \onslide<3->{= \sum_x P_{f{x}} (E) P_m({x})} \\
-\onslide<4->{\expect_{m \mbind f}[ g ]  & = & \int_{m} \expect_{f(x)} [g] \, dx}
-\end{eqnarray*}
-\end{block}
-%\end{frame}
-%\begin{frame}
-%\frametitle{Giry Monad II}
-\end{frame}
-\againframe<5>{algsplit}
-\begin{frame}
-\frametitle{Our algorithm in monadic notation}
-\begin{eqnarray*}
-  \mathrm{init} & \mbind & \mathrm{step}_1 a_1 \mbind \mathrm{step}_2 \mbind \mathrm{step}_1 a_2 \mbind \mathrm{step}_2 \cdots \\
-  & \mbind & \mathrm{step_1} a_l \mbind \mathrm{step}_2 \mbind \mathrm{estimate} \\
-  \mathrm{init} & = & \mathrm{return} \, (1, \emptyset)
-\end{eqnarray*}
-\pause
-Key invariant:
-\[
-\expect \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1
-\]
-for all $x$ that are present in the stream.
-%\end{block}
-\end{frame}
-
-\begin{frame}
-\frametitle{Verifying the invariant: Induction step for Step 2}
-\begin{block}{Assumption}
-$\expect_m \left[ \frac{\indicat(x \in \chi)}{p} \right] = 1$
-\end{block}
-\begin{block}{Goal}
-$ L := \expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]= 1$
-\end{block}
-\[
-L =
-\only<1>{\expect_{m \mbind \mathrm{step}_2} \left[ \frac{\indicat(x \in \chi)}{p} \right]}
-\only<2>{\integral{m}
-      {\left(\ift{|\chi|=n}{\integral{\mathrm{subs.}(\chi)}{\frac{\indicat(s \in \tau)}{p/2}}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
-      {\sigma}}
-\only<3>{\integral{m}
-      {\left(\ift{|\chi|=n}{\frac{p}{2}\integral{\mathrm{subs.}(\chi)}{\indicat(s \in \tau)}{\tau}}{\frac{\indicat(s \in \chi)}{p}} \right)}
-      {\sigma}}
-\only<4>{\integral{m}
-      {\left(\ift{|\chi|=n}{\frac{p}{2}\frac{\indicat(s \in \chi)}{2}}{\frac{\indicat(s \in \chi)}{p}} \right)}
-      {\sigma}}
-\only<5>{\integral{m}
-      {\left(\ift{|\chi|=n}{\frac{\indicat(s \in \chi)}{p}}{\frac{\indicat(s \in \chi)}{p}} \right)}
-      {\sigma}}
-\only<6>{\integral{m}
-      {\left(\frac{\indicat(s \in \chi)}{p} \right)}
-      {\sigma}}
-\only<7>{1}
-\]
-\end{frame}
-
-\begin{frame}<1>[label=observations]
-\frametitle{Observations}
-\begin{itemize}
-\item Expected result is the count of distinct elements.
-\only<2->{\item Can extend the above recursive analysis to more complex expressions: }
-\only<3>{
-\[
-  \expect \left[ h \left( \frac{\indicat(s \in \chi)}{p} \right)\right] \leq h(1)
-\]
-for every non-negative concave function $h$ and stream elements $s$.}
-\only<4->{
-\[
-  \expect \left[ \prod_{s \in S}  h \left( \frac{\indicat(s \in \chi)}{p} \right)\right] \leq h(1)^{|S|}
-\]
-for every non-negative concave function $h$ and subset of stream elements $S$.}
-\only<5->{\item $\Rightarrow$ Approximation Guarantee
-\[
-  \prob\left(|X - |A|| \geq \varepsilon\right) \leq \delta 
-\]
-}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Concentration}
-\begin{tikzpicture}[scale=0.5]
-\begin{axis}[
-    xmin=0, xmax = 6.5,
-    ymin=0, ymax=0.7,
-    axis lines = left,
-    xlabel = {},
-    xtick={2.75,3.75,4.75},
-    xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
-    ytick=\empty,
-    ylabel = {pdf},
-    height = \textwidth,
-    width = \textwidth,
-]
-\addplot [
-    domain=0:6, 
-    samples=1000, 
-]{max(0,35/2*(x/6)^4*(1-x/6)^2)};
-\addplot [
-    domain=4.75:6, 
-    samples=1000, 
-    fill=gray,
-]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
-\addplot [
-    domain=0:2.75, 
-    samples=1000, 
-    fill=gray,
-]{max(0,35/2*(x/6)^4*(1-x/6)^2)} \closedcycle;
-\draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
-\end{axis}
-\end{tikzpicture}
-\begin{tikzpicture}[scale=0.5]
-\begin{axis}[
-    xmin=0, xmax = 6.5,
-    ymin=0, ymax=0.7,
-    axis lines = left,
-    xlabel = {},
-    xtick={2.75,3.75,4.75},
-    xticklabels={$\mu-\varepsilon$,$\mu$,$\mu+\varepsilon$},
-    ytick=\empty,
-    ylabel = {pdf},
-    height = \textwidth,
-    width = \textwidth,
-]
-\addplot [
-    domain=0:6, 
-    samples=1000, 
-]{0.65*exp(-(2*(x-3.75))^2)};
-\addplot [
-    domain=4.75:6, 
-    samples=1000, 
-    fill=gray,
-]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
-\addplot [
-    domain=0:2.75, 
-    samples=1000, 
-    fill=gray,
-]{0.65*exp(-(2*(x-3.75))^2)} \closedcycle;
-\draw[color=red, dashed] (axis cs:3.75, 0) -- (axis cs:3.75, 0.7);
-\end{axis}
-\end{tikzpicture}
-\end{frame}
-
-\againframe<2->{observations}
-
-\begin{frame}
-\frametitle{Observations II}
-\begin{itemize}
-\item The new proof is much shorter than the original proof (1003 lines vs. 2634 lines.)
-\only<2->{\item \emph{As far as we can tell,} this technique is new.}
-\only<3->{\item The new proof opens up the design-space of the algorithm.}
-\only<4>{\item For example we can select the subsampling ratio dynamically.}
-\only<6->{\item More interestingly: It is possible to use a different subsampling operation.}
-\only<7->{\item Select a random $nf$-subset (where $\frac{1}{2} \leq f < 1$, $nf \in \mathbb Z$).}
-\only<8->{\item $\Rightarrow$ Unbiased algorithm}
-\end{itemize}
-\end{frame}
-\againframe<4>{algsplit}
-
-\section{Sketch of the Original Proof}
-\begin{frame}
-\frametitle{Original Proof}
-\begin{tikzpicture}[node distance=2.7cm, auto,scale=0.8,every node/.style={scale=0.8}]
-
-    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center] {CVM};
-    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=A] {CVM w/o\\ Step 3};
-    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=B] {Eager Alg.};
-    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=C] {Random \\ Process};
-    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=D] {Chernoff \\ Bounds};
-
-    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=B, node distance = 2.0cm] {Functional\\Invariant};
-
-    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=E, node distance = 2.0cm] {CVM PAC\\ Guarantee};
-
-    \draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(E.south east) + (0.3,-0.6)$);
-    \node[anchor=south] at ($(D.south) + (0.0,-0.6)$) {Transformation-based Proof};
-
-    \draw[dashed,thick] ($(A.north west) + (-0.5,0.4)$) rectangle ($(C'.south east) + (0.2,-0.2)$);
-    \node[anchor=south,align=center] at ($(C'.west) + (-1.5,-0.45)$) {Probabilistic\\Invariant Proof};
-
-    \draw[->] (A) -- (B);
-    \draw[->] (B) -- (C);
-    \draw[->] (B) -- (C');
-    \draw[->] (C) -- (D);
-    \draw[->] (D) -- (E);
-    \draw[->] (E) -- (pac);
-    \draw[->] (C') -- (pac);
-
-\end{tikzpicture}
-
-\end{frame}
-
-\againframe{cvm_algorithm}
-
-\begin{frame}
-\frametitle{CVM Algorithm II}
-\begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
-  \For{$i \gets 1$ to $l$}
-    \State $b \getsr \Ber(2^{-k})$ \Comment random bit $b$ from the Bernoulli distribution
-    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
-      \State $\chi \gets \chi \cup \{a_i\}$
-    \Else \Comment remove $a_i$ otherwise
-      \State $\chi \gets \chi - \{a_i\}$
-    \EndIf
-    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
-      \State $\chi \getsr \mathrm{subsample}(\chi)$
-      \State $k \gets k + 1$
-    \EndIf
-  \EndFor
-  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
-\end{algorithmic}
-\end{frame}
-
-\begin{frame}
-\frametitle{Eager Algorithm}
-\begin{algorithmic}
-  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
-  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ %such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
-  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
-  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
-  \For{$i \gets 1$ to $l$}
-    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} % \Comment insert $a_i$ if first $k$ flips are $1$s.
-      \State $\chi \gets \chi \cup \{a_i\}$
-    \Else
-      \State $\chi \gets \chi - \{a_i\}$
-    \EndIf
-    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
-      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ % \Comment keep elems.~whose $k+1$-th flip is $1$
-      \State $k \gets k+1$
-    \EndIf
-  \EndFor
-  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
-  \end{algorithmic}
-\end{frame}
-
-\begin{frame}
-\frametitle{Random Process}
-Previous algorithm preserves the invariant:
-\[
-  s \in \chi \leftrightarrow \textrm{b[ last{\textunderscore}index}(s), 0] = \cdots = \textrm{b[ last{\textunderscore}index}(s), k-1] = 1
-\]
-Establish Chernoff bounds using independence of the coin flips of $b$ for all possible outcomes of $k$.
-\end{frame}
-
-\begin{frame}
-\frametitle{Negative Association}
-\begin{itemize}
-\item Generalization of the concept of Independence
-\pause
-\item Obey rules similar to independent random variables
-\pause
-\item Ober Chernoff-Bounds similar to independent random variables
-\pause
-\item Formalized in a separate AFP entry: Negative{\textunderscore}Association
-\pause
-\item Example: The sequence of bits in an idealized Bloom Filter form an negative dependent set
-\end{itemize}
-
-\section*{References}
-\begin{frame}[allowframebreaks]
-        \frametitle{References}
-        \bibliographystyle{plain}
-        \bibliography{main.bib}
-\end{frame}
+\input{sections/conclusion.tex}
 

### 2025-09-27 presentation/sections/conclusion.tex
Minor edits and split into sections.

\begin{frame}
\frametitle{Conclusion/Summary}
\begin{itemize}
\item Verification of the CVM Algorithm
\pause
\item New Technique for the Verfication of Randomized Algorithms
\pause
\item New unbiased/total variant of the CVM Algorithm
\pause
\item Development of a Library for Negative Association
\pause
\begin{itemize}
\item A generalization of the concept of independence
\pause
\item Example: Indicator random variables of an idealized Bloom-Filter
\end{itemize}
\pause
\item Thank You!
\pause
\item Artwork: \raisebox{-0.3cm}{\robot{1cm}} By Vecteezy.com.
\end{itemize}
\end{frame}

\section*{References}
\begin{frame}[allowframebreaks]
        \frametitle{References}
        \bibliographystyle{plain}
        \bibliography{main.bib}
\end{frame}

### 2025-09-27 presentation/sections/intro.tex
Minor edits and split into sections.

\section{Introduction to the CVM algorithm}
\begin{frame}
\frametitle{What is the CVM algorithm?}
\begin{itemize}
\item Name after initals of S. Chakraborty, N. V. Vinodchandran, K. S. Meel
\pause
\item Streaming Algorithm (one sequential pass): $a_1, \ldots, a_l$
\pause
\item Algorithm has only limited mutable state: $\bigo( \varepsilon^{-2} \ln(\delta^{-1} l ) b )$
\pause
\item Approximation of the count of distinct elements
\pause
\item Randomized
\pause
\item Small error with high probability
\[
  \prob\left(|X - |A|| \geq \varepsilon |A| \right) \leq \delta 
\]
for choosable $\varepsilon > 0$, $\delta > 0$
\pause
\item Suprisingly simple compared to previous algorithms for the same problem \pause (see e.g. article in Quanta Magazine\cite{quanta})
\pause
\item No hashing
\end{itemize}
\end{frame}
\begin{frame}
\frametitle{Illustration of Streaming Algorithms}
\begin{figure}
\begin{center}
\begin{tikzpicture}[
  streamcell/.style={rectangle, draw=black, fill=black!15, thick, text centered,inner sep=3pt,minimum width=0.6cm},
]
\node[streamcell] (cell1) at (-3,2) {$a_1$};
\node[streamcell] (cell1) at (-2.4,2) {$a_1$};
\node[streamcell] (cell1) at (-1.8,2) {$a_2$};
\node at (0,2) {$\cdots$};
\node[streamcell] (cell1) at (2,2) {$a_{l}$};
\node (robot) at (0,0) {\robot{.25\textwidth}};
\end{tikzpicture}
\end{center}
\end{figure}
\end{frame}

\begin{frame}[label=cvm_algorithm]
\frametitle{The CVM Algorithm}
\begin{algorithmic}
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ % such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$ \Comment random bit $b$ from the Bernoulli distribution
    \If{$b$} \Comment insert $a_i$ if $b$ is true (with prob. $p$)
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else \Comment remove $a_i$ otherwise
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment subsample if buffer $\chi$ is full
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$ \Comment fail if $\chi$ remains full
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$ \Comment estimate cardinality of $A$
\end{algorithmic}
\end{frame}

\input{sections/algorithm_animation}

\againframe{cvm_algorithm}

\begin{frame}
\frametitle{Comparison with Previous Algorithms}
\alt<2>{\definecolor{colorrowa}{rgb}{0,0.8,0.6}}{\definecolor{colorrowa}{rgb}{1,1,1}}
\alt<3>{\definecolor{colorrowb}{rgb}{0,0.8,0.6}}{\definecolor{colorrowb}{rgb}{1,1,1}}
\alt<4>{\definecolor{colorrowc}{rgb}{0,0.8,0.6}}{\definecolor{colorrowc}{rgb}{1,1,1}}
\alt<5>{\definecolor{colorrowd}{rgb}{0,0.8,0.6}}{\definecolor{colorrowd}{rgb}{1,1,1}}
\alt<6>{\definecolor{colorrowe}{rgb}{0,0.8,0.6}}{\definecolor{colorrowe}{rgb}{1,1,1}}
\alt<7>{\definecolor{colorrowf}{rgb}{0,0.8,0.6}}{\definecolor{colorrowf}{rgb}{1,1,1}}
\begin{table}
\begin{tabular}{l | l | l | l}
\hline
Year & Author(s) & Space & Remarks \\
\hline
\rowcolor{colorrowa} 2001 & Bar-Yossef et al.~\cite{baryossef2002} & $\tilde{O}( \ln(\delta^{-1}) (\varepsilon^{-2}  + b ))$ &  M. H. \\
\rowcolor{colorrowb} 2007 & HLL~\cite{flajolet2007} &  $O(\ln(\delta^{-1} \varepsilon^{-2} \ln b)$ & O. M. H. \\
\rowcolor{colorrowc} 2010 & Kane et al.~\cite{kane2010} & $O( \ln(\delta^-1) (\varepsilon^{-2} + b) )$ & H. \\
\rowcolor{colorrowd} 2020 & B\l{}asiok~\cite{blasiok2020} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. \\
\rowcolor{colorrowe} 2022 & Karayel~\cite{karayel2022} & $O( \varepsilon^{-2} \ln (\delta^{-1}) + b )$ & H. E. M. \\
\rowcolor{colorrowf} 2023 & Chakraborty et al.~\cite{chakraborty2023} & $O( \varepsilon^{-2} \ln( \delta^{-1} l) b )$ & - \\
\end{tabular}
\end{table}
Abbreviations: 
\begin{itemize}
\item M: Supports merging of sketches.
\item H: Relies on hash functions.
\item E: Relies on expander graphs.
\item O: Random oracle model.
\end{itemize}
\end{frame}

\section{Summary of Results}
\begin{frame}
\frametitle{Summary of Results}
\begin{itemize}
\item <2-> Verification of Correctness \only<6->{using the original and new method}
\item <3-> Discovery of a new more succinct proof\only<6->{ (1003 lines vs.\ 2634 lines)}\only<4->{, which is based on a \emph{new} proof technique for randomized algorithms.}
\item <5-> We call it: \emph{Functional Probabilistic Invariants}
\item <7-> Discovery of an unbiased total variant of the algorithm. (Application of our new technique.)
\end{itemize}
\end{frame}

### 2025-09-27 presentation/sections/original_proof.tex
Minor edits and split into sections.

\section{Sketch of the Original Proof}
\begin{frame}
\frametitle{Original Proof}
\begin{tikzpicture}[node distance=2.7cm, auto,scale=0.8,every node/.style={scale=0.8}]

    \node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center] {CVM};
    \node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=A] {CVM w/o\\ Step 3};
    \node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=B] {Eager Alg.};
    \node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=C] {Random \\ Process};
    \node (E) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, right of=D] {Chernoff \\ Bounds};

    \node (C') [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=B, node distance = 2.0cm] {Functional\\Invariant};

    \node (pac) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, below of=E, node distance = 2.0cm] {CVM PAC\\ Guarantee};

    \draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(E.south east) + (0.3,-0.6)$);
    \node[anchor=south] at ($(D.south) + (0.0,-0.6)$) {Transformation-based Proof};

    \draw[dashed,thick] ($(A.north west) + (-0.5,0.4)$) rectangle ($(C'.south east) + (0.2,-0.2)$);
    \node[anchor=south,align=center] at ($(C'.west) + (-1.5,-0.45)$) {Probabilistic\\Invariant Proof};

    \draw[->] (A) -- (B);
    \draw[->] (B) -- (C);
    \draw[->] (B) -- (C');
    \draw[->] (C) -- (D);
    \draw[->] (D) -- (E);
    \draw[->] (E) -- (pac);
    \draw[->] (C') -- (pac);

\end{tikzpicture}

\end{frame}

\newcommand{\overlaytransform}[1]{
\begin{tikzpicture}[node distance=2.7cm, auto,scale=0.8,every node/.style={scale=0.8}]

\ifthenelse{#1=1}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
\node (A) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor] {CVM};
\ifthenelse{#1=2}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
\node (B) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=A] {CVM w/o\\ Step 3};
\ifthenelse{#1=3}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
\node (C) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=B] {Eager Alg.};
\ifthenelse{#1=4}{\colorlet{stagecolor}{gray}}{\colorlet{stagecolor}{white}}
\node (D) [draw, rectangle, minimum width=2.1cm, minimum height=1.2cm, align=center, fill=stagecolor, right of=C] {Random \\ Process};
\draw[->] (A) -- (B);
\draw[->] (B) -- (C);
\draw[->] (C) -- (D);
\draw[thick] ($(A.north west) + (-0.3,0.2)$) rectangle ($(D.south east) + (0.3,-0.3)$);
\end{tikzpicture}
}

\begin{frame}
\frametitle{CVM}
\overlaytransform{1}
\begin{algorithmic}
  \State $\chi \gets \{\}, p \gets 1, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6 l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(p)$
    \If{$b$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$}
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $p \gets \frac{p}{2}$
    \EndIf
    \If{$|\chi| = n$}
      \Return $\bot$
    \EndIf
  \EndFor
  \State \Return $\frac{|\chi|}{p}$
\end{algorithmic}
\end{frame}

\begin{frame}
\frametitle{CVM Algorithm II}
\overlaytransform{2}
\begin{algorithmic}
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \For{$i \gets 1$ to $l$}
    \State $b \getsr \Ber(2^{-k})$ 
    \If{$b$} 
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else 
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} 
      \State $\chi \getsr \mathrm{subsample}(\chi)$
      \State $k \gets k + 1$
    \EndIf
  \EndFor
  \State \Return $2^k |\chi|$ 
\end{algorithmic}
\end{frame}

\begin{frame}
\frametitle{Eager Algorithm}
\overlaytransform{3}
\begin{algorithmic}
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \only<2->{\color{red}}
  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
  \only<2->{\color{black}}
  \For{$i \gets 1$ to $l$}
    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$}
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} % \Comment if buffer $\chi$ is full
      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ % \Comment keep elems.~whose $k+1$-th flip is $1$
      \State $k \gets k+1$
    \EndIf
  \EndFor
  \State \Return $2^k |\chi|$ 
  \end{algorithmic}
\end{frame}

\begin{frame}
\frametitle{Random Process}
\overlaytransform{4}

\begin{block}{Observation}
The eager algorithm preserves the invariant:
\[
  s \in \chi \leftrightarrow \text{For all $j < k$:} b[ \textrm{last{\textunderscore}index}(s), j] = 1
\]
\end{block}
\pause
Non-deterministic algorithm:
\begin{algorithmic}
  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ % \Comment perform $l^2$ unbiased independent coin flips
  \State $k \gets \{0,\dots,k_\mathrm{max}\}$ \Comment Non-deterministc step.
  \State $\chi \gets \{ s \in A \mid b[ \textrm{last{\textunderscore}index}(s), j] = 1 \textrm { for all } j < k \}$
  \State \Return $2^k |\chi|$ 
\end{algorithmic}
\end{frame}
